### 2023-08-19 notes/most_likely_diffusion.tex
add notes on diffusion and ancestral positions

\documentclass[aps,rmp, onecolumn]{revtex4}
%\documentclass[a4paper,10pt]{scrartcl}
%\documentclass[aps,rmp,twocolumn]{revtex4}

\usepackage[utf8]{inputenc}
\usepackage{amsmath,graphicx}
\usepackage{color}
%\usepackage{cite}

\newcommand{\bq}{\begin{equation}}
\newcommand{\eq}{\end{equation}}
\newcommand{\bn}{\begin{eqnarray}}
\newcommand{\en}{\end{eqnarray}}
\newcommand{\Richard}[1]{{\color{red}Richard: #1}}
\newcommand{\LH}{\mathcal{L}}

\begin{document}
\title{Inference of ancestral locations}
\author{Richard Neher}
\date{\today}
\maketitle
A common phylogeographic model is diffusion.
Given the sampling locations $r_i$ of the tips $i$ of the tree, the ancestral locations $r_n$ of internal nodes of the tree have a likelihood function
\begin{equation}
    \log L = -\sum_{n\neq root} \frac{(r_n - r_p)^2}{4Dt_n}
\end{equation}
where $r_p$ is the location of the parent of $n$ and $t_n$ is the length of the branch leading to node $n$.
The ancestral locations are not observed and to calculate the overall likelihood need to be integrated over.

This can be done recursively, where the likelihood of a subtree $n$ given the position of node $r_n$ is given by
\begin{equation}
    \begin{split}
        L_n(r_n) & = \prod_{c\in n}\frac{1}{4\pi D t_c}\int  d\, r_c e^{- \frac{(r_c - r_n)^2}{4Dt_c}}P(r_c) \\
        & = \prod_{c\in n} \frac{\sqrt{d_c}}{\sqrt{\pi}}\int d\, r_c e^{- d_c r_n^2 + 2d_c r_n r_c - d_c r_c^2 - a_c r_c^2 + 2 b_c r_c - c_c} \\
        & = \prod_{c\in n} \frac{\sqrt{d_c}}{\sqrt{\pi}}\int d\, r_c e^{- d_c r_n^2 + 2r_c(b_c + r_n d_c) - r_c^2 \left(d_c + a_c\right) - c_c} \\
        & = \prod_{c\in n} \frac{\sqrt{d_c}}{\sqrt{\pi}}\int d\, r_c e^{- d_c r_n^2 - (d_c + a_c)(r_c^2 - 2r_c\frac{b_c + d_c r_n}{d_c + a_c} + \frac{(b_c + d_c r_n)^2}{(d_c + a_c)^2}) + \frac{(b_c + d_c r_n)^2}{d_c + a_c} - c_c} \\
        & = \prod_{c\in n} \frac{\sqrt{d_c}}{\sqrt{\pi}}\int d\, r_c e^{- d_c r_n^2 + (d_c + a_c)(r_c - \frac{b_c + d_c r_n}{d_c + a_c})^2 + \frac{(b_c + d_c r_n)^2}{d_c + a_c} - c_c} \\
        & = \frac{\sqrt{d_c}}{\sqrt{a_c + d_c}} e^{- d_c r_n^2 + (b_c^2 + 2r_n b_c d_c + d_c^2 r_n^2 )/\left(d_c + a_c\right) - c_c} \\
        & = \frac{\sqrt{d_c}}{\sqrt{a_c + d_c}} e^{- d_c\left(1 - \frac{d_c}{a_c+d_c}\right) r_n^2 +  2\frac{b_c d_c}{d_c+a_c} r_n  - c_c + \frac{b_c^2}{d_c + a_c}} \\
    \end{split}
\end{equation}
This allows calculation of the parameters $a_n$, $b_n$, and $c_n$ of node $n$ from the children that are not terminal nodes as.
\begin{equation}
    a_n = \sum_{c\in n}d_c \left(1-\frac{d_c}{a_c+d_c}\right) = \sum_{c\in n} \frac{d_c a_c}{a_c+d_c}
\end{equation}
\begin{equation}
    b_n = \sum_{c\in n}\frac{b_c d_c}{a_c + d_c}
\end{equation}
\begin{equation}
    c_n = \sum_{c\in n} c_c  + \frac{b_c^2}{d_c + a_c} + \frac{\log(d_c) - \log(a_c+d_c)}{2}
\end{equation}
If a child is a terminal node, the terms in the sum need to be replaced by
\begin{equation}
    a_n = \sum_{c\in n} d_c
\end{equation}
\begin{equation}
    b_n = \sum_{c\in n} d_c r_c
\end{equation}
\begin{equation}
    c_n = \sum_{c\in n} d_c r_c^2 - \log(2\pi/d_c)/2
\end{equation}
Note that for a single child, the variances add ($a_n^{-1} = a_c^{-1} + d_c^{-1}$) and the most likely positions don't change ($b_n/a_n = b_c/a_c$).

The same propagation can be used up the tree
\begin{equation}
    a'_n  =  \frac{d_p a'_p}{a'_p+d_c} + \sum_{c\in p, c\neq n} \frac{d_c a_c}{a_c+d_c}
\end{equation}
\begin{equation}
    b'_n = \frac{b'_p d_p}{a'_p + d_p} + \sum_{c\in p, c\neq n}\frac{b_c d_c}{a_c + d_c}
\end{equation}
\begin{equation}
    c'_n = c'_p + + \frac{b'_p^2}{d_p + a'_p} + \frac{\log(d_p) - \log(a'_p+d_p)}{2} +  \sum_{c\in p, c\neq n} c_c  + \frac{b_c^2}{d_c + a_c} + \frac{\log(d_c) - \log(a_c+d_c)}{2}
\end{equation}

\end{document}

### 2023-08-19 src/estimate_diffusion_from_tree.py
clean transitive edges from tree

@@ -1,29 +1,41 @@
 import numpy as np
 
-def branch_contribution(b, x0, y0, t, non_trivial_nodes):
+def next_nontrivial_child(node):
+    b = node
     while len([c for c in b['children'] if c['alive']])==1:
         i = np.argmax([c['alive'] for c in b['children']])
         b = b['children'][i]
-    return {'dx': b['x']-x0, 'dy':b['y'] - y0, 'dt': b['time']-t, 'next_node':b}
+    return b
 
-def estimate_diffusion_rec(node, displacements, non_trivial_nodes=False):
-    for child in node['children']:
-        data = branch_contribution(child, node['x'], node['y'], node['time'], non_trivial_nodes)
+def clean_tree(tree):
+    def recursively_bridge(node):
+        node["clades"] = [next_nontrivial_child(c) for c in node['children'] if c['alive']]
+        for c in node['clades']:
+            recursively_bridge(c)
+
+    recursively_bridge(tree)
+
+
+def branch_contribution(b, x0, y0, t):
+    return {'dx': b['x']-x0, 'dy':b['y'] - y0, 'dt': b['time']-t}
+
+
+def estimate_diffusion_rec(node, displacements):
+    for child in node['clades']:
+        data = branch_contribution(child, node['x'], node['y'], node['time'])
         displacements.append(data)
-        estimate_diffusion_rec(data['next_node'], displacements, non_trivial_nodes)
+        estimate_diffusion_rec(child, displacements)
 
 # produce empirical estimates of the diffusion constant from the simulated tree annotated with
 # positions and time.
-def estimate_diffusion(tree, include_root=False, non_trivial_nodes=False):
+def estimate_diffusion(tree, include_root=False):
     displacements = []
 
     # handle root-parent-branch contribution separately
-    if len(tree['children'])==1:
+    if len(tree['clades'])==1 and include_root:
         #advance to first furcating node
-        data = branch_contribution(tree, tree['x'], tree['y'], tree['time'], non_trivial_nodes)
-        c = data['next_node']
-        if include_root: # otherwise ignore
-            displacements.append(data)
+        data = branch_contribution(tree, tree['x'], tree['y'], tree['time'])
+        displacements.append(data)
     else:
         c=tree
 

### 2023-08-19 src/habitat_shifts.py

@@ -2,7 +2,7 @@ import numpy as np
 from density_regulation import make_node, evolve
 from heterogeneity import get_2d_hist
 from inflated_diffusion import subsample_tree
-from estimate_diffusion_from_tree import estimate_diffusion
+from estimate_diffusion_from_tree import estimate_diffusion, clean_tree
 
 def generate_target_density(N, Lx, Ly, period, wave_length):
     def f(x,y,t):
@@ -30,6 +30,7 @@ def diffusion_in_changing_habitats(D, interaction_radius, density_reg, N, subsam
             density_variation.append(np.std(H)/N*np.prod(H.shape))
             for sample in range(5):
                 subsample_tree(terminal_nodes, tree, p=subsampling)
+                clean_tree(tree)
                 D_res = estimate_diffusion(tree)
                 D_est.extend([D_res['Dx_total'], D_res['Dy_total']])
 

### 2023-08-19 src/inflated_diffusion.py

@@ -2,7 +2,7 @@ import numpy as np
 from itertools import product
 from density_regulation import make_node, evolve
 from heterogeneity import get_2d_hist, get_granularity
-from estimate_diffusion_from_tree import estimate_diffusion
+from estimate_diffusion_from_tree import estimate_diffusion, clean_tree
 import matplotlib.pyplot as plt
 
 def set_alive_rec(node):
@@ -44,6 +44,7 @@ def estimate_inflated_diffusion(D, interaction_radius, density_reg, N, subsampli
             density_variation.append(np.std(H)/N*np.prod(H.shape))
             for sample in range(5):
                 subsample_tree(terminal_nodes, tree, p=subsampling)
+                clean_tree(tree)
                 D_res = estimate_diffusion(tree)
                 D_est.extend([D_res['Dx_total'], D_res['Dy_total']])
 

### 2023-08-19 src/plot_habitat_diffusion.py

@@ -42,8 +42,11 @@ if __name__=="__main__":
         area = np.minimum(1,D_array*np.pi**2)
         for i, (ir, dr, T, p) in enumerate(product(interaction_radius, density_reg,
                                                    period, subsampling)):
-            plt.plot(D_array*N/Lx/Ly, density_variation[ir, dr, :, N, T, p],
-                    label=f'r={ir}, a={dr} N={N}, T={T} p={p}', ls=ls[i%len(ls)], c=f"C{i//len(ls)}")
+            try:
+                plt.plot(D_array*N/Lx/Ly, density_variation[ir, dr, :, N, T, p],
+                        label=f'r={ir}, a={dr} N={N}, T={T} p={p}', ls=ls[i%len(ls)], c=f"C{i//len(ls)}")
+            except:
+                pass
 
         free_diffusion_heterogeneity = free_diffusion(D_array*N/Lx/Ly, N, linear_bins=linear_bins)
         plt.plot(D_array*N/Lx/Ly, free_diffusion_heterogeneity, label='diffusion', c='k')
@@ -62,9 +65,12 @@ if __name__=="__main__":
         print(D_array[0])
         for i, (ir, dr, T, p) in enumerate(product(interaction_radius, density_reg,
                                                    period, subsampling)):
-            plt.errorbar(D_array*N/Lx/Ly, diffusion_mean[ir, dr, :, N, T, p]/D_array,
+            try:
+                plt.errorbar(D_array*N/Lx/Ly, diffusion_mean[ir, dr, :, N, T, p]/D_array,
                                     diffusion_std[ir, dr, :, N, T, p]/D_array/np.sqrt(10), marker=m,
                     label=f'r={ir}, a={dr} N={N}, T={T} p={p}', ls=ls[i%len(ls)], c=f"C{i//len(ls)}")
+            except:
+                pass
 
         plt.plot(N*D_array/Lx/Ly, np.ones_like(D_array), c='k')
     plt.legend()

### 2023-08-20 src/estimate_diffusion_from_tree.py
add estimation of positions

@@ -15,7 +15,6 @@ def clean_tree(tree):
 
     recursively_bridge(tree)
 
-
 def branch_contribution(b, x0, y0, t):
     return {'dx': b['x']-x0, 'dy':b['y'] - y0, 'dt': b['time']-t}
 
@@ -50,3 +49,79 @@ def estimate_diffusion(tree, include_root=False):
             'vx_total': 0.5*np.sum(np.abs([d['dx'] for d in displacements])/np.sum([d['dt'] for d in displacements])),
             'vy_total': 0.5*np.sum(np.abs([d['dy'] for d in displacements])/np.sum([d['dt'] for d in displacements]))
             }
+
+
+def estimate_ancestral_positions(tree, D):
+    def preorder(node):
+
+        node['dis_to_parent'] = {'x':{'a':0, 'b':0, 'c':0}, 'y':{'a':0, 'b':0, 'c':0}}
+        for c in node['clades']:
+            dc = 1.0/(4*D*(c['time']-node['time']))
+            print(dc)
+            if 'clades' in c and len(c['clades']):
+                preorder(c)
+                for x in ['x', 'y']:
+                    child_d = c['dis_to_parent'][x]
+                    node['dis_to_parent'][x]['a'] += dc*child_d['a']/(dc + child_d['a'])
+                    node['dis_to_parent'][x]['b'] += dc*child_d['b']/(dc + child_d['a'])
+                    node['dis_to_parent'][x]['c'] += child_d['c'] \
+                                        + child_d['b']**2/(dc + child_d['a']) \
+                                        + 0.5*np.log(dc/(dc + child_d['a']))
+            else:
+                for x in ['x', 'y']:
+                    node['dis_to_parent'][x]['a'] += dc
+                    node['dis_to_parent'][x]['b'] += dc*c[x]
+                    node['dis_to_parent'][x]['c'] += dc*c[x]**2 - 0.5*np.log(np.pi/dc)
+
+
+    preorder(tree)
+    tree['dis_from_parent'] = {'x':{'a':0, 'b':0, 'c':0}, 'y':{'a':0, 'b':0, 'c':0}, 'dt': np.inf}
+    tree['position'] = {}
+    for x in ['x', 'y']:
+        d = tree['dis_to_parent'][x]
+        tree['position'][x] = {'var': 2.0/d['a'], 'mean': d['b']/d['a']}
+
+    def postorder(node):
[…]

### 2023-08-20 src/free_diffusion.py

@@ -5,7 +5,7 @@ def random_tree(n, yule=False):
     Generate a random tree either with kingman or yule branching properties
     '''
     def node(t, children=None, node_type="internal", name=None):
-        return {"time": t, "children": children or [], "type":node_type, "name":name}
+        return {"time": t, "children": children or [], "type":node_type, "name":name, "alive":True }
 
     t = 0
     tree = [node(t, node_type="terminal", name=i) for i in range(n)]

### 2023-08-20 src/sample_size_dependence.py

@@ -4,7 +4,7 @@ from estimate_diffusion_from_tree import *
 from itertools import product
 
 if __name__=="__main__":
-    replicates = 1000
+    replicates = 10
     results = {}
     fields = ['Dx_branch', 'Dy_branch', 'vx_branch', 'vy_branch', 'Dx_total', 'Dy_total', 'vx_total', 'vy_total']
     for n, yule in product([10,30, 100, 300, 1000], [True, False]):
@@ -12,6 +12,7 @@ if __name__=="__main__":
         for i in range(replicates):
             tree = random_tree(n, yule=yule)['tree']
             add_positions(tree, 1)
+            clean_tree(tree)
             D = estimate_diffusion(tree)
             res.append(D)
 

### 2023-08-20 src/density_regulation.py
fix errors in diffusion calc

@@ -72,14 +72,14 @@ def dict_to_phylo_tree(d):
 
     def clade_from_dict(d, parent=None):
         clade = Clade()
-        clade.name = 'root'
+        clade.name = f'{d["t"]:1.2f}_{d["x"]:1.2f}_{d["y"]:1.2f}'
         clade.branch_length = d['time'] - parent.t if parent else 0.001
         clade.t = d['time']
         clade.pos = [d['x'], d['y']]
         return clade
 
     def add_clades(clade, d):
-        for k in d['children']:
+        for k in d['clades']:
             new_clade = clade_from_dict(k, clade)
             clade.clades.append(new_clade)
             add_clades(new_clade, k)

### 2023-08-20 src/estimate_diffusion_from_tree.py

@@ -57,7 +57,6 @@ def estimate_ancestral_positions(tree, D):
         node['dis_to_parent'] = {'x':{'a':0, 'b':0, 'c':0}, 'y':{'a':0, 'b':0, 'c':0}}
         for c in node['clades']:
             dc = 1.0/(4*D*(c['time']-node['time']))
-            print(dc)
             if 'clades' in c and len(c['clades']):
                 preorder(c)
                 for x in ['x', 'y']:
@@ -72,6 +71,9 @@ def estimate_ancestral_positions(tree, D):
                     node['dis_to_parent'][x]['a'] += dc
                     node['dis_to_parent'][x]['b'] += dc*c[x]
                     node['dis_to_parent'][x]['c'] += dc*c[x]**2 - 0.5*np.log(np.pi/dc)
+            for x in ['x', 'y']:
+                node['dis_to_parent'][x]['var'] = 2.0/node['dis_to_parent'][x]['a']
+                node['dis_to_parent'][x]['mean'] = 0.5*node['dis_to_parent'][x]['b']*node['dis_to_parent'][x]['var']
 
 
     preorder(tree)
@@ -90,7 +92,7 @@ def estimate_ancestral_positions(tree, D):
                 if 'clades' in c1 and len(c1['clades']):
                     c1_d = c1['dis_to_parent'][x]
                     c1['dis_from_parent'][x]['a'] = node['dis_to_parent'][x]['a'] - dc*c1_d['a']/(dc + c1_d['a'])
-                    c1['dis_from_parent'][x]['b'] = node['dis_to_parent'][x]['b'] - dc*c_d['b']/(dc + c1_d['c'])
+                    c1['dis_from_parent'][x]['b'] = node['dis_to_parent'][x]['b'] - dc*c1_d['b']/(dc + c1_d['a'])
                     c1['dis_from_parent'][x]['c'] = node['dis_to_parent'][x]['c'] - c1_d['c'] - c1_d['b']**2/(dc + c1_d['a']) - 0.5*np.log(dc/(dc + c1_d['a']))
                 else:
                     c1['dis_from_parent'][x]['a'] = node['dis_to_parent'][x]['a'] - dc
@@ -107,9 +109,9 @@ def estimate_ancestral_positions(tree, D):
 
     def marginal_positions(node):
         node['position'] = {}
+        dn = 1/(4*D*node['dis_from_parent']['dt'])
         for x in ['x', 'y']:
[…]

### 2023-09-10 manuscript/phylogeo.tex
started ms

\documentclass[aps,rmp, onecolumn]{revtex4}
%\documentclass[a4paper,10pt]{scrartcl}
%\documentclass[aps,rmp,twocolumn]{revtex4}

\usepackage[utf8]{inputenc}
\usepackage{amsmath,graphicx}
\usepackage{color}
%\usepackage{cite}

\newcommand{\Richard}[1]{{\color{red}Richard: #1}}
\newcommand{\LH}{\mathcal{L}}
\newcommand{\rvec}{\mathbf{r}}

\bibliography{bib}

\begin{document}
\title{When to trust phylogeographic inferences?}
\author{Richard Neher}
\date{\today}
\begin{abstract}
    Sometimes.
\end{abstract}

\maketitle
As organisms replicate, they also disperse in space, potentially exploring new habitats.
The reconstruction of ancestral locations and past migrations from sampling locations of extant individuals, typically in combination with genome sequence information to infer the phylogeny, are known as phylogeography.
Phylogeographic methods are implemented in popular evolutionary analysis software such as BEAST \citep{pybus_unifying_2012}.
The underlying model of dispersal is typically diffusive, that is the probability of sampling a descendent at position $\rvec_c$ after time $t$ when the parent was at position $\rvec_p$ is given by
\begin{equation}
    P(\rvec_c| \rvec_p, t) = \frac{1}{4\pi D t}e^{\frac{|r_c - r_p|^2}{4Dt}}
\end{equation}
where $D$ is a diffusion constant with dimensions $\mathrm{length}^2/\mathrm{time}$ and we have assumed a two dimensional space.
Each branch in the phylogeny is adorned with such a diffusive propagator.
Such a diffusive process is the simplest and most natural choice in absence of directed motion or long range migration.
It is also mathematically convenient as unknown ancestral locations can be integrated in closed form and the marginal distribution of each position can be calculated exactly for a fixed tree.
However, it has one major shortcoming: tree topology and branching rates of the tree are independent of the spatial process.
However, these two interact in important ways, for example when a population expands into an uninhabited fertile habitat or a disease spreads in a susceptible population.

At the same time, information in two dimensional sampling locations is low compared to that from high-dimensional sequence space and processes not captured by the diffusive model can potentially result in wrong inferences.
The purpose of this note is to explore how sensitive phylogeographic inferences are to model assumptions and what quantities can be robustly estimated.
My focus here is only on the ability and robustness to estimate dispersal parameter and ancestral locations and I will assume perfect knowledge of the tree and the sampling locations.
So no tree-sampling will be necessary.

In a typically Bayesian phylogeographic inference problem as discussed in \citet{pybus_unifying_2012}, the input data are typically viral genome sequences as well as their sampling times and sampling locations.
Software like BEAST will then sample the posterior distribution of the phylogenetic tree topology, the sampling times and locations of ancestral nodes in the tree, and model parameters like the diffusion constant $D$ and the evolutionary rate $\mu$.
Results can be estimates of the ancestral location of a particular groups of samples, or summary statistics of the dispersal process.
Popular summaries include empirical estimates of the diffusion constant and a so called \emph{dispersal rate}. These are calculated from the displacements along branch $i$ with parent $p_i$ given by $d_i = |\rvec_{p_i} - \rvec_{i}|$ and the elapsed time $\Delta t_i$ as
\begin{equation}
    D_w = \frac{\sum_{i\in B}d_i^2}{4\sum_{i\in B} \Delta t_i} \quad \mathrm{and}  \quad v_w = \frac{\sum_{i\in B} d_i}{\sum_{i\in B} \Delta t_i}
\end{equation}
where $B$ is the set of all branches of the tree.
Effectively $D_w$ divides the sum of all observed squared displacements by the total time elapsed on the tree to obtain and estimate of $D$.
The dispersal rate estimate compares the sum of absolute values of observed displacements to the total time, which has dimensions of a velocity.

Fig.~\ref{fig:D_and_v} shows empirical estimates of $D_w$ and $v_w$ as defined above from simulated data using freely diffusing locations along Kingman or Yule trees.
The diffusion constant $D$ can be estimated robustly from these data, but $v_w$ is not a well defined quantity.
Values of $v_w$ depends strongly on the sample size and there is not ground truth to compare to as there is no parameter with dimensions of a velocity in the model.
That $v_w$ is ill-defined is not surprising: By changing the sample size, the average length of branches changes. Diffusion along each branch results in a displacement of $d_i \sim \sqrt{4Dt_i}$ such that a quantity that depends on $\Delta d_i / t_i \sim 1/t_i^{1/2}$ will change with the time $t_i$ that elapsed along the branch.
Values of $v_w$ are meaningless in this model.

\begin{figure}
    \includegraphics*[width=0.48\textwidth]{figures/kingman_total_dispersal.png}
    \includegraphics*[width=0.48\textwidth]{figures/yule_total_dispersal.png}
    \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal rate'' for different sample sizes.}
    While diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ is not a well defined quantity. Its numerical value depends strongly on sample size and there is no ground truth to compare to. This sample size dependence is stronger for Yule trees (right) than Kingman trees (left). }
\end{figure}

\section*{Is the diffusion model consistent?}
Phylogeographic assume that the spatial diffusion process is independent of the branching pattern of the tree and ignores coupling between the two that might arise from density dependent growth rates.
If one simulates a diffusive spatial process in a square of length $L$, the population density shows strong fluctuations whenever the diffusion constant $D<L^2/N$, see Fig.~\ref{fig:heterogeneity}.
$N$ here plays the role of the time to the MRCA of the population.
At $D=0.1L^2/N$, the population distribution is patchy with large areas un-inhabited (Fig.~\ref{fig:heterogeneity} left).
This heterogeneity increases substantially for even lower $D$, when the population is essentially a cloud of width $\sqrt{4DN}$.

\begin{figure}
    \includegraphics*[width=\textwidth]{figures/heterogeneity_free_diffusion.pdf}
    \caption[short]{\label{fig:heterogeneity}{\bf Population density variation in free diffusion models.}
    When spatial motion is independent of population density, the realized population density in the population will be much larger than biologically plausible. The left panel shows the population density for the case $D = 0.1 L^2/N$ in linear bins of size $L/20$. The right panels shows the standard deviation of the population density for different bin sizes as a function of the diffusion coefficient. For $D>L^2/N$, the heterogeneity approaches the expected value in the well mixed limit. }
\end{figure}

Biological populations spread into areas that support them.
The simplest model of this process are FKPP waves that model the population density $\phi(x,t)$ as
\begin{equation}
    \frac{\partial \phi(x,t)}{\partial t} = D\frac{\partial^2 \phi(x,t)}{\partial x^2} + \alpha phi(x,t)(1-\phi(x,t)/\phi_0)
\end{equation}
where $\phi_0$ is the carrying capacity and $\alpha$ is the population growth rate at zero density.
This equation admits solutions with a traveling front, that the population "invades" the empty territory with velocity $v = \sqrt{D \alpha}$.
Note that the model itself does not have an intrinsic velocity.
The speed at which the population invades space is given as compound of diffusion and accelerated growth in regions of low density.

This dependence of growth rate (and thus tree structure) on density is one key element that phylogeographic models ignore because it is very difficult to capture mathematically.
It is, however, fairly straightforward to include this in simulation and these simulations can be used to investigate the nature of phylogeographic inferences when some of the assumptions are violated.

\end{document}

### 2023-09-10 src/density_regulation.py

@@ -127,7 +127,9 @@ def dict_to_phylo_tree(d, child_attr='children'):
         clade.name = f'{d["time"]:1.2f}_{d["x"]:1.2f}_{d["y"]:1.2f}'
         clade.branch_length = d['time'] - parent.t if parent else 0.001
         clade.t = d['time']
-        clade.pos = [d['x'], d['y']]
+        clade.pos = {'x': d['x'], 'y': d['y']}
+        if 'position' in d:
+            clade.inferred_pos = d['position']
         return clade
 
     def add_clades(clade, d):
@@ -142,6 +144,20 @@ def dict_to_phylo_tree(d, child_attr='children'):
 
     return tree
 
+def add_as_color(tree, quantity = 'x', cmap = None):
+    if cmap is None:
+        from matplotlib.cm import RdBu
+        cmap = RdBu
+
+    def value(n):
+        if quantity=='x':
+            return n.pos['x']
+        elif quantity == 'devx':
+            return 0.5 + 0.2*(n.pos['x']-n.inferred_pos['x']['mean'])/n.inferred_pos['x']['var']**0.5
+
+
+    for n in tree.find_clades():
+        n.color = [int(255*x) for x in  cmap(value(n))[:3]]
 
 if __name__=="__main__":
     import matplotlib.pyplot as plt

### 2023-09-10 src/estimate_diffusion_from_tree.py

@@ -98,13 +98,13 @@ def estimate_ancestral_positions(tree, D):
             d = node['dis_from_parent'][x]
             if "clades" in node and len(node['clades']):
                 n = node['dis_to_parent'][x]
-                node['position'][x] = {'var': 2.0/(n['a'] + d['a']*dn/(d['a'] + dn))}
-                node['position'][x]['mean'] =  0.5*(n['b'] + d['b']*dn/(d['a'] + dn))*node['position'][x]['var']
+                node['position'][x] = {'var': 0.5/(n['a'] + d['a']*dn/(d['a'] + dn))}
+                node['position'][x]['mean'] =  2.0*(n['b'] + d['b']*dn/(d['a'] + dn))*node['position'][x]['var']
             else:
                 if (d['a']*dn/(d['a'] + dn))==0:
                     import ipdb; ipdb.set_trace()
-                node['position'][x] = {'var': 2.0/(d['a']*dn/(d['a'] + dn))}
-                node['position'][x]['mean'] =  0.5*(d['b']*dn/(d['a'] + dn))*node['position'][x]['var']
+                node['position'][x] = {'var': 0.5/(d['a']*dn/(d['a'] + dn))}
+                node['position'][x]['mean'] =  2.0*(d['b']*dn/(d['a'] + dn))*node['position'][x]['var']
 
     def assign_positions(node):
         if node!=tree:

### 2023-09-10 src/heterogeneity.py

@@ -12,7 +12,7 @@ def get_granularity(ntips, D, yule=False, L=1, bins=10):
     T = random_tree(ntips, yule=yule)
     add_positions(T['tree'], D)
     H, bx, by = get_2d_hist(T['terminals'], L, L, bins)
-    return np.std(H)/ntips*np.prod(H.shape)
+    return np.std(H)/ntips*np.prod(H.shape), H
 
 
 if __name__=="__main__":
@@ -20,19 +20,21 @@ if __name__=="__main__":
     ntips = 1000
     D_array = np.logspace(-2,2,101)
 
-    plt.figure()
+    fig, axs = plt.subplots(1,2, figsize=(12,6))
 
+    g, H = get_granularity(ntips, 0.1, bins=20)
+    axs[0].matshow(H)
 
     for line_style, linear_bins in zip(['-', '--', '-.'], [3,5,10]):
         nbins=linear_bins**2
         res = []
         for D in D_array:
-            res.append([get_granularity(ntips, D, bins=linear_bins) for i in range(n_iter)])
+            res.append([get_granularity(ntips, D, bins=linear_bins)[0] for i in range(n_iter)])
 
         res = np.array(res)
         res_mean = res.mean(axis=1)
-        plt.plot(D_array, res_mean, label=f'free diffusion, dx=L/{linear_bins}', c='C0', ls=line_style)
-        plt.plot(D_array, np.ones_like(D_array)*np.sqrt(nbins/ntips), label=f'well mixed limit dx=L/{linear_bins}', c='C1', ls=line_style)
+        axs[1].plot(D_array, res_mean, label=f'free diffusion, dx=L/{linear_bins}', c='C0', ls=line_style)
+        axs[1].plot(D_array, np.ones_like(D_array)*np.sqrt(nbins/ntips), label=f'well mixed limit dx=L/{linear_bins}', c='C1', ls=line_style)
 
     # area = np.minimum(1,D_array*np.pi**2)
     # pred_heterogeneity = np.sqrt(1/area - 1)  # p*(1/p - 1)**2 + (1-p) = p*(1/p^2 - 2/p + 1) + (1-p) = 1/p - 2 + p -1 + p = 1/p - 1
@@ -42,4 +44,5 @@ if __name__=="__main__":
     plt.xlabel('diffusion constant [habitat_size^2/T_c]')
     plt.ylabel('heterogeneity')
     plt.legend()
[…]

### 2023-09-10 src/inflated_diffusion.py

@@ -15,13 +15,14 @@ def estimate_inflated_diffusion(D, interaction_radius, density_reg, N, subsampli
     density_variation = []
     D_est = []
     zscores = []
-    for t in range((n_iter+2)*N):
+    Tmrca = []
+    for t in range((n_iter+5)*N):
         terminal_nodes = evolve(terminal_nodes, t, Lx=Lx, Ly=Ly, interaction_radius=interaction_radius,
                                 density_reg=density_reg, D=D, target_density=N)
         if len(terminal_nodes)<10:
             print("population nearly extinct")
             continue
-        if t%(N//5)==0 and t>2*N: # take samples after burnin every 5 Tc
+        if t%(N//5)==0 and t>5*N: # take samples after burnin every Tc//5
             clean_tree(tree)
             H, bx, by = get_2d_hist(terminal_nodes, Lx, Ly, linear_bins)
             density_variation.append(np.std(H)/N*np.prod(H.shape))
@@ -30,11 +31,15 @@ def estimate_inflated_diffusion(D, interaction_radius, density_reg, N, subsampli
                 D_res = estimate_diffusion(tree)
                 estimate_ancestral_positions(tree, D)
                 z = collect_zscore(tree)
+                if len(tree['clades'])==1:
+                    Tmrca.append(t-tree['clades'][0]['time'])
+                else:
+                    Tmrca.append(t)
                 D_est.extend([D_res['Dx_total'], D_res['Dy_total']])
                 zscores.extend([np.mean(z.loc[z.nonterminal, 'zx']**2),
                                 np.mean(z.loc[z.nonterminal, 'zy']**2)])
 
-    return {"density_variation": density_variation, "D_est": D_est, "zscores": zscores}
+    return {"density_variation": density_variation, "D_est": D_est, "zscores": zscores, "Tmrca":Tmrca}
 
 if __name__=="__main__":
     import sys
@@ -54,8 +59,8 @@ if __name__=="__main__":
     res_density = {}
     res_density_mean = {}
     D_est = []
-    D_array_dens = np.logspace(-3,0,21)*Lx*Ly*2/N
-    n_iter = 30
+    D_array_dens = np.logspace(-1,0,1)*Lx*Ly*2/N
+    n_iter = 1
     linear_bins=5
     interaction_radius, density_reg = args.interaction_radius, args.density_reg
     nsub = 1 if args.subsampling>0.9 else 5
@@ -73,7 +78,8 @@ if __name__=="__main__":
[…]

### 2023-09-10 src/plot_diffusion.py

@@ -1,0 +1,45 @@
+import numpy as np
+from density_regulation import make_node, evolve, clean_tree, dict_to_phylo_tree, add_as_color, subsample_tree
+from heterogeneity import get_2d_hist
+from estimate_diffusion_from_tree import estimate_diffusion, estimate_ancestral_positions, collect_zscore
+from Bio import Phylo
+
+N=1000
+tmax = 10000
+density_reg = 0.1
+interaction_radius = 0.1
+Lx=1
+Ly=1
+D=0.1*Lx*Ly/N
+
+tree = make_node(Lx/2,Ly/2,-2, None)
+tree['children'] = [make_node(np.random.random()*Lx, np.random.random()*Ly, -1, tree)
+                    for i in range(N)]
+terminal_nodes = tree['children']
+
+
+for t in range(tmax):
+    terminal_nodes = evolve(terminal_nodes, t, Lx=Lx, Ly=Ly, interaction_radius=interaction_radius,
+                            density_reg=density_reg, D=D, target_density=N)
+
+clean_tree(tree)
+subsample_tree(terminal_nodes, tree, p=1.0)
+D_res = estimate_diffusion(tree)
+estimate_ancestral_positions(tree, D)
+z = collect_zscore(tree)
+
+import matplotlib.pyplot as plt
+plt.ion()
+
+T = dict_to_phylo_tree(tree)
+from matplotlib.cm import RdBu, Spectral
+from Bio import Phylo
+add_as_color(T, quantity='x', cmap=RdBu)
+Phylo.draw(T)
+
+add_as_color(T, quantity='devx', cmap=Spectral)
+def label_func(n):
+    return f"{n.pos['x']:1.2f}_{n.inferred_pos['x']['mean']:1.2f}_{n.inferred_pos['x']['var']**0.5:1.2f}"
+
+Phylo.draw(T, label_func=label_func)
+

### 2023-09-10 src/plot_habitat_diffusion.py

@@ -8,7 +8,7 @@ def free_diffusion(D_array, N, linear_bins=5):
     n_iter = 10
     res = []
     for D in D_array:
-        res.append([get_granularity(N, D, bins=linear_bins) for i in range(n_iter)])
+        res.append([get_granularity(N, D, bins=linear_bins)[0] for i in range(n_iter)])
 
     res = np.array(res)
     return res.mean(axis=1)
@@ -38,7 +38,7 @@ if __name__=="__main__":
     N_vals = data.N.unique()
 
 
-    ls = ['-', '-.', "--", ":", '-'] #, "-", "--"]
+    ls = ['-', '-.', "--"] #, ":", '-'] #, "-", "--"]
     plt.figure()
     for N in N_vals:
         D_array = data.loc[data.N==N].D.unique()
@@ -62,26 +62,26 @@ if __name__=="__main__":
         plt.savefig(args.output_heterogeneity)
 
 
-    for m, N in zip(['o', 'd'], N_vals):
-        for T in period:
-            plt.figure()
+    plt.figure()
+    for T in period:
+        for m, N in zip(['o', 'd'], N_vals):
             D_array = data.loc[data.N==N].D.unique()
             print(D_array[0])
             for i, (ir, dr, p) in enumerate(product(interaction_radius, density_reg,
-                                                    subsampling)):
+                                                    [1])):
                 try:
-                    plt.errorbar(D_array*N/Lx/Ly, diffusion_mean[ir, dr, :, N, T, p]/D_array,
+                    plt.errorbar(D_array*T/Lx/Ly, diffusion_mean[ir, dr, :, N, T, p]/D_array,
                                         diffusion_std[ir, dr, :, N, T, p]/D_array/np.sqrt(10), marker=m,
                         label=f'r={ir}, a={dr} N={N}, T={T} p={p}', ls=ls[i%len(ls)], c=f"C{i//len(ls)}")
                 except:
                     pass
 
-            plt.plot(N*D_array/Lx/Ly, np.ones_like(D_array), c='k')
-            plt.legend()
-            plt.xlabel('true N*D')
-            plt.xlabel('estimated N*D')
-            plt.yscale('log')
-            plt.xscale('log')
+        #plt.plot(D_array/Lx/Ly, np.ones_like(D_array), c='k')
+        plt.xlabel('true N*D')
+        plt.xlabel('estimated N*D')
+        plt.yscale('log')
+        plt.xscale('log')
[…]

### 2023-09-10 src/plot_inflated_diffusion.py

@@ -8,7 +8,7 @@ def free_diffusion(D_array, N, linear_bins=5):
     n_iter = 10
     res = []
     for D in D_array:
-        res.append([get_granularity(N, D, bins=linear_bins) for i in range(n_iter)])
+        res.append([get_granularity(N, D, bins=linear_bins)[0] for i in range(n_iter)])
 
     res = np.array(res)
     return res.mean(axis=1)

### 2023-09-10 src/habitat_shifts.py
extend tmrca to habitat shifts

@@ -21,15 +21,16 @@ def diffusion_in_changing_habitats(D, interaction_radius, density_reg, N, subsam
     density_variation = []
     D_est = []
     zscores = []
+    Tmrca = []
 
-    for t in range((n_iter+2)*N):
+    for t in range((n_iter+5)*N):
         target_density = generate_target_density(N, Lx, Ly, period=period, wave_length=wave_length)
         terminal_nodes = evolve(terminal_nodes, t, Lx=Lx, Ly=Ly, interaction_radius=interaction_radius,
                                 density_reg=density_reg, D=D, target_density=target_density)
         if len(terminal_nodes)<10:
             print("population nearly extinct")
             continue
-        if t%(N//5)==0 and t>2*N: # take samples after burnin every 5 Tc
+        if t%(N//5)==0 and t>5*N: # take samples after burnin every Tc//5
             clean_tree(tree)
             H, bx, by = get_2d_hist(terminal_nodes, Lx, Ly, linear_bins)
             density_variation.append(np.std(H)/N*np.prod(H.shape))
@@ -38,11 +39,15 @@ def diffusion_in_changing_habitats(D, interaction_radius, density_reg, N, subsam
                 D_res = estimate_diffusion(tree)
                 estimate_ancestral_positions(tree, D)
                 z = collect_zscore(tree)
+                if len(tree['clades'])==1:
+                    Tmrca.append(t-tree['clades'][0]['time'])
+                else:
+                    Tmrca.append(t)
                 D_est.extend([D_res['Dx_total'], D_res['Dy_total']])
                 zscores.extend([np.mean(z.loc[z.nonterminal, 'zx']**2),
                                 np.mean(z.loc[z.nonterminal, 'zy']**2)])
 
-    return {"density_variation": density_variation, "D_est": D_est, 'zscores':zscores}
+    return {"density_variation": density_variation, "D_est": D_est, 'zscores':zscores, "Tmrca":Tmrca}
 
 def test_density(Lx, Ly, period, wave_length):
     d = generate_target_density(1, Lx, Ly, period, wave_length)
@@ -91,7 +96,8 @@ if __name__=="__main__":
                       "N": N, "n": len(res["D_est"]), "period": args.period, "subsampling": args.subsampling,
                       "D":D, "meanD": tmpD, "stdD": tmpStdD,
                       "meanZsq": tmpZ, "stdZsq": tmpStdZ, 'observations': nobs,
[…]

### 2023-09-10 src/inflated_diffusion.py

@@ -60,7 +60,7 @@ if __name__=="__main__":
     res_density_mean = {}
     D_est = []
     D_array_dens = np.logspace(-1,0,1)*Lx*Ly*2/N
-    n_iter = 1
+    n_iter = 30
     linear_bins=5
     interaction_radius, density_reg = args.interaction_radius, args.density_reg
     nsub = 1 if args.subsampling>0.9 else 5

### 2023-09-10 src/inflated_diffusion.py
extend tmrca to habitat shifts

@@ -59,7 +59,7 @@ if __name__=="__main__":
     res_density = {}
     res_density_mean = {}
     D_est = []
-    D_array_dens = np.logspace(-1,0,1)*Lx*Ly*2/N
+    D_array_dens = np.logspace(-3,0,21)*Lx*Ly*2/N
     n_iter = 30
     linear_bins=5
     interaction_radius, density_reg = args.interaction_radius, args.density_reg

### 2023-09-11 manuscript/phylogeo.tex
updated ms and snake

@@
 It is, however, fairly straightforward to include this in simulation and these simulations can be used to investigate the nature of phylogeographic inferences when some of the assumptions are violated.
+We calculate the population density using a Gaussian kernel with interaction radius $r$
+\begin{equation}
+    \phi(\rvec) = \sum_{i\in N} \frac{e^{-|\rvec_i - \rvec|^2/2r^2}}{N\sqrt{2\pi r^2}}
+\end{equation}
+where the sum is over all individuals $i$ in the population of size $N$.
+As soon as the interaction radius $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}.
+Diffusion constants estimated from the population remain accurate, but with decreasing diffusion, the population fragments into smaller than smaller subpopulations that don't interact.
+This manifests in an increasing time to the MRCA.
+
+\begin{figure}
+    \includegraphics[width=0.5\textwidth]{figures/density_reg_heterogeneity.pdf}
+    \includegraphics[width=0.5\textwidth]{figures/density_reg_tmrca.pdf}
+    \caption{\label{fig:density_reg} {\bf Density regulation reduces fluctuations in density regulation, but results in population fragmentation at low diffusion constants.}}
+\end{figure}
 

### 2023-09-11 src/plot_inflated_diffusion.py

@@ -20,6 +20,7 @@ if __name__=="__main__":
     parser.add_argument('--output-diffusion', type=str)
     parser.add_argument('--output-heterogeneity', type=str)
     parser.add_argument('--output-zscore', type=str)
+    parser.add_argument('--output-tmrca', type=str)
     args = parser.parse_args()
     data = pd.read_csv(args.data)
 
@@ -30,13 +31,15 @@ if __name__=="__main__":
     density_variation = data.groupby(['interaction_radius', 'density_reg', 'D'])['density_variation'].mean()
     diffusion_mean = data.groupby(['interaction_radius', 'density_reg', 'D'])['meanD'].mean()
     diffusion_std = data.groupby(['interaction_radius', 'density_reg', 'D'])['stdD'].mean()
+    tmrca_mean = data.groupby(['interaction_radius', 'density_reg', 'D'])['meanTmrca'].mean()
+    tmrca_std = data.groupby(['interaction_radius', 'density_reg', 'D'])['stdTmrca'].mean()
     z_mean = data.groupby(['interaction_radius', 'density_reg', 'D'])['meanZsq'].mean()
     z_std = data.groupby(['interaction_radius', 'density_reg', 'D'])['stdZsq'].mean()
     D_array = data.D.unique()
     interaction_radius = data.interaction_radius.unique()
     density_reg = data.density_reg.unique()
     N_vals = [500] #data.N.unique()
-
+    nobs = 10
     area = np.minimum(1,D_array*np.pi**2)
 
     ls = ['-', '-.', "--", ":"] #, "-", "--"]
@@ -61,7 +64,7 @@ if __name__=="__main__":
     for N in N_vals:
         for i, (ir, dr) in enumerate(product(interaction_radius[1:], density_reg[:-2])):
             plt.errorbar(D_array*N, diffusion_mean[ir, dr, :]/D_array,
-                                    diffusion_mean[ir, dr, :]/D_array/np.sqrt(10),
+                                    diffusion_std[ir, dr, :]/D_array/np.sqrt(nobs),
                  label=f'r={ir}, a={dr}', ls=ls[i%len(ls)], c=f"C{i//len(ls)}")
 
     plt.legend()
@@ -77,7 +80,7 @@ if __name__=="__main__":
     for N in N_vals:
         for i, (ir, dr) in enumerate(product(interaction_radius[1:], density_reg[:-2])):
[…]

### 2023-09-16 manuscript/phylogeo.tex
work on text and figures

@@
 At $D=0.1L^2/N$, the population distribution is patchy with large areas un-inhabited (Fig.~\ref{fig:heterogeneity} left).
-This heterogeneity increases substantially for even lower $D$, when the population is essentially a cloud of width $\sqrt{4DN}$.
+This heterogeneity increases substantially for even lower $D$, when the population is essentially a cloud of width $\sqrt{4DN}\ll L$.
 
@@
 
-Biological populations spread into areas that support them.
-The simplest model of this process are FKPP waves that model the population density $\phi(x,t)$ as
+Biological populations spread into areas that support them and will quickly populate fertile areas that are below their carrying capacity.
+The simplest model of this process are FKPP waves that model the population density $\phi(x,t)$ as \citep{fisher_wave_1937,KPP1937,hallatschek_life_2010}
 \begin{equation}
@@
 where $\phi_0$ is the carrying capacity and $\alpha$ is the population growth rate at zero density.
-This equation admits solutions with a traveling front, that the population "invades" the empty territory with velocity $v = \sqrt{D \alpha}$.
+This equation admits solutions with a traveling front, that the population ``invades" the empty territory with velocity $v = 2\sqrt{D \alpha}$.
 Note that the model itself does not have an intrinsic velocity.
-The speed at which the population invades space is given as compound of diffusion and accelerated growth in regions of low density.
+The speed at which the population invades space is given as compound of diffusion and accelerated growth in regions of low density and only grows like the square root of diffusion constant.
+It is thus not clear a priori how estimates of $D$ will be affected by density regulation, even in a constant habitat.
 
-This dependence of growth rate (and thus tree structure) on density is one key element that phylogeographic models ignore because it is very difficult to capture mathematically.
-It is, however, fairly straightforward to include this in simulation and these simulations can be used to investigate the nature of phylogeographic inferences when some of the assumptions are violated.
+Dependence of growth rate (and thus tree structure) on density is one key element that phylogeographic models ignore because it is very difficult to capture mathematically.
+It is, however, fairly straightforward to include this in simulations and these simulations can be used to investigate the nature of phylogeographic inferences when some of the assumptions are violated.
 We calculate the population density using a Gaussian kernel with interaction radius $r$
@@
 As soon as the interaction radius $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}.
-Diffusion constants estimated from the population remain accurate, but with decreasing diffusion, the population fragments into smaller than smaller subpopulations that don't interact.
-This manifests in an increasing time to the MRCA.
+Diffusion constants estimated from the population remain accurate with moderate overestimation of small diffusion constants.
+But with decreasing diffusion, the population fragments into smaller than smaller subpopulations that don't interact or coalesce.
+This manifests in an increasing time to the MRCA, see Fig.~\ref{fig:density_reg}.
+Spatial location and the coalescence process are thus strongly coupled, in conflict with model assumptions and estimates of population processes are likely biased.
 
 \begin{figure}
-    \includegraphics[width=0.5\textwidth]{figures/density_reg_heterogeneity.pdf}
-    \includegraphics[width=0.5\textwidth]{figures/density_reg_tmrca.pdf}
+    \includegraphics[width=0.48\textwidth]{figures/density_reg_heterogeneity.pdf}
+    \includegraphics[width=0.48\textwidth]{figures/density_reg_tmrca.pdf}
     \caption{\label{fig:density_reg} {\bf Density regulation reduces fluctuations in density regulation, but results in population fragmentation at low diffusion constants.}}
@@
 
+The above example assumed a habitat that is constant over times comparable to the time to the most recent common ancestor of the population.
+
 \end{document}

### 2023-09-16 src/inflated_diffusion.py

@@ -16,13 +16,13 @@ def estimate_inflated_diffusion(D, interaction_radius, density_reg, N, subsampli
     D_est = []
     zscores = []
     Tmrca = []
-    for t in range((n_iter+5)*N):
+    for t in range((n_iter+10)*N):
         terminal_nodes = evolve(terminal_nodes, t, Lx=Lx, Ly=Ly, interaction_radius=interaction_radius,
                                 density_reg=density_reg, D=D, target_density=N)
         if len(terminal_nodes)<10:
             print("population nearly extinct")
             continue
-        if t%(N//5)==0 and t>5*N: # take samples after burnin every Tc//5
+        if t%(N//5)==0 and t>10*N: # take samples after burnin every Tc//5
             clean_tree(tree)
             H, bx, by = get_2d_hist(terminal_nodes, Lx, Ly, linear_bins)
             density_variation.append(np.std(H)/N*np.prod(H.shape))
@@ -60,7 +60,7 @@ if __name__=="__main__":
     res_density_mean = {}
     D_est = []
     D_array_dens = np.logspace(-3,0,21)*Lx*Ly*2/N
-    n_iter = 30
+    n_iter = 50
     linear_bins=5
     interaction_radius, density_reg = args.interaction_radius, args.density_reg
     nsub = 1 if args.subsampling>0.9 else 5

### 2023-09-16 src/plot_habitat_diffusion.py

@@ -20,6 +20,7 @@ if __name__=="__main__":
     parser.add_argument('--output-diffusion', type=str)
     parser.add_argument('--output-heterogeneity', type=str)
     parser.add_argument('--output-zscore', type=str)
+    parser.add_argument('--output-tmrca', type=str)
     args = parser.parse_args()
     data = pd.read_csv(args.data)
     linear_bins=5
@@ -31,12 +32,14 @@ if __name__=="__main__":
     diffusion_std = data.groupby(['interaction_radius', 'density_reg', 'D', 'N', 'period', 'subsampling'])['stdD'].mean()
     z_mean = data.groupby(['interaction_radius', 'density_reg', 'D', 'N', 'period', 'subsampling'])['meanZsq'].mean()
     z_std = data.groupby(['interaction_radius', 'density_reg', 'D', 'N', 'period', 'subsampling'])['stdZsq'].mean()
+    tmrca_mean = data.groupby(['interaction_radius', 'density_reg', 'D', 'N', 'period', 'subsampling'])['meanTmrca'].mean()
+    tmrca_std = data.groupby(['interaction_radius', 'density_reg', 'D', 'N', 'period', 'subsampling'])['stdTmrca'].mean()
     interaction_radius = data.interaction_radius.unique()
     period = data.period.unique()
     subsampling = data.subsampling.unique()
     density_reg = data.density_reg.unique()
     N_vals = data.N.unique()
-
+    nobs = 10
 
     ls = ['-', '-.', "--"] #, ":", '-'] #, "-", "--"]
     plt.figure()
@@ -62,8 +65,8 @@ if __name__=="__main__":
         plt.savefig(args.output_heterogeneity)
 
 
-    plt.figure()
     for T in period:
+        plt.figure()
         for m, N in zip(['o', 'd'], N_vals):
             D_array = data.loc[data.N==N].D.unique()
             print(D_array[0])
@@ -71,7 +74,7 @@ if __name__=="__main__":
                                                     [1])):
                 try:
                     plt.errorbar(D_array*T/Lx/Ly, diffusion_mean[ir, dr, :, N, T, p]/D_array,
-                                        diffusion_std[ir, dr, :, N, T, p]/D_array/np.sqrt(10), marker=m,
+                                        diffusion_std[ir, dr, :, N, T, p]/D_array/np.sqrt(nobs), marker=m,
[…]

### 2023-09-16 src/plot_inflated_diffusion.py

@@ -28,30 +28,33 @@ if __name__=="__main__":
     Lx, Ly = 1, 1
     nbins=linear_bins**2
 
-    density_variation = data.groupby(['interaction_radius', 'density_reg', 'D'])['density_variation'].mean()
-    diffusion_mean = data.groupby(['interaction_radius', 'density_reg', 'D'])['meanD'].mean()
-    diffusion_std = data.groupby(['interaction_radius', 'density_reg', 'D'])['stdD'].mean()
-    tmrca_mean = data.groupby(['interaction_radius', 'density_reg', 'D'])['meanTmrca'].mean()
-    tmrca_std = data.groupby(['interaction_radius', 'density_reg', 'D'])['stdTmrca'].mean()
-    z_mean = data.groupby(['interaction_radius', 'density_reg', 'D'])['meanZsq'].mean()
-    z_std = data.groupby(['interaction_radius', 'density_reg', 'D'])['stdZsq'].mean()
-    D_array = data.D.unique()
+    density_variation = data.groupby(['interaction_radius', 'density_reg', 'N', 'D'])['density_variation'].mean()
+    nobs = data.groupby(['interaction_radius', 'density_reg', 'N'])['n'].mean()/25
+    diffusion_mean = data.groupby(['interaction_radius', 'density_reg', 'N', 'D'])['meanD'].mean()
+    diffusion_std = data.groupby(['interaction_radius', 'density_reg', 'N', 'D'])['stdD'].mean()
+    tmrca_mean = data.groupby(['interaction_radius', 'density_reg', 'N', 'D'])['meanTmrca'].mean()
+    tmrca_std = data.groupby(['interaction_radius', 'density_reg', 'N', 'D'])['stdTmrca'].mean()
+    z_mean = data.groupby(['interaction_radius', 'density_reg', 'N', 'D'])['meanZsq'].mean()
+    z_std = data.groupby(['interaction_radius', 'density_reg', 'N', 'D'])['stdZsq'].mean()
     interaction_radius = data.interaction_radius.unique()
     density_reg = data.density_reg.unique()
-    N_vals = [500] #data.N.unique()
-    nobs = 10
-    area = np.minimum(1,D_array*np.pi**2)
+    N_vals = data.N.unique()
 
-    ls = ['-', '-.', "--", ":"] #, "-", "--"]
+    ir_to_plot = interaction_radius[1:]
[…]

### 2023-09-17 src/habitat_shifts.py
plotting

@@ -23,14 +23,14 @@ def diffusion_in_changing_habitats(D, interaction_radius, density_reg, N, subsam
     zscores = []
     Tmrca = []
 
-    for t in range((n_iter+5)*N):
+    for t in range((n_iter+10)*N):
         target_density = generate_target_density(N, Lx, Ly, period=period, wave_length=wave_length)
         terminal_nodes = evolve(terminal_nodes, t, Lx=Lx, Ly=Ly, interaction_radius=interaction_radius,
                                 density_reg=density_reg, D=D, target_density=target_density)
         if len(terminal_nodes)<10:
             print("population nearly extinct")
             continue
-        if t%(N//5)==0 and t>5*N: # take samples after burnin every Tc//5
+        if t%(N//5)==0 and t>10*N: # take samples after burnin every Tc//5
             clean_tree(tree)
             H, bx, by = get_2d_hist(terminal_nodes, Lx, Ly, linear_bins)
             density_variation.append(np.std(H)/N*np.prod(H.shape))
@@ -79,7 +79,7 @@ if __name__=="__main__":
     res_density_mean = {}
     D_est = []
     D_array_dens = np.logspace(-3,0,21)*Lx*Ly*2/N
-    n_iter = 30
+    n_iter = 50
     linear_bins=5
     interaction_radius, density_reg = args.interaction_radius, args.density_reg
     print(f"{interaction_radius=:1.3f}, {density_reg=:1.3f}")

### 2023-09-17 src/plot_habitat_diffusion.py

@@ -1,17 +1,20 @@
 import numpy as np
 import matplotlib.pyplot as plt
 import pandas as pd
-from heterogeneity import get_granularity
+from plot_inflated_diffusion import free_diffusion
 from itertools import product
+from habitat_shifts import generate_target_density
 
-def free_diffusion(D_array, N, linear_bins=5):
-    n_iter = 10
-    res = []
-    for D in D_array:
-        res.append([get_granularity(N, D, bins=linear_bins)[0] for i in range(n_iter)])
-
-    res = np.array(res)
-    return res.mean(axis=1)
+def make_figure(fname=None):
+    fig, axs = plt.subplots(1,10, sharex=True, sharey=True, figsize=(15,2))
+    f = generate_target_density(1, 1, 1, period=9, wave_length=1.0)
+    for i, ax in enumerate(axs.flatten()):
+        grid = np.meshgrid(np.linspace(0,1,30), np.linspace(0,1,30))
+        ax.matshow(f(grid[0], grid[1], i))
+        ax.set_axis_off()
+    plt.tight_layout()
+    if fname:
+        plt.savefig(fname)
 
 if __name__=="__main__":
     import argparse
@@ -21,6 +24,7 @@ if __name__=="__main__":
     parser.add_argument('--output-heterogeneity', type=str)
     parser.add_argument('--output-zscore', type=str)
     parser.add_argument('--output-tmrca', type=str)
+    parser.add_argument('--illustration', type=str)
     args = parser.parse_args()
     data = pd.read_csv(args.data)
     linear_bins=5
@@ -28,6 +32,7 @@ if __name__=="__main__":
     nbins=linear_bins**2
 
     density_variation = data.groupby(['interaction_radius', 'density_reg', 'D', 'N', 'period', 'subsampling'])['density_variation'].mean()
+    nobs = data.groupby(['interaction_radius', 'density_reg', 'N', 'period', 'subsampling'])['n'].mean()/25
     diffusion_mean = data.groupby(['interaction_radius', 'density_reg', 'D', 'N', 'period', 'subsampling'])['meanD'].mean()
     diffusion_std = data.groupby(['interaction_radius', 'density_reg', 'D', 'N', 'period', 'subsampling'])['stdD'].mean()
     z_mean = data.groupby(['interaction_radius', 'density_reg', 'D', 'N', 'period', 'subsampling'])['meanZsq'].mean()
[…]

### 2023-09-19 manuscript/phylogeo.tex
plotting
add Z for time windows

@@
 \newcommand{\rvec}{\mathbf{r}}
-
-\bibliography{bib}
 
@@
 The above example assumed a habitat that is constant over times comparable to the time to the most recent common ancestor of the population.
+If instead the habitat shifts on shorter time scales, the population has to move and such population movement will skew estimates of dispersal.
+We simulated such changing habitats by assuming the carrying capacity of the habitat varies periodically in time and space
+\begin{equation}
+    \phi_0(\rvec, t) = N\left(1 + \sin(2\pi (r_x/L_x + t/T)\cos(2\pi (r_y/L_y + t/2T)))
+\end{equation}
+This results in moving peaks and troughs in carrying capacity (and thus population density) as illustrated in
+
+\begin{figure}
+    \includegraphics[width=\textwidth]{figures/habitats.png}
+    \caption{\label{fig:moving_habitats}}
+\end{figure}
+
+\bibliography{bib}
 

### 2023-09-19 src/habitat_shifts.py

@@ -31,6 +31,7 @@ def diffusion_in_changing_habitats(D, interaction_radius, density_reg, N, subsam
             print("population nearly extinct")
             continue
         if t%(N//5)==0 and t>10*N: # take samples after burnin every Tc//5
+            tbins = sorted([0] + [t - i*N/10 for i in range(4)])
             clean_tree(tree)
             H, bx, by = get_2d_hist(terminal_nodes, Lx, Ly, linear_bins)
             density_variation.append(np.std(H)/N*np.prod(H.shape))
@@ -44,8 +45,8 @@ def diffusion_in_changing_habitats(D, interaction_radius, density_reg, N, subsam
                 else:
                     Tmrca.append(t)
                 D_est.extend([D_res['Dx_total'], D_res['Dy_total']])
-                zscores.extend([np.mean(z.loc[z.nonterminal, 'zx']**2),
-                                np.mean(z.loc[z.nonterminal, 'zy']**2)])
+                zscores.extend([[np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'zx']**2) for i in range(len(tbins)-1)],
+                                [np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'zy']**2) for i in range(len(tbins)-1)]])
 
     return {"density_variation": density_variation, "D_est": D_est, 'zscores':zscores, "Tmrca":Tmrca}
 
@@ -89,8 +90,8 @@ if __name__=="__main__":
                                           Lx=Lx, Ly=Ly, linear_bins=linear_bins, n_iter=n_iter, period=args.period)
         tmpD = np.mean(res["D_est"], axis=0)
         tmpStdD = np.std(res["D_est"], axis=0)
-        tmpZ = np.mean(res["zscores"], axis=0)
-        tmpStdZ = np.std(res["zscores"], axis=0)
+        tmpZ =    f"[{' '.join(str(x) for x in np.ma.mean(np.ma.masked_invalid(res['zscores']), axis=0).filled(fill_value=np.nan))}]"
+        tmpStdZ = f"[{' '.join(str(x) for x in np.ma.std(np.ma.masked_invalid(res['zscores']), axis=0).filled(fill_value=np.nan))}]"
         nobs = len(res["D_est"])
         D_est.append({"interaction_radius":interaction_radius, "density_reg": density_reg,
                       "N": N, "n": len(res["D_est"]), "period": args.period, "subsampling": args.subsampling,

### 2023-09-19 src/plot_habitat_diffusion.py

@@ -5,6 +5,11 @@ from plot_inflated_diffusion import free_diffusion
 from itertools import product
 from habitat_shifts import generate_target_density
 
+plt.rcParams.update({
+    "text.usetex": True,
+    "font.family": "Helvetica"
+})
+
 def make_figure(fname=None):
     fig, axs = plt.subplots(1,10, sharex=True, sharey=True, figsize=(15,2))
     f = generate_target_density(1, 1, 1, period=9, wave_length=1.0)
@@ -48,6 +53,7 @@ if __name__=="__main__":
     make_figure(fname = args.illustration)
 
     ls = ['-', ':', "--", ".-"][len(interaction_radius)]
+    markers = ['o', '<', '>', 's', 'd', '^', 'v']
     plt.figure()
     for N in N_vals:
         D_array = data.loc[data.N==N].D.unique()
@@ -64,7 +70,7 @@ if __name__=="__main__":
         plt.plot(D_array*N/Lx/Ly, free_diffusion_heterogeneity, label='diffusion', c='k')
         plt.plot(D_array*N/Lx/Ly, np.ones_like(D_array)*np.sqrt(nbins/N), label='well mixed limit', c='k', ls='--')
     plt.xscale('log')
-    plt.xlabel('diffusion constant')
+    plt.xlabel('diffusion constant $D$')
     plt.ylabel('heterogeneity')
     plt.legend()
     if args.output_heterogeneity:
@@ -73,24 +79,28 @@ if __name__=="__main__":
 
     plt.figure()
     #plt.title(f"{T}")
-    for T in period:
+    for ti,T in enumerate(period):
         for m, N in zip(['o', '<'], N_vals):
             D_array = data.loc[data.N==N].D.unique()
             for i, (ir, dr, p) in enumerate(product(interaction_radius, density_reg,
                                                     [1])):
                 try:
-                    plt.errorbar(np.sqrt(D_array*dr)*T/Lx, diffusion_mean[ir, dr, :, N, T, p]/D_array,
-                                        diffusion_std[ir, dr, :, N, T, p]/D_array/np.sqrt(nobs[ir, dr, N, T, p]), marker=m,
-                        label=f'r={ir}, a={dr} N={N}, T={T} p={p}', ls=ls[i%len(ls)], c=f"C{i//len(ls)}")
[…]

### 2023-09-19 src/plot_inflated_diffusion.py

@@ -40,8 +40,8 @@ if __name__=="__main__":
     density_reg = data.density_reg.unique()
     N_vals = data.N.unique()
 
-    ir_to_plot = interaction_radius[1:]
-    density_reg_to_plot = density_reg[:-2:2]
+    ir_to_plot = interaction_radius[:]
+    density_reg_to_plot = density_reg[:-2]
 
     ls = ['-', '-.', "--", ":"][:len(density_reg_to_plot)]
     plt.figure()

### 2023-09-19 src/inflated_diffusion.py
add Z for inflation

@@ -23,6 +23,7 @@ def estimate_inflated_diffusion(D, interaction_radius, density_reg, N, subsampli
             print("population nearly extinct")
             continue
         if t%(N//5)==0 and t>10*N: # take samples after burnin every Tc//5
+            tbins = sorted([0] + [t - i*N/10 for i in range(4)])
             clean_tree(tree)
             H, bx, by = get_2d_hist(terminal_nodes, Lx, Ly, linear_bins)
             density_variation.append(np.std(H)/N*np.prod(H.shape))
@@ -36,8 +37,8 @@ def estimate_inflated_diffusion(D, interaction_radius, density_reg, N, subsampli
                 else:
                     Tmrca.append(t)
                 D_est.extend([D_res['Dx_total'], D_res['Dy_total']])
-                zscores.extend([np.mean(z.loc[z.nonterminal, 'zx']**2),
-                                np.mean(z.loc[z.nonterminal, 'zy']**2)])
+                zscores.extend([[np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'zx']**2) for i in range(len(tbins)-1)],
+                                [np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'zy']**2) for i in range(len(tbins)-1)]])
 
     return {"density_variation": density_variation, "D_est": D_est, "zscores": zscores, "Tmrca":Tmrca}
 
@@ -71,8 +72,8 @@ if __name__=="__main__":
                                           Lx=Lx, Ly=Ly, linear_bins=linear_bins, n_iter=n_iter, n_subsamples=nsub)
         tmpD = np.mean(res["D_est"], axis=0)
         tmpStdD = np.std(res["D_est"], axis=0)
-        tmpZ = np.mean(res["zscores"], axis=0)
-        tmpStdZ = np.std(res["zscores"], axis=0)
+        tmpZ =    f"[{' '.join(str(x) for x in np.ma.mean(np.ma.masked_invalid(res['zscores']), axis=0).filled(fill_value=np.nan))}]"
+        tmpStdZ = f"[{' '.join(str(x) for x in np.ma.std(np.ma.masked_invalid(res['zscores']), axis=0).filled(fill_value=np.nan))}]"
         nobs = len(res["D_est"])
         D_est.append({"interaction_radius":interaction_radius, "density_reg": density_reg,
                       "N": N, "n": len(res["D_est"]), "subsampling": args.subsampling,

### 2024-03-24 manuscript/phylogeo.tex
big commit

@@
 As organisms replicate, they also disperse in space, potentially exploring new habitats.
-The reconstruction of ancestral locations and past migrations from sampling locations of extant individuals, typically in combination with genome sequence information to infer the phylogeny, are known as phylogeography.
+The reconstruction of ancestral locations and past migrations from sampling locations of extant individuals, typically in combination with genome sequence information to infer the phylogeny, is known as phylogeography.
 Phylogeographic methods are implemented in popular evolutionary analysis software such as BEAST \citep{pybus_unifying_2012}.
-The underlying model of dispersal is typically diffusive, that is the probability of sampling a descendent at position $\rvec_c$ after time $t$ when the parent was at position $\rvec_p$ is given by
+The underlying model of dispersal is typically diffusive, that is the probability of sampling a descendant at position $\rvec_c$ after time $t$ when the parent was at position $\rvec_p$ is given by
 \begin{equation}
@@
 \end{equation}
-where $D$ is a diffusion constant with dimensions $\mathrm{length}^2/\mathrm{time}$ and we have assumed a two dimensional space.
-Each branch in the phylogeny is adorned with such a diffusive propagator.
+where $D$ is a diffusion constant with dimensions $\mathrm{length}^2/\mathrm{time}$, and we have assumed a two-dimensional space.
+Each branch in the phylogeny is decorated with such a diffusive propagator.
 Such a diffusive process is the simplest and most natural choice in absence of directed motion or long range migration.
-It is also mathematically convenient as unknown ancestral locations can be integrated in closed form and the marginal distribution of each position can be calculated exactly for a fixed tree.
-However, it has one major shortcoming: tree topology and branching rates of the tree are independent of the spatial process.
+It is also mathematically convenient as unknown ancestral locations can be integrated in closed form and the marginal distribution of each position can be calculated exactly for a fixed tree along with the likelihood.
+The total likelihood of the configuration is then assumed to be the product of the spatial likelihood and the phylogenetic likelihood, i.e.
+tree topology and branching rates of the tree are assumed to be independent of the spatial process.
 However, these two interact in important ways, for example when a population expands into an uninhabited fertile habitat or a disease spreads in a susceptible population.
-
-At the same time, information in two dimensional sampling locations is low compared to that from high-dimensional sequence space and processes not captured by the diffusive model can potentially result in wrong inferences.
-The purpose of this note is to explore how sensitive phylogeographic inferences are to model assumptions and what quantities can be robustly estimated.
-My focus here is only on the ability and robustness to estimate dispersal parameter and ancestral locations and I will assume perfect knowledge of the tree and the sampling locations.
-So no tree-sampling will be necessary.
 
@@
 Results can be estimates of the ancestral location of a particular groups of samples, or summary statistics of the dispersal process.
-Popular summaries include empirical estimates of the diffusion constant and a so called \emph{dispersal rate}.
-These are calculated from the displacements along branch $i$ with parent $p_i$ given by $d_i = |\rvec_{p_i} - \rvec_{i}|$ and the elapsed time $\Delta t_i$ as
+Popular summaries include empirical estimates of the diffusion constant  \citep{pybus_unifying_2012} and a so-called \emph{dispersal velocity} \citep{dellicour_using_2017}.
+These are calculated from the estimated displacements along branch $i$ with parent $p_i$ given by $d_i = |\rvec_{p_i} - \rvec_{i}|$ and the elapsed time $\Delta t_i$ as
 \begin{equation}
@@
 Effectively $D_w$ divides the sum of all observed squared displacements by the total time elapsed on the tree to obtain and estimate of $D$.
-The dispersal rate estimate compares the sum of absolute values of observed displacements to the total time, which has dimensions of a velocity.
+The dispersal velocity estimate compares the sum of absolute values of observed displacements to the total time, which has dimensions of a velocity.
+Alternative formulations use the mean of fractions $d_i^2/4t_i$ or $d_i/t_i$ instead of the ratio of sums, but these tend to be less robust.
+
+The purpose of this note is to explore how sensitive phylogeographic inferences are to model assumptions and what quantities can be robustly estimated.
+My focus here is only on the ability and robustness to estimate dispersal parameter and ancestral locations and I will assume perfect knowledge of the tree and the sampling locations.
+So no tree-sampling will be necessary.
 
 Fig.~\ref{fig:D_and_v} shows empirical estimates of $D_w$ and $v_w$ as defined above from simulated data using freely diffusing locations along Kingman or Yule trees.
-The diffusion constant $D$ can be estimated robustly from these data, but $v_w$ is not a well defined quantity.
-Values of $v_w$ depends strongly on the sample size and there is not ground truth to compare to as there is no parameter with dimensions of a velocity in the model.
+The diffusion constant $D$ can be estimated robustly from these data, but $v_w$ is not a well-defined quantity.
+Values of $v_w$ depends strongly on the sample size and there is no ground truth to compare to as there is no parameter with dimensions of a velocity in the model.
 That $v_w$ is ill-defined is not surprising: By changing the sample size, the average length of branches changes.
 Diffusion along each branch results in a displacement of $d_i \sim \sqrt{4Dt_i}$ such that a quantity that depends on $\Delta d_i / t_i \sim 1/t_i^{1/2}$ will change with the time $t_i$ that elapsed along the branch.
-Values of $v_w$ are meaningless in this model.
+If the underlying dispersal process is diffusive, estimates of $v_w$ are not meaningful.
 
-\begin{figure}
+\begin{figure}[tb]
+    \includegraphics*[width=0.48\textwidth]{figures/yule_total_dispersal.png}
     \includegraphics*[width=0.48\textwidth]{figures/kingman_total_dispersal.png}
-    \includegraphics*[width=0.48\textwidth]{figures/yule_total_dispersal.png}
-    \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal rate'' for different sample sizes.}
-    While diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ is not a well defined quantity.
+    \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
+    While diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ is not a well-defined quantity.
 Its numerical value depends strongly on sample size and there is no ground truth to compare to.
-This sample size dependence is stronger for Yule trees (right) than Kingman trees (left).
+This sample size dependence is present for Yule trees (left) and Kingman trees (right).
 }
@@
 
-\section*{Is the diffusion model consistent?}
-Phylogeographic assume that the spatial diffusion process is independent of the branching pattern of the tree and ignores coupling between the two that might arise from density dependent growth rates.
-If one simulates a diffusive spatial process in a square of length $L$, the population density shows strong fluctuations whenever the diffusion constant $D<L^2/N$, see Fig.~\ref{fig:heterogeneity}.
-$N$ here plays the role of the time to the MRCA of the population.
-At $D=0.1L^2/N$, the population distribution is patchy with large areas un-inhabited (Fig.~\ref{fig:heterogeneity} left).
-This heterogeneity increases substantially for even lower $D$, when the population is essentially a cloud of width $\sqrt{4DN}\ll L$.
+\section*{How does density regulation affect estimates phylogeographic estimates?}
+Phylogeographic inference typically assumes that the spatial diffusion process is independent of the branching pattern of the tree and ignores coupling between the two that might arise from density dependent growth rates.
+If one simulates a diffusive spatial process in a square of length $L$, the population density shows strong fluctuations whenever the diffusion constant $D<L^2/T_c$, see Fig.~\ref{fig:heterogeneity} where $T_c$ is the coalescent times scale of the population.
+In a well-mixed neutral model, $T_c$ is equal to the population size $N$.
 
-\begin{figure}
+For $D<L^2/2T_c$, the time to the most recent common ancestor $T_{MRCA}$ is too small for the population to spread evenly in the habitat.
+This is illustrated in Fig.~\ref{fig:heterogeneity}A that shows the population distribution across the habitat for $D=0.1L^2/T_c$. 
+The population distribution is patchy with large uninhabited areas.
+This heterogeneity increases substantially for even lower $D$, when the population is essentially a cloud of width $\sqrt{4DT_c}\ll L$.
+Fig.~\ref{fig:heterogeneity}B quantifies this heterogeneity as the standard deviation of population density in two-dimensional bins of different size  relative to the expectation in the case of a spatially well-mixed population.
+
+\begin{figure}[tb]
     \includegraphics*[width=\textwidth]{figures/heterogeneity_free_diffusion.pdf}
     \caption[short]{\label{fig:heterogeneity}{\bf Population density variation in free diffusion models.}
-    When spatial motion is independent of population density, the realized population density in the population will be much larger than biologically plausible.
+    When spatial motion is independent of population density and coalesce is sufficiently rapid, the realized population density fluctuates strongly across the habitat.
 The left panel shows the population density for the case $D = 0.1 L^2/N$ in linear bins of size $L/20$.
@@
 \begin{equation}
-    \frac{\partial \phi(x,t)}{\partial t} = D\frac{\partial^2 \phi(x,t)}{\partial x^2} + \alpha phi(x,t)(1-\phi(x,t)/\phi_0)
+    \frac{\partial \phi(x,t)}{\partial t} = D\frac{\partial^2 \phi(x,t)}{\partial x^2} + \alpha \phi(x,t)(1-\phi(x,t)/\phi_0)
 \end{equation}
@@
 \begin{equation}
-    \phi_0(\rvec, t) = N\left(1 + \sin(2\pi (r_x/L_x + t/T)\cos(2\pi (r_y/L_y + t/2T)))
+    \phi_0(\rvec, t) = N\left(1 + \sin(2\pi (r_x/L_x + t/T)\cos(2\pi (r_y/L_y + t/2T))) \right)
 \end{equation}

### 2024-03-24 src/heterogeneity.py

@@ -25,7 +25,7 @@ if __name__=="__main__":
     g, H = get_granularity(ntips, 0.1, bins=20)
     axs[0].matshow(H)
 
-    for line_style, linear_bins in zip(['-', '--', '-.'], [3,5,10]):
+    for line_style, linear_bins in zip(['-', '--', '-.'], [3,10,30]):
         nbins=linear_bins**2
         res = []
         for D in D_array:
@@ -33,16 +33,16 @@ if __name__=="__main__":
 
         res = np.array(res)
         res_mean = res.mean(axis=1)
-        axs[1].plot(D_array, res_mean, label=f'free diffusion, dx=L/{linear_bins}', c='C0', ls=line_style)
-        axs[1].plot(D_array, np.ones_like(D_array)*np.sqrt(nbins/ntips), label=f'well mixed limit dx=L/{linear_bins}', c='C1', ls=line_style)
+        axs[1].plot(D_array, res_mean/np.sqrt(nbins/ntips), label=f'bin size=L/{linear_bins}') #, c='C0', ls=line_style)
+        # axs[1].plot(D_array, np.ones_like(D_array)*np.sqrt(nbins/ntips), label=f'well mixed limit dx=L/{linear_bins}', c='C1', ls=line_style)
 
     # area = np.minimum(1,D_array*np.pi**2)
     # pred_heterogeneity = np.sqrt(1/area - 1)  # p*(1/p - 1)**2 + (1-p) = p*(1/p^2 - 2/p + 1) + (1-p) = 1/p - 2 + p -1 + p = 1/p - 1
 
     #plt.plot(D_array, pred_heterogeneity, label='well mixed limit')
     plt.xscale('log')
-    plt.xlabel('diffusion constant [habitat_size^2/T_c]')
-    plt.ylabel('heterogeneity')
+    plt.yscale('log')
+    plt.xlabel('diffusion constant [L^2/T_c]')
+    plt.ylabel('heterogeneity (relative to well-mixed case)')
     plt.legend()
-    plt.savefig('figures/heterogeneity_free_diffusion.pdf')
     plt.savefig('figures/heterogeneity_free_diffusion.pdf')

### 2024-03-24 src/inflated_diffusion.py

@@ -4,6 +4,7 @@ from heterogeneity import get_2d_hist
 from estimate_diffusion_from_tree import estimate_diffusion, estimate_ancestral_positions, collect_zscore
 
 
+
 def estimate_inflated_diffusion(D, interaction_radius, density_reg, N, subsampling=1.0,
                                 Lx=1, Ly=1, linear_bins=5, n_iter=10, n_subsamples=1):
     # set up tree and initial population uniformly in space

### 2024-03-24 src/plot_habitat_diffusion.py

@@ -1,7 +1,7 @@
 import numpy as np
 import matplotlib.pyplot as plt
 import pandas as pd
-from plot_inflated_diffusion import free_diffusion
+from plot_inflated_diffusion import free_diffusion, parse_data
 from itertools import product
 from habitat_shifts import generate_target_density
 
@@ -21,6 +21,7 @@ def make_figure(fname=None):
     if fname:
         plt.savefig(fname)
 
+
 if __name__=="__main__":
     import argparse
     parser = argparse.ArgumentParser()
@@ -36,14 +37,7 @@ if __name__=="__main__":
     Lx, Ly = 1, 1
     nbins=linear_bins**2
 
-    density_variation = data.groupby(['interaction_radius', 'density_reg', 'D', 'N', 'period', 'subsampling'])['density_variation'].mean()
-    nobs = data.groupby(['interaction_radius', 'density_reg', 'N', 'period', 'subsampling'])['n'].mean()/25
-    diffusion_mean = data.groupby(['interaction_radius', 'density_reg', 'D', 'N', 'period', 'subsampling'])['meanD'].mean()
-    diffusion_std = data.groupby(['interaction_radius', 'density_reg', 'D', 'N', 'period', 'subsampling'])['stdD'].mean()
-    z_mean = data.groupby(['interaction_radius', 'density_reg', 'D', 'N', 'period', 'subsampling'])['meanZsq'].mean()
-    z_std = data.groupby(['interaction_radius', 'density_reg', 'D', 'N', 'period', 'subsampling'])['stdZsq'].mean()
-    tmrca_mean = data.groupby(['interaction_radius', 'density_reg', 'D', 'N', 'period', 'subsampling'])['meanTmrca'].mean()
-    tmrca_std = data.groupby(['interaction_radius', 'density_reg', 'D', 'N', 'period', 'subsampling'])['stdTmrca'].mean()
+    res = parse_data(data, groupby=['interaction_radius', 'density_reg', 'N', 'period', 'subsampling'])
     interaction_radius = data.interaction_radius.unique()
     period = data.period.unique()
     subsampling = data.subsampling.unique()
@@ -61,7 +55,7 @@ if __name__=="__main__":
         for i, (ir, dr, T, p) in enumerate(product(interaction_radius, density_reg,
                                                    period, subsampling)):
             try:
[…]

### 2024-03-24 src/plot_inflated_diffusion.py

@@ -4,6 +4,30 @@ import pandas as pd
 from heterogeneity import get_granularity
 from itertools import product
 
+
+def parse_data(data, groupby=None):
+    density_variation = data.groupby(groupby + ['D'])['density_variation'].mean()
+    nobs = data.groupby(groupby)['n'].mean()/25
+    diffusion_mean = data.groupby(groupby + ['D'])['meanD'].mean()
+    diffusion_std = data.groupby(groupby + ['D'])['stdD'].mean()
+    tmrca_mean = data.groupby(groupby + ['D'])['meanTmrca'].mean()
+    tmrca_std = data.groupby(groupby + ['D'])['stdTmrca'].mean()
+
+    z_mean = {}
+    for g, d in data.groupby(groupby + ['D'])['meanZsq']:
+        z_mean[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
+
+    z_std = {}
+    for g, d in data.groupby(groupby + ['D'])['stdZsq']:
+        z_std[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
+
+    return {"density_variation": density_variation, "nobs":nobs,
+            "diffusion_mean":diffusion_mean, "diffusion_std":diffusion_std,
+            "tmrca_mean":tmrca_mean, "tmrca_std":tmrca_std,
+            "z_mean": pd.DataFrame(z_mean).T, "z_std":pd.DataFrame(z_std).T,
+            }
+
+
 def free_diffusion(D_array, N, linear_bins=5):
     n_iter = 10
     res = []
@@ -27,15 +51,8 @@ if __name__=="__main__":
     linear_bins=5
     Lx, Ly = 1, 1
     nbins=linear_bins**2
+    res = parse_data(data, groupby=['interaction_radius', 'density_reg', 'N'])
 
-    density_variation = data.groupby(['interaction_radius', 'density_reg', 'N', 'D'])['density_variation'].mean()
-    nobs = data.groupby(['interaction_radius', 'density_reg', 'N'])['n'].mean()/25
-    diffusion_mean = data.groupby(['interaction_radius', 'density_reg', 'N', 'D'])['meanD'].mean()
-    diffusion_std = data.groupby(['interaction_radius', 'density_reg', 'N', 'D'])['stdD'].mean()
[…]

### 2024-03-24 src/sample_size_dependence.py

@@ -4,34 +4,40 @@ from estimate_diffusion_from_tree import *
 from itertools import product
 
 if __name__=="__main__":
-    replicates = 10
+    replicates = 100
     results = {}
     fields = ['Dx_branch', 'Dy_branch', 'vx_branch', 'vy_branch', 'Dx_total', 'Dy_total', 'vx_total', 'vy_total']
-    for n, yule in product([10,30, 100, 300, 1000], [True, False]):
-        res = []
-        for i in range(replicates):
-            tree = random_tree(n, yule=yule)['tree']
-            add_positions(tree, 1)
-            D = estimate_diffusion(tree)
-            res.append(D)
+    for coal in ['yule', 'kingman']:
+        results[coal] = {}
+        for n in [10,30, 100, 300, 1000, 3000]:
+            res = []
+            for i in range(replicates):
+                tree = random_tree(n, yule=coal=='yule')['tree']
+                add_positions(tree, 1)
+                D = estimate_diffusion(tree)
+                res.append(D)
 
-        results[n] = {k: np.mean([x[k] for x in res]) for k in fields}
-        results[n].update({f"{k}_std": np.std([x[k] for x in res]) for k in fields})
+            results[coal][n] = {k: np.mean([x[k] for x in res]) for k in fields}
+            results[coal][n].update({f"{k}_std": np.std([x[k] for x in res]) for k in fields})
 
 
     import matplotlib.pyplot as plt
 
-    for suffix in ['_total', '_branch']:
+    for coal in ['yule', 'kingman']:
+        res = results[coal]
         plt.figure()
-        for p in ['Dx', 'vx']:
-            q = p+suffix
-            label = f"{q} [{'mean(dx_i^2)/mean(2t_i)' if 'D' in q else 'mean(dx_i)/mean(t_i)'}]" if suffix=='_total' else f"{q} [{'mean(dx_i^2/2t_i)' if 'D' in q else 'mean(dx_i/t_i)'}]"
-            plt.errorbar([n for n in results], [results[n][q] for n in results], [results[n][q+"_std"] for n in results], ls = '-', marker='o',
-                        label=label)
+        for suffix in ['_total', '_branch']:
+            for p in ['Dx', 'vx']:
+                q = p+suffix
+                label = f"{q} [{'mean(dx_i^2)/mean(2t_i)' if 'D' in q \
+                                else 'mean(dx_i)/mean(t_i)'}]" if suffix=='_total' \
[…]

### 2024-03-24 manuscript/phylogeo.tex
edits

@@
-\documentclass[aps,rmp, onecolumn]{revtex4}
+\documentclass[aps,rmp, twocolumn]{revtex4}
 %\documentclass[a4paper,10pt]{scrartcl}
@@
 Effectively $D_w$ divides the sum of all observed squared displacements by the total time elapsed on the tree to obtain and estimate of $D$.
-The dispersal velocity estimate compares the sum of absolute values of observed displacements to the total time, which has dimensions of a velocity.
-Alternative formulations use the mean of fractions $d_i^2/4t_i$ or $d_i/t_i$ instead of the ratio of sums, but these tend to be less robust.
+It is known as the `weighted' estimate of the diffusion constant.
+The weighted dispersal velocity estimate compares the sum of absolute values of observed displacements to the total time, which has dimensions of a velocity.
+Alternative formulations use the mean of fractions $D_b = |B|^{-1} \sum_{i\in B}d_i^2/4t_i$ or $v_b = |B|^{-1} \sum_{i\in B}d_i/t_i$ instead of the ratio of sums. \
+We refer to these here as `by branch' estimates. 
+The latter tend to be dominated by short branches and thus noisier.
 
@@
 So no tree-sampling will be necessary.
+To explore qualitatively different behaviors, I will consider Kingman and Yule tree ensembles.
+Kingman trees are the classical neutral model of a constant size population, while Yule trees capture features of growing populations (see Fig.~\ref{fig:illustration_tree}).
 
-Fig.~\ref{fig:D_and_v} shows empirical estimates of $D_w$ and $v_w$ as defined above from simulated data using freely diffusing locations along Kingman or Yule trees.
-The diffusion constant $D$ can be estimated robustly from these data, but $v_w$ is not a well-defined quantity.
-Values of $v_w$ depends strongly on the sample size and there is no ground truth to compare to as there is no parameter with dimensions of a velocity in the model.
-That $v_w$ is ill-defined is not surprising: By changing the sample size, the average length of branches changes.
+\begin{figure*}[tb]
+    \includegraphics*[width=0.7\textwidth]{figures/illustration_tree.pdf}
+    \caption{\label{fig:illustration_tree}{\bf Illustration Yule and Kingman trees and the spatial location of the tips.}
+    The total tree length of Yule trees is proportional to the number of tips $n$ and the time to the MRCA is $\log(n)$.
+    Kingman trees have a total tree length that is proportional to $\log(n)$ and are characterized by rapid merging close to the present. }
+\end{figure*}
+
+Fig.~\ref{fig:D_and_v} shows empirical estimates of diffusion constants and velocity as defined above from simulated data using freely diffusing locations along Kingman trees.
+The diffusion constant $D$ can be estimated robustly from these data and are compatible with the diffusion constant used to simulate the data $D=1$.
+Values of $v_w$ and $v_b$, however, depend strongly on the sample size. 
+That the velocity estimates are ill-defined is not surprising: By changing the sample size, the average length of branches changes.
 Diffusion along each branch results in a displacement of $d_i \sim \sqrt{4Dt_i}$ such that a quantity that depends on $\Delta d_i / t_i \sim 1/t_i^{1/2}$ will change with the time $t_i$ that elapsed along the branch.
-If the underlying dispersal process is diffusive, estimates of $v_w$ are not meaningful.
+With more samples in the tree, the average branch length decreases and the velocity estimates increase. 
+This expected scaling $\sim \sqrt{n}$ is indicated by the dashed line in Fig.~\ref{fig:D_and_v}.
+For Yule trees, the average branch length is independent of the number of samples and the velocity estimates are more stable. 
+But estimates for $v_w$ and $v_b$ are not consistent with each other and their numerical values have no interpretation within the model.
+
+In either case, estimates of $v_w$ or $v_b$ are not meaningful. 
+The underlying model is diffusive and has no parameters that individually or combined would have dimensions of a velocity. 
+There is no ground truth to compare the estimates to and $v_w$ and $v_b$ are just descriptive summaries of spatial spread with unclear interpretation.
 
 \begin{figure}[tb]
-    \includegraphics*[width=0.48\textwidth]{figures/yule_total_dispersal.png}
-    \includegraphics*[width=0.48\textwidth]{figures/kingman_total_dispersal.png}
+    \includegraphics*[width=0.48\textwidth]{figures/kingman_dispersal.pdf}
     \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
-    While diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ is not a well-defined quantity.
-Its numerical value depends strongly on sample size and there is no ground truth to compare to.
-This sample size dependence is present for Yule trees (left) and Kingman trees (right).
-}
+    While diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ (weighted) and $v_b$ (by branch) are not well-defined quantities.
+Their numerical value depends strongly on sample size and there is no ground truth to compare to.
+The expected scaling $\sim \sqrt{n}$ of $v$ estimates is indicated by the dashed line.}
 \end{figure}
@@
 \section*{How does density regulation affect estimates phylogeographic estimates?}
-Phylogeographic inference typically assumes that the spatial diffusion process is independent of the branching pattern of the tree and ignores coupling between the two that might arise from density dependent growth rates.
-If one simulates a diffusive spatial process in a square of length $L$, the population density shows strong fluctuations whenever the diffusion constant $D<L^2/T_c$, see Fig.~\ref{fig:heterogeneity} where $T_c$ is the coalescent times scale of the population.
+Phylogeographic inference typically assumes that the spatial diffusion process is independent of the branching pattern of the tree and ignores coupling between the two. 
+Such a coupling could, for example, arise when growth rates depend on local population density, and it has been known for a long time that ignoring such coupling leads to very unrealistic population densities \citep{felsenstein_pain_1975}.
+If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D<L^2/T_c$, where $T_c$ is the coalescent times scale of the population (see Fig.~\ref{fig:heterogeneity}).
 In a well-mixed neutral model, $T_c$ is equal to the population size $N$.
@@
 
-\begin{figure}[tb]
+\begin{figure*}[tb]
     \includegraphics*[width=\textwidth]{figures/heterogeneity_free_diffusion.pdf}
@@
 }
-\end{figure}
+\end{figure*}
 
 Biological populations spread into areas that support them and will quickly populate fertile areas that are below their carrying capacity.
-The simplest model of this process are FKPP waves that model the population density $\phi(x,t)$ as \citep{fisher_wave_1937,KPP1937,hallatschek_life_2010}
+A simple model for the population density $\phi(x,t)$ would be 
 \begin{equation}
+    \label{eq:FKPP}
     \frac{\partial \phi(x,t)}{\partial t} = D\frac{\partial^2 \phi(x,t)}{\partial x^2} + \alpha \phi(x,t)(1-\phi(x,t)/\phi_0)
@@
 where $\phi_0$ is the carrying capacity and $\alpha$ is the population growth rate at zero density.
-This equation admits solutions with a traveling front, that the population ``invades" the empty territory with velocity $v = 2\sqrt{D \alpha}$.
-Note that the model itself does not have an intrinsic velocity.
-The speed at which the population invades space is given as compound of diffusion and accelerated growth in regions of low density and only grows like the square root of diffusion constant.
-It is thus not clear a priori how estimates of $D$ will be affected by density regulation, even in a constant habitat.
+This model has a stable fixed point at $\phi=\phi_0$.
+In a static environment, we thus expect the population density to settle at a constant density $\phi_0$ and density fluctuations will decay with rate $\alpha$.
 
-Dependence of growth rate (and thus tree structure) on density is one key element that phylogeographic models ignore because it is very difficult to capture mathematically.
-It is, however, fairly straightforward to include this in simulations and these simulations can be used to investigate the nature of phylogeographic inferences when some of the assumptions are violated.
+While dependence of growth (and thus tree structure) on density is a key component of real-world population dynamics, it is difficult to implemented in phylogeographic inference frameworks.
+It is, however, fairly straightforward to include density dependence in forward simulations and these simulations can be used to investigate the effect of such coupling between growth and spatial location on phylogeographic inferences.
 We calculate the population density using a Gaussian kernel with interaction radius $r$
@@
 where the sum is over all individuals $i$ in the population of size $N$.
+
 As soon as the interaction radius $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}.
@@
 This manifests in an increasing time to the MRCA, see Fig.~\ref{fig:density_reg}.
-Spatial location and the coalescence process are thus strongly coupled, in conflict with model assumptions and estimates of population processes are likely biased.
+Spatial location and the coalescence process are thus strongly coupled, though this coupling manifests itself in population structure rather than estimates of the diffusion coefficients.
 
-\begin{figure}
+\begin{figure*}
     \includegraphics[width=0.48\textwidth]{figures/density_reg_heterogeneity.pdf}
     \includegraphics[width=0.48\textwidth]{figures/density_reg_tmrca.pdf}
-    \caption{\label{fig:density_reg} {\bf Density regulation reduces fluctuations in density regulation, but results in population fragmentation at low diffusion constants.}}
-\end{figure}
+    \caption{\label{fig:density_reg} {\bf Density regulation reduces fluctuations in density regulation, but results in population fragmentation at low diffusion constants.} The left panel shows population heterogeneity measured using bins of linear dimension $L/5$ for $\alpha =0.05$ for different values of the interaction radius $r$.
+For $r\ll L$, density fluctuations are strongly suppressed.
+At the same time, the time to the most recent common ancestor increases dramatically, indicating that the population fragments in subpopulations that don't interact.}
+\end{figure*}
+
+The simplest model of this process are FKPP waves that model the population density $\phi(x,t)$ as \citep{fisher_wave_1937,KPP1937,hallatschek_life_2010}
+
+This equation admits solutions with a traveling front, that the population ``invades'' the empty territory with velocity $v = 2\sqrt{D \alpha}$.
+Note that the model itself does not have an intrinsic velocity.
+The speed at which the population invades space is given as compound of diffusion and accelerated growth in regions of low density and only grows like the square root of diffusion constant.
+It is thus not clear a priori how estimates of $D$ will be affected by density regulation, even in a constant habitat.
 

### 2024-03-24 src/estimate_diffusion_from_tree.py
work on plotting scripts

@@ -26,12 +26,17 @@ def estimate_diffusion(tree, include_root=False):
 
     return {'Dx_branch': 0.5*np.mean([d['dx']**2/d['dt'] for d in displacements]),
             'Dy_branch': 0.5*np.mean([d['dy']**2/d['dt'] for d in displacements]),
-            'vx_branch': 0.5*np.mean([np.abs(d['dx'])/d['dt'] for d in displacements]),
-            'vy_branch': 0.5*np.mean([np.abs(d['dy'])/d['dt'] for d in displacements]),
+            'vx_branch': np.mean([np.abs(d['dx'])/d['dt'] for d in displacements]),
+            'vy_branch': np.mean([np.abs(d['dy'])/d['dt'] for d in displacements]),
             'Dx_total': 0.5*np.sum([d['dx']**2 for d in displacements])/np.sum([d['dt'] for d in displacements]),
             'Dy_total': 0.5*np.sum([d['dy']**2 for d in displacements])/np.sum([d['dt'] for d in displacements]),
-            'vx_total': 0.5*np.sum(np.abs([d['dx'] for d in displacements])/np.sum([d['dt'] for d in displacements])),
-            'vy_total': 0.5*np.sum(np.abs([d['dy'] for d in displacements])/np.sum([d['dt'] for d in displacements]))
+            'vx_total': np.sum(np.abs([d['dx'] for d in displacements])/np.sum([d['dt'] for d in displacements])),
+            'vy_total': np.sum(np.abs([d['dy'] for d in displacements])/np.sum([d['dt'] for d in displacements])),
+            'Dxy_total': 0.25*np.sum([(d['dx']**2+d['dy']**2) for d in displacements])/np.sum([d['dt'] for d in displacements]),
+            'vxy_total': np.sum(np.sqrt([(d['dx']**2+d['dy']**2) for d in displacements])/np.sum([d['dt'] for d in displacements])),
+            'Dxy_branch': 0.25*np.mean([(d['dx']**2+d['dy']**2)/d['dt'] for d in displacements]),
+            'vxy_branch': np.mean([np.sqrt(d['dx']**2+d['dy']**2)/d['dt'] for d in displacements]),
+            't_total': np.sum([d['dt'] for d in displacements])
             }
 
 

### 2024-03-24 src/free_diffusion.py

@@ -14,7 +14,7 @@ def random_tree(n, yule=False):
     while len(tree)>=2:
         i,j = np.random.choice(len(tree), size=2, replace=False)
         if yule:
-            dt = np.random.exponential()/(len(tree)-1)/n*2
+            dt = np.random.exponential()/(len(tree)-1)
         else:
             dt = np.random.exponential()/(len(tree)-1)/len(tree)*2
 

### 2024-03-24 src/illustration.py

@@ -1,0 +1,33 @@
+import numpy as np
+from free_diffusion import *
+from density_regulation import dict_to_phylo_tree
+import matplotlib.pyplot as plt
+from Bio import Phylo
+from matplotlib.cm import viridis_r 
+
+if __name__=="__main__":
+    n_leaves = 100
+
+    fig, axs = plt.subplots(2,2, figsize=(12,8))
+    for row, yule in [(0,True), (1,False)]:
+        scale = np.log(n_leaves) if yule else 2
+        tree = random_tree(n_leaves, yule=yule)
+        add_positions(tree['tree'], 1)
+        phylo_tree = dict_to_phylo_tree(tree['tree'], child_attr='clades')
+        for n in phylo_tree.find_clades():
+            n.color = [int(x) for x in viridis_r(-n.t/scale, bytes=True)[:3]]
+
+        Phylo.draw(phylo_tree, axes=axs[row,0], label_func=lambda x: '')
+        axs[row,0].set_title(f"{'Yule' if yule else 'Kingman'} tree")
+        axs[row,0].axis('off')
+        for n in phylo_tree.get_nonterminals():
+            for c in n.clades:
+                axs[row,1].plot([n.pos['x'], c.pos['x']], [n.pos['y'], c.pos['y']], c='k', lw=0.5, alpha=0.5)
+
+        axs[row,1].scatter([n.pos['x'] for n in phylo_tree.find_clades()], [n.pos['y'] for n in phylo_tree.find_clades()], 
+                c=[n.t for n in phylo_tree.find_clades()], s=30)
+        axs[row,1].axis('equal')
+        axs[row,1].scatter([0], [0], c='r', s=100)
+    
+    plt.savefig('figures/illustration_tree.pdf')
+    

### 2024-03-24 src/inflated_diffusion.py

@@ -24,7 +24,6 @@ def estimate_inflated_diffusion(D, interaction_radius, density_reg, N, subsampli
             print("population nearly extinct")
             continue
         if t%(N//5)==0 and t>10*N: # take samples after burnin every Tc//5
-            tbins = sorted([0] + [t - i*N/10 for i in range(4)])
             clean_tree(tree)
             H, bx, by = get_2d_hist(terminal_nodes, Lx, Ly, linear_bins)
             density_variation.append(np.std(H)/N*np.prod(H.shape))
@@ -37,9 +36,12 @@ def estimate_inflated_diffusion(D, interaction_radius, density_reg, N, subsampli
                     Tmrca.append(t-tree['clades'][0]['time'])
                 else:
                     Tmrca.append(t)
+
+                tbins = np.linspace(Tmrca[-1], t, 5)
                 D_est.extend([D_res['Dx_total'], D_res['Dy_total']])
-                zscores.extend([[np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'zx']**2) for i in range(len(tbins)-1)],
-                                [np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'zy']**2) for i in range(len(tbins)-1)]])
+                # calculate the mean squared z-scores for the root node and each time bin
+                zscores.extend([[z.iloc[2].zx**2]+[np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'zx']**2) for i in range(len(tbins)-1)],
+                                [z.iloc[2].zy**2]+[np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'zy']**2) for i in range(len(tbins)-1)]])
 
     return {"density_variation": density_variation, "D_est": D_est, "zscores": zscores, "Tmrca":Tmrca}
 
@@ -49,7 +51,6 @@ if __name__=="__main__":
     import argparse
     parser = argparse.ArgumentParser()
     parser.add_argument('--N', type=int, default=500)
-    parser.add_argument('--D', type=float, default=0.1)
     parser.add_argument('--interaction-radius', type=float, default=0.1)
     parser.add_argument('--density-reg', type=float, default=0.1)
     parser.add_argument('--subsampling', type=float, default=1.0)

### 2024-03-24 src/plot_inflated_diffusion.py

@@ -29,7 +29,7 @@ def parse_data(data, groupby=None):
 
 
 def free_diffusion(D_array, N, linear_bins=5):
-    n_iter = 10
+    n_iter = 100
     res = []
     for D in D_array:
         res.append([get_granularity(N, D, bins=linear_bins)[0] for i in range(n_iter)])
@@ -58,28 +58,28 @@ if __name__=="__main__":
     N_vals = data.N.unique()
 
     ir_to_plot = interaction_radius[:]
-    density_reg_to_plot = density_reg[:-2]
+    density_reg_to_plot = density_reg[1:2]
 
     ls = ['-', '-.', "--", ":"][:len(density_reg_to_plot)]
     plt.figure()
-    for N in N_vals:
+    for N in N_vals[1:]:
         for i, (ir, dr) in enumerate(product(ir_to_plot, density_reg_to_plot)):
             D_array = np.array(res["density_variation"][ir, dr, N, :].index)
-            label = f'r={ir}, a={dr}' if N==N_vals[0] else ''
-            plt.plot(D_array*N/Lx/Ly, res["density_variation"][ir, dr, N, :],
+            label = f'r={ir}' if N==N_vals[1] else ''
+            plt.plot(D_array*N/Lx/Ly, res["density_variation"][ir, dr, N, :]/np.sqrt(nbins/N),
                     label=label, ls=ls[i%len(ls)], c=f"C{i//len(ls)}")
 
         free_diffusion_heterogeneity = free_diffusion(D_array*N/Lx/Ly, N, linear_bins=linear_bins)
-    plt.plot(D_array*N/Lx/Ly, free_diffusion_heterogeneity, label='diffusion', c='k')
-    plt.plot(D_array*N/Lx/Ly, np.ones_like(D_array)*np.sqrt(nbins/N), label='well mixed limit', c='k', ls='--')
+    plt.plot(D_array*N/Lx/Ly, free_diffusion_heterogeneity/np.sqrt(nbins/N), label='diffusion', c='k')
+    plt.plot(D_array*N/Lx/Ly, np.ones_like(D_array), c='k', ls='--')
     plt.xscale('log')
-    plt.xlabel('diffusion constant')
-    plt.ylabel('heterogeneity')
+    plt.xlabel('diffusion constant [L^2/N]')
+    plt.ylabel('heterogeneity (relative to well-mixed case)')
     plt.legend()
     if args.output_heterogeneity:
         plt.savefig(args.output_heterogeneity)
 
-
[…]

### 2024-03-24 src/sample_size_dependence.py

@@ -6,7 +6,7 @@ from itertools import product
 if __name__=="__main__":
     replicates = 100
     results = {}
-    fields = ['Dx_branch', 'Dy_branch', 'vx_branch', 'vy_branch', 'Dx_total', 'Dy_total', 'vx_total', 'vy_total']
+    fields = ['Dxy_branch', 'Dxy_total', 'vxy_branch', 'vxy_total', 't_total']
     for coal in ['yule', 'kingman']:
         results[coal] = {}
         for n in [10,30, 100, 300, 1000, 3000]:
@@ -16,7 +16,7 @@ if __name__=="__main__":
                 add_positions(tree, 1)
                 D = estimate_diffusion(tree)
                 res.append(D)
-
+            print(n, np.mean([x['t_total'] for x in res]))
             results[coal][n] = {k: np.mean([x[k] for x in res]) for k in fields}
             results[coal][n].update({f"{k}_std": np.std([x[k] for x in res]) for k in fields})
 
@@ -26,18 +26,22 @@ if __name__=="__main__":
     for coal in ['yule', 'kingman']:
         res = results[coal]
         plt.figure()
-        for suffix in ['_total', '_branch']:
-            for p in ['Dx', 'vx']:
-                q = p+suffix
-                label = f"{q} [{'mean(dx_i^2)/mean(2t_i)' if 'D' in q \
-                                else 'mean(dx_i)/mean(t_i)'}]" if suffix=='_total' \
-                                else f"{q} [{'mean(dx_i^2/2t_i)' if 'D' in q else 'mean(dx_i/t_i)'}]"
-                plt.errorbar([n for n in res], [res[n][q] for n in res], [res[n][q+"_std"] for n in res], ls = '-', marker='o',
+        offset = 1.0/1.02
+        for suffix, suffix_label in [('_total', 'weighted'), ('_branch', 'by branch')]:
+            for p, quantity in [('Dxy', 'diffusion'), ('vxy', 'velocity')]:
+                q = f"{p}{suffix}"
+                label = f"{quantity}, {suffix_label}"
+                plt.errorbar([n*offset for n in res], [res[n][q] for n in res], [res[n][q+"_std"] for n in res], ls = '-', marker='o',
                             label=label)
+                offset *= 1.02**2
+        plt.plot([10, 3000], [1,1], ls='--', color='k')
+        plt.plot([10, 3000], [10,10*300**0.5], ls='-.', color='k')
 
[…]

### 2024-03-24 src/inflated_diffusion.py
calculate z-scores in quantiles

@@ -7,6 +7,7 @@ from estimate_diffusion_from_tree import estimate_diffusion, estimate_ancestral_
 
 def estimate_inflated_diffusion(D, interaction_radius, density_reg, N, subsampling=1.0,
                                 Lx=1, Ly=1, linear_bins=5, n_iter=10, n_subsamples=1):
+    from scipy.stats import scoreatpercentile
     # set up tree and initial population uniformly in space
     tree = make_node(Lx/2,Ly/2,-2, None)
     tree['children'] = [make_node(np.random.random()*Lx, np.random.random()*Ly, -1, tree)
@@ -37,7 +38,8 @@ def estimate_inflated_diffusion(D, interaction_radius, density_reg, N, subsampli
                 else:
                     Tmrca.append(t)
 
-                tbins = np.linspace(Tmrca[-1], t, 5)
+                internal_node_times= sorted(z.loc[z.nonterminal, 't'])
+                tbins = scoreatpercentile(internal_node_times, [0, 20, 40, 60, 80, 100])
                 D_est.extend([D_res['Dx_total'], D_res['Dy_total']])
                 # calculate the mean squared z-scores for the root node and each time bin
                 zscores.extend([[z.iloc[2].zx**2]+[np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'zx']**2) for i in range(len(tbins)-1)],

### 2024-03-25 src/stable_density.py
rename inflated to stable density

@@ -4,8 +4,7 @@ from heterogeneity import get_2d_hist
 from estimate_diffusion_from_tree import estimate_diffusion, estimate_ancestral_positions, collect_zscore
 
 
-
-def estimate_inflated_diffusion(D, interaction_radius, density_reg, N, subsampling=1.0,
+def evolve_stable_density(D, interaction_radius, density_reg, N, subsampling=1.0,
                                 Lx=1, Ly=1, linear_bins=5, n_iter=10, n_subsamples=1):
     from scipy.stats import scoreatpercentile
     # set up tree and initial population uniformly in space
@@ -35,15 +34,17 @@ def estimate_inflated_diffusion(D, interaction_radius, density_reg, N, subsampli
                 z = collect_zscore(tree)
                 if len(tree['clades'])==1:
                     Tmrca.append(t-tree['clades'][0]['time'])
+                    root_index = 1
                 else:
                     Tmrca.append(t)
+                    root_index = 0
 
                 internal_node_times= sorted(z.loc[z.nonterminal, 't'])
                 tbins = scoreatpercentile(internal_node_times, [0, 20, 40, 60, 80, 100])
                 D_est.extend([D_res['Dx_total'], D_res['Dy_total']])
                 # calculate the mean squared z-scores for the root node and each time bin
-                zscores.extend([[z.iloc[2].zx**2]+[np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'zx']**2) for i in range(len(tbins)-1)],
-                                [z.iloc[2].zy**2]+[np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'zy']**2) for i in range(len(tbins)-1)]])
+                zscores.extend([[z.iloc[root_index].zx**2]+[np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'zx']**2) for i in range(len(tbins)-1)],
+                                [z.iloc[root_index].zy**2]+[np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'zy']**2) for i in range(len(tbins)-1)]])
 
     return {"density_variation": density_variation, "D_est": D_est, "zscores": zscores, "Tmrca":Tmrca}
 
@@ -59,6 +60,8 @@ if __name__=="__main__":
     parser.add_argument('--output', type=str)
     args = parser.parse_args()
 
[…]

### 2024-03-25 src/plot_stable_density.py
rename plotting script

@@ -40,7 +40,7 @@ def free_diffusion(D_array, N, linear_bins=5):
 if __name__=="__main__":
     import argparse
     parser = argparse.ArgumentParser()
-    parser.add_argument('--data', type=str, default='data/inflated_diffusion.csv')
+    parser.add_argument('--data', type=str, default='data/stable_density.csv')
     parser.add_argument('--output-diffusion', type=str)
     parser.add_argument('--output-heterogeneity', type=str)
     parser.add_argument('--output-zscore', type=str)
@@ -57,9 +57,11 @@ if __name__=="__main__":
     density_reg = data.density_reg.unique()
     N_vals = data.N.unique()
 
+    # select the interaction radius and density regulation strength to plot
     ir_to_plot = interaction_radius[:]
     density_reg_to_plot = density_reg[1:2]
 
+    # Plot the heterogeneity
     ls = ['-', '-.', "--", ":"][:len(density_reg_to_plot)]
     plt.figure()
     for N in N_vals[1:]:

### 2024-04-05 src/estimate_diffusion_from_tree.py
add seasaw

@@ -136,16 +136,20 @@ def collect_positioning(tree):
     collect_positioning_rec(tree, res)
     return res
 
-def collect_zscore(tree):
+def collect_errors(tree):
     import pandas as pd
     res = []
-    def collect_zscore_rec(node):
+    def collect_errors_rec(node):
         nonterminal = 'clades' in node and len(node['clades'])>0
-        res.append({'zx': (node['x']-node['position']['x']['mean'])/node['position']['x']['var']**0.5,
-                    'zy': (node['y']-node['position']['y']['mean'])/node['position']['y']['var']**0.5,
+        x_err = (node['x']-node['position']['x']['mean'])
+        y_err = (node['y']-node['position']['y']['mean'])
+        x_std = node['position']['x']['var']**0.5
+        y_std = node['position']['y']['var']**0.5
+        res.append({'x_err': x_err, 'x_std': x_std, 'xz':x_err/x_std,
+                    'y_err': y_err, 'y_std': y_std, 'xz':y_err/y_std,
                     't': node['time'], 'nonterminal':nonterminal})
         if nonterminal:
             for c in node['clades']:
-                collect_zscore_rec(c)
-    collect_zscore_rec(tree)
+                collect_errors_rec(c)
+    collect_errors_rec(tree)
     return pd.DataFrame(res)

### 2024-04-05 src/habitat_shifts.py

@@ -2,7 +2,7 @@ import numpy as np
 from density_regulation import make_node, evolve, clean_tree, subsample_tree
 from heterogeneity import get_2d_hist
 from density_regulation import subsample_tree
-from estimate_diffusion_from_tree import estimate_diffusion, estimate_ancestral_positions, collect_zscore
+from estimate_diffusion_from_tree import estimate_diffusion, estimate_ancestral_positions, collect_errors
 
 def cycling_patches(N, Lx, Ly, period, wave_length):
     def f(x,y,t):
@@ -23,6 +23,10 @@ def breathing(N, Lx, Ly, period, width):
         return N*(np.exp(-prefactor*(min((pos - x%Lx)%Lx, (x%Lx - pos)%Lx)**2)))
     return f
 
+def seasaw(N, Lx, Ly, period):
+    def f(x,y,t):
+        return N*np.maximum(0.01,np.minimum(1,0.5*(1+0.8*np.cos(2*np.pi*t/period)*np.cos(2*np.pi*x/Lx)**1)))
+    return f
 
 def diffusion_in_changing_habitats(D, interaction_radius, density_reg, N, subsampling=1.0,
                                 Lx=1, Ly=1, linear_bins=5, n_iter=10, n_subsamples=1, gtd=None, habitat_params=None):
@@ -39,6 +43,8 @@ def diffusion_in_changing_habitats(D, interaction_radius, density_reg, N, subsam
     v_est_y = []
     zscores_x = []
     zscores_y = []
+    x_err = []
+    y_err = []
     Tmrca = []
 
     for t in range((n_iter+10)*N):
@@ -58,7 +64,7 @@ def diffusion_in_changing_habitats(D, interaction_radius, density_reg, N, subsam
                 subsample_tree(terminal_nodes, tree, p=subsampling, subtree_attr='clades')
                 D_res = estimate_diffusion(tree)
                 estimate_ancestral_positions(tree, D)
-                z = collect_zscore(tree)
+                z = collect_errors(tree)
                 if len(tree['clades'])==1:
                     Tmrca.append(t-tree['clades'][0]['time'])
                 else:
@@ -69,10 +75,12 @@ def diffusion_in_changing_habitats(D, interaction_radius, density_reg, N, subsam
[…]

### 2024-04-05 src/plot_diffusion.py

@@ -1,7 +1,7 @@
 import numpy as np
 from density_regulation import make_node, evolve, clean_tree, dict_to_phylo_tree, add_as_color, subsample_tree
 from heterogeneity import get_2d_hist
-from estimate_diffusion_from_tree import estimate_diffusion, estimate_ancestral_positions, collect_zscore
+from estimate_diffusion_from_tree import estimate_diffusion, estimate_ancestral_positions, collect_errors
 from Bio import Phylo
 
 N=1000
@@ -26,7 +26,7 @@ clean_tree(tree)
 subsample_tree(terminal_nodes, tree, p=1.0)
 D_res = estimate_diffusion(tree)
 estimate_ancestral_positions(tree, D)
-z = collect_zscore(tree)
+z = collect_errors(tree)
 
 import matplotlib.pyplot as plt
 plt.ion()

### 2024-04-05 src/plot_waves.py

@@ -40,9 +40,9 @@ def parse_data(data, groupby=None):
 
 
 
-def make_figure(fname=None):
+def make_figure(fname=None, width=0.5):
     fig, ax = plt.subplots(1,1, sharex=True, sharey=True, figsize=(4,2))
-    f = waves(1, 3, 1, velocity=1, width=1.0)
+    f = waves(1, 3, 1, velocity=1, width=width)
     grid = np.meshgrid(np.linspace(0,3,31), np.linspace(0,1,11))
     ax.matshow(f(grid[0], grid[1], 0))
     ax.set_axis_off()
@@ -58,13 +58,14 @@ if __name__=="__main__":
     parser = argparse.ArgumentParser()
     parser.add_argument('--data', type=str, default='data/waves.csv')
     parser.add_argument('--output-diffusion', type=str)
+    parser.add_argument('--output-velocity', type=str)
     parser.add_argument('--output-zscore', type=str)
     parser.add_argument('--output-tmrca', type=str)
     parser.add_argument('--illustration', type=str)
     args = parser.parse_args()
 
     # make an illustration of the carrying capacity at different times
-    make_figure(fname = args.illustration)
+    make_figure(fname = args.illustration, width=0.5)
 
 
     # read and organize data
@@ -75,101 +76,74 @@ if __name__=="__main__":
 
 
     res = parse_data(data, groupby=['interaction_radius', 'density_reg', 'N', 'velocity', 'subsampling'])
-    interaction_radius = data.interaction_radius.unique()
     velocity = sorted(data.velocity.unique())
-    subsampling = data.subsampling.unique()
-    density_reg = data.density_reg.unique()
-    N_vals = data.N.unique()
+    interaction_radius = sorted(data.interaction_radius.unique())
+    subsampling = sorted(data.subsampling.unique())
+    density_reg = sorted(data.density_reg.unique())
+    N_vals = sorted(data.N.unique())
 
     ls = ['-', ':', "--", ".-"][len(interaction_radius)]
     markers = ['o', '<', '>', 's', 'd', '^', 'v']
 
 
     ## Diffusion constant estimates
-    for c in ['x', 'y']:
-        plt.figure()
+    for c in ['x']:
+        plt.figure(figsize=(6,3))
         #plt.title(f"{T}")
         for vi,v in enumerate(velocity):
+            labeled=False
             for m, N in zip(['o', '<'], N_vals):
                 D_array = data.loc[data.N==N].D.unique()
[…]

### 2024-04-05 src/seasaw.py

@@ -1,5 +1,5 @@
 import numpy as np
-from habitat_shifts import waves, diffusion_in_changing_habitats
+from habitat_shifts import seasaw, diffusion_in_changing_habitats
 
 
 if __name__=="__main__":
@@ -8,9 +8,10 @@ if __name__=="__main__":
     import argparse
     parser = argparse.ArgumentParser()
     parser.add_argument('--N', type=int, default=500)
+    parser.add_argument('--D', type=float, default=0.1)
     parser.add_argument('--interaction-radius', type=float, default=0.1)
     parser.add_argument('--density-reg', type=float, default=0.1)
-    parser.add_argument('--velocity', type=float, default=0.01)
+    parser.add_argument('--period', type=int, default=100)
     parser.add_argument('--subsampling', type=float, default=1.0)
     parser.add_argument('--output', type=str)
     args = parser.parse_args()
@@ -30,7 +31,7 @@ if __name__=="__main__":
         print(f"{di} out of {len(D_array_dens)}: D={D:1.3e}")
         res = diffusion_in_changing_habitats(D, interaction_radius, density_reg, N, subsampling=args.subsampling,
                                           Lx=Lx, Ly=Ly, linear_bins=linear_bins, n_iter=n_iter, 
-                                          gtd=waves, habitat_params={'velocity':args.velocity, 'width':width})
+                                          gtd=seasaw, habitat_params={'period':args.period})
 
         tmpD_x = np.mean(res["D_est_x"], axis=0)
         tmpD_y = np.mean(res["D_est_y"], axis=0)
@@ -65,7 +66,7 @@ if __name__=="__main__":
         import os
         if not os.path.exists('data'):
             os.makedirs('data')
-        fname = f'data/habitats_diffusion_{N=}_ir={interaction_radius}_dr={density_reg}.csv'
+        fname = f'data/seasaw_{N=}_ir={interaction_radius}_dr={density_reg}.csv'
 
     pd.DataFrame(D_est).to_csv(fname, index=False)
 

### 2024-04-05 src/stable_density.py

@@ -1,7 +1,7 @@
 import numpy as np
 from density_regulation import make_node, evolve, clean_tree, subsample_tree
 from heterogeneity import get_2d_hist
-from estimate_diffusion_from_tree import estimate_diffusion, estimate_ancestral_positions, collect_zscore
+from estimate_diffusion_from_tree import estimate_diffusion, estimate_ancestral_positions, collect_errors
 
 
 def evolve_stable_density(D, interaction_radius, density_reg, N, subsampling=1.0,
@@ -31,7 +31,7 @@ def evolve_stable_density(D, interaction_radius, density_reg, N, subsampling=1.0
                 subsample_tree(terminal_nodes, tree, p=subsampling, subtree_attr='clades')
                 D_res = estimate_diffusion(tree)
                 estimate_ancestral_positions(tree, D)
-                z = collect_zscore(tree)
+                z = collect_errors(tree)
                 if len(tree['clades'])==1:
                     Tmrca.append(t-tree['clades'][0]['time'])
                     root_index = 1

### 2024-04-05 src/waves_example.py

@@ -1,0 +1,73 @@
+import numpy as np
+from habitat_shifts import waves, seasaw
+from density_regulation import make_node, evolve, subsample_tree, dict_to_phylo_tree
+from estimate_diffusion_from_tree import estimate_diffusion, estimate_ancestral_positions
+import matplotlib.pyplot as plt
+from Bio import Phylo
+from matplotlib.cm import viridis_r 
+
+
+if __name__=="__main__":
+    import sys
+    sys.setrecursionlimit(10000)
+    import argparse
+    parser = argparse.ArgumentParser()
+    parser.add_argument('--N', type=int, default=500)
+    parser.add_argument('--D', type=float, default=0.1)
+    parser.add_argument('--interaction-radius', type=float, default=0.1)
+    parser.add_argument('--density-reg', type=float, default=0.1)
+    parser.add_argument('--velocity', type=float, default=0.01)
+    parser.add_argument('--subsampling', type=float, default=1.0)
+    parser.add_argument('--output', type=str)
+    args = parser.parse_args()
+
+    N = args.N
+    width = 0.5
+    Lx, Ly = 3, 1
+    D = args.D
+    interaction_radius, density_reg = args.interaction_radius, args.density_reg
+    #target_density = waves(N, Lx, Ly, **{'velocity':args.velocity, "width":width})
+    period = 100
+    target_density = seasaw(N, Lx, Ly, **{'period':period})
+
+    v_FKPP = 2*np.sqrt(args.velocity*D)
+    print(f"{interaction_radius=:1.3f}, {density_reg=:1.3f}")
+
+    tree = make_node(Lx/2,Ly/2,-2, None)
+    tree['children'] = [make_node(np.random.random()*Lx, np.random.random()*Ly, -1, tree)
+                        for i in range(N)]
+    terminal_nodes = tree['children']
+
+    t = 0
+    for tmax in [5*N + i*period/3.0 for i in range(5)]:
+        while t<tmax:
+            terminal_nodes = evolve(terminal_nodes, t, Lx=Lx, Ly=Ly, interaction_radius=interaction_radius,
+                                density_reg=density_reg, D=D, target_density=target_density, total_population=N)
+            t+=1
+
+        subsample_tree(terminal_nodes, tree, p=1.0, subtree_attr='clades')
+        D_res = estimate_diffusion(tree)
[…]

### 2024-04-05 src/breathing.py
fix file name

@@ -45,6 +45,8 @@ if __name__=="__main__":
         tmpStdZ_x = f"[{' '.join(str(x) for x in np.ma.std(np.ma.masked_invalid(res['zscores_x']), axis=0).filled(fill_value=np.nan))}]"
         tmpZ_y =    f"[{' '.join(str(x) for x in np.ma.mean(np.ma.masked_invalid(res['zscores_y']), axis=0).filled(fill_value=np.nan))}]"
         tmpStdZ_y = f"[{' '.join(str(x) for x in np.ma.std(np.ma.masked_invalid(res['zscores_y']), axis=0).filled(fill_value=np.nan))}]"
+        tmp_x_err =  f"[{' '.join(str(x) for x in np.ma.mean(np.ma.masked_invalid(res['x_err']), axis=0).filled(fill_value=np.nan))}]"
+        tmp_y_err =  f"[{' '.join(str(x) for x in np.ma.mean(np.ma.masked_invalid(res['y_err']), axis=0).filled(fill_value=np.nan))}]"
         nobs = len(res["D_est"])
         D_est.append({"interaction_radius":interaction_radius, "density_reg": density_reg,
                       "N": N, "n": len(res["D_est"]), "period": args.period, "subsampling": args.subsampling, "D":D, 
@@ -52,6 +54,7 @@ if __name__=="__main__":
                       "meanvx": tmpv_x, "stdv_x": tmpStdv_x, "meanvy": tmpv_y, "stdv_y": tmpStdv_y,
                       "meanZsq_x": tmpZ_x, "stdZsq_x": tmpStdZ_x, 
                       "meanZsq_y": tmpZ_y, "stdZsq_y": tmpStdZ_y, 
+                      "x_err": tmp_x_err, "y_err": tmp_y_err, 
                       'observations': nobs,
                       "density_variation": np.mean(res['density_variation']),
                       "meanTmrca":np.mean(res["Tmrca"]), "stdTmrca":np.std(res["Tmrca"])})

### 2024-04-05 src/estimate_diffusion_from_tree.py
fix zx/zy keys, store absolute errors

@@ -145,8 +145,8 @@ def collect_errors(tree):
         y_err = (node['y']-node['position']['y']['mean'])
         x_std = node['position']['x']['var']**0.5
         y_std = node['position']['y']['var']**0.5
-        res.append({'x_err': x_err, 'x_std': x_std, 'xz':x_err/x_std,
-                    'y_err': y_err, 'y_std': y_std, 'xz':y_err/y_std,
+        res.append({'x_err': x_err, 'x_std': x_std, 'zx':x_err/x_std,
+                    'y_err': y_err, 'y_std': y_std, 'zy':y_err/y_std,
                     't': node['time'], 'nonterminal':nonterminal})
         if nonterminal:
             for c in node['clades']:

### 2024-04-05 src/habitat_shifts.py

@@ -75,8 +75,10 @@ def diffusion_in_changing_habitats(D, interaction_radius, density_reg, N, subsam
                 v_est_y.append(D_res['vy_total'])
                 zscores_x.append([np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'zx']**2) for i in range(len(tbins)-1)])
                 zscores_y.append([np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'zy']**2) for i in range(len(tbins)-1)])
-                x_err.append([np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'x_err']) for i in range(len(tbins)-1)])
-                y_err.append([np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'y_err']) for i in range(len(tbins)-1)])
+                x_err.append([np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'x_err']) for i in range(len(tbins)-1)] + 
+                             [np.mean(np.abs(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'x_err'])) for i in range(len(tbins)-1)])
+                y_err.append([np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'y_err']) for i in range(len(tbins)-1)] + 
+                             [np.mean(np.abs(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'y_err'])) for i in range(len(tbins)-1)])
 
     return {"density_variation": density_variation, "D_est_x": D_est_x, "D_est_y": D_est_y, "D_est": D_est_x + D_est_y,
             "v_est_x": v_est_x, "v_est_y": v_est_y, "v_est": v_est_x + v_est_y,

### 2024-04-05 src/seasaw.py
fix fields in table

@@ -49,7 +49,7 @@ if __name__=="__main__":
         tmpStdZ_y = f"[{' '.join(str(x) for x in np.ma.std(np.ma.masked_invalid(res['zscores_y']), axis=0).filled(fill_value=np.nan))}]"
         nobs = len(res["D_est"])
         D_est.append({"interaction_radius":interaction_radius, "density_reg": density_reg,
-                      "N": N, "n": len(res["D_est"]), "velocity": args.velocity, "subsampling": args.subsampling, "D":D, 
+                      "N": N, "n": len(res["D_est"]), "period": args.period, "subsampling": args.subsampling, "D":D, 
                       "meanDx": tmpD_x, "stdD_x": tmpStdD_x, "meanDy": tmpD_y, "stdD_y": tmpStdD_y,
                       "meanvx": tmpv_x, "stdv_x": tmpStdv_x, "meanvy": tmpv_y, "stdv_y": tmpStdv_y,
                       "meanZsq_x": tmpZ_x, "stdZsq_x": tmpStdZ_x, 

### 2024-04-05 manuscript/phylogeo.tex
add to ms

@@
 Spatial location and the coalescence process are thus strongly coupled, though this coupling manifests itself in population structure rather than estimates of the diffusion coefficients.
+In such a spatially structured population, estimates of historical populations sizes (e.g.~skyline estimates) will be strongly affected.
 
@@
 
-The simplest model of this process are FKPP waves that model the population density $\phi(x,t)$ as \citep{fisher_wave_1937,KPP1937,hallatschek_life_2010}
+As discussed above, a static carrying capacity in an equilibrium population has limited effects on estimates of ancestral locations and diffusion coefficients. 
+In reality, carrying capacities vary through time. 
+Such variation happens on all time scales, ranging from environmental changes over millennia to seasonal fluctuations. 
+The circulation of seasonal influenza viruses, for example, varies by many orders of magnitude between summer and winter in temperate climates.
+Fluctuations in carrying capacity mean that populations not just spread because individuals move, but also because populations grow in regions where the population density is below the carrying capacity.
+In such situations, the accuracy and interpretation of phylogeographic inferences are unclear.
 
-This equation admits solutions with a traveling front, that the population ``invades'' the empty territory with velocity $v = 2\sqrt{D \alpha}$.
-Note that the model itself does not have an intrinsic velocity.
+A simple model of the interplay between growth and spatial spread is again given by the Fisher-KPP equation (Eq.~\ref{eq:FKPP}) above. 
+This equation admits solutions with a traveling front, that the population ``invades'' the empty territory with velocity $v_{FKPP} = 2\sqrt{D \alpha}$ \citep{fisher_wave_1937,KPP1937,hallatschek_life_2010}.
+Note that the model itself does not have an explicit parameter that has dimensions of a velocity.
 The speed at which the population invades space is given as compound of diffusion and accelerated growth in regions of low density and only grows like the square root of diffusion constant.
-It is thus not clear a priori how estimates of $D$ will be affected by density regulation, even in a constant habitat.
+
+To explore such the result of such changing habitats on phylogeographic inferences, we simulated a habitat with a Gaussian carrying capacity that moves at a constant velocity $v$ along the $x$-axis (Fig.~\ref{fig:traveling_wave}).
+When varying the microscopic dispersal parameter $D$ and the velocity of the carrying capacity $v$, I observe diverse behavior with often dramatically overestimated $\hat{D}$.
+These behaviors are explained comparing the speed $v_{FKPP}$ at which the population can invade new territory with the external speed $v$, the ratio of which is used to plot the simulation data in Fig.~\ref{fig:traveling_wave}.
+For $v_{FKPP} \ll v$, the population cannot keep up with the moving carrying capacity and is saved from extinction only because selection is soft.
+In this regime, the estimated velocity of lineages is below $v$ and both the estimated $v$ and $D$ increase with $v_{FKPP}$. 
+Once $v\sim v_{FKPP}$, the estimated lineage velocity coincides with $v_{FKPP}$, while the estimated $D$ peaks at values orders of magnitude above the true value. 
+For larger $v_{FKPP}$, the estimated velocity is stable for some time, before increasing again with $v_{FKKP}$ is a sampling dependent manner. 
+At larger $v_{FKPP}$, $D$ is eventually underestimates.
+
+In this extreme example of a population dragged along by a moving carrying capacity, estimates of dispersal rates are meaningless, and estimates of velocities are only compatible with the speed of the shifting habitat in a narrow range when the habitat speed matches the FKPP velocity.
+Similarly, estimates of ancestral locations are biased to the current position of the habitat.
+
+Less extreme spatio-temporal dynamics of the habitat results in less extreme biases, but still illustrate that caution is required when interpreting phylogeographic inferences in environment that might change between the present and the $T_{MRCA}$ of the population.
+I simulated two additional scenarios: one habitat that expands and contracts to a core region periodically, and one that oscillates back and forth between two regions with a homogeneous intermediate state.
+
+\begin{figure}
+    \includegraphics[width=0.48\textwidth]{figures/traveling_wave}
+    \includegraphics[width=0.48\textwidth]{figures/waves_D}
+    \includegraphics[width=0.48\textwidth]{figures/waves_v}
+    \caption{\label{fig:traveling_wave} {\bf Phylogeographic estimates in rapidly changing environments.} Panel A shows an illustration of the simulated set-up: a carrying capacity with a Gaussian profile with $\sigma=0.5$ that moves along the $x$-axis with velocity $v$ on a patch with $L_x=3$, $L_y=1$, and periodic boundary conditions.
+    Panels B\&C show estimates of diffusion coefficients and velocities, relative to their true values, as a function of the ratio of the FKPP velocity to the external velocity $v$. }
+\end{figure}
+
+\begin{itemize}
+    \item periodic seasaws and other patters
+    \item can we trust ancestral location inferences? with true D, with estimated D? significant biases?
+    \item can we say something general? coalescence time versus diffusion time across the habitat, habitat changes vs fkpp speed
+    \item typical FKPP speed: $D=km^2/day$, $\alpha = 0.2/day$. $v_{FKPP} = 2\sqrt{D\alpha} = km/day$.
+\end{itemize}
+
+\section*{Discussion}
+
+\bibliography{bib}
+
+\end{document}
 
@@
 \end{figure}
-
-\bibliography{bib}
-
-\end{document}

### 2024-04-22 notes/most_likely_diffusion.tex
add estimation of positions
big commit
clean up notes

@@
         & = \prod_{c\in n} \frac{\sqrt{d_c}}{\sqrt{\pi}}\int d\, r_c e^{- d_c r_n^2 + (d_c + a_c)(r_c - \frac{b_c + d_c r_n}{d_c + a_c})^2 + \frac{(b_c + d_c r_n)^2}{d_c + a_c} - c_c} \\
-        & = \frac{\sqrt{d_c}}{\sqrt{a_c + d_c}} e^{- d_c r_n^2 + (b_c^2 + 2r_n b_c d_c + d_c^2 r_n^2 )/\left(d_c + a_c\right) - c_c} \\
-        & = \frac{\sqrt{d_c}}{\sqrt{a_c + d_c}} e^{- d_c\left(1 - \frac{d_c}{a_c+d_c}\right) r_n^2 +  2\frac{b_c d_c}{d_c+a_c} r_n  - c_c + \frac{b_c^2}{d_c + a_c}} \\
+        & = \prod_{c\in n}\frac{\sqrt{d_c}}{\sqrt{a_c + d_c}} e^{- d_c r_n^2 + (b_c^2 + 2r_n b_c d_c + d_c^2 r_n^2 )/\left(d_c + a_c\right) - c_c} \\
+        & = \prod_{c\in n}\frac{\sqrt{d_c}}{\sqrt{a_c + d_c}} e^{- d_c\left(1 - \frac{d_c}{a_c+d_c}\right) r_n^2 +  2\frac{b_c d_c}{d_c+a_c} r_n  - c_c + \frac{b_c^2}{d_c + a_c}} \\
     \end{split}
 \end{equation}
-This allows calculation of the parameters $a_n$, $b_n$, and $c_n$ of node $n$ from the children that are not terminal nodes as.
+This allows calculation of the parameters $a_n$, $b_n$, and $c_n$ of node $n$ from the children its children, though the contributions are different for terminal and non-terminal children. 
+For each internal child $c$, we add
 \begin{equation}
-    a_n = \sum_{c\in n}d_c \left(1-\frac{d_c}{a_c+d_c}\right) = \sum_{c\in n} \frac{d_c a_c}{a_c+d_c}
+    a_n += d_c \left(1-\frac{d_c}{a_c+d_c}\right) = \frac{d_c a_c}{a_c+d_c}
 \end{equation}
 \begin{equation}
-    b_n = \sum_{c\in n}\frac{b_c d_c}{a_c + d_c}
+    b_n += \frac{b_c d_c}{a_c + d_c}
 \end{equation}
 \begin{equation}
-    c_n = \sum_{c\in n} c_c  + \frac{b_c^2}{d_c + a_c} + \frac{\log(d_c) - \log(a_c+d_c)}{2}
+    c_n += c_c  + \frac{b_c^2}{d_c + a_c} + \frac{\log(d_c) - \log(a_c+d_c)}{2}
 \end{equation}
@@
 \begin{equation}
-    a_n = \sum_{c\in n} d_c
+    a_n +=  d_c
 \end{equation}
 \begin{equation}
-    b_n = \sum_{c\in n} d_c r_c
+    b_n += d_c r_c
 \end{equation}
 \begin{equation}
-    c_n = \sum_{c\in n} d_c r_c^2 - \log(2\pi/d_c)/2
+    c_n += d_c r_c^2 - \log(2\pi/d_c)/2
 \end{equation}
-Note that for a single child, the variances add ($a_n^{-1} = a_c^{-1} + d_c^{-1}$) and the most likely positions don't change ($b_n/a_n = b_c/a_c$).
+Note that for a single child, the variances due to uncertainty in the child and the transmission along the branch add ($a_n^{-1} = a_c^{-1} + d_c^{-1}$) and the most likely positions don't change ($b_n/a_n = b_c/a_c$).
 
-The same propagation can be used up the tree
+The same propagation can be used up the tree, where we use the index $p$ for the parent of $n$ and label coefficients of the distribution that is passed up a branch with a prime.
 \begin{equation}
-    a'_n  =  \frac{d_p a'_p}{a'_p+d_c} + \sum_{c\in p, c\neq n} \frac{d_c a_c}{a_c+d_c}
+    a'_n  =  \frac{d_n a'_p}{a'_p + d_n} + \sum_{c\in p, c\neq n} \frac{d_c a_c}{a_c+d_c}
 \end{equation}
 \begin{equation}
-    b'_n = \frac{b'_p d_p}{a'_p + d_p} + \sum_{c\in p, c\neq n}\frac{b_c d_c}{a_c + d_c}
+    b'_n = \frac{b'_p d_n}{a'_p + d_n} + \sum_{c\in p, c\neq n}\frac{b_c d_c}{a_c + d_c}
 \end{equation}
 \begin{equation}
-    c'_n = c'_p + + \frac{b'_p^2}{d_p + a'_p} + \frac{\log(d_p) - \log(a'_p+d_p)}{2} +  \sum_{c\in p, c\neq n} c_c  + \frac{b_c^2}{d_c + a_c} + \frac{\log(d_c) - \log(a_c+d_c)}{2}
+    c'_n = c'_p + \frac{{b'_p}^2}{a'_p + d_n} + \frac{\log(d_n) - \log(a'_p+d_n)}{2} +  \sum_{c\in p, c\neq n} c_c  + \frac{b_c^2}{d_c + a_c} + \frac{\log(d_c) - \log(a_c+d_c)}{2}
+\end{equation}
+
+The distribution of positions at an internal node is then
+\begin{equation}
+    \log P(x) = -a_n r_n^2 + 2 b_n r_n - a'_n d_n/(d_n + a'_n) r_n^2 + b'_n d_n/(d_n + a'_n) + C
+\end{equation}
+This can be rearranged into a standard Gaussian form
+\begin{equation}
+    \log P(x) = -(a_n + a'_n d_n/(d_n + a'_n)) (r_n - (b_n + b'_n d_n/(d_n + a'_n))/(a_n + a'_n d_n/(d_n + a'_n)))^2 + C
+\end{equation}
+such that the mean is
+\begin{equation}
+    x_n = b_n/a_n + b'_n d_n/(d_n + a'_n)
+\end{equation}
+and the variance ($() = 1/2\sigma^2$)
+\begin{equation}
+    \sigma^2_n = \frac{1}{2(a_n + a'_n d_n/(d_n + a'_n))} = \frac{(d_n + a'_n)}{2(a_n(d_n + a'_n) + a'_n d_n)}
 \end{equation}

### 2024-04-22 src/breathing.py
catch small subpopulation and prevent failure with just one sampled node

@@ -32,7 +32,9 @@ if __name__=="__main__":
         res = diffusion_in_changing_habitats(D, interaction_radius, density_reg, N, subsampling=args.subsampling,
                                           Lx=Lx, Ly=Ly, linear_bins=linear_bins, n_iter=n_iter, 
                                           gtd=breathing, habitat_params={'period':args.period, 'width':width})
-
+        if len(res['D_est_x'])==0:
+            print("no data")
+            continue
         tmpD_x = np.mean(res["D_est_x"], axis=0)
         tmpD_y = np.mean(res["D_est_y"], axis=0)
         tmpStdD_x = np.std(res["D_est_x"], axis=0)

### 2024-04-22 src/density_regulation.py

@@ -93,11 +93,15 @@ def make_subtree_rec(node, subtree_attr='clades'):
 
 def subsample_tree(terminal_nodes, tree, p=0.1, subtree_attr='clades'):
     sampled = np.random.random(len(terminal_nodes))<p
+    if sampled.sum()<2:
+        print(f"not enough samples: {sampled.sum()} out of {len(terminal_nodes)}")
+
     for state, n in zip(sampled, terminal_nodes):
         n['sampled'] = state
 
     set_sampled_rec(tree, subtree_attr=subtree_attr)
     make_subtree_rec(tree, subtree_attr=subtree_attr)
+    return sampled.sum()
 
 def next_nontrivial_child(node, subtree_attr='children'):
     b = node

### 2024-04-22 src/estimate_diffusion_from_tree.py

@@ -41,13 +41,13 @@ def estimate_diffusion(tree, include_root=False):
 
 
 def estimate_ancestral_positions(tree, D):
-    def preorder(node):
+    def postorder(node):
 
         node['dis_to_parent'] = {'x':{'a':0, 'b':0, 'c':0}, 'y':{'a':0, 'b':0, 'c':0}}
         for c in node['clades']:
             dc = 1.0/(4*D*(c['time']-node['time']))
             if 'clades' in c and len(c['clades']):
-                preorder(c)
+                postorder(c)
                 for x in ['x', 'y']:
                     child_d = c['dis_to_parent'][x]
                     node['dis_to_parent'][x]['a'] += dc*child_d['a']/(dc + child_d['a'])
@@ -60,19 +60,19 @@ def estimate_ancestral_positions(tree, D):
                     node['dis_to_parent'][x]['a'] += dc
                     node['dis_to_parent'][x]['b'] += dc*c[x]
                     node['dis_to_parent'][x]['c'] += dc*c[x]**2 - 0.5*np.log(np.pi/dc)
-            for x in ['x', 'y']:
-                node['dis_to_parent'][x]['var'] = 2.0/node['dis_to_parent'][x]['a']
-                node['dis_to_parent'][x]['mean'] = 0.5*node['dis_to_parent'][x]['b']*node['dis_to_parent'][x]['var']
+        for x in ['x', 'y']:
+            node['dis_to_parent'][x]['var'] = 2.0/node['dis_to_parent'][x]['a']
+            node['dis_to_parent'][x]['mean'] = 0.5*node['dis_to_parent'][x]['b']*node['dis_to_parent'][x]['var']
 
 
-    preorder(tree)
+    postorder(tree)
     tree['dis_from_parent'] = {'x':{'a':0, 'b':0, 'c':0}, 'y':{'a':0, 'b':0, 'c':0}, 'dt': np.inf}
     tree['position'] = {}
     for x in ['x', 'y']:
         d = tree['dis_to_parent'][x]
         tree['position'][x] = {'var': 2.0/d['a'], 'mean': d['b']/d['a']}
 
-    def postorder(node):
+    def preorder(node):
         dn = 1.0/(4*D*node['dis_from_parent']['dt'])
         for c1 in node['clades']:
             c1['dis_from_parent'] = {'x':{'a':0, 'b':0, 'c':0}, 'y':{'a':0, 'b':0, 'c':0}, 'dt':c1['time']-node['time']}
[…]

### 2024-04-22 src/habitat_shifts.py

@@ -61,9 +61,15 @@ def diffusion_in_changing_habitats(D, interaction_radius, density_reg, N, subsam
             H, bx, by = get_2d_hist(terminal_nodes, Lx, Ly, linear_bins)
             density_variation.append(np.std(H)/N*np.prod(H.shape))
             for sample in range(n_subsamples):
-                subsample_tree(terminal_nodes, tree, p=subsampling, subtree_attr='clades')
-                D_res = estimate_diffusion(tree)
-                estimate_ancestral_positions(tree, D)
+                nsamples = subsample_tree(terminal_nodes, tree, p=subsampling, subtree_attr='clades')
+                if nsamples<3:
+                    print("population too small")
+                    continue
+                try:
+                    D_res = estimate_diffusion(tree)
+                    estimate_ancestral_positions(tree, D)
+                except:
+                    import ipdb; ipdb.set_trace()
                 z = collect_errors(tree)
                 if len(tree['clades'])==1:
                     Tmrca.append(t-tree['clades'][0]['time'])
@@ -79,6 +85,7 @@ def diffusion_in_changing_habitats(D, interaction_radius, density_reg, N, subsam
                              [np.mean(np.abs(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'x_err'])) for i in range(len(tbins)-1)])
                 y_err.append([np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'y_err']) for i in range(len(tbins)-1)] + 
                              [np.mean(np.abs(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'y_err'])) for i in range(len(tbins)-1)])
+
 
     return {"density_variation": density_variation, "D_est_x": D_est_x, "D_est_y": D_est_y, "D_est": D_est_x + D_est_y,
             "v_est_x": v_est_x, "v_est_y": v_est_y, "v_est": v_est_x + v_est_y,

### 2024-05-07 src/heterogeneity.py
update scripts

@@ -1,5 +1,6 @@
 from free_diffusion import *
 import matplotlib.pyplot as plt
+from parse_and_plot import *
 
 def get_2d_hist(terminal_nodes, Lx, Ly, bins):
     x_pos = (np.array([c['x'] for c in terminal_nodes]))%Lx
@@ -24,6 +25,7 @@ if __name__=="__main__":
 
     g, H = get_granularity(ntips, 0.1, bins=20)
     axs[0].matshow(H)
+    axs[0].set_axis_off()
 
     for line_style, linear_bins in zip(['-', '--', '-.'], [3,10,30]):
         nbins=linear_bins**2
@@ -42,7 +44,8 @@ if __name__=="__main__":
     #plt.plot(D_array, pred_heterogeneity, label='well mixed limit')
     plt.xscale('log')
     plt.yscale('log')
-    plt.xlabel('diffusion constant [L^2/T_c]')
+    plt.xlabel(r'diffusion constant $[L^2/T_c]$')
     plt.ylabel('heterogeneity (relative to well-mixed case)')
     plt.legend()
+    plt.tight_layout()
     plt.savefig('figures/heterogeneity_free_diffusion.pdf')

### 2024-05-07 src/parse_and_plot.py

@@ -1,0 +1,94 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+
+plt.rcParams.update({
+    "text.usetex": True,
+    "font.family": "Helvetica"
+})
+
+def parse_data(data, groupby=None):
+    dgb = data.groupby(groupby+['D'])
+
+    density_variation = dgb['density_variation'].mean()
+    nobs = data.groupby(groupby)['n'].mean()/25
+    symmetric = False
+    try:
+        diffusion_mean_x = dgb['meanDx'].mean()
+        diffusion_mean_y = dgb['meanDy'].mean()
+        v_mean_x = dgb['meanvx'].mean()
+        v_mean_y = dgb['meanvy'].mean()
+        diffusion_std_x = dgb['stdD_x'].mean()
+        diffusion_std_y = dgb['stdD_y'].mean()
+    except:
+        symmetric = True
+        diffusion_mean = dgb['meanD'].mean()
+        diffusion_std = dgb['stdD'].mean()
+
+    tmrca_mean = dgb['meanTmrca'].mean()
+    tmrca_std = dgb['stdTmrca'].mean()
+
+    x_err = {}
+    for g, d in dgb['x_err']:
+        x_err[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])[:5]), axis=0)
+
+    y_err = {}
+    for g, d in dgb['y_err']:
+        y_err[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])[:5]), axis=0)
+
+    x_err_abs = {}
+    for g, d in dgb['x_err']:
+        x_err_abs[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()][5:])), axis=0)
+
+    y_err_abs = {}
+    for g, d in dgb['y_err']:
+        y_err_abs[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])[5:]), axis=0)
+
+    if symmetric:
+        z_mean = {}
+        for g, d in dgb['meanZsq']:
+            z_mean[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
+
+        z_std = {}
+        for g, d in dgb['stdZsq']:
+            z_std[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
+    else:
[…]

### 2024-05-07 src/plot_periodic.py

@@ -1,29 +1,10 @@
 import numpy as np
 import matplotlib.pyplot as plt
 import pandas as pd
-from plot_stable_density import free_diffusion, parse_data
-from plot_waves import parse_data
+from plot_stable_density import free_diffusion
+from parse_and_plot import parse_data, make_figure
 from itertools import product
 from habitat_shifts import cycling_patches,breathing, seasaw
-
-plt.rcParams.update({
-    "text.usetex": True,
-    "font.family": "Helvetica"
-})
-
-def make_figure(fname=None):
-    fig, axs = plt.subplots(1,10, sharex=True, sharey=True, figsize=(15,2))
-    Lx, Ly = 3,1
-    f = seasaw(1, Lx, Ly, period=9, amplitude=1.1)
-    # f = breathing(1, Lx, Ly, period=9, width=0.25)
-    for i, ax in enumerate(axs.flatten()):
-        grid = np.meshgrid(np.linspace(0,Lx,30*Lx), np.linspace(0,Ly,30))
-        ax.matshow(f(grid[0], grid[1], i))
-        ax.set_axis_off()
-    plt.tight_layout()
-    if fname:
-        plt.savefig(fname)
-
 
 if __name__=="__main__":
     import argparse
@@ -40,6 +21,11 @@ if __name__=="__main__":
     Lx, Ly = 1, 1
     nbins=linear_bins**2
 
+    # make an illustration of the carrying capacity at different times
+    make_figure(seasaw, {'period':9, 'amplitude':1.1}, fname=args.illustration)
+    # make an illustration of the carrying capacity at different times
+    make_figure(breathing, {'period':9, 'width':0.5}, fname=args.illustration)
+
     res = parse_data(data, groupby=['interaction_radius', 'density_reg', 'N', 'period', 'subsampling'])
     interaction_radius = data.interaction_radius.unique()
     period = data.period.unique()
@@ -47,8 +33,6 @@ if __name__=="__main__":
     density_reg = data.density_reg.unique()
     N_vals = data.N.unique()
 
-    # make an illustration of the carrying capacity at different times
-    make_figure(fname = args.illustration)
 
     ## Population heterogeneity
     ls = ['-', ':', "--", ".-"][len(interaction_radius)]
@@ -141,30 +125,30 @@ if __name__=="__main__":
         plt.legend()
         plt.plot(N*D_array, np.zeros_like(D_array), c='k')
[…]

### 2024-05-07 src/plot_stable_density.py

@@ -3,30 +3,7 @@ import matplotlib.pyplot as plt
 import pandas as pd
 from heterogeneity import get_granularity
 from itertools import product
-
-
-def parse_data(data, groupby=None):
-    density_variation = data.groupby(groupby + ['D'])['density_variation'].mean()
-    nobs = data.groupby(groupby)['n'].mean()/25
-    diffusion_mean = data.groupby(groupby + ['D'])['meanD'].mean()
-    diffusion_std = data.groupby(groupby + ['D'])['stdD'].mean()
-    tmrca_mean = data.groupby(groupby + ['D'])['meanTmrca'].mean()
-    tmrca_std = data.groupby(groupby + ['D'])['stdTmrca'].mean()
-
-    z_mean = {}
-    for g, d in data.groupby(groupby + ['D'])['meanZsq']:
-        z_mean[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
-
-    z_std = {}
-    for g, d in data.groupby(groupby + ['D'])['stdZsq']:
-        z_std[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
-
-    return {"density_variation": density_variation, "nobs":nobs,
-            "diffusion_mean":diffusion_mean, "diffusion_std":diffusion_std,
-            "tmrca_mean":tmrca_mean, "tmrca_std":tmrca_std,
-            "z_mean": pd.DataFrame(z_mean).T, "z_std":pd.DataFrame(z_std).T,
-            }
-
+from parse_and_plot import parse_data
 
 def free_diffusion(D_array, N, linear_bins=5):
     n_iter = 100
@@ -75,7 +52,7 @@ if __name__=="__main__":
     plt.plot(D_array*N/Lx/Ly, free_diffusion_heterogeneity/np.sqrt(nbins/N), label='diffusion', c='k')
     plt.plot(D_array*N/Lx/Ly, np.ones_like(D_array), c='k', ls='--')
     plt.xscale('log')
-    plt.xlabel('diffusion constant [L^2/N]')
+    plt.xlabel('diffusion constant $[L^2/N]$')
     plt.ylabel('heterogeneity (relative to well-mixed case)')
     plt.legend()
     if args.output_heterogeneity:
@@ -112,7 +89,7 @@ if __name__=="__main__":
 
     plt.legend()
[…]

### 2024-05-07 src/plot_waves.py

@@ -3,74 +3,7 @@ import matplotlib.pyplot as plt
 import pandas as pd
 from itertools import product
 from habitat_shifts import waves
-
-plt.rcParams.update({
-    "text.usetex": True,
-    "font.family": "Helvetica"
-})
-
-def parse_data(data, groupby=None):
-    dgb = data.groupby(groupby+['D'])
-
-    density_variation = dgb['density_variation'].mean()
-    nobs = data.groupby(groupby)['n'].mean()/25
-    diffusion_mean_x = dgb['meanDx'].mean()
-    diffusion_mean_y = dgb['meanDy'].mean()
-    v_mean_x = dgb['meanvx'].mean()
-    v_mean_y = dgb['meanvy'].mean()
-    diffusion_std_x = dgb['stdD_x'].mean()
-    diffusion_std_y = dgb['stdD_y'].mean()
-    tmrca_mean = dgb['meanTmrca'].mean()
-    tmrca_std = dgb['stdTmrca'].mean()
-
-    x_err = {}
-    for g, d in dgb['x_err']:
-        x_err[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])[:5]), axis=0)
-
-    y_err = {}
-    for g, d in dgb['y_err']:
-        y_err[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])[:5]), axis=0)
-
-    x_err_abs = {}
-    for g, d in dgb['x_err']:
-        x_err_abs[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()][5:])), axis=0)
-
-    y_err_abs = {}
-    for g, d in dgb['y_err']:
-        y_err_abs[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])[5:]), axis=0)
-
-    z_mean = {}
-    for g, d in dgb['meanZsq_x']:
-        z_mean[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
-
-    z_std = {}
-    for g, d in dgb['stdZsq_x']:
-        z_std[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
-
-    return {"density_variation": density_variation, "nobs":nobs,
-            "diffusion_mean_x":diffusion_mean_x, "diffusion_std_x":diffusion_std_x,
[…]

### 2024-05-07 src/stable_density.py
make periodicBC optional

@@ -4,7 +4,7 @@ from heterogeneity import get_2d_hist
 from estimate_diffusion_from_tree import estimate_diffusion, estimate_ancestral_positions, collect_errors
 
 
-def evolve_stable_density(D, interaction_radius, density_reg, N, subsampling=1.0,
+def evolve_stable_density(D, interaction_radius, density_reg, N, subsampling=1.0, periodic=False,
                                 Lx=1, Ly=1, linear_bins=5, n_iter=10, n_subsamples=1):
     from scipy.stats import scoreatpercentile
     # set up tree and initial population uniformly in space
@@ -21,7 +21,7 @@ def evolve_stable_density(D, interaction_radius, density_reg, N, subsampling=1.0
     y_err = []
     for t in range((n_iter+10)*N):
         terminal_nodes = evolve(terminal_nodes, t, Lx=Lx, Ly=Ly, interaction_radius=interaction_radius,
-                                density_reg=density_reg, D=D, target_density=N, total_population=N, periodic=False)
+                                density_reg=density_reg, D=D, target_density=N, total_population=N, periodic=periodic)
         if len(terminal_nodes)<10:
             print("population nearly extinct")
             continue
@@ -48,9 +48,9 @@ def evolve_stable_density(D, interaction_radius, density_reg, N, subsampling=1.0
                 # calculate the mean squared z-scores for the root node and each time bin
                 zscores.extend([[z.iloc[root_index].zx**2]+[np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'zx']**2) for i in range(len(tbins)-1)],
                                 [z.iloc[root_index].zy**2]+[np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'zy']**2) for i in range(len(tbins)-1)]])
-                x_err.append([np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'x_err']) for i in range(len(tbins)-1)] + 
+                x_err.append([np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'x_err']) for i in range(len(tbins)-1)] +
                              [np.mean(np.abs(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'x_err'])) for i in range(len(tbins)-1)])
[…]

### 2024-05-07 src/estimate_diffusion_from_tree.py
fix root inclusion

@@ -15,10 +15,12 @@ def estimate_diffusion(tree, include_root=False):
     displacements = []
 
     # handle root-parent-branch contribution separately
-    if len(tree['clades'])==1 and include_root:
-        #advance to first furcating node
-        data = branch_contribution(tree, tree['x'], tree['y'], tree['time'])
-        displacements.append(data)
+    if len(tree['clades'])==1:
+        c=tree['clades'][0]
+        if include_root:
+            #advance to first furcating node
+            data = branch_contribution(tree['clades'][0], tree['x'], tree['y'], tree['time'])
+            displacements.append(data)
     else:
         c=tree
 
@@ -153,4 +155,3 @@ def collect_errors(tree):
                 collect_errors_rec(c)
     collect_errors_rec(tree)
     return pd.DataFrame(res)
-    

### 2024-05-07 manuscript/phylogeo.tex
manuscript edits

@@
 Phylogeographic methods are implemented in popular evolutionary analysis software such as BEAST \citep{pybus_unifying_2012}.
-The underlying model of dispersal is typically diffusive, that is the probability of sampling a descendant at position $\rvec_c$ after time $t$ when the parent was at position $\rvec_p$ is given by
+The underlying model of dispersal is typically assumed to be diffusive, that is the probability of sampling a descendant at position $\rvec_c$ after time $t$ when the parent was at position $\rvec_p$ is given by
 \begin{equation}
@@
 where $D$ is a diffusion constant with dimensions $\mathrm{length}^2/\mathrm{time}$, and we have assumed a two-dimensional space.
-Each branch in the phylogeny is decorated with such a diffusive propagator.
 Such a diffusive process is the simplest and most natural choice in absence of directed motion or long range migration.
@@
 tree topology and branching rates of the tree are assumed to be independent of the spatial process.
-However, these two interact in important ways, for example when a population expands into an uninhabited fertile habitat or a disease spreads in a susceptible population.
+
+However, population dynamics is often tightly coupled to spatial location, for example when a population expands into an uninhabited fertile habitat or a disease spreads in a susceptible population, and the independence assumption of spatial and phylogenetic processes is not justified.
+The extent to which violations of this assumption affect phylogeographic inferences is not well understood.
+To accommodate variation not captured by the simple diffusion model, ``relaxed random walk'' models allow the diffusion constant $D$ to vary across the tree (the diffusion constant is integrated over a prior distribution) \citep{dellicour_relax_2021}.
 
@@
 \begin{equation}
+    \label{eq:disperal_parameters}
     D_w = \frac{\sum_{i\in B}d_i^2}{4\sum_{i\in B} \Delta t_i} \quad \mathrm{and}  \quad v_w = \frac{\sum_{i\in B} d_i}{\sum_{i\in B} \Delta t_i}
@@
 
-Fig.~\ref{fig:D_and_v} shows empirical estimates of diffusion constants and velocity as defined above from simulated data using freely diffusing locations along Kingman trees.
-The diffusion constant $D$ can be estimated robustly from these data and are compatible with the diffusion constant used to simulate the data $D=1$.
+Fig.~\ref{fig:D_and_v} shows empirical estimates of diffusion constants and velocity as defined in Eq.~\ref{eq:disperal_parameters} from simulated data using freely diffusing locations along Kingman trees.
+The diffusion constant $D$ can be estimated robustly from these data and the estimates are compatible with the diffusion constant used to simulate the data $D=1$.
 Values of $v_w$ and $v_b$, however, depend strongly on the sample size.
@@
 Diffusion along each branch results in a displacement of $d_i \sim \sqrt{4Dt_i}$ such that a quantity that depends on $\Delta d_i / t_i \sim 1/t_i^{1/2}$ will change with the time $t_i$ that elapsed along the branch.
-With more samples in the tree, the average branch length decreases and the velocity estimates increase. 
+With more samples in the tree, the average branch length decreases, and the velocity estimate increases.
 This expected scaling $\sim \sqrt{n}$ is indicated by the dashed line in Fig.~\ref{fig:D_and_v}.
 For Yule trees, the average branch length is independent of the number of samples and the velocity estimates are more stable.
-But estimates for $v_w$ and $v_b$ are not consistent with each other and their numerical values have no interpretation within the model.
+But estimates for $v_w$ and $v_b$ are not consistent with each other.
 
@@
 The underlying model is diffusive and has no parameters that individually or combined would have dimensions of a velocity.
-There is no ground truth to compare the estimates to and $v_w$ and $v_b$ are just descriptive summaries of spatial spread with unclear interpretation.
+There is no ground truth to compare the estimates to and $v_w$ and $v_b$ are just descriptive summaries of spatial spread that can not be compared between datasets.
 
@@
 Phylogeographic inference typically assumes that the spatial diffusion process is independent of the branching pattern of the tree and ignores coupling between the two.
-Such a coupling could, for example, arise when growth rates depend on local population density, and it has been known for a long time that ignoring such coupling leads to very unrealistic population densities \citep{felsenstein_pain_1975}.
-If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D<L^2/T_c$, where $T_c$ is the coalescent times scale of the population (see Fig.~\ref{fig:heterogeneity}).
+Such a coupling, however, is expected since growth rates naturally depend on local population density when organisms compete for resources.
+It has been known for a long time that ignoring such coupling leads to unrealistically heterogeneous population densities \citep{felsenstein_pain_1975}.
+If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D<L^2/T_c$, where $T_c$ is the coalescent timescale of the population (see Fig.~\ref{fig:heterogeneity}).
 In a well-mixed neutral model, $T_c$ is equal to the population size $N$.
@@
 The left panel shows the population density for the case $D = 0.1 L^2/N$ in linear bins of size $L/20$.
-The right panels shows the standard deviation of the population density for different bin sizes as a function of the diffusion coefficient.
+The right panel shows the standard deviation of the population density for different bin sizes as a function of the diffusion coefficient.
 For $D>L^2/N$, the heterogeneity approaches the expected value in the well mixed limit.
@@
 As soon as the interaction radius $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}.
-Diffusion constants estimated from the population remain accurate with moderate overestimation of small diffusion constants.
+Diffusion constants estimated from the population using Eq.~\ref{eq:dispersal} remain accurate with moderate overestimation of small diffusion constants.
+If instead of periodic boundary conditions reflective boundary conditions are used, diffusion constants are underestimated once $\sqrt{2DT_c}$ becomes comparable to the habitat size.
+This underestimation will then result in over-confident inference of ancestral locations, in particular for recent ancestral nodes.
+
 But with decreasing diffusion, the population fragments into smaller than smaller subpopulations that don't interact or coalesce.
@@
 
-To explore such the result of such changing habitats on phylogeographic inferences, we simulated a habitat with a Gaussian carrying capacity that moves at a constant velocity $v$ along the $x$-axis (Fig.~\ref{fig:traveling_wave}).
-When varying the microscopic dispersal parameter $D$ and the velocity of the carrying capacity $v$, I observe diverse behavior with often dramatically overestimated $\hat{D}$.
+To explore such the result of such changing habitats on phylogeographic inferences, we simulated a habitat with a Gaussian carrying capacity that moves at a constant velocity $v$ along the $x$-axis (Fig.~\ref{fig:traveling_wave}) with periodic boundary conditions.
+When varying the microscopic dispersal parameter $D$ and the velocity $v$ of the carrying capacity profile, I observe diverse behavior with often dramatically overestimated $\hat{D}$.
 These behaviors are explained comparing the speed $v_{FKPP}$ at which the population can invade new territory with the external speed $v$, the ratio of which is used to plot the simulation data in Fig.~\ref{fig:traveling_wave}.

### 2024-05-08 src/estimate_diffusion_from_tree.py
further root exclusion

@@ -135,7 +135,7 @@ def collect_positioning(tree):
             res[-1]['subtree_y_std'] = node['dis_to_parent']['y']['var']**0.5
             for c in node['clades']:
                 collect_positioning_rec(c, res)
-    collect_positioning_rec(tree, res)
+    collect_positioning_rec(tree['clades'][0] if len(tree['clades'])==1 else tree, res)
     return res
 
 def collect_errors(tree):
@@ -153,5 +153,5 @@ def collect_errors(tree):
         if nonterminal:
             for c in node['clades']:
                 collect_errors_rec(c)
-    collect_errors_rec(tree)
+    collect_errors_rec(tree['clades'][0] if len(tree['clades'])==1 else tree)
     return pd.DataFrame(res)

### 2024-05-08 src/plot_stable_density.py

@@ -35,21 +35,21 @@ if __name__=="__main__":
     N_vals = data.N.unique()
 
     # select the interaction radius and density regulation strength to plot
-    ir_to_plot = interaction_radius[:]
+    ir_to_plot = interaction_radius[1:]
     density_reg_to_plot = density_reg[1:2]
 
     # Plot the heterogeneity
-    ls = ['-', '-.', "--", ":"][:len(density_reg_to_plot)]
+    ls = ['-', '-.', "--", ":"] #[:len(density_reg_to_plot)]
     plt.figure()
-    for N in N_vals[1:]:
+    for N in N_vals[1:2]:
         for i, (ir, dr) in enumerate(product(ir_to_plot, density_reg_to_plot)):
             D_array = np.array(res["density_variation"][ir, dr, N, :].index)
             label = f'r={ir}' if N==N_vals[1] else ''
             plt.plot(D_array*N/Lx/Ly, res["density_variation"][ir, dr, N, :]/np.sqrt(nbins/N),
-                    label=label, ls=ls[i%len(ls)], c=f"C{i//len(ls)}")
+                    label=label, ls=ls[i%len(ls)], c=f"C{i%10}")
 
         free_diffusion_heterogeneity = free_diffusion(D_array*N/Lx/Ly, N, linear_bins=linear_bins)
-    plt.plot(D_array*N/Lx/Ly, free_diffusion_heterogeneity/np.sqrt(nbins/N), label='diffusion', c='k')
+    plt.plot(D_array*N/Lx/Ly, free_diffusion_heterogeneity/np.sqrt(nbins/N), label='free diffusion', c='k')
     plt.plot(D_array*N/Lx/Ly, np.ones_like(D_array), c='k', ls='--')
     plt.xscale('log')
     plt.xlabel('diffusion constant $[L^2/N]$')
@@ -60,13 +60,13 @@ if __name__=="__main__":
 
     ## Plot the diffusion estiamtes
     plt.figure()
-    for N in N_vals:
+    for ni,N in enumerate(N_vals):
         for i, (ir, dr) in enumerate(product(ir_to_plot, density_reg_to_plot)):
             D_array = np.array(res["diffusion_mean"][ir, dr, N, :].index)
             label = f'r={ir}, a={dr}' if N==N_vals[0] else ''
             plt.errorbar(D_array*N, res["diffusion_mean"][ir, dr, N, :]/D_array,
[…]

### 2024-05-08 src/stable_density.py

@@ -15,10 +15,15 @@ def evolve_stable_density(D, interaction_radius, density_reg, N, subsampling=1.0
 
     density_variation = []
     D_est = []
-    zscores = []
+    zscores_x = []
+    zscores_y = []
     Tmrca = []
     x_err = []
     y_err = []
+    x_err_abs = []
+    y_err_abs = []
+    x_err_sq = []
+    y_err_sq = []
     for t in range((n_iter+10)*N):
         terminal_nodes = evolve(terminal_nodes, t, Lx=Lx, Ly=Ly, interaction_radius=interaction_radius,
                                 density_reg=density_reg, D=D, target_density=N, total_population=N, periodic=periodic)
@@ -34,27 +39,25 @@ def evolve_stable_density(D, interaction_radius, density_reg, N, subsampling=1.0
                 subsample_tree(terminal_nodes, tree, p=subsampling, subtree_attr='clades')
                 D_res = estimate_diffusion(tree)
                 estimate_ancestral_positions(tree, D)
-                z = collect_errors(tree)
-                if len(tree['clades'])==1:
-                    Tmrca.append(t-tree['clades'][0]['time'])
-                    root_index = 1
-                else:
-                    Tmrca.append(t)
-                    root_index = 0
+                z = collect_errors(tree) # this excludes the founding node, first node should be tree root
+                root_index = 0
 
                 internal_node_times= sorted(z.loc[z.nonterminal, 't'])
                 tbins = scoreatpercentile(internal_node_times, [0, 20, 40, 60, 80, 100])
                 D_est.extend([D_res['Dx_total'], D_res['Dy_total']])
                 # calculate the mean squared z-scores for the root node and each time bin
-                zscores.extend([[z.iloc[root_index].zx**2]+[np.mean(z.loc[(z.t >= tbins[i]) & (z.t<tbins[i+1]), 'zx']**2) for i in range(len(tbins)-1)],
[…]

### 2024-05-08 src/breathing.py
refactor

@@ -1,5 +1,6 @@
 import numpy as np
-from habitat_shifts import breathing, diffusion_in_changing_habitats
+from habitat_shifts import breathing
+from density_regulation import run_simulation, serialize_result
 
 
 if __name__=="__main__":
@@ -19,9 +20,7 @@ if __name__=="__main__":
     N = args.N
     width = 0.5
     Lx, Ly = 3, 1
-    res_density = {}
-    res_density_mean = {}
-    D_est = []
+    results = []
     D_array_dens = np.logspace(-3,0,21)*Lx*Ly*2/N
     n_iter = 50
     linear_bins=5
@@ -29,37 +28,17 @@ if __name__=="__main__":
     print(f"{interaction_radius=:1.3f}, {density_reg=:1.3f}")
     for di, D in enumerate(D_array_dens):
         print(f"{di} out of {len(D_array_dens)}: D={D:1.3e}")
-        res = diffusion_in_changing_habitats(D, interaction_radius, density_reg, N, subsampling=args.subsampling,
+        res = run_simulation(D, interaction_radius, density_reg, N, subsampling=args.subsampling,
                                           Lx=Lx, Ly=Ly, linear_bins=linear_bins, n_iter=n_iter, periodic=False,
                                           gtd=breathing, habitat_params={'period':args.period, 'width':width})
         if len(res['D_est_x'])==0:
             print("no data")
             continue
-        tmpD_x = np.mean(res["D_est_x"], axis=0)
-        tmpD_y = np.mean(res["D_est_y"], axis=0)
-        tmpStdD_x = np.std(res["D_est_x"], axis=0)
-        tmpStdD_y = np.std(res["D_est_y"], axis=0)
-        tmpv_x = np.mean(res["v_est_x"], axis=0)
-        tmpv_y = np.mean(res["v_est_y"], axis=0)
-        tmpStdv_x = np.std(res["v_est_x"], axis=0)
-        tmpStdv_y = np.std(res["v_est_y"], axis=0)
-        tmpZ_x =    f"[{' '.join(str(x) for x in np.ma.mean(np.ma.masked_invalid(res['zscores_x']), axis=0).filled(fill_value=np.nan))}]"
-        tmpStdZ_x = f"[{' '.join(str(x) for x in np.ma.std(np.ma.masked_invalid(res['zscores_x']), axis=0).filled(fill_value=np.nan))}]"
[…]

### 2024-05-08 src/waves.py

@@ -1,5 +1,5 @@
 import numpy as np
-from habitat_shifts import waves, diffusion_in_changing_habitats
+from habitat_shifts import waves, run_simulation, serialize_result
 
 
 if __name__=="__main__":
@@ -18,9 +18,7 @@ if __name__=="__main__":
     N = args.N
     width = 0.5
     Lx, Ly = 3, 1
-    res_density = {}
-    res_density_mean = {}
-    D_est = []
+    results = []
     D_array_dens = np.logspace(-3,0,21)*Lx*Ly*2/N
     n_iter = 50
     linear_bins=5
@@ -28,35 +26,14 @@ if __name__=="__main__":
     print(f"{interaction_radius=:1.3f}, {density_reg=:1.3f}")
     for di, D in enumerate(D_array_dens):
         print(f"{di} out of {len(D_array_dens)}: D={D:1.3e}")
-        res = diffusion_in_changing_habitats(D, interaction_radius, density_reg, N, subsampling=args.subsampling,
-                                          Lx=Lx, Ly=Ly, linear_bins=linear_bins, n_iter=n_iter, 
-                                          gtd=waves, habitat_params={'velocity':args.velocity, 'width':width})
+        res = run_simulation(D, interaction_radius, density_reg, N, subsampling=args.subsampling,
+                            Lx=Lx, Ly=Ly, linear_bins=linear_bins, n_iter=n_iter,
+                            gtd=waves, habitat_params={'velocity':args.velocity, 'width':width})
+        processed_res = serialize_result(res)
+        processed_res.update({"interaction_radius":interaction_radius, "density_reg": density_reg,
+                            "N": N, "subsampling": args.subsampling, "D":D, "velocity": args.velocity})
 
-        tmpD_x = np.mean(res["D_est_x"], axis=0)
-        tmpD_y = np.mean(res["D_est_y"], axis=0)
-        tmpStdD_x = np.std(res["D_est_x"], axis=0)
-        tmpStdD_y = np.std(res["D_est_y"], axis=0)
-        tmpv_x = np.mean(res["v_est_x"], axis=0)
-        tmpv_y = np.mean(res["v_est_y"], axis=0)
-        tmpStdv_x = np.std(res["v_est_x"], axis=0)
-        tmpStdv_y = np.std(res["v_est_y"], axis=0)
[…]

### 2024-05-10 manuscript/phylogeo.tex
add to ms

@@
 
-\begin{figure*}[tb]
-    \includegraphics*[width=\textwidth]{figures/heterogeneity_free_diffusion.pdf}
-    \caption[short]{\label{fig:heterogeneity}{\bf Population density variation in free diffusion models.}
+\begin{figure*}
+    \includegraphics[width=0.48\textwidth]{figures/heterogeneity_free_diffusion}
+    \includegraphics[width=0.48\textwidth]{figures/stable_density}
+    \caption{\label{fig:density_reg}  {\bf Population density variation in free diffusion models.}
     When spatial motion is independent of population density and coalesce is sufficiently rapid, the realized population density fluctuates strongly across the habitat.
@@
 For $D>L^2/N$, the heterogeneity approaches the expected value in the well mixed limit.
-}
+    {\bf Density regulation reduces fluctuations in density regulation, but results in population fragmentation at low diffusion constants.} The left panel shows population heterogeneity measured using bins of linear dimension $L/5$ for $\alpha =0.05$ for different values of the interaction radius $r$.
+For $r\ll L$, density fluctuations are strongly suppressed.
+At the same time, the time to the most recent common ancestor increases dramatically, indicating that the population fragments in subpopulations that don't interact.}
 \end{figure*}
@@
 As soon as the interaction radius $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}.
-Diffusion constants estimated from the population using Eq.~\ref{eq:dispersal} remain accurate with moderate overestimation of small diffusion constants.
+Despite the substantial reduction in density fluctuations, the estimates of dispersal rates using Eq.~\ref{eq:disperal_parameters} are not substantially affected, at least when boundary conditions are periodic.
 If instead of periodic boundary conditions reflective boundary conditions are used, diffusion constants are underestimated once $\sqrt{2DT_c}$ becomes comparable to the habitat size.
@@
 Spatial location and the coalescence process are thus strongly coupled, though this coupling manifests itself in population structure rather than estimates of the diffusion coefficients.
-In such a spatially structured population, estimates of historical populations sizes (e.g.~skyline estimates) will be strongly affected.
+Hence both at high and at low diffusion constants, assumptions of phylogeographic inference can be problematic: if diffusion constants are much smaller than then $L^2/T_{mrca}$, the population is fragmented into effective subpopulations that violate the typical tree-priors, if $D\geq L^/T_{mrca}$, the boundaries of the domain restrict free diffusion.
 
-\begin{figure*}
-    \includegraphics[width=0.48\textwidth]{figures/density_reg_heterogeneity.pdf}
-    \includegraphics[width=0.48\textwidth]{figures/density_reg_tmrca.pdf}
-    \caption{\label{fig:density_reg} {\bf Density regulation reduces fluctuations in density regulation, but results in population fragmentation at low diffusion constants.} The left panel shows population heterogeneity measured using bins of linear dimension $L/5$ for $\alpha =0.05$ for different values of the interaction radius $r$.
-For $r\ll L$, density fluctuations are strongly suppressed.
-At the same time, the time to the most recent common ancestor increases dramatically, indicating that the population fragments in subpopulations that don't interact.}
-\end{figure*}
-
-As discussed above, a static carrying capacity in an equilibrium population has limited effects on estimates of ancestral locations and diffusion coefficients.
+\section*{How do changing habitats affect phylogeographic inferences?}
+As discussed above, a static carrying capacity in an equilibrium population has limited effects on estimates of ancestral locations and diffusion coefficients beyond boundary effects.
 In reality, carrying capacities vary through time.
@@
 
-A simple model of the interplay between growth and spatial spread is again given by the Fisher-KPP equation (Eq.~\ref{eq:FKPP}) above.
+In Fig.~\ref{fig:seasaw} we consider a case where the region with highest carrying capacity shifts from the left to the right in a periodic manner.
+The carrying capacity is illustrated in Fig.~\ref{fig:seasaw}A for 5 time points covering half a period.
+To avoid extinction at low dispersal rates, selection is soft meaning that the overall population size is kept approximately constant.
+If the dispersal rate is sufficiently high, the population follows the shifting carrying capacity and lineage in the tree undergo directed motion.
+When the population sampled at a time when the population has just contracted to the left (see Fig.~\ref{fig:seasaw}B), the position of older nodes is not estimated correctly.
+Their true positions are to the right of the current population, but the inferred positions are central in the shifted population.
+
+This bias depends on the period at which the carrying capacity oscillates and the dispersal rate, which determines how quickly the population can follow the shifting carrying capacity.
+Fig.~\ref{fig:seasaw}C shows the estimated diffusion constant (using Eq.~\ref{eq:disperal_parameters}) in $x$ and $y$ direction relative to the true value used in the simulations.
+At very high diffusion constants, both $D_x$ and $D_y$ are underestimated due to spatial constraint of carrying capacity.
+With decreasing $D_x$, it starts to be overestimated until at some point the population no longer follows the shifting carrying capacity.
+The panel Fig.~\ref{fig:seasaw}D shows $z_x^2 = \langle (\hat{x} - x)^2/\sigma_x^2 \rangle$ and $z_y^2=\langle (\hat{y} - y)^2/\sigma_y^2\rangle$ for the oldest 20\% of internal nodes.
+For free diffusion, $z_x^2$ should be 1, but for $x$ it is substantially larger, indicating that inferences are over-confident.
+Conversely, $z_y^2$ is smaller than 1 since positions are constrained by the habitat.
+
+\begin{figure*}
+    \includegraphics[width=\textwidth]{figures/seasaw_illustration}
+    \includegraphics[width=\textwidth]{figures/seasaw_example}
+    \includegraphics[width=\textwidth]{figures/seasaw}
+    \caption{\label{fig:seasaw} {\bf Phylogeographic estimates in rapidly changing environments.} Panel A shows an illustration of the simulated set-up: The carrying capacity first peaks on the left, becomes flat, and then peaks on the right before shifting back.
+The habitat size is $L_x=3$, $L_y=1$.
+    Panel B shows a tree from a simulation under these conditions with $D=0.0004$ and $N=500$.
+The panel on the right shows the inferred (triangles) and true positions (circles) of internal nodes, highlight the root with large markers.
+Blue and green colors correspond to older nodes.
+    Panels C show estimates of diffusion coefficients if $x$ (solid lines) and $y$ (dashed lines) relative to their true values as a function of the $D$.
+The right panel shows the average $z_x^2$ and $z_y^2$ for the oldest 20\% of internal nodes.}
+\end{figure*}
+
+The above example and summaries show that behavior of a population in shifting habitats and the interpretation of phylogeographic inferences depends strongly one whether dispersal is fast enough that the population can follow the shifting carrying capacity.
+If the population can not follow the situation is similar to the static case often leading to fragmented populations.
+Once the population starts following, inferred ancestral locations of deep nodes become unreliable.
+The dispersal rates necessary to keep up with the shifting carrying capacity can be estimated from Fisher-KPP equation (Eq.~\ref{eq:FKPP}) above.
 This equation admits solutions with a traveling front, that the population ``invades'' the empty territory with velocity $v_{FKPP} = 2\sqrt{D \alpha}$ \citep{fisher_wave_1937,KPP1937,hallatschek_life_2010}.
@@
 
-To explore such the result of such changing habitats on phylogeographic inferences, we simulated a habitat with a Gaussian carrying capacity that moves at a constant velocity $v$ along the $x$-axis (Fig.~\ref{fig:traveling_wave}) with periodic boundary conditions.
-When varying the microscopic dispersal parameter $D$ and the velocity $v$ of the carrying capacity profile, I observe diverse behavior with often dramatically overestimated $\hat{D}$.
+To explore such the result of such changing habitats on phylogeographic inferences, I simulated a habitat with a Gaussian carrying capacity that moves at a constant velocity $v$ along the $x$-axis (Fig.~\ref{fig:traveling_wave}) with periodic boundary conditions.
+Unsurprisingly, estimates of $\hat{D}$ using Eq.~\ref{eq:dispersal_parameters} are often dramatically overestimated (by up to factors of 1000).
+In contrast to the above examples with no persistent directionality, in this example there is a sense in which a velocity estimate might be meaningful.
+Fig.~\ref{fig:traveling_wave}C shows estimates of $v$ using Eq.~\ref{eq:dispersal_parameters} relative to the true velocity.
+This ratio is monotonically increasing agrees with the diffusion coefficient and shows a range where it close to 1, i.e.~where the estimated velocity of lineages agrees with the external velocity.
+
 These behaviors are explained comparing the speed $v_{FKPP}$ at which the population can invade new territory with the external speed $v$, the ratio of which is used to plot the simulation data in Fig.~\ref{fig:traveling_wave}.
@@
 For larger $v_{FKPP}$, the estimated velocity is stable for some time, before increasing again with $v_{FKKP}$ is a sampling dependent manner.
-At larger $v_{FKPP}$, $D$ is eventually underestimates.
-
-In this extreme example of a population dragged along by a moving carrying capacity, estimates of dispersal rates are meaningless, and estimates of velocities are only compatible with the speed of the shifting habitat in a narrow range when the habitat speed matches the FKPP velocity.
-Similarly, estimates of ancestral locations are biased to the current position of the habitat.
-
-Less extreme spatio-temporal dynamics of the habitat results in less extreme biases, but still illustrate that caution is required when interpreting phylogeographic inferences in environment that might change between the present and the $T_{MRCA}$ of the population.
-I simulated two additional scenarios: one habitat that expands and contracts to a core region periodically, and one that oscillates back and forth between two regions with a homogeneous intermediate state.
+At larger $v_{FKPP}$, $D$ is eventually underestimated since the Gaussian population profile hinders dispersal.
 
 \begin{figure}
-    \includegraphics[width=0.48\textwidth]{figures/traveling_wave}
+    \includegraphics[width=0.3\textwidth]{figures/traveling_wave}
     \includegraphics[width=0.48\textwidth]{figures/waves_D}
@@
 \end{figure}
+
+\section*{Discussion}
 
@@
 \end{itemize}
-
-\section*{Discussion}
 

### 2024-05-10 src/density_regulation.py

@@ -177,7 +177,8 @@ def evolve(terminals, t, Lx=1, Ly=1, D=0.1, target_density = 100.0, density_reg 
         target_density_vals = target_density
 
     # calculate fitness of each extant individual
-    fitness = np.maximum(0.1,1 + density_reg*(1-dens_array/target_density_vals))
+    scale_factor = np.mean(target_density_vals)/np.mean(dens_array)
+    fitness = np.maximum(0.1,1 + density_reg*(1-scale_factor*dens_array/target_density_vals))
     #print(fitness.mean(), dens_array.mean(), target_density, len(terminals))
     if global_pop_reg: # add global density regulation (set average fitness to one, add density independent term)
         fitness += (1 - fitness.mean()) + 0.1*(1-len(terminals)/total_population)

### 2024-05-10 src/habitat_shifts.py

@@ -22,7 +22,8 @@ def breathing(N, Lx, Ly, period, width):
 
 def seasaw(N, Lx, Ly, period, amplitude=0.9):
     def f(x,y,t):
-        return N*np.maximum(0.01,np.minimum(1,0.5*(0.1+amplitude*np.cos(2*np.pi*t/period)*np.cos(np.pi*x/Lx)**1)))
+        v = np.maximum(0.03,np.minimum(1,0.5*(0.1+amplitude*np.cos(2*np.pi*t/period)*np.cos(np.pi*x/Lx)**1)))
+        return N*v
     return f
 
 def test_density(Lx, Ly, tmax, gtd=None, habitat_params=None):

### 2024-05-10 src/heterogeneity.py

@@ -21,7 +21,7 @@ if __name__=="__main__":
     ntips = 1000
     D_array = np.logspace(-2,2,101)
 
-    fig, axs = plt.subplots(1,2, figsize=(12,6))
+    fig, axs = plt.subplots(2,1, figsize=(6,9))
 
     g, H = get_granularity(ntips, 0.1, bins=20)
     axs[0].matshow(H)

### 2024-05-10 src/parse_and_plot.py

@@ -11,7 +11,7 @@ def parse_data(data, groupby=None):
     dgb = data.groupby(groupby+['D'])
 
     density_variation = dgb['density_variation'].mean()
-    nobs = data.groupby(groupby)['n'].mean()/25
+    nobs = data.groupby(groupby)['observations'].mean()/25
     symmetric = False
     try:
         diffusion_mean_x = dgb['meanDx'].mean()
@@ -52,22 +52,21 @@ def parse_data(data, groupby=None):
     for g, d in dgb['y_err_sq']:
         y_err_sq[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
 
-    if symmetric:
-        z_mean = {}
-        for g, d in dgb['meanZsq']:
-            z_mean[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
+    z_mean_x = {}
+    for g, d in dgb['meanZsq_x']:
+        z_mean_x[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
 
-        z_std = {}
-        for g, d in dgb['stdZsq']:
-            z_std[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
-    else:
-        z_mean = {}
-        for g, d in dgb['meanZsq_x']:
-            z_mean[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
+    z_std_x = {}
+    for g, d in dgb['stdZsq_x']:
+        z_std_x[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
 
-        z_std = {}
-        for g, d in dgb['stdZsq_x']:
-            z_std[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
+    z_mean_y = {}
+    for g, d in dgb['meanZsq_y']:
+        z_mean_y[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
+
+    z_std_y = {}
+    for g, d in dgb['stdZsq_y']:
+        z_std_y[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
[…]

### 2024-05-10 src/plot_periodic.py

@@ -18,13 +18,13 @@ if __name__=="__main__":
     args = parser.parse_args()
     data = pd.read_csv(args.data)
     linear_bins=5
-    Lx, Ly = 1, 1
+    Lx, Ly = 3, 1
     nbins=linear_bins**2
 
     # make an illustration of the carrying capacity at different times
-    make_figure(seasaw, {'period':9, 'amplitude':1.1}, fname=args.illustration)
+    make_figure(seasaw, {'period':8, 'amplitude':1.1}, fname=args.illustration)
     # make an illustration of the carrying capacity at different times
-    make_figure(breathing, {'period':9, 'width':0.5}, fname=args.illustration)
+    make_figure(breathing, {'period':8, 'width':0.5}, fname=args.illustration)
 
     res = parse_data(data, groupby=['interaction_radius', 'density_reg', 'N', 'period', 'subsampling'])
     interaction_radius = data.interaction_radius.unique()
@@ -34,73 +34,77 @@ if __name__=="__main__":
     N_vals = data.N.unique()
 
 
-    ## Population heterogeneity
     ls = ['-', ':', "--", ".-"][len(interaction_radius)]
     markers = ['o', '<', '>', 's', 'd', '^', 'v']
-    plt.figure()
-    for N in N_vals:
-        D_array = data.loc[data.N==N].D.unique()
-        area = np.minimum(1,D_array*np.pi**2)
-        for i, (ir, dr, T, p) in enumerate(product(interaction_radius, density_reg,
-                                                   period, subsampling)):
-            try:
-                plt.plot(D_array*N/Lx/Ly, res["density_variation"][ir, dr, N, T, p,:],
-                        label=f'r={ir}, a={dr} N={N}, T={T} p={p}', ls=ls[i%len(ls)], c=f"C{i//len(ls)}")
-            except:
-                pass
 
-        free_diffusion_heterogeneity = free_diffusion(D_array*N/Lx/Ly, N, linear_bins=linear_bins)
-        plt.plot(D_array*N/Lx/Ly, free_diffusion_heterogeneity, label='diffusion', c='k')
-        plt.plot(D_array*N/Lx/Ly, np.ones_like(D_array)*np.sqrt(nbins/N), label='well mixed limit', c='k', ls='--')
-    plt.xscale('log')
-    plt.xlabel('diffusion constant $D$')
-    plt.ylabel('heterogeneity')
[…]

### 2024-05-10 src/plot_seasaw.py

@@ -1,0 +1,87 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+from heterogeneity import get_granularity
+from itertools import product
+from parse_and_plot import parse_data, make_figure
+from habitat_shifts import seasaw
+
+def free_diffusion(D_array, N, linear_bins=5):
+    n_iter = 100
+    res = []
+    for D in D_array:
+        res.append([get_granularity(N, D, bins=linear_bins)[0] for i in range(n_iter)])
+
+    res = np.array(res)
+    return res.mean(axis=1)
+
+if __name__=="__main__":
+    import argparse
+    parser = argparse.ArgumentParser()
+    parser.add_argument('--data', type=str, default='data/seasaw.csv')
+    args = parser.parse_args()
+    data = pd.read_csv(args.data)
+    Lx, Ly = 3, 1
+
+    make_figure(seasaw, {'period':8, 'amplitude':1.1}, fname="figures/seasaw_illustration.pdf")
+
+    res = parse_data(data, groupby=['interaction_radius', 'density_reg', 'N', 'period', 'subsampling'])
+    interaction_radius = data.interaction_radius.unique()
+    period = data.period.unique()
+    subsampling = data.subsampling.unique()
+    density_reg = data.density_reg.unique()
+    N_vals = data.N.unique()
+
+    fig, axs = plt.subplots(1,2, figsize=(12,3))
+
+    ## Diffusion constant estimates
+    ir = interaction_radius[1]
+    dr = density_reg[2]
+    p = 1.0
+    ax=axs[0]
+    for ti,T in enumerate([200, 500]):
+        for m, N in zip(['o', '<'], N_vals):
+            D_array = data.loc[data.N==N].D.unique()
+            x_vals = D_array*N/Ly**2
+            ax.errorbar(x_vals, res['diffusion_mean_x'][ir, dr, N, T, p, :]/D_array,
+                                res['diffusion_std_x'][ir, dr, N, T, p, :]/D_array/np.sqrt(res['nobs'][ir, dr, N, T, p]), marker=m,
+                                label=f'N={N}, T={T}', ls='-', c=f"C{ti%10}")
[…]

### 2024-05-10 src/stable_density_figure.py

@@ -1,0 +1,131 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+from heterogeneity import get_granularity
+from itertools import product
+from parse_and_plot import parse_data
+
+def free_diffusion(D_array, N, linear_bins=5):
+    n_iter = 100
+    res = []
+    for D in D_array:
+        res.append([get_granularity(N, D, bins=linear_bins)[0] for i in range(n_iter)])
+
+    res = np.array(res)
+    return res.mean(axis=1)
+
+if __name__=="__main__":
+    import argparse
+    parser = argparse.ArgumentParser()
+    parser.add_argument('--data', type=str, default='data/stable_density.csv')
+    parser.add_argument('--dataPBC', type=str, default='data/stable_density_periodicBC.csv')
+    args = parser.parse_args()
+
+    linear_bins=5
+    Lx, Ly = 1, 1
+    nbins=linear_bins**2
+
+    data = pd.read_csv(args.data)
+    res = parse_data(data, groupby=['interaction_radius', 'density_reg', 'N'])
+
+    dataPBC = pd.read_csv(args.dataPBC)
+    resPBC = parse_data(dataPBC, groupby=['interaction_radius', 'density_reg', 'N'])
+
+    interaction_radius = data.interaction_radius.unique()
+    density_reg = data.density_reg.unique()
+    N_vals = data.N.unique()
+
+    # figure
+    fig, axs = plt.subplots(4, 1, figsize=(6, 9), sharex=True)
+
+    # Plot the heterogeneity
+    ls = ['-', '-.', "--", ":"] #[:len(density_reg_to_plot)]
+    N=500
+    dr = density_reg[1]
+    ax=axs[0]
+    for i, ir in enumerate(interaction_radius[1:]):
+        D_array = np.array(resPBC["density_variation"][ir, dr, N, :].index)
+        label = f'r={ir}' if N==N_vals[1] else ''
+        ax.plot(D_array*N/Lx/Ly, resPBC["density_variation"][ir, dr, N, :]/np.sqrt(nbins/N),
+                label=label, ls='-', c=f"C{i%10}")
+    free_diffusion_heterogeneity = free_diffusion(D_array*N/Lx/Ly, N, linear_bins=linear_bins)
[…]

### 2024-05-10 src/waves_example.py

@@ -4,7 +4,7 @@ from density_regulation import make_node, evolve, subsample_tree, dict_to_phylo_
 from estimate_diffusion_from_tree import estimate_diffusion, estimate_ancestral_positions
 import matplotlib.pyplot as plt
 from Bio import Phylo
-from matplotlib.cm import viridis_r 
+from matplotlib.cm import viridis_r
 
 
 if __name__=="__main__":
@@ -27,10 +27,10 @@ if __name__=="__main__":
     D = args.D
     interaction_radius, density_reg = args.interaction_radius, args.density_reg
     #target_density = waves(N, Lx, Ly, **{'velocity':args.velocity, "width":width})
-    # period = 500
-    # target_density = seasaw(N, Lx, Ly, **{'period':period, 'amplitude':1.1})
     period = 500
-    target_density = breathing(N, Lx, Ly, **{'period':period, 'width':0.5})
+    target_density = seasaw(N, Lx, Ly, **{'period':period, 'amplitude':1.1})
+    # period = 1000
+    #target_density = breathing(N, Lx, Ly, **{'period':period, 'width':0.5})
 
     v_FKPP = 2*np.sqrt(args.velocity*D)
     print(f"{interaction_radius=:1.3f}, {density_reg=:1.3f}")
@@ -41,7 +41,7 @@ if __name__=="__main__":
     terminal_nodes = tree['children']
 
     t = 0
-    for tmax in [5*N + i*period/6.0 for i in range(7)]:
+    for tmax in [5*N + i*period/8.0 for i in range(9)]:
         while t<tmax:
             terminal_nodes = evolve(terminal_nodes, t, Lx=Lx, Ly=Ly, interaction_radius=interaction_radius,
                                 density_reg=density_reg, D=D, target_density=target_density, total_population=N, periodic=False)
@@ -49,29 +49,44 @@ if __name__=="__main__":
 
         subsample_tree(terminal_nodes, tree, p=1.0, subtree_attr='clades')
         D_res = estimate_diffusion(tree)
-        estimate_ancestral_positions(tree, D)
-        phylo_tree = dict_to_phylo_tree(tree['clades'][0], child_attr='clades')
+        estimate_ancestral_positions(tree, D_res['Dxy_total'])
+        phylo_tree = dict_to_phylo_tree(tree['clades'][0] if len(tree['clades'])==1 else tree,
[…]

### 2024-06-01 manuscript/phylogeo.tex
add to ms

@@
 \begin{figure*}
-    \includegraphics[width=0.48\textwidth]{figures/heterogeneity_free_diffusion}
-    \includegraphics[width=0.48\textwidth]{figures/stable_density}
+    \includegraphics[width=0.46\textwidth]{figures/heterogeneity_free_diffusion}
+    \includegraphics[width=0.53\textwidth]{figures/stable_density}
     \caption{\label{fig:density_reg}  {\bf Population density variation in free diffusion models.}
@@
 
-As soon as the interaction radius $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}.
-Despite the substantial reduction in density fluctuations, the estimates of dispersal rates using Eq.~\ref{eq:disperal_parameters} are not substantially affected, at least when boundary conditions are periodic.
+As soon as the interaction radius $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}C.
+Despite the substantial reduction in density fluctuations, the estimates of dispersal rates using Eq.~\ref{eq:disperal_parameters} are not substantially affected, at least when boundary conditions are periodic, see Fig.~\ref{fig:density_reg}D.
 If instead of periodic boundary conditions reflective boundary conditions are used, diffusion constants are underestimated once $\sqrt{2DT_c}$ becomes comparable to the habitat size.
-This underestimation will then result in over-confident inference of ancestral locations, in particular for recent ancestral nodes.
+This underestimation of diffusion constants will then result in over-confident inference of ancestral locations, in particular for recent ancestral nodes.
 
-But with decreasing diffusion, the population fragments into smaller than smaller subpopulations that don't interact or coalesce.
-This manifests in an increasing time to the MRCA, see Fig.~\ref{fig:density_reg}.
-Spatial location and the coalescence process are thus strongly coupled, though this coupling manifests itself in population structure rather than estimates of the diffusion coefficients.
+But with decreasing diffusion, the population fragments into smaller and smaller subpopulations that don't interact or coalesce.
+This manifests in an increasing time to the MRCA, see Fig.~\ref{fig:density_reg}E.
+Spatial location and the coalescence process are thus strongly coupled, though this coupling manifests itself in population structure rather than biased estimates of the diffusion coefficients.
 Hence both at high and at low diffusion constants, assumptions of phylogeographic inference can be problematic: if diffusion constants are much smaller than then $L^2/T_{mrca}$, the population is fragmented into effective subpopulations that violate the typical tree-priors, if $D\geq L^/T_{mrca}$, the boundaries of the domain restrict free diffusion.
@@
 In reality, carrying capacities vary through time.
-Such variation happens on all time scales, ranging from environmental changes over millennia to seasonal fluctuations.
+Such variation happens on all time scales, ranging from geological timescales over millions of years, environmental changes over millennia to seasonal fluctuations.
 The circulation of seasonal influenza viruses, for example, varies by many orders of magnitude between summer and winter in temperate climates.
@@
 The carrying capacity is illustrated in Fig.~\ref{fig:seasaw}A for 5 time points covering half a period.
-To avoid extinction at low dispersal rates, selection is soft meaning that the overall population size is kept approximately constant.
-If the dispersal rate is sufficiently high, the population follows the shifting carrying capacity and lineage in the tree undergo directed motion.
-When the population sampled at a time when the population has just contracted to the left (see Fig.~\ref{fig:seasaw}B), the position of older nodes is not estimated correctly.
+To avoid extinction at low dispersal rates, selection is soft, meaning that the overall population size is kept approximately constant even if the entire population is stuck in the unfavorable region.
+If the dispersal rate is sufficiently high, however, the population follows the shifting carrying capacity and lineages in the tree undergo directed back-and-forth motion.
+Inference of ancestral locations in such situations can be misleading as illustrated in Fig.~\ref{fig:seasaw}B.
+The population was sampled at a time when it has just contracted to the left and the positions of older nodes is not estimated correctly.
 Their true positions are to the right of the current population, but the inferred positions are central in the shifted population.
 
-This bias depends on the period at which the carrying capacity oscillates and the dispersal rate, which determines how quickly the population can follow the shifting carrying capacity.
-Fig.~\ref{fig:seasaw}C shows the estimated diffusion constant (using Eq.~\ref{eq:disperal_parameters}) in $x$ and $y$ direction relative to the true value used in the simulations.
-At very high diffusion constants, both $D_x$ and $D_y$ are underestimated due to spatial constraint of carrying capacity.
-With decreasing $D_x$, it starts to be overestimated until at some point the population no longer follows the shifting carrying capacity.
-The panel Fig.~\ref{fig:seasaw}D shows $z_x^2 = \langle (\hat{x} - x)^2/\sigma_x^2 \rangle$ and $z_y^2=\langle (\hat{y} - y)^2/\sigma_y^2\rangle$ for the oldest 20\% of internal nodes.
-For free diffusion, $z_x^2$ should be 1, but for $x$ it is substantially larger, indicating that inferences are over-confident.
-Conversely, $z_y^2$ is smaller than 1 since positions are constrained by the habitat.
+The severity of such biases depends on the period at which the carrying capacity oscillates and on the dispersal rate, see  Fig.~\ref{fig:seasaw}C.
+Fig.~\ref{fig:seasaw}D shows the estimated diffusion constant (using Eq.~\ref{eq:disperal_parameters}) in $x$ (solid lines) and $y$ (dashed) direction relative to the true value used in the simulations.
+At very high diffusion constants, both $D_x$ and $D_y$ are underestimated due to spatial constraint of carrying capacity as also discussed above (see Fig.~\ref{fig:stable_density}).
+With decreasing $D$, $D_x$ starts to be overestimated and peaks at a value that depends on the period $T$, before decreasing to (noisy) estimates that are broadly compatible with the true value.
+In this range of dispersal rates, the population can no longer follow the shifting carrying capacity.
+This decoupling happens at lower values of $D$ for longer periods of environmental change.
+With the parameter used in the simulation, the estimates of $D_x$ deviate by a factor of 2 from its true value, which is not dramatic compared to the variation of dispersal rates in nature.
+
+However, along with problematic inference of diffusion coefficients, estimates of ancestral locations are over-confident.
+Panel Fig.~\ref{fig:seasaw}E shows $z_x^2 = \langle (\hat{x} - x)^2/\sigma_x^2 \rangle$ and $z_y^2=\langle (\hat{y} - y)^2/\sigma_y^2\rangle$ for the oldest 20\% of internal nodes.
+If the model estimated ancestral locations consistently, $z_x^2$ should be 1 on average.
+Instead, it is as high as 20 for the simulations shown and this over-confident and biased estimations depend in non-monotonic ways on the true diffusion constant and the period of environmental change.
+
+At high dispersal rates, the average $z_x^2$ and $z_y$ are both smaller than one since the estimated uncertainty is larger than the habitat.
+With decreasing $D$, $z_x^2$ increases largely because the expected uncertainty decreases, before peaking when the population can no longer follow the shifting carrying capacity.
+A peak value of 10-20 indicates that the typical ancestral location of the deep nodes is 2-4 standard deviations away from the true location, an example of which is shown in Fig.~\ref{fig:seasaw}C.
 
@@
 The habitat size is $L_x=3$, $L_y=1$.
-    Panel B shows a tree from a simulation under these conditions with $D=0.0004$ and $N=500$.
-The panel on the right shows the inferred (triangles) and true positions (circles) of internal nodes, highlight the root with large markers.
-Blue and green colors correspond to older nodes.
-    Panels C show estimates of diffusion coefficients if $x$ (solid lines) and $y$ (dashed lines) relative to their true values as a function of the $D$.
-The right panel shows the average $z_x^2$ and $z_y^2$ for the oldest 20\% of internal nodes.}
+    Panel B shows a tree from a simulation under these conditions with $D=0.0008$ and $N=500$, while panel C right shows the inferred (triangles) and true positions (circles) of internal nodes, highlight the root with large markers.
+Blue and green colors correspond to older nodes, yellow to recent nodes.
+    Panels D show estimates of diffusion coefficients in directions $x$ (solid lines) and $y$ (dashed lines) relative to their true values as a function of the true value of $D$.
+Panel E shows the average $z_x^2$ and $z_y^2$ for the oldest 20\% of internal nodes.
+Inference of the x-coordinate of ancestral nodes is drastically overconfident for some values of $D$.
+}
 \end{figure*}
 
-The above example and summaries show that behavior of a population in shifting habitats and the interpretation of phylogeographic inferences depends strongly one whether dispersal is fast enough that the population can follow the shifting carrying capacity.
-If the population can not follow the situation is similar to the static case often leading to fragmented populations.
-Once the population starts following, inferred ancestral locations of deep nodes become unreliable.
-The dispersal rates necessary to keep up with the shifting carrying capacity can be estimated from Fisher-KPP equation (Eq.~\ref{eq:FKPP}) above.
+The above results show that behavior of a population in shifting habitats and the interpretation of phylogeographic inferences depend strongly on the rate of environmental change and the dispersal rate.
+If the population can not follow shifting habitat, the situation is similar to the static case often leads to fragmented populations.
+Once the population starts following, inferred ancestral locations of deep nodes become unreliable, while at high dispersal rates expected uncertainties are larger than the habitat size and there is little information ancestral locations.
+
+To study the interplay of moving habitats and phylogeographic inference more systematically, I will now consider a situation where the population is constrained to a stripe with a Gaussian profile that moves at a constant velocity $v$ along the $x$-axis (Fig.~\ref{fig:traveling_wave}).
+The dispersal rates necessary to keep up with the shifting carrying capacity can be estimated from the Fisher-KPP equation (Eq.~\ref{eq:FKPP}) above.
 This equation admits solutions with a traveling front, that the population ``invades'' the empty territory with velocity $v_{FKPP} = 2\sqrt{D \alpha}$ \citep{fisher_wave_1937,KPP1937,hallatschek_life_2010}.
@@
 The speed at which the population invades space is given as compound of diffusion and accelerated growth in regions of low density and only grows like the square root of diffusion constant.
-
-To explore such the result of such changing habitats on phylogeographic inferences, I simulated a habitat with a Gaussian carrying capacity that moves at a constant velocity $v$ along the $x$-axis (Fig.~\ref{fig:traveling_wave}) with periodic boundary conditions.
-Unsurprisingly, estimates of $\hat{D}$ using Eq.~\ref{eq:dispersal_parameters} are often dramatically overestimated (by up to factors of 1000).
-In contrast to the above examples with no persistent directionality, in this example there is a sense in which a velocity estimate might be meaningful.
-Fig.~\ref{fig:traveling_wave}C shows estimates of $v$ using Eq.~\ref{eq:dispersal_parameters} relative to the true velocity.
-This ratio is monotonically increasing agrees with the diffusion coefficient and shows a range where it close to 1, i.e.~where the estimated velocity of lineages agrees with the external velocity.
-
-These behaviors are explained comparing the speed $v_{FKPP}$ at which the population can invade new territory with the external speed $v$, the ratio of which is used to plot the simulation data in Fig.~\ref{fig:traveling_wave}.
-For $v_{FKPP} \ll v$, the population cannot keep up with the moving carrying capacity and is saved from extinction only because selection is soft.
-In this regime, the estimated velocity of lineages is below $v$ and both the estimated $v$ and $D$ increase with $v_{FKPP}$.
-Once $v\sim v_{FKPP}$, the estimated lineage velocity coincides with $v_{FKPP}$, while the estimated $D$ peaks at values orders of magnitude above the true value.
-For larger $v_{FKPP}$, the estimated velocity is stable for some time, before increasing again with $v_{FKKP}$ is a sampling dependent manner.
-At larger $v_{FKPP}$, $D$ is eventually underestimated since the Gaussian population profile hinders dispersal.
 
@@
 
+Fig \ref{fig:traveling_wave}B\&C show simulation results for this traveling habitat.
+In Panel B, estimates of $D_x$ using Eq.~\ref{eq:dispersal_parameters} are shown as a function of the ratio of the FKPP velocity to the external velocity $v$.
+As expected, estimates of $\hat{D}_x$ are often dramatically too high (by up to factors of 1000).
+They peak right around when the external velocity matches the FKPP velocity, corresponding to the situation where the population dynamics is maximally effected by the moving carrying capacity.
+If the external velocity was lower, undirected dispersal contributes more and more to the displacement of lineages, while the population would be unable to follow the moving carrying capacity if the external velocity was higher.
+
+In contrast to the above examples with no persistent directionality, one can now compare velocity estimates to a velocity $v_{FKPP}$ defined by the model.
+Fig.~\ref{fig:traveling_wave}C shows estimates of $v$ using Eq.~\ref{eq:dispersal_parameters} relative to the true velocity.
+This ratio is monotonically increasing with the diffusion coefficient and shows a range where it close to 1, i.e.~where the estimated velocity of lineages agrees with the external velocity.
+
+These behaviors are explained comparing the speed $v_{FKPP}$ at which the population can invade new territory with the external speed $v$, the ratio of which is used to plot the simulation data in Fig.~\ref{fig:traveling_wave}.
+For $v_{FKPP} \ll v$, the population cannot keep up with the moving carrying capacity and is saved from extinction only because selection is soft.
+In this regime, the estimated velocity of lineages is below $v$ and both the estimated $v$ and $D$ increase with $v_{FKPP}$.
+Once $v\sim v_{FKPP}$, the estimated lineage velocity coincides with $v_{FKPP}$, while the estimated $D$ peaks at values orders of magnitude above the true value.
+For larger $v_{FKPP}$, the estimated velocity is stable for some time, before increasing again with $v_{FKKP}$.
+In this latter regime, diffusive motions dominates over directed motion and the estimates of $v$ suffer from the problems shown in Fig.~\ref{fig:D_and_v}.
+
 \section*{Discussion}
+Phylogeography aims to reconstruct location and migration of ancestors from the spatial distribution of extant individuals and their genome sequences.
+The latter allow the reconstruction of the phylogenetic tree which encodes how closely different individuals are related to each other.
+Together with the spatial location of the samples -- the leafs of the tree -- one can reconstruct the likely locations of ancestors and estimate parameters of dispersal.
+Popular tools for such inference, however, make strong assumptions about the dispersal process, including assuming that dispersal of individuals is diffusive (i.e.~characterized by small displacements in random directions) and that spatial location is independent of the replication process.
+In this note, I have explored the robustness of such inference to violations of the latter assumption while keeping microscopic dispersal strictly diffusive, that is without any long-range migration.
+I showed that common summary statistics of dispersal rates are often problematic and that interactions between dispersal and replication can affect estimates of ancestral locations in ways that are hard to control.
 
-\begin{itemize}
-    \item periodic seasaws and other patters
-    \item can we trust ancestral location inferences? with true D, with estimated D? significant biases?
-    \item can we say something general? coalescence time versus diffusion time across the habitat, habitat changes vs fkpp speed
-    \item typical FKPP speed: $D=km^2/day$, $\alpha = 0.2/day$. $v_{FKPP} = 2\sqrt{D\alpha} = km/day$.
-\end{itemize}
+The common practice to estimates ``velocities'' of lineages using displacements of lineages along the tree \citep{dellicour_relax_2021} is problematic.
+Firstly, the numerical value of such estimates depends strongly on the sample size and tree ensemble.
+Secondly, the underlying model is diffusive and has no parameters that have dimensions of a velocity and even if the quantity could be estimated robustly, there is no interpretation of the quantity within the model.
+
+A more complex set of questions concerns how coupling between growth rate of the population and spatial density affects phylogeographic inferences and how robust these inferences are to shifting habitats.
+As has long been known, ignoring the coupling between population growth and spatial location can lead to unrealistic population densities \citep{felsenstein_pain_1975}.
+Simple density regulation leads to homogeneous population densities, but inferences of dispersal rates and ancestral locations using models that ignore this coupling are not substantially affected.
+However, the population dynamics strongly depends on whether dispersal is fast enough to explore the entire habitat during the coalescence time $T_c$ of the population in absence of spatial structure.
+If $D\ll L^2/T_c$, the population fragments into subpopulations and the time to the most recent common ancestor increases dramatically.
+The resulting strong coupling between spatial location and coalescence process violates typical tree priors used in such inference.
+In the other case of $D\gg L^2/T_c$, dispersal is rapid enough that ancestral locations of deep nodes are very uncertain compared to the size of the habitat.
+
+Interpretation of such inference becomes even more challenging when habitats shift in time.
+Depending on parameters, dispersal rates $D$ can be overestimated or underestimated, sometimes by large factors, and estimates of ancestral locations can be overconfident or too uncertain.
+Simulations reveal that this behavior depends on the relative magnitude of the rate at which the habitat shifts and the FKPP velocity which depends on the diffusion constant $D$ and the growth rate in empty regions $\alpha$, as well as curtailed dispersal by the boundaries of the habitat.
+The parameter range where both ancestral locations and dispersal parameters are inferred correctly is narrow.
+
+Inference using phylogeographic methods should thus be interpreted with caution.
+In many cases the habitat of a population has undergone dramatic and recurrent changes since the $MRCA$ and such changes will affect inferences in ways that are not captured by the models.
+The ability to infer ancestral locations is further limited by long-range dispersal \citep{hallatschek_acceleration_2014}.
+To capture such deviations from simple diffusive motions, many inference models assume a ``relaxed random walk'', where diffusion constant can vary across the tree according to a broad prior distribution \citep{dellicour_relax_2021}.
+However, while such models will often generate a better fit, they don't capture complex interactions between spatial location and population dynamics.
 

### 2024-06-01 manuscript/phylogeo.tex
move figures around

@@
 \begin{document}
-\title{When to trust phylogeographic inferences?}
-\author{Richard Neher}
+\title{On the robustness of phylogeographic inference to model assumptions}
+\author{Richard A.~Neher}
 \date{\today}
 \begin{abstract}
-    Sometimes.
+Phylogeographic inference in continuous space is a popular method to reconstruct the spatial distribution of ancestral populations and estimate parameters of the dispersal process.
+While the underlying probabilistic models can be complex and computationally demanding to infer, they typically ignore that replication and population growth are tightly coupled to spatial location.
+Such coupling, in particular when habitats change over time, can result in biased and overconfident estimates of ancestral locations and dispersal parameters.
+These distortions depend in complicated ways the past dynamics of habitats and underlying population dynamics and dispersal processes such that validity of phylogeographic inferences is hard to assess.
 \end{abstract}
@@
 To accommodate variation not captured by the simple diffusion model, ``relaxed random walk'' models allow the diffusion constant $D$ to vary across the tree (the diffusion constant is integrated over a prior distribution) \citep{dellicour_relax_2021}.
+
+\begin{figure*}[tb]
+    \includegraphics*[width=0.7\textwidth]{figures/illustration_tree.pdf}
+    \caption{\label{fig:illustration_tree}{\bf Illustration Yule and Kingman trees and the spatial location of the tips.}
+    The total tree length of Yule trees is proportional to the number of tips $n$ and the time to the MRCA is $\log(n)$.
+    Kingman trees have a total tree length that is proportional to $\log(n)$ and are characterized by rapid merging close to the present. }
+\end{figure*}
 
@@
 
-\begin{figure*}[tb]
-    \includegraphics*[width=0.7\textwidth]{figures/illustration_tree.pdf}
-    \caption{\label{fig:illustration_tree}{\bf Illustration Yule and Kingman trees and the spatial location of the tips.}
-    The total tree length of Yule trees is proportional to the number of tips $n$ and the time to the MRCA is $\log(n)$.
-    Kingman trees have a total tree length that is proportional to $\log(n)$ and are characterized by rapid merging close to the present. }
-\end{figure*}
+\begin{figure}[tb]
+    \includegraphics*[width=0.48\textwidth]{figures/kingman_dispersal.pdf}
+    \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
+    While diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ (weighted) and $v_b$ (by branch) are not well-defined quantities.
+Their numerical value depends strongly on sample size and there is no ground truth to compare to.
+The expected scaling $\sim \sqrt{n}$ of $v$ estimates is indicated by the dashed line.}
+\end{figure}
 
@@
 There is no ground truth to compare the estimates to and $v_w$ and $v_b$ are just descriptive summaries of spatial spread that can not be compared between datasets.
-
-\begin{figure}[tb]
-    \includegraphics*[width=0.48\textwidth]{figures/kingman_dispersal.pdf}
-    \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
-    While diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ (weighted) and $v_b$ (by branch) are not well-defined quantities.
-Their numerical value depends strongly on sample size and there is no ground truth to compare to.
-The expected scaling $\sim \sqrt{n}$ of $v$ estimates is indicated by the dashed line.}
-\end{figure}
 
@@
 
+\begin{figure*}
+    \includegraphics[width=\textwidth]{figures/seasaw_illustration}
+    \includegraphics[width=\textwidth]{figures/seasaw_example}
+    \includegraphics[width=\textwidth]{figures/seasaw}
+    \caption{\label{fig:seasaw} {\bf Phylogeographic estimates in rapidly changing environments.} Panel A shows an illustration of the simulated set-up: The carrying capacity first peaks on the left, becomes flat, and then peaks on the right before shifting back.
+The habitat size is $L_x=3$, $L_y=1$.
+    Panel B shows a tree from a simulation under these conditions with $D=0.0008$ and $N=500$, while panel C right shows the inferred (triangles) and true positions (circles) of internal nodes, highlight the root with large markers.
+Blue and green colors correspond to older nodes, yellow to recent nodes.
+    Panels D show estimates of diffusion coefficients in directions $x$ (solid lines) and $y$ (dashed lines) relative to their true values as a function of the true value of $D$.
+Panel E shows the average $z_x^2$ and $z_y^2$ for the oldest 20\% of internal nodes.
+Inference of the x-coordinate of ancestral nodes is drastically overconfident for some values of $D$.
+}
+\end{figure*}
+
 \section*{How do changing habitats affect phylogeographic inferences?}
@@
 A peak value of 10-20 indicates that the typical ancestral location of the deep nodes is 2-4 standard deviations away from the true location, an example of which is shown in Fig.~\ref{fig:seasaw}C.
-
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/seasaw_illustration}
-    \includegraphics[width=\textwidth]{figures/seasaw_example}
-    \includegraphics[width=\textwidth]{figures/seasaw}
-    \caption{\label{fig:seasaw} {\bf Phylogeographic estimates in rapidly changing environments.} Panel A shows an illustration of the simulated set-up: The carrying capacity first peaks on the left, becomes flat, and then peaks on the right before shifting back.
-The habitat size is $L_x=3$, $L_y=1$.
-    Panel B shows a tree from a simulation under these conditions with $D=0.0008$ and $N=500$, while panel C right shows the inferred (triangles) and true positions (circles) of internal nodes, highlight the root with large markers.
-Blue and green colors correspond to older nodes, yellow to recent nodes.
-    Panels D show estimates of diffusion coefficients in directions $x$ (solid lines) and $y$ (dashed lines) relative to their true values as a function of the true value of $D$.
-Panel E shows the average $z_x^2$ and $z_y^2$ for the oldest 20\% of internal nodes.
-Inference of the x-coordinate of ancestral nodes is drastically overconfident for some values of $D$.
-}
-\end{figure*}
 

### 2024-06-01 manuscript/phylogeo.tex
add refs

@@
 Phylogeographic inference in continuous space is a popular method to reconstruct the spatial distribution of ancestral populations and estimate parameters of the dispersal process.
-While the underlying probabilistic models can be complex and computationally demanding to infer, they typically ignore that replication and population growth are tightly coupled to spatial location.
+While the underlying probabilistic models can be complex and computationally demanding to infer, they typically ignore that replication and population growth are tightly coupled to spatial location: populations expand into fertile uninhabited areas and contract in regions with limited resources.
 Such coupling, in particular when habitats change over time, can result in biased and overconfident estimates of ancestral locations and dispersal parameters.
-These distortions depend in complicated ways the past dynamics of habitats and underlying population dynamics and dispersal processes such that validity of phylogeographic inferences is hard to assess.
+These distortions depend in complicated ways on the past dynamics of habitats and underlying population dynamics and dispersal processes such that validity of phylogeographic inferences is hard to assess.
 \end{abstract}
@@
 The reconstruction of ancestral locations and past migrations from sampling locations of extant individuals, typically in combination with genome sequence information to infer the phylogeny, is known as phylogeography.
-Phylogeographic methods are implemented in popular evolutionary analysis software such as BEAST \citep{pybus_unifying_2012}.
+Phylogeographic methods are implemented in popular evolutionary analysis software such as BEAST \citep{pybus_unifying_2012,lemey_phylogeography_2010}.
 The underlying model of dispersal is typically assumed to be diffusive, that is the probability of sampling a descendant at position $\rvec_c$ after time $t$ when the parent was at position $\rvec_p$ is given by
@@
 \end{equation}
-where $D$ is a diffusion constant with dimensions $\mathrm{length}^2/\mathrm{time}$, and we have assumed a two-dimensional space.
+where $D$ is a diffusion constant with dimensions $\mathrm{length}^2/\mathrm{time}$, and we have assumed a two-dimensional space and ignored anisotropic dispersal.
 Such a diffusive process is the simplest and most natural choice in absence of directed motion or long range migration.
@@
 However, population dynamics is often tightly coupled to spatial location, for example when a population expands into an uninhabited fertile habitat or a disease spreads in a susceptible population, and the independence assumption of spatial and phylogenetic processes is not justified.
-The extent to which violations of this assumption affect phylogeographic inferences is not well understood.
-To accommodate variation not captured by the simple diffusion model, ``relaxed random walk'' models allow the diffusion constant $D$ to vary across the tree (the diffusion constant is integrated over a prior distribution) \citep{dellicour_relax_2021}.
+Such coupling can be accommodated in structured coalescent models where the population is divided into discrete subpopulations that exchange migrants at a rate $m$ \citep{vaughan_efficient_2014}.
+But the extent to which violations of the independence assumption affect continuous phylogeographic inferences is not well understood.
+To accommodate variation not captured by the simple diffusion model, ``relaxed random walk'' models allow the diffusion constant $D$ to vary across the tree (the diffusion constant is integrated over a prior distribution) \citep{dellicour_relax_2021,lemey_phylogeography_2010}.
 
@@
 The weighted dispersal velocity estimate compares the sum of absolute values of observed displacements to the total time, which has dimensions of a velocity.
-Alternative formulations use the mean of fractions $D_b = |B|^{-1} \sum_{i\in B}d_i^2/4t_i$ or $v_b = |B|^{-1} \sum_{i\in B}d_i/t_i$ instead of the ratio of sums. \
+Alternative formulations use the mean of fractions $D_b = |B|^{-1} \sum_{i\in B}d_i^2/4t_i$ or $v_b = |B|^{-1} \sum_{i\in B}d_i/t_i$ instead of the ratio of sums.
 We refer to these here as `by branch' estimates.
-The latter tend to be dominated by short branches and thus noisier.
+The latter tend to be dominated by short branches and thus noisier \citep{trovao_bayesian_2015}.
 
@@
 There is no ground truth to compare the estimates to and $v_w$ and $v_b$ are just descriptive summaries of spatial spread that can not be compared between datasets.
+
+In cases where a population invades a novel habitat, e.g.~the invasion of North America by the West-Nile virus \citep{pybus_unified_2010}, the moving front of the population defines a bona-fide velocity.
+However, as discussed below, this velocity is determined by the interplay of dispersal and population growth at the front.
+The velocity is better determined by modeling the position of the front, rather than by tracking individual lineages along the tree.
 
@@
 In many cases the habitat of a population has undergone dramatic and recurrent changes since the $MRCA$ and such changes will affect inferences in ways that are not captured by the models.
+In addition, uneven sampling, or complete lack of samples from some regions, can undermine phylogeographic estimates \citep{kalkauskas_sampling_2021}.
 The ability to infer ancestral locations is further limited by long-range dispersal \citep{hallatschek_acceleration_2014}.
@@
 However, while such models will often generate a better fit, they don't capture complex interactions between spatial location and population dynamics.
+
+In absence of a good understanding of the dispersal properties and how population dynamics varies and time and space, simple models that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models \citep{layan_impact_2023}.
 

### 2024-06-01 manuscript/phylogeo.tex
edits

@@
 \begin{document}
-\title{On the robustness of phylogeographic inference to model assumptions}
+\title{On the robustness and consistency of phylogeographic inference}
 \author{Richard A.~Neher}
@@
 \begin{equation}
-    \label{eq:disperal_parameters}
+    \label{eq:dispersal_parameters}
     D_w = \frac{\sum_{i\in B}d_i^2}{4\sum_{i\in B} \Delta t_i} \quad \mathrm{and}  \quad v_w = \frac{\sum_{i\in B} d_i}{\sum_{i\in B} \Delta t_i}
@@
 The weighted dispersal velocity estimate compares the sum of absolute values of observed displacements to the total time, which has dimensions of a velocity.
-Alternative formulations use the mean of fractions $D_b = |B|^{-1} \sum_{i\in B}d_i^2/4t_i$ or $v_b = |B|^{-1} \sum_{i\in B}d_i/t_i$ instead of the ratio of sums.
+Alternative formulations use the mean of fractions $D_b = |B|^{-1} \sum_{i\in B}d_i^2/4\Delta t_i$ or $v_b = |B|^{-1} \sum_{i\in B}d_i/\Delta t_i$ instead of the ratio of sums.
 We refer to these here as `by branch' estimates.
@@
 
-Fig.~\ref{fig:D_and_v} shows empirical estimates of diffusion constants and velocity as defined in Eq.~\ref{eq:disperal_parameters} from simulated data using freely diffusing locations along Kingman trees.
+Fig.~\ref{fig:D_and_v} shows empirical estimates of diffusion constants and velocity as defined in Eq.~\ref{eq:dispersal_parameters} from simulated data using freely diffusing locations along Kingman trees.
 The diffusion constant $D$ can be estimated robustly from these data and the estimates are compatible with the diffusion constant used to simulate the data $D=1$.
@@
 That the velocity estimates are ill-defined is not surprising: By changing the sample size, the average length of branches changes.
-Diffusion along each branch results in a displacement of $d_i \sim \sqrt{4Dt_i}$ such that a quantity that depends on $\Delta d_i / t_i \sim 1/t_i^{1/2}$ will change with the time $t_i$ that elapsed along the branch.
+Diffusion along each branch results in a displacement of $d_i \sim \sqrt{4D\Delta t_i}$ such that a quantity that depends on $d_i / \Delta t_i \sim 1/\Delta t_i^{1/2}$ will change with the time $\Delta t_i$ that elapsed along the branch.
 With more samples in the tree, the average branch length decreases, and the velocity estimate increases.
@@
 
-In cases where a population invades a novel habitat, e.g.~the invasion of North America by the West-Nile virus \citep{pybus_unified_2010}, the moving front of the population defines a bona-fide velocity.
+In cases where a population invades a novel habitat, e.g.~the invasion of North America by the West-Nile virus \citep{pybus_unifying_2012}, the moving front of the population defines a bona-fide velocity.
 However, as discussed below, this velocity is determined by the interplay of dispersal and population growth at the front.
@@
 It has been known for a long time that ignoring such coupling leads to unrealistically heterogeneous population densities \citep{felsenstein_pain_1975}.
-If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D<L^2/T_c$, where $T_c$ is the coalescent timescale of the population (see Fig.~\ref{fig:heterogeneity}).
+If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D<L^2/T_c$, where $T_c$ is the coalescent timescale of the population (see Fig.~\ref{fig:density_reg}).
 In a well-mixed neutral model, $T_c$ is equal to the population size $N$.
@@
 For $D<L^2/2T_c$, the time to the most recent common ancestor $T_{MRCA}$ is too small for the population to spread evenly in the habitat.
-This is illustrated in Fig.~\ref{fig:heterogeneity}A that shows the population distribution across the habitat for $D=0.1L^2/T_c$.
+This is illustrated in Fig.~\ref{fig:density_reg}A that shows the population distribution across the habitat for $D=0.1L^2/T_c$.
 The population distribution is patchy with large uninhabited areas.
 This heterogeneity increases substantially for even lower $D$, when the population is essentially a cloud of width $\sqrt{4DT_c}\ll L$.
-Fig.~\ref{fig:heterogeneity}B quantifies this heterogeneity as the standard deviation of population density in two-dimensional bins of different size  relative to the expectation in the case of a spatially well-mixed population.
+Fig.~\ref{fig:density_reg}B quantifies this heterogeneity as the standard deviation of population density in two-dimensional bins of different size  relative to the expectation in the case of a spatially well-mixed population.
 
@@
 As soon as the interaction radius $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}C.
-Despite the substantial reduction in density fluctuations, the estimates of dispersal rates using Eq.~\ref{eq:disperal_parameters} are not substantially affected, at least when boundary conditions are periodic, see Fig.~\ref{fig:density_reg}D.
+Despite the substantial reduction in density fluctuations, the estimates of dispersal rates using Eq.~\ref{eq:dispersal_parameters} are not substantially affected, at least when boundary conditions are periodic, see Fig.~\ref{fig:density_reg}D.
 If instead of periodic boundary conditions reflective boundary conditions are used, diffusion constants are underestimated once $\sqrt{2DT_c}$ becomes comparable to the habitat size.
@@
 Spatial location and the coalescence process are thus strongly coupled, though this coupling manifests itself in population structure rather than biased estimates of the diffusion coefficients.
-Hence both at high and at low diffusion constants, assumptions of phylogeographic inference can be problematic: if diffusion constants are much smaller than then $L^2/T_{mrca}$, the population is fragmented into effective subpopulations that violate the typical tree-priors, if $D\geq L^/T_{mrca}$, the boundaries of the domain restrict free diffusion.
+Hence, both at high and at low diffusion constants, assumptions of phylogeographic inference can be problematic: if diffusion constants are much smaller than then $L^2/T_{mrca}$, the population is fragmented into effective subpopulations and this fragmentation violates the assumptions of phylodynamic models; if $D\gg L^2/T_{mrca}$, the boundaries of the habitat restrict free diffusion, resulting in underestimated dispersal parameters.
 
@@
 The severity of such biases depends on the period at which the carrying capacity oscillates and on the dispersal rate, see  Fig.~\ref{fig:seasaw}C.
-Fig.~\ref{fig:seasaw}D shows the estimated diffusion constant (using Eq.~\ref{eq:disperal_parameters}) in $x$ (solid lines) and $y$ (dashed) direction relative to the true value used in the simulations.
-At very high diffusion constants, both $D_x$ and $D_y$ are underestimated due to spatial constraint of carrying capacity as also discussed above (see Fig.~\ref{fig:stable_density}).
+Fig.~\ref{fig:seasaw}D shows the estimated diffusion constant (using Eq.~\ref{eq:dispersal_parameters}) in $x$ (solid lines) and $y$ (dashed) direction relative to the true value used in the simulations.
+At very high diffusion constants, both $D_x$ and $D_y$ are underestimated due to spatial constraint of carrying capacity as also discussed above (see Fig.~\ref{fig:density_reg}).
 With decreasing $D$, $D_x$ starts to be overestimated and peaks at a value that depends on the period $T$, before decreasing to (noisy) estimates that are broadly compatible with the true value.
@@
 
+Why does phylogeography then often generate sensible results?
+On short time scales, a diffusive model will mostly reduce to a parsimonious reconstruction of ancestral locations which is robust to details of the model.
+It is longer time scales, that is deeper nodes in the phylogeny, for which ancestral inferences can get problematic as the chance that the environment has changed and shifted, or that the ancestral locations are in poorly sampled regions, increases.
+Inferences, and in particular confidence statements, can be highly unreliable and misleading.
 In absence of a good understanding of the dispersal properties and how population dynamics varies and time and space, simple models that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models \citep{layan_impact_2023}.

### 2024-06-02 manuscript/phylogeo.tex
edits

@@
 \maketitle
-As organisms replicate, they also disperse in space, potentially exploring new habitats.
+As organisms replicate, they also disperse in space, resulting in mixing of the population and exploration of new habitats.
 The reconstruction of ancestral locations and past migrations from sampling locations of extant individuals, typically in combination with genome sequence information to infer the phylogeny, is known as phylogeography.
-Phylogeographic methods are implemented in popular evolutionary analysis software such as BEAST \citep{pybus_unifying_2012,lemey_phylogeography_2010}.
+Phylogeographic methods are implemented in popular evolutionary analysis software such as BEAST \citep{lemey_bayesian_2009,lemey_phylogeography_2010}.
 The underlying model of dispersal is typically assumed to be diffusive, that is the probability of sampling a descendant at position $\rvec_c$ after time $t$ when the parent was at position $\rvec_p$ is given by
@@
 It is also mathematically convenient as unknown ancestral locations can be integrated in closed form and the marginal distribution of each position can be calculated exactly for a fixed tree along with the likelihood.
-The total likelihood of the configuration is then assumed to be the product of the spatial likelihood and the phylogenetic likelihood, i.e.
-tree topology and branching rates of the tree are assumed to be independent of the spatial process.
+The total likelihood of the configuration is then assumed to be the product of the spatial likelihood and the phylogenetic likelihood, that is the tree topology and branching rates of the tree are assumed to be independent of the spatial process.
+This independence assumption is related to the common assumption that all individuals in the population are equally fit.
 
-However, population dynamics is often tightly coupled to spatial location, for example when a population expands into an uninhabited fertile habitat or a disease spreads in a susceptible population, and the independence assumption of spatial and phylogenetic processes is not justified.
-Such coupling can be accommodated in structured coalescent models where the population is divided into discrete subpopulations that exchange migrants at a rate $m$ \citep{vaughan_efficient_2014}.
+However, population dynamics is often tightly coupled to spatial location.
+Environmental change could mean that a habitat of a population is shifting, resulting in population growth and rapid branching at the expanding edge of the habitat and/or contraction in other areas.
+Similarly, invasive species or pathogen might spread into previously areas that support high population densities but where previously empty.
+Examples of such shifting habitats are common and range from geological time scales, over decades (climate change), to months (seasonal fluctuations).
+In all such cases, the replication process determining the phylogeny of the sample and the spatial location are strongly coupled.
+
+Such coupling can be accommodated in structured coalescent models where the population is divided into discrete subpopulations that exchange migrants and have different population dynamics \citep{vaughan_efficient_2014}.
 But the extent to which violations of the independence assumption affect continuous phylogeographic inferences is not well understood.
@@
 
-In a typically Bayesian phylogeographic inference problem as discussed in \citet{pybus_unifying_2012}, the input data are typically viral genome sequences as well as their sampling times and sampling locations.
+In a typically Bayesian phylogeographic inference problem, the input data are typically viral genome sequences as well as their sampling times and sampling locations.
 Software like BEAST will then sample the posterior distribution of the phylogenetic tree topology, the sampling times and locations of ancestral nodes in the tree, and model parameters like the diffusion constant $D$ and the evolutionary rate $\mu$.
 Results can be estimates of the ancestral location of a particular groups of samples, or summary statistics of the dispersal process.
-Popular summaries include empirical estimates of the diffusion constant  \citep{pybus_unifying_2012} and a so-called \emph{dispersal velocity} \citep{dellicour_using_2017}.
+Popular summaries include empirical estimates of the diffusion constant  \citep{pybus_unifying_2012,trovao_bayesian_2015} and a so-called \emph{dispersal velocity} \citep{dellicour_using_2017}.
 These are calculated from the estimated displacements along branch $i$ with parent $p_i$ given by $d_i = |\rvec_{p_i} - \rvec_{i}|$ and the elapsed time $\Delta t_i$ as
@@
 where $B$ is the set of all branches of the tree.
-Effectively $D_w$ divides the sum of all observed squared displacements by the total time elapsed on the tree to obtain and estimate of $D$.
+Effectively $D_w$ divides the sum of all observed squared displacements by the total time elapsed on the tree to obtain an estimate of $D$.
 It is known as the `weighted' estimate of the diffusion constant.
@@
 The purpose of this note is to explore how sensitive phylogeographic inferences are to model assumptions and what quantities can be robustly estimated.
-My focus here is only on the ability and robustness to estimate dispersal parameter and ancestral locations and I will assume perfect knowledge of the tree and the sampling locations.
+My focus here is only on the ability and robustness to estimate dispersal parameters and ancestral locations and I will assume perfect knowledge of the tree and the sampling locations.
 So no tree-sampling will be necessary.
-To explore qualitatively different behaviors, I will consider Kingman and Yule tree ensembles.
+
+\section*{Consistency of summary statistics of dispersal parameters.}
+
+The estimators in Eq.~\ref{eq:dispersal_parameters} are commonly used to summarize dispersal of lineages.
+While the estimator $D_w$ directly estimates a parameter of the model and its behavior has been studied in simulations \citep{pybus_unifying_2012}, it is unclear what $v_w$ is measuring and how it behaves.
+I simulated diffusion along trees from a Kingman \citep{kingman_coalescent_1982} and Yule \citep{yule_iimathematical_1997} tree ensembles and evaluated the different estimators.
 Kingman trees are the classical neutral model of a constant size population, while Yule trees capture features of growing populations (see Fig.~\ref{fig:illustration_tree}).
@@
 The diffusion constant $D$ can be estimated robustly from these data and the estimates are compatible with the diffusion constant used to simulate the data $D=1$.
-Values of $v_w$ and $v_b$, however, depend strongly on the sample size.
+Values of $v_w$ and $v_b$, however, depend strongly on the sample size and are incompatible with each other.
 That the velocity estimates are ill-defined is not surprising: By changing the sample size, the average length of branches changes.
@@
 
-In cases where a population invades a novel habitat, e.g.~the invasion of North America by the West-Nile virus \citep{pybus_unifying_2012}, the moving front of the population defines a bona-fide velocity.
-However, as discussed below, this velocity is determined by the interplay of dispersal and population growth at the front.
-The velocity is better determined by modeling the position of the front, rather than by tracking individual lineages along the tree.
-
-\section*{How does density regulation affect estimates phylogeographic estimates?}
+\section*{Effects of density regulation on population structure and dispersal}
 Phylogeographic inference typically assumes that the spatial diffusion process is independent of the branching pattern of the tree and ignores coupling between the two.
@@
 It has been known for a long time that ignoring such coupling leads to unrealistically heterogeneous population densities \citep{felsenstein_pain_1975}.
-If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D<L^2/T_c$, where $T_c$ is the coalescent timescale of the population (see Fig.~\ref{fig:density_reg}).
-In a well-mixed neutral model, $T_c$ is equal to the population size $N$.
+If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D<L^2/T_c$, where $T_c$ is the coalescent timescale of the population.
+This is seen in the examples in Fig.~\ref{fig:illustration_tree} where the population is distributed in patches corresponding to clades in the tree with large uninhabited areas.
 
 For $D<L^2/2T_c$, the time to the most recent common ancestor $T_{MRCA}$ is too small for the population to spread evenly in the habitat.
-This is illustrated in Fig.~\ref{fig:density_reg}A that shows the population distribution across the habitat for $D=0.1L^2/T_c$.
-The population distribution is patchy with large uninhabited areas.
 This heterogeneity increases substantially for even lower $D$, when the population is essentially a cloud of width $\sqrt{4DT_c}\ll L$.
-Fig.~\ref{fig:density_reg}B quantifies this heterogeneity as the standard deviation of population density in two-dimensional bins of different size  relative to the expectation in the case of a spatially well-mixed population.
+Fig.~\ref{fig:density_reg}A (black line) quantifies this heterogeneity as the standard deviation of population density in two-dimensional bins of different size relative to the expectation in the case of a spatially well-mixed population.
 
-\begin{figure*}
-    \includegraphics[width=0.46\textwidth]{figures/heterogeneity_free_diffusion}
+\begin{figure}
     \includegraphics[width=0.53\textwidth]{figures/stable_density}
-    \caption{\label{fig:density_reg}  {\bf Population density variation in free diffusion models.}
-    When spatial motion is independent of population density and coalesce is sufficiently rapid, the realized population density fluctuates strongly across the habitat.
-The left panel shows the population density for the case $D = 0.1 L^2/N$ in linear bins of size $L/20$.
-The right panel shows the standard deviation of the population density for different bin sizes as a function of the diffusion coefficient.
-For $D>L^2/N$, the heterogeneity approaches the expected value in the well mixed limit.
-    {\bf Density regulation reduces fluctuations in density regulation, but results in population fragmentation at low diffusion constants.} The left panel shows population heterogeneity measured using bins of linear dimension $L/5$ for $\alpha =0.05$ for different values of the interaction radius $r$.
+    \caption{\label{fig:density_reg}  {\bf Effects of population density regulation.}
+    When spatial motion is independent of population density and coalescence is sufficiently rapid, the realized population density fluctuates strongly across the habitat.
+Panel A shows the standard deviation of the population density in bins of linear dimension $L/5$ as a function of the diffusion coefficient relative to the expected value if individuals where independently distributed across the habitat.
+The thick black line shows the case of free diffusion, the colored lines density regulation with different interaction radii.
 For $r\ll L$, density fluctuations are strongly suppressed.
-At the same time, the time to the most recent common ancestor increases dramatically, indicating that the population fragments in subpopulations that don't interact.}
-\end{figure*}
+    At the same time, the time to the most recent common ancestor increases dramatically (panel B), indicating that the population fragments in subpopulations that don't interact.
+    Panel C shows the fold-error of the dispersal estimator $D_w$ for periodic boundary conditions (solid lines, like panels A\&B) and reflecting boundaries (dashed lines).
+    }
+\end{figure}
 
@@
 
-As soon as the interaction radius $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}C.
-Despite the substantial reduction in density fluctuations, the estimates of dispersal rates using Eq.~\ref{eq:dispersal_parameters} are not substantially affected, at least when boundary conditions are periodic, see Fig.~\ref{fig:density_reg}D.
+As soon as the interaction radius $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}A.
+But with decreasing diffusion, the population fragments into smaller and smaller subpopulations that interact only weakly and coalesce slowly.
+This manifests in an increasing time to the MRCA, see Fig.~\ref{fig:density_reg}B, which at low $D$ is limited by the time the population was simulated for.
+Spatial location and the coalescence process are thus strongly coupled.
+
+Despite the substantial reduction in density fluctuations, the estimates of dispersal rates using Eq.~\ref{eq:dispersal_parameters} are not dramatically affected, see Fig.~\ref{fig:density_reg}C.
+If boundary conditions are periodic, that is the habitat has the topology of a torus, estimates $D_w$ of the diffusion constant are accurate for most values of $D$ but slightly inflated when the population is heavily fragmented.
 If instead of periodic boundary conditions reflective boundary conditions are used, diffusion constants are underestimated once $\sqrt{2DT_c}$ becomes comparable to the habitat size.
@@
 
-But with decreasing diffusion, the population fragments into smaller and smaller subpopulations that don't interact or coalesce.
-This manifests in an increasing time to the MRCA, see Fig.~\ref{fig:density_reg}E.
-Spatial location and the coalescence process are thus strongly coupled, though this coupling manifests itself in population structure rather than biased estimates of the diffusion coefficients.
-Hence, both at high and at low diffusion constants, assumptions of phylogeographic inference can be problematic: if diffusion constants are much smaller than then $L^2/T_{mrca}$, the population is fragmented into effective subpopulations and this fragmentation violates the assumptions of phylodynamic models; if $D\gg L^2/T_{mrca}$, the boundaries of the habitat restrict free diffusion, resulting in underestimated dispersal parameters.
+These simulations reveal that both at high and at low diffusion constants, assumptions of phylogeographic inference can be problematic: if diffusion constants are much smaller than then $L^2/T_{mrca}$, the population is fragmented into effective subpopulations and this fragmentation violates the assumptions of phylodynamic models; if $D\gg L^2/T_{mrca}$, the boundaries of the habitat restrict free diffusion, resulting in underestimated dispersal parameters.
 
@@
 \section*{How do changing habitats affect phylogeographic inferences?}
-As discussed above, a static carrying capacity in an equilibrium population has limited effects on estimates of ancestral locations and diffusion coefficients beyond boundary effects.
+As discussed above, even a static carrying capacity in an equilibrium population can strongly effect population structure, though this structure has limited effects on estimates of ancestral locations and diffusion coefficients beyond boundary effects.
 In reality, carrying capacities vary through time.
@@
 The carrying capacity is illustrated in Fig.~\ref{fig:seasaw}A for 5 time points covering half a period.
+In between the two extreme left/right concentration of the carrying capacity, carrying capacity is uniform.
 To avoid extinction at low dispersal rates, selection is soft, meaning that the overall population size is kept approximately constant even if the entire population is stuck in the unfavorable region.
@@
 The population was sampled at a time when it has just contracted to the left and the positions of older nodes is not estimated correctly.
-Their true positions are to the right of the current population, but the inferred positions are central in the shifted population.
+Their true locations are to the right of the current population, but the inferred location are central in the shifted population.
 
@@
 This decoupling happens at lower values of $D$ for longer periods of environmental change.
-With the parameter used in the simulation, the estimates of $D_x$ deviate by a factor of 2 from its true value, which is not dramatic compared to the variation of dispersal rates in nature.
+With the parameters used in the simulation, the estimates of $D_x$ deviate by a factor of 2 from its true value, which is not dramatic compared to the variation of dispersal rates in nature.
 
 However, along with problematic inference of diffusion coefficients, estimates of ancestral locations are over-confident.
-Panel Fig.~\ref{fig:seasaw}E shows $z_x^2 = \langle (\hat{x} - x)^2/\sigma_x^2 \rangle$ and $z_y^2=\langle (\hat{y} - y)^2/\sigma_y^2\rangle$ for the oldest 20\% of internal nodes.
+Panel Fig.~\ref{fig:seasaw}E shows $z_x^2 = \langle (\hat{x} - x)^2/\sigma_x^2 \rangle$ and $z_y^2=\langle (\hat{y} - y)^2/\sigma_y^2\rangle$ for the oldest 20\% of internal nodes, where $\hat{x}$ and $\sigma_x$ are the mean and standard deviation of the posterior of $x$, respectively (and analogously for $y$).
 If the model estimated ancestral locations consistently, $z_x^2$ should be 1 on average.
 Instead, it is as high as 20 for the simulations shown and this over-confident and biased estimations depend in non-monotonic ways on the true diffusion constant and the period of environmental change.
-
-At high dispersal rates, the average $z_x^2$ and $z_y$ are both smaller than one since the estimated uncertainty is larger than the habitat.
 With decreasing $D$, $z_x^2$ increases largely because the expected uncertainty decreases, before peaking when the population can no longer follow the shifting carrying capacity.
 A peak value of 10-20 indicates that the typical ancestral location of the deep nodes is 2-4 standard deviations away from the true location, an example of which is shown in Fig.~\ref{fig:seasaw}C.
+Conversely, at high dispersal rates, the average $z_x^2$ and $z_y$ are both smaller than one since the estimated uncertainty is larger than the true uncertainty since the diffusion is constrained by the boundaries of the habitat.
 
@@
 If the population can not follow shifting habitat, the situation is similar to the static case often leads to fragmented populations.
-Once the population starts following, inferred ancestral locations of deep nodes become unreliable, while at high dispersal rates expected uncertainties are larger than the habitat size and there is little information ancestral locations.
+Once the population starts following, inferred ancestral locations of deep nodes become unreliable, while at high dispersal rates expected uncertainties are larger than the habitat size and there is little information on deep ancestral locations.
 
@@
     \includegraphics[width=0.3\textwidth]{figures/traveling_wave}
-    \includegraphics[width=0.48\textwidth]{figures/waves_D}
-    \includegraphics[width=0.48\textwidth]{figures/waves_v}
-    \caption{\label{fig:traveling_wave} {\bf Phylogeographic estimates in rapidly changing environments.} Panel A shows an illustration of the simulated set-up: a carrying capacity with a Gaussian profile with $\sigma=0.5$ that moves along the $x$-axis with velocity $v$ on a patch with $L_x=3$, $L_y=1$, and periodic boundary conditions.
+    \includegraphics[width=0.48\textwidth]{figures/waves}
+    % \includegraphics[width=0.48\textwidth]{figures/waves_v}
+    \caption{\label{fig:traveling_wave} Panel A shows an illustration of the simulated set-up: a carrying capacity with a Gaussian profile with $\sigma=0.5$ that moves along the $x$-axis with velocity $v$ on a patch with $L_x=3$, $L_y=1$, and periodic boundary conditions.
     Panels B\&C show estimates of diffusion coefficients and velocities, relative to their true values, as a function of the ratio of the FKPP velocity to the external velocity $v$. }
@@
 As expected, estimates of $\hat{D}_x$ are often dramatically too high (by up to factors of 1000).
-They peak right around when the external velocity matches the FKPP velocity, corresponding to the situation where the population dynamics is maximally effected by the moving carrying capacity.
+They peak right around when the external velocity matches the FKPP velocity, corresponding to the situation where the population dynamics is maximally affected by the moving carrying capacity.
 If the external velocity was lower, undirected dispersal contributes more and more to the displacement of lineages, while the population would be unable to follow the moving carrying capacity if the external velocity was higher.
 
-In contrast to the above examples with no persistent directionality, one can now compare velocity estimates to a velocity $v_{FKPP}$ defined by the model.
 Fig.~\ref{fig:traveling_wave}C shows estimates of $v$ using Eq.~\ref{eq:dispersal_parameters} relative to the true velocity.
@@
 
-These behaviors are explained comparing the speed $v_{FKPP}$ at which the population can invade new territory with the external speed $v$, the ratio of which is used to plot the simulation data in Fig.~\ref{fig:traveling_wave}.
+These behaviors are explained comparing the speed $v_{FKPP}$ at which the population can invade new territory with the external speed $v$, the ratio of which is used as the $x$-axis in Fig.~\ref{fig:traveling_wave} B\&C.
 For $v_{FKPP} \ll v$, the population cannot keep up with the moving carrying capacity and is saved from extinction only because selection is soft.
@@
 In this note, I have explored the robustness of such inference to violations of the latter assumption while keeping microscopic dispersal strictly diffusive, that is without any long-range migration.
-I showed that common summary statistics of dispersal rates are often problematic and that interactions between dispersal and replication can affect estimates of ancestral locations in ways that are hard to control.
 
-The common practice to estimates ``velocities'' of lineages using displacements of lineages along the tree \citep{dellicour_relax_2021} is problematic.
+The common practice to estimate dispersal ``velocities'' using displacements of lineages along the tree \citep{dellicour_relax_2021} is problematic.
 Firstly, the numerical value of such estimates depends strongly on the sample size and tree ensemble.
 Secondly, the underlying model is diffusive and has no parameters that have dimensions of a velocity and even if the quantity could be estimated robustly, there is no interpretation of the quantity within the model.
+In cases where a population invades a novel habitat, e.g.~the invasion of North America by the West-Nile virus \citep{pybus_unifying_2012}, the moving front of the population defines a bona fide velocity.
+However, as discussed above, this velocity is determined by the interplay of dispersal and population growth at the front.
+The velocity is better determined by modeling the position of the front, rather than by tracking individual lineages along the tree.
 
@@
 As has long been known, ignoring the coupling between population growth and spatial location can lead to unrealistic population densities \citep{felsenstein_pain_1975}.
-Simple density regulation leads to homogeneous population densities, but inferences of dispersal rates and ancestral locations using models that ignore this coupling are not substantially affected.
+Density regulation leads to homogeneous population densities.
 However, the population dynamics strongly depends on whether dispersal is fast enough to explore the entire habitat during the coalescence time $T_c$ of the population in absence of spatial structure.

### 2024-06-02 src/heterogeneity.py
changes to plotting scripts

@@ -21,33 +21,34 @@ if __name__=="__main__":
     ntips = 1000
     D_array = np.logspace(-2,2,101)
 
-    fig, axs = plt.subplots(2,1, figsize=(5,9), gridspec_kw={'height_ratios': [2, 1]})
-
+    fig, ax = plt.subplots(1,1, figsize=(6,4))
+    # fig, axs = plt.subplots(2,1, figsize=(5,9), gridspec_kw={'height_ratios': [2, 1]})
+    # ax = axs[0]
     g, H = get_granularity(ntips, 0.1, bins=20)
-    axs[0].matshow(H)
-    axs[0].set_axis_off()
-    add_panel_label(axs[0], 'A')
+    ax.matshow(H)
+    ax.set_axis_off()
+    add_panel_label(ax, 'A')
 
-    for line_style, linear_bins in zip(['-', '--', '-.'], [3,10,30]):
-        nbins=linear_bins**2
-        res = []
-        for D in D_array:
-            res.append([get_granularity(ntips, D, bins=linear_bins)[0] for i in range(n_iter)])
+    # for line_style, linear_bins in zip(['-', '--', '-.'], [3,10,30]):
+    #     nbins=linear_bins**2
+    #     res = []
+    #     for D in D_array:
+    #         res.append([get_granularity(ntips, D, bins=linear_bins)[0] for i in range(n_iter)])
 
-        res = np.array(res)
-        res_mean = res.mean(axis=1)
-        axs[1].plot(D_array, res_mean/np.sqrt(nbins/ntips), label=f'bin size=L/{linear_bins}') #, c='C0', ls=line_style)
-        # axs[1].plot(D_array, np.ones_like(D_array)*np.sqrt(nbins/ntips), label=f'well mixed limit dx=L/{linear_bins}', c='C1', ls=line_style)
+    #     res = np.array(res)
+    #     res_mean = res.mean(axis=1)
+    #     axs[1].plot(D_array, res_mean/np.sqrt(nbins/ntips), label=f'bin size=L/{linear_bins}') #, c='C0', ls=line_style)
+    #     # axs[1].plot(D_array, np.ones_like(D_array)*np.sqrt(nbins/ntips), label=f'well mixed limit dx=L/{linear_bins}', c='C1', ls=line_style)
 
-    # area = np.minimum(1,D_array*np.pi**2)
[…]

### 2024-06-02 src/parse_and_plot.py

@@ -42,21 +42,29 @@ def parse_data(data, groupby=None):
     for g, d in dgb['y_err']:
         y_err[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
 
-    x_err_abs = {}
-    for g, d in dgb['x_err_abs']:
-        x_err_abs[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
+    try:
+        x_err_abs = {}
+        for g, d in dgb['x_err_abs']:
+            x_err_abs[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
 
-    y_err_abs = {}
-    for g, d in dgb['y_err_abs']:
-        y_err_abs[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
+        y_err_abs = {}
+        for g, d in dgb['y_err_abs']:
+            y_err_abs[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
 
-    x_err_sq = {}
-    for g, d in dgb['x_err_sq']:
-        x_err_sq[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
+        x_err_sq = {}
+        for g, d in dgb['x_err_sq']:
+            x_err_sq[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
 
-    y_err_sq = {}
-    for g, d in dgb['y_err_sq']:
-        y_err_sq[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
+        y_err_sq = {}
+        for g, d in dgb['y_err_sq']:
+            y_err_sq[g] = np.mean(d.apply(lambda x:np.array([float(y) for y in x[1:-1].split()])), axis=0)
+    except:
+        print("failed to parse error information")
+        x_err_abs = pd.DataFrame()
+        y_err_abs = pd.DataFrame()
+        x_err_sq = pd.DataFrame()
+        y_err_sq = pd.DataFrame()
+
 
     z_mean_x = {}
     for g, d in dgb['meanZsq_x']:
@@ -97,7 +105,7 @@ def parse_data(data, groupby=None):
                 }
 
 
[…]

### 2024-06-02 src/plot_seasaw.py

@@ -23,7 +23,7 @@ if __name__=="__main__":
     data = pd.read_csv(args.data)
     Lx, Ly = 3, 1
 
-    make_figure(seasaw, {'period':8, 'amplitude':1.1}, fname="figures/seasaw_illustration.pdf")
+    make_figure(seasaw, {'period':8, 'amplitude':1.1}, fname="figures/seasaw_illustration.pdf", panel_label='A')
 
     res = parse_data(data, groupby=['interaction_radius', 'density_reg', 'N', 'period', 'subsampling'])
     interaction_radius = data.interaction_radius.unique()
@@ -59,7 +59,7 @@ if __name__=="__main__":
     #ax.colorbar()
     ax.legend()
     ax.plot(ax.get_xlim(), [1,1], ls='-', c='k', lw=3, alpha=0.3)
-    add_panel_label(ax, 'A')
+    add_panel_label(ax, 'D')
 
     ## Diffusion constant estimates
     ax=axs[1]
@@ -83,7 +83,7 @@ if __name__=="__main__":
     #ax.colorbar()
     # ax.legend()
     ax.plot(ax.get_xlim(), [1,1], ls='-', c='k', lw=3, alpha=0.3)
-    add_panel_label(ax, 'A')
+    add_panel_label(ax, 'E')
 
     plt.tight_layout()
     plt.savefig("figures/seasaw.pdf")

### 2024-06-02 src/plot_waves.py

@@ -3,7 +3,7 @@ import matplotlib.pyplot as plt
 import pandas as pd
 from itertools import product
 from habitat_shifts import waves
-from parse_and_plot import parse_data, make_figure
+from parse_and_plot import parse_data, make_figure, add_panel_label
 
 
 if __name__=="__main__":
@@ -18,7 +18,7 @@ if __name__=="__main__":
     args = parser.parse_args()
 
     # make an illustration of the carrying capacity at different times
-    make_figure(func=waves, params={'width':0.5, 'velocity':1}, fname = args.illustration)
+    make_figure(func=waves, params={'width':0.5, 'velocity':1}, fname = args.illustration, panel_label='A')
 
 
     # read and organize data
@@ -37,11 +37,10 @@ if __name__=="__main__":
 
     ls = ['-', ':', "--", ".-"][len(interaction_radius)]
     markers = ['o', '<', '>', 's', 'd', '^', 'v']
-
-
+    fig, axs = plt.subplots(2,1, figsize=(6,6), sharex=True)
+    ax=axs[0]
     ## Diffusion constant estimates
     for c in ['x']:
-        plt.figure(figsize=(6,3))
         #plt.title(f"{T}")
         for vi,v in enumerate(velocity):
             labeled=False
@@ -50,28 +49,24 @@ if __name__=="__main__":
                 for i, (ir, dr, p) in enumerate(product(interaction_radius, density_reg[:2], [1.0])):
                     v_array = 2*np.sqrt(dr*D_array)
                     try:
-                        plt.plot(v_array/v, res["diffusion_mean_"+c][ir, dr, N, v, p,:]/D_array**1.0,
+                        ax.plot(v_array/v, res["diffusion_mean_"+c][ir, dr, N, v, p,:]/D_array**1.0,
                                             c=f'C{vi}', marker=markers[density_reg.index(dr)], ls='-',
                             label='' if labeled else f'v={v}')
                         labeled=True
                     except:
                         pass
 
-            plt.ylabel(r'$\hat{D}_x / D$')
-            plt.xlabel(r'$\frac{2\sqrt{\alpha D}}{v} = \frac{v_{FKPP}}{v}$')
-            plt.yscale('log')
-            plt.xscale('log')
+            ax.set_ylabel(r'$\hat{D}_x / D$')
[…]

### 2024-06-02 src/seasaw_example.py

@@ -31,7 +31,7 @@ if __name__=="__main__":
     period = 500
     target_density = seasaw(N, Lx, Ly, **{'period':period, 'amplitude':1.1})
 
-    v_FKPP = 2*np.sqrt(args.velocity*D)
+    v_FKPP = 2*np.sqrt(args.density_reg*D)
     print(f"{interaction_radius=:1.3f}, {density_reg=:1.3f}")
 
     tree = make_node(Lx/2,Ly/2,-2, None)
@@ -52,9 +52,12 @@ if __name__=="__main__":
         estimate_ancestral_positions(tree, D_res['Dxy_total'])
         phylo_tree = dict_to_phylo_tree(tree['clades'][0] if len(tree['clades'])==1 else tree,
                                         child_attr='clades')
+        phylo_tree.ladderize()
 
         fig, axs = plt.subplots(1,2, figsize=(12,3))
         Phylo.draw(phylo_tree, axes=axs[0], label_func=lambda x: '')
+        add_panel_label(axs[0], 'B', x=-0.05, y=1.1)
+        axs[0].set_axis_off()
         print(f"N={len(terminal_nodes)}")
 
         # plot true and inferred positions of nodes
@@ -68,8 +71,8 @@ if __name__=="__main__":
 
         # highlight true and inferred root positions
         n = phylo_tree.root
-        axs[1].scatter([n.pos['x']], [n.pos['y']], c='r', s=100)
-        axs[1].scatter([n.inferred_pos['x']['mean']], [n.inferred_pos['y']['mean']], c='g', s=100, marker='^')
+        axs[1].scatter([n.pos['x']], [n.pos['y']], c='g', s=100, label='True root')
+        axs[1].scatter([n.inferred_pos['x']['mean']], [n.inferred_pos['y']['mean']], c='r', s=100, marker='^', label='Inferred root')
 #                c=[n.t for n in phylo_tree.find_clades()], s=30, marker='d', ls=None)
         # z = [(n.inferred_pos['x']['mean'] - n.pos['x'])/n.inferred_pos['x']['var']**0.5 for n in phylo_tree.get_nonterminals()]
         # tps = [n.t for n in phylo_tree.get_nonterminals()]
@@ -83,8 +86,12 @@ if __name__=="__main__":
         # indicate density
         x = np.linspace(0, Lx,101)
         dens = target_density(x, 0, t)
[…]

### 2024-06-02 src/stable_density_figure.py

@@ -36,7 +36,7 @@ if __name__=="__main__":
     N_vals = data.N.unique()
 
     # figure
-    fig, axs = plt.subplots(3, 1, figsize=(6, 9), sharex=True)
+    fig, axs = plt.subplots(3, 1, figsize=(5, 7.5), sharex=True)
 
     # Plot the heterogeneity
     ls = ['-', '-.', "--", ":"] #[:len(density_reg_to_plot)]
@@ -48,36 +48,39 @@ if __name__=="__main__":
         label = f'r={ir}' if N==N_vals[1] else ''
         ax.plot(D_array*N/Lx/Ly, resPBC["density_variation"][ir, dr, N, :]/np.sqrt(nbins/N),
                 label=label, ls='-', c=f"C{i%10}")
+        # ax.plot(D_array*N/Lx/Ly, resPBC["density_variation"][ir, dr, N, :]/np.sqrt(nbins/N),
+        #         label=label, ls='--', c=f"C{i%10}")
     free_diffusion_heterogeneity = free_diffusion(D_array*N/Lx/Ly, N, linear_bins=linear_bins)
-    ax.plot(D_array*N/Lx/Ly, free_diffusion_heterogeneity/np.sqrt(nbins/N), label='free diffusion', c='k')
+    ax.plot(D_array*N/Lx/Ly, free_diffusion_heterogeneity/np.sqrt(nbins/N), label='free diffusion', c='k', lw=3)
     ax.plot(D_array*N/Lx/Ly, np.ones_like(D_array), c='k', ls='--')
     ax.set_xscale('log')
 #    ax.set_xlabel('diffusion constant $[L^2/N]$')
     ax.set_ylabel('rel. heterogeneity')
-    ax.legend()
+    ax.legend(fontsize=8)
     ax.set_xlim(0)
-    add_panel_label(ax, 'C')
+    add_panel_label(ax, 'A')
     #ax.text(f'N={N}, alpha={dr}, periodic boundary conditions')
 
 
     ## Plot the diffusion estiamtes
-    ax=axs[1]
+    ax=axs[2]
     for i, ir in enumerate(interaction_radius[1:]):
-        D_array = np.array(res["diffusion_mean"][ir, dr, N, :].index)
+        D_array = np.array(resPBC["diffusion_mean"][ir, dr, N, :].index)
         label = f'r={ir}, a={dr}' if N==N_vals[0] else ''
[…]

### 2024-06-02 manuscript/phylogeo.tex
edits

@@
 \begin{abstract}
-Phylogeographic inference in continuous space is a popular method to reconstruct the spatial distribution of ancestral populations and estimate parameters of the dispersal process.
-While the underlying probabilistic models can be complex and computationally demanding to infer, they typically ignore that replication and population growth are tightly coupled to spatial location: populations expand into fertile uninhabited areas and contract in regions with limited resources.
+Continuous phylogeographic inference is a popular method to reconstruct the spatial distribution of ancestral populations and estimate parameters of the dispersal process.
+While the underlying probabilistic models can be complex and their parameters are often computationally demanding to infer, these models typically ignore that replication and population growth are tightly coupled to spatial location: populations expand into fertile uninhabited areas and contract in regions with limited resources.
 Such coupling, in particular when habitats change over time, can result in biased and overconfident estimates of ancestral locations and dispersal parameters.
@@
 
-Inference using phylogeographic methods should thus be interpreted with caution.
+Phylogeographic methods are often studied using West Nile virus and Rabies virus.
+West-Nile virus was first detected on the east coast of the US in 1999 and reached the West Coast five years later in 2004 \citep{pybus_unifying_2012}, translating to a wave front velocity of about 1000 km/year.
+\citet{pybus_unifying_2012} estimated a diffusion coefficient of between 200 and 5000 km$^2$/day, where the latter refers to branches with particularly high diffusion constant.
+Using $v_{FKPP} = 2\sqrt{D\alpha}$ and equating it to 1000 km/year $\approx$ 3km/day, we obtain ranges of $\alpha=\frac{9 km^2}{4 D day^2} = 0.012/day \ldots 0.00045/day$ or between $0.003$ and $0.08$/week.
+A pathogen that produces seasonal outbreaks with many orders of magnitude in prevalence between peak and troughs typically requires a growth rate of 1/week, considerably more than what the estimates of $D$ and the FKPP equation imply.
+Alternatively, if we assume $\alpha = 0.1/day$, we would expect $D = \frac{10 km^2}{0.4 day} = 25 km^2/day$, implying a daily exploration radius of about 5 km.
+This picture is also consistent with the apparent ``slowing down'' of lineages after the initial expansion across North America \citep{dellicour_epidemiological_2020}: Once the habitat was fully explored, directed invasion with $v_{FKPP}$ ceases.
+Furthermore, rare long range dispersal could have strongly influenced the spread of the virus in ways not captured by Brownian or relaxed random walks \citep{hallatschek_acceleration_2014}.
+
+Dispersal rates of rabies virus are typically estimated to be between 500 and 1500 km$^2$/year \citep{dellicour_using_2017}.
+The most recent common ancestor of various populations is between 30 and 100 years in the past.
+This would translate into rather limited spread of 150 to 500km of individual clades from their common ancestor, suggesting that these populations are highly structured (isolated by distance) and that their long range dispersal is dominated by rare introductions of the virus into new populations.
+
+The results presented here and the discussion of the examples above suggested that phylogeographic methods should be used and interpreted with caution.
 In many cases the habitat of a population has undergone dramatic and recurrent changes since the $MRCA$ and such changes will affect inferences in ways that are not captured by the models.
-In addition, uneven sampling, or complete lack of samples from some regions, can undermine phylogeographic estimates \citep{kalkauskas_sampling_2021}.
+In addition, uneven sampling, or complete lack of samples from some regions, can undermine phylogeographic estimates \citep{kalkauskas_sampling_2021,layan_impact_2023}.
 The ability to infer ancestral locations is further limited by long-range dispersal \citep{hallatschek_acceleration_2014}.
@@
 Inferences, and in particular confidence statements, can be highly unreliable and misleading.
-In absence of a good understanding of the dispersal properties and how population dynamics varies and time and space, simple models that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models \citep{layan_impact_2023}.
+In absence of a good understanding of the dispersal properties and how population dynamics varies and time and space, simple models that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models.
 

### 2024-06-21 manuscript/phylogeo.tex
edit manuscript

@@
 Phylogeographic methods are implemented in popular evolutionary analysis software such as BEAST \citep{lemey_bayesian_2009,lemey_phylogeography_2010}.
-The underlying model of dispersal is typically assumed to be diffusive, that is the probability of sampling a descendant at position $\rvec_c$ after time $t$ when the parent was at position $\rvec_p$ is given by
+
+In phylogeographic inference, one can either assume that individuals migrate between discrete locations or disperse in continuous space.
+Here, we consider migration in continuous space, where the underlying model is typically assumed to be diffusive, that is the probability of sampling a descendant at position $\rvec_c$ after time $t$ when the parent was at position $\rvec_p$ is given by
 \begin{equation}
@@
 \end{equation}
-where $D$ is a diffusion constant with dimensions $\mathrm{length}^2/\mathrm{time}$, and we have assumed a two-dimensional space and ignored anisotropic dispersal.
+where $D$ is a diffusivity with dimensions $\mathrm{length}^2/\mathrm{time}$, we have assumed a two-dimensional space, and we have assumed dispersal is isotropic.
 Such a diffusive process is the simplest and most natural choice in absence of directed motion or long range migration.
-It is also mathematically convenient as unknown ancestral locations can be integrated in closed form and the marginal distribution of each position can be calculated exactly for a fixed tree along with the likelihood.
+It is also mathematically convenient as unknown ancestral locations can be integrated in closed form and the marginal distribution of each position can be calculated exactly for a fixed tree along with the spatial likelihood.
 The total likelihood of the configuration is then assumed to be the product of the spatial likelihood and the phylogenetic likelihood, that is the tree topology and branching rates of the tree are assumed to be independent of the spatial process.
-This independence assumption is related to the common assumption that all individuals in the population are equally fit.
+This independence assumption is for example manifest in coalescent models assuming ``exchangeability'' of individuals, or birth-death models assuming identical rates on contemporary branches of the tree.
 
-However, population dynamics is often tightly coupled to spatial location.
-Environmental change could mean that a habitat of a population is shifting, resulting in population growth and rapid branching at the expanding edge of the habitat and/or contraction in other areas.
-Similarly, invasive species or pathogen might spread into previously areas that support high population densities but where previously empty.
-Examples of such shifting habitats are common and range from geological time scales, over decades (climate change), to months (seasonal fluctuations).
+However, population dynamics are often tightly coupled to spatial location.
+Changes of the environment could mean that a habitat of a population is shifting, resulting in population growth and rapid branching at the expanding edge of the habitat and/or contraction in other areas.
+Similarly, invasive species or pathogens spread into areas that support high population densities but were previously unoccupied.
+Examples of such shifting habitats are common and range from geological time scales, over decades (climate change), to weeks (seasonal fluctuations).
+Such changes are not restricted to the physical environment, but ecological shifts can similarly affect the habitat.
 In all such cases, the replication process determining the phylogeny of the sample and the spatial location are strongly coupled.
 
-Such coupling can be accommodated in structured coalescent models where the population is divided into discrete subpopulations that exchange migrants and have different population dynamics \citep{vaughan_efficient_2014}.
+Such coupling can be accommodated in inference methods based on structured models that divide the population into discrete demes that could correspond to spatial locations that exchange migrants.
+Such models are flexible enough to allow for different population dynamics in different demes \citep{vaughan_efficient_2014}.
 But the extent to which violations of the independence assumption affect continuous phylogeographic inferences is not well understood.
-To accommodate variation not captured by the simple diffusion model, ``relaxed random walk'' models allow the diffusion constant $D$ to vary across the tree (the diffusion constant is integrated over a prior distribution) \citep{dellicour_relax_2021,lemey_phylogeography_2010}.
 
-\begin{figure*}[tb]
-    \includegraphics*[width=0.7\textwidth]{figures/illustration_tree.pdf}
-    \caption{\label{fig:illustration_tree}{\bf Illustration Yule and Kingman trees and the spatial location of the tips.}
+\begin{figure}[tb]
+    \includegraphics*[width=\columnwidth]{figures/illustration_tree.pdf}
+    \caption{\label{fig:illustration_tree}{\bf Illustration Yule and Kingman trees with $N=100$ tips and the spatial location of the tips.}
     The total tree length of Yule trees is proportional to the number of tips $n$ and the time to the MRCA is $\log(n)$.
     Kingman trees have a total tree length that is proportional to $\log(n)$ and are characterized by rapid merging close to the present. }
-\end{figure*}
+\end{figure}
 
-In a typically Bayesian phylogeographic inference problem, the input data are typically viral genome sequences as well as their sampling times and sampling locations.
+In a typical phylogeographic inference problem, the input data are genome sequences as well as their sampling times and sampling locations.
 Software like BEAST will then sample the posterior distribution of the phylogenetic tree topology, the sampling times and locations of ancestral nodes in the tree, and model parameters like the diffusion constant $D$ and the evolutionary rate $\mu$.
-Results can be estimates of the ancestral location of a particular groups of samples, or summary statistics of the dispersal process.
-Popular summaries include empirical estimates of the diffusion constant  \citep{pybus_unifying_2012,trovao_bayesian_2015} and a so-called \emph{dispersal velocity} \citep{dellicour_using_2017}.
-These are calculated from the estimated displacements along branch $i$ with parent $p_i$ given by $d_i = |\rvec_{p_i} - \rvec_{i}|$ and the elapsed time $\Delta t_i$ as
+The inferred quantities of interest could be estimates of the ancestral location of a particular groups of samples, or summary statistics of the dispersal process.
+Here, I explore what dispersal quantities can be robustly estimated and how sensitive inferences are to the fact that coupling between population dynamics and spatial location is ignored.
+My focus here is only on the ability and robustness to estimate dispersal parameters and ancestral locations and I will use simulations with perfect knowledge of the tree and the sampling locations.
+So no tree-sampling will be necessary.
+This simplification is justified since the tree tends to depend much more strongly on sequence information and spatial information and will allow stronger conclusions.
+Furthermore, I will assume strictly diffusive dispersal without any long-range migrations.
+
+\section*{Consistency of summary statistics of dispersal parameters.}
+Before investigating the more complex issues of coupling between spatial locations and reproduction, I will first explore the properties of summary statistics in simple models where spatial location and population dynamics are independent.
+Popular summaries of the dispersal process are empirical estimates of the diffusion constant $D_w$ \citep{pybus_unifying_2012,trovao_bayesian_2015} and a so-called \emph{dispersal velocity} $v_w$ \citep{dellicour_using_2017}.
+These are calculated from the displacements along branches of the trees sampled from the posterior.
+For each branch $i$ with parent $p_i$, the displacement $d_i = |\rvec_{p_i} - \rvec_{i}|$ and the elapsed time $\Delta t_i$ are used to calculate
 \begin{equation}
@@
 
-The purpose of this note is to explore how sensitive phylogeographic inferences are to model assumptions and what quantities can be robustly estimated.
-My focus here is only on the ability and robustness to estimate dispersal parameters and ancestral locations and I will assume perfect knowledge of the tree and the sampling locations.
-So no tree-sampling will be necessary.
-
-\section*{Consistency of summary statistics of dispersal parameters.}
-
 The estimators in Eq.~\ref{eq:dispersal_parameters} are commonly used to summarize dispersal of lineages.
 While the estimator $D_w$ directly estimates a parameter of the model and its behavior has been studied in simulations \citep{pybus_unifying_2012}, it is unclear what $v_w$ is measuring and how it behaves.
-I simulated diffusion along trees from a Kingman \citep{kingman_coalescent_1982} and Yule \citep{yule_iimathematical_1997} tree ensembles and evaluated the different estimators.
-Kingman trees are the classical neutral model of a constant size population, while Yule trees capture features of growing populations (see Fig.~\ref{fig:illustration_tree}).
+I simulated diffusion along trees from a Kingman coalescent\citep{kingman_coalescent_1982} and Yule model \citep{yule_iimathematical_1925} tree ensembles and evaluated the different estimators using the known true ancestral locations.
+Kingman trees are the classical neutral model of a constant size population, while Yule trees capture features of exponentially growing populations (see Fig.~\ref{fig:illustration_tree}).
+Examples for each ensemble are shown in Fig.~\ref{fig:illustration_tree}.
 
-\begin{figure}[tb]
-    \includegraphics*[width=0.48\textwidth]{figures/kingman_dispersal.pdf}
+\begin{figure*}[tb]
+    \includegraphics*[width=0.99\textwidth]{figures/dispersal_stats.pdf}
+    % \includegraphics*[width=0.48\textwidth]{figures/yule_dispersal.pdf}
     \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
@@
 The expected scaling $\sim \sqrt{n}$ of $v$ estimates is indicated by the dashed line.}
-\end{figure}
+\end{figure*}
 
-Fig.~\ref{fig:D_and_v} shows empirical estimates of diffusion constants and velocity as defined in Eq.~\ref{eq:dispersal_parameters} from simulated data using freely diffusing locations along Kingman trees.
+Fig.~\ref{fig:D_and_v} shows empirical estimates of diffusion constants and velocity as defined in Eq.~\ref{eq:dispersal_parameters} from simulated data using freely diffusing locations along Yule and Kingman trees.
 The diffusion constant $D$ can be estimated robustly from these data and the estimates are compatible with the diffusion constant used to simulate the data $D=1$.
-Values of $v_w$ and $v_b$, however, depend strongly on the sample size and are incompatible with each other.
-That the velocity estimates are ill-defined is not surprising: By changing the sample size, the average length of branches changes.
+Values of $v_w$ and $v_b$ estimated from Kingman trees, however, depend strongly on the sample size and are incompatible with each other.
+That the velocity estimates are ill-defined is not surprising: By changing the sample size, the average length of branches in Kingman coalescent trees changes.
 Diffusion along each branch results in a displacement of $d_i \sim \sqrt{4D\Delta t_i}$ such that a quantity that depends on $d_i / \Delta t_i \sim 1/\Delta t_i^{1/2}$ will change with the time $\Delta t_i$ that elapsed along the branch.
 With more samples in the tree, the average branch length decreases, and the velocity estimate increases.
-This expected scaling $\sim \sqrt{n}$ is indicated by the dashed line in Fig.~\ref{fig:D_and_v}.
+This expected scaling $\sim \sqrt{n}$ is indicated by the dashed line in right panel of Fig.~\ref{fig:D_and_v}.
 For Yule trees, the average branch length is independent of the number of samples and the velocity estimates are more stable.
-But estimates for $v_w$ and $v_b$ are not consistent with each other.
+But as for the Kingman case, the estimates for $v_w$ and $v_b$ are not consistent with each other.
 
@@
 
+While the estimator $D_w$ as defined in Eq.~\ref{eq:dispersal_parameters} is well-behaved when the model is exactly diffusive, long range dispersal can affect $D_w$ dramatically.
+If the probability of long jumps with distance $x<d<x+dx$ decays more slowly than $\sim dx/x^3$, the expectation value of the $D_w$ diverges.
+This may mean that even $D_w$ becomes problematic when very long-tailed prior distributions for $D$ are used in relaxed random walk models.
+
 \section*{Effects of density regulation on population structure and dispersal}
 Phylogeographic inference typically assumes that the spatial diffusion process is independent of the branching pattern of the tree and ignores coupling between the two.
-Such a coupling, however, is expected since growth rates naturally depend on local population density when organisms compete for resources.
-It has been known for a long time that ignoring such coupling leads to unrealistically heterogeneous population densities \citep{felsenstein_pain_1975}.
+Such a coupling, however, is expected since organisms compete for resources and growth rates naturally depend on local population density.
+It has been known for a long time that ignoring such coupling leads to unrealistically heterogeneous population densities \citep{felsenstein_pain_1975}, as is evident in the example of the Kingman tree in Fig.~\ref{fig:illustration_tree}.
 If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D<L^2/T_c$, where $T_c$ is the coalescent timescale of the population.
-This is seen in the examples in Fig.~\ref{fig:illustration_tree} where the population is distributed in patches corresponding to clades in the tree with large uninhabited areas.
-
-For $D<L^2/2T_c$, the time to the most recent common ancestor $T_{MRCA}$ is too small for the population to spread evenly in the habitat.
+Fig.~\ref{fig:density_reg}A (black line) quantifies this heterogeneity as the standard deviation of population density in two-dimensional bins of size $L/5$ relative to the expectation in the case of a spatially well-mixed population when each individual picks its location at random.
+For $D<L^2/2T_c$, the time to the most recent common ancestor $T_{mrca}$ is too small for the population to spread evenly in the habitat.
 This heterogeneity increases substantially for even lower $D$, when the population is essentially a cloud of width $\sqrt{4DT_c}\ll L$.
-Fig.~\ref{fig:density_reg}A (black line) quantifies this heterogeneity as the standard deviation of population density in two-dimensional bins of different size relative to the expectation in the case of a spatially well-mixed population.
 
@@
 Biological populations spread into areas that support them and will quickly populate fertile areas that are below their carrying capacity.
-A simple model for the population density $\phi(x,t)$ would be
+A simple model for the population density $\phi(\rvec,t)$ would be
 \begin{equation}
     \label{eq:FKPP}
-    \frac{\partial \phi(x,t)}{\partial t} = D\frac{\partial^2 \phi(x,t)}{\partial x^2} + \alpha \phi(x,t)(1-\phi(x,t)/\phi_0)
+    \frac{\partial \phi(\rvec,t)}{\partial t} = D \Delta_{\rvec} \phi(\rvec,t) + \alpha \phi(\rvec,t)(1-\phi(\rvec,t)/\phi_0)
 \end{equation}
-where $\phi_0$ is the carrying capacity and $\alpha$ is the population growth rate at zero density.
-This model has a stable fixed point at $\phi=\phi_0$.
-In a static environment, we thus expect the population density to settle at a constant density $\phi_0$ and density fluctuations will decay with rate $\alpha$.
+where $ \Delta_{\rvec}$ is the two-dimensional Laplace operator, $\phi_0$ is the carrying capacity, and $\alpha$ is the population growth rate at zero density.
+Such logistic density regulation was proposed by \citet{bolker_using_1997} and further analyzed by \citet{etheridge_survival_2004}.
+The density has a stable fixed point at $\phi=\phi_0$ we thus expect the population density to settle at a constant density $\phi_0$ and density fluctuations will decay with rate $\alpha$.
 
-While dependence of growth (and thus tree structure) on density is a key component of real-world population dynamics, it is difficult to implemented in phylogeographic inference frameworks.
+While the dependence of growth (and thus tree structure) on density is a key component of real-world population dynamics, it is difficult to implement in phylogeographic inference frameworks.
 It is, however, fairly straightforward to include density dependence in forward simulations and these simulations can be used to investigate the effect of such coupling between growth and spatial location on phylogeographic inferences.
-We calculate the population density using a Gaussian kernel with interaction radius $r$
+
+Instead of the coalescent trees used above, I now used an individual based simulation forward in time where individuals are assigned a fitness $f=\alpha(1-\hat{\phi}/\phi_0) + (1-N/N_0)$ and produce a Poisson distributed number of offspring with mean $1+f$.
+The first term determines the fitness due to the local population density and has the tendency to stabilize this density at $\phi_0$, the second term $(1-N/N_0)$ keeps the overall population close to the target population size $N_0$.
+The population density is calculated using a Gaussian kernel with interaction radius $r$
 \begin{equation}
-    \phi(\rvec) = \sum_{i\in N} \frac{e^{-|\rvec_i - \rvec|^2/2r^2}}{N\sqrt{2\pi r^2}}
+    \ref{eq:density_estimator}
+    \hat{\phi}(\rvec, t) = \sum_{i\in N} \frac{e^{-|\rvec_i - \rvec|^2/2r^2}}{\sqrt{2\pi r^2}}
 \end{equation}
-where the sum is over all individuals $i$ in the population of size $N$.
+where the sum is over all individuals $i$ in the population of size $N$ and $\rvec_i$ is location of individual $i$ at time $t$.
+The spatial locations of the offspring are chosen from a 2-D Gaussian centered on the parent with a variance to match the diffusion constant.
+The simulation records the tree of population, including their ancestral location and times.
+In addition to the true location of ancestral nodes, I also infer their locations from the locations of the extant samples.
+Since the model is purely diffusive, the marginal distributions of ancestral locations and their mean and variance can be calculated exactly using recursion relations.
 
-As soon as the interaction radius $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}A.
+The radius $r$ of the density estimate $\hat{\phi}(\rvec, t)$ in Eq.~\ref{eq:density_estimator} determines the distance over which different individuals compete for resources.
+As soon as $r$ is much smaller than the habitat size, this density regulation leads to a more homogeneous population density, see Fig.~\ref{fig:density_reg}A.
 But with decreasing diffusion, the population fragments into smaller and smaller subpopulations that interact only weakly and coalesce slowly.
@@
 
-Despite the substantial reduction in density fluctuations, the estimates of dispersal rates using Eq.~\ref{eq:dispersal_parameters} are not dramatically affected, see Fig.~\ref{fig:density_reg}C.
+Despite the substantial reduction in density fluctuations, the estimates of dispersal rates using Eq.~\ref{eq:dispersal_parameters} are not dramatically affected by density regulation, see Fig.~\ref{fig:density_reg}C.
 If boundary conditions are periodic, that is the habitat has the topology of a torus, estimates $D_w$ of the diffusion constant are accurate for most values of $D$ but slightly inflated when the population is heavily fragmented.
@@
 
-These simulations reveal that both at high and at low diffusion constants, assumptions of phylogeographic inference can be problematic: if diffusion constants are much smaller than then $L^2/T_{mrca}$, the population is fragmented into effective subpopulations and this fragmentation violates the assumptions of phylodynamic models; if $D\gg L^2/T_{mrca}$, the boundaries of the habitat restrict free diffusion, resulting in underestimated dispersal parameters.
+These simulations reveal that both at high and at low diffusion constants, assumptions of phylogeographic inference can be problematic: if diffusion constants are much smaller than then $L^2/T_{mrca}$, the population is fragmented into effective subpopulations and this fragmentation violates the assumptions of phylodynamic models.
+A structured model would be more appropriate in this case.
+If $D\gg L^2/T_{mrca}$, the boundaries of the habitat restrict free diffusion, resulting in underestimated dispersal parameters, while estimates of deeper ancestral locations have uncertainties comparable to the habitat size.
 
@@
 Inference of ancestral locations in such situations can be misleading as illustrated in Fig.~\ref{fig:seasaw}B.
-The population was sampled at a time when it has just contracted to the left and the positions of older nodes is not estimated correctly.
+The population was sampled at a time when it had just contracted to the left and the positions of older nodes are not estimated correctly.
 Their true locations are to the right of the current population, but the inferred location are central in the shifted population.
+This is expected as there is no information in the current sample to suggest otherwise, but it highlights the potential to mislead.
 
@@
 If the model estimated ancestral locations consistently, $z_x^2$ should be 1 on average.
-Instead, it is as high as 20 for the simulations shown and this over-confident and biased estimations depend in non-monotonic ways on the true diffusion constant and the period of environmental change.
+Instead, it is as high as 20 for the simulations shown and these over-confident and biased estimates depend in non-monotonic ways on the true diffusion constant and the period of environmental change.
 With decreasing $D$, $z_x^2$ increases largely because the expected uncertainty decreases, before peaking when the population can no longer follow the shifting carrying capacity.

### 2024-06-21 manuscript/phylogeo.tex
edits

@@
     When spatial motion is independent of population density and coalescence is sufficiently rapid, the realized population density fluctuates strongly across the habitat.
-Panel A shows the standard deviation of the population density in bins of linear dimension $L/5$ as a function of the diffusion coefficient relative to the expected value if individuals where independently distributed across the habitat.
+Panel A shows the standard deviation of the population density in bins of linear dimension $L/5$ as a function of the diffusion coefficient relative to the expected value if individuals were independently distributed across the habitat.
 The thick black line shows the case of free diffusion, the colored lines density regulation with different interaction radii.
@@
 
-These simulations reveal that both at high and at low diffusion constants, assumptions of phylogeographic inference can be problematic: if diffusion constants are much smaller than then $L^2/T_{mrca}$, the population is fragmented into effective subpopulations and this fragmentation violates the assumptions of phylodynamic models.
+These simulations reveal that both at high and at low diffusion constants, assumptions of phylogeographic inference can be problematic: if diffusion constants are much smaller than $L^2/T_{mrca}$, the population is fragmented into effective subpopulations and this fragmentation violates the assumptions of phylodynamic models.
 A structured model would be more appropriate in this case.
@@
 \section*{How do changing habitats affect phylogeographic inferences?}
-As discussed above, even a static carrying capacity in an equilibrium population can strongly effect population structure, though this structure has limited effects on estimates of ancestral locations and diffusion coefficients beyond boundary effects.
+As discussed above, even a static carrying capacity in an equilibrium population can strongly affect population structure, though this structure has limited effects on estimates of ancestral locations and diffusion coefficients beyond boundary effects.
 In reality, carrying capacities vary through time.
@@
 The circulation of seasonal influenza viruses, for example, varies by many orders of magnitude between summer and winter in temperate climates.
-Fluctuations in carrying capacity mean that populations not just spread because individuals move, but also because populations grow in regions where the population density is below the carrying capacity.
+Fluctuations in carrying capacity mean that populations not only spread because individuals move, but also because populations grow in regions where the population density is below the carrying capacity.
 In such situations, the accuracy and interpretation of phylogeographic inferences are unclear.
@@
 The carrying capacity is illustrated in Fig.~\ref{fig:seasaw}A for 5 time points covering half a period.
-In between the two extreme left/right concentration of the carrying capacity, carrying capacity is uniform.
+In between the two extreme left/right concentration of the carrying capacity, the carrying capacity is uniform.
 To avoid extinction at low dispersal rates, selection is soft, meaning that the overall population size is kept approximately constant even if the entire population is stuck in the unfavorable region.
@@
 The population was sampled at a time when it had just contracted to the left and the positions of older nodes are not estimated correctly.
-Their true locations are to the right of the current population, but the inferred location are central in the shifted population.
-This is expected as there is no information in the current sample to suggest otherwise, but it highlights the potential to mislead.
+Their true locations are to the right of the current population, but the inferred locations are central in the shifted population.
+This is expected as there is no information in the current sample to suggest otherwise, but it highlights the potential of misleading inference, in particular when extrapolating far into the past.
 
@@
 Fig.~\ref{fig:seasaw}D shows the estimated diffusion constant (using Eq.~\ref{eq:dispersal_parameters}) in $x$ (solid lines) and $y$ (dashed) direction relative to the true value used in the simulations.
-At very high diffusion constants, both $D_x$ and $D_y$ are underestimated due to spatial constraint of carrying capacity as also discussed above (see Fig.~\ref{fig:density_reg}).
+At very high diffusion constants, both $D_x$ and $D_y$ are underestimated due to the spatial constraints of carrying capacity as also discussed above (see Fig.~\ref{fig:density_reg}).
 With decreasing $D$, $D_x$ starts to be overestimated and peaks at a value that depends on the period $T$, before decreasing to (noisy) estimates that are broadly compatible with the true value.
-In this range of dispersal rates, the population can no longer follow the shifting carrying capacity.
+In this range of dispersal rates, the population can no longer follow the shifting carrying capacity and the $T_{mrca}$ increases sharply (not shown).
 This decoupling happens at lower values of $D$ for longer periods of environmental change.
-With the parameters used in the simulation, the estimates of $D_x$ deviate by a factor of 2 from its true value, which is not dramatic compared to the variation of dispersal rates in nature.
+With the parameters used in the simulation, the estimates of $D_x$ deviate by a factor of 2 from its true value, which is not dramatic compared to the variation of dispersal rates in nature, but clearly shows that spatial inference can be substantially affected if the habitat changes on time scales comparable to the $T_{mrca}$.
 
-However, along with problematic inference of diffusion coefficients, estimates of ancestral locations are over-confident.
+Along with problematic inference of diffusion coefficients, estimates of ancestral locations are over-confident.
 Panel Fig.~\ref{fig:seasaw}E shows $z_x^2 = \langle (\hat{x} - x)^2/\sigma_x^2 \rangle$ and $z_y^2=\langle (\hat{y} - y)^2/\sigma_y^2\rangle$ for the oldest 20\% of internal nodes, where $\hat{x}$ and $\sigma_x$ are the mean and standard deviation of the posterior of $x$, respectively (and analogously for $y$).
@@
 Instead, it is as high as 20 for the simulations shown and these over-confident and biased estimates depend in non-monotonic ways on the true diffusion constant and the period of environmental change.
-With decreasing $D$, $z_x^2$ increases largely because the expected uncertainty decreases, before peaking when the population can no longer follow the shifting carrying capacity.
-A peak value of 10-20 indicates that the typical ancestral location of the deep nodes is 2-4 standard deviations away from the true location, an example of which is shown in Fig.~\ref{fig:seasaw}C.
-Conversely, at high dispersal rates, the average $z_x^2$ and $z_y$ are both smaller than one since the estimated uncertainty is larger than the true uncertainty since the diffusion is constrained by the boundaries of the habitat.
+With decreasing $D$, $z_x^2$ increases largely because the expected uncertainty decreases, before peaking before the population can no longer follow the shifting carrying capacity.
+A peak value of $z_x^2\approx 10-20$ indicates that the typical ancestral location of the deep nodes is 3-5 standard deviations away from the true location, an example of which is shown in Fig.~\ref{fig:seasaw}C.
+Conversely, at high dispersal rates, the average $z_x^2$ and $z_y^2$ are both smaller than one, i.e.~the inference is not confident enough.
+In this range, the estimated uncertainty is larger than the true uncertainty since the diffusion is constrained by the boundaries of the habitat.
 
-The above results show that behavior of a population in shifting habitats and the interpretation of phylogeographic inferences depend strongly on the rate of environmental change and the dispersal rate.
+The above results show that behavior of a population in shifting habitats and the interpretation of phylogeographic inferences can depend strongly on the rate of environmental change and how it compares to the dispersal rate.
 If the population can not follow shifting habitat, the situation is similar to the static case often leads to fragmented populations.
@@
 The dispersal rates necessary to keep up with the shifting carrying capacity can be estimated from the Fisher-KPP equation (Eq.~\ref{eq:FKPP}) above.
-This equation admits solutions with a traveling front, that the population ``invades'' the empty territory with velocity $v_{FKPP} = 2\sqrt{D \alpha}$ \citep{fisher_wave_1937,KPP1937,hallatschek_life_2010}.
+This equation admits solutions with a traveling front, i.e.~the population ``invades'' the empty territory with velocity $v_{FKPP} = 2\sqrt{D \alpha}$ \citep{fisher_wave_1937,KPP1937,hallatschek_life_2010}.
 Note that the model itself does not have an explicit parameter that has dimensions of a velocity.
@@
     % \includegraphics[width=0.48\textwidth]{figures/waves_v}
-    \caption{\label{fig:traveling_wave} Panel A shows an illustration of the simulated set-up: a carrying capacity with a Gaussian profile with $\sigma=0.5$ that moves along the $x$-axis with velocity $v$ on a patch with $L_x=3$, $L_y=1$, and periodic boundary conditions.
+    \caption{\label{fig:traveling_wave}{\bf Phylogeographic inference in a moving habitat.}
+    Panel A shows an illustration of the simulated set-up: a carrying capacity with a Gaussian profile with $\sigma=0.5$ that moves along the $x$-axis with velocity $v$ on a patch with $L_x=3$, $L_y=1$, and periodic boundary conditions.
     Panels B\&C show estimates of diffusion coefficients and velocities, relative to their true values, as a function of the ratio of the FKPP velocity to the external velocity $v$. }
@@
 Fig.~\ref{fig:traveling_wave}C shows estimates of $v$ using Eq.~\ref{eq:dispersal_parameters} relative to the true velocity.
-This ratio is monotonically increasing with the diffusion coefficient and shows a range where it close to 1, i.e.~where the estimated velocity of lineages agrees with the external velocity.
+This ratio is monotonically increasing with the diffusion coefficient and shows a range where it is close to 1, i.e.~where the estimated velocity of lineages agrees with the external velocity.
+This range coincides with the location of the peak of $D_w$.
 
@@
 Once $v\sim v_{FKPP}$, the estimated lineage velocity coincides with $v_{FKPP}$, while the estimated $D$ peaks at values orders of magnitude above the true value.
-For larger $v_{FKPP}$, the estimated velocity is stable for some time, before increasing again with $v_{FKKP}$.
+For values of $v_{FKPP}$ that are comparable the external velocity $v$, the estimated velocity is stable for some time and broadly agrees with $v$, before increasing again with $v_{FKKP}$.
 In this latter regime, diffusive motions dominates over directed motion and the estimates of $v$ suffer from the problems shown in Fig.~\ref{fig:D_and_v}.
@@
 The latter allow the reconstruction of the phylogenetic tree which encodes how closely different individuals are related to each other.
-Together with the spatial location of the samples -- the leafs of the tree -- one can reconstruct the likely locations of ancestors and estimate parameters of dispersal.
+Together with the spatial location of the samples -- the leaves of the tree -- one can reconstruct the likely locations of ancestors and estimate parameters of dispersal.
 Popular tools for such inference, however, make strong assumptions about the dispersal process, including assuming that dispersal of individuals is diffusive (i.e.~characterized by small displacements in random directions) and that spatial location is independent of the replication process.

### 2024-06-21 src/illustration.py
adapt plotting script

@@ -3,12 +3,12 @@ from free_diffusion import *
 from density_regulation import dict_to_phylo_tree
 import matplotlib.pyplot as plt
 from Bio import Phylo
-from matplotlib.cm import viridis_r 
+from matplotlib.cm import viridis_r
 
 if __name__=="__main__":
     n_leaves = 100
 
-    fig, axs = plt.subplots(2,2, figsize=(12,8))
+    fig, axs = plt.subplots(2,2, figsize=(6,8))
     for row, yule in [(0,True), (1,False)]:
         scale = np.log(n_leaves) if yule else 2
         tree = random_tree(n_leaves, yule=yule)
@@ -24,10 +24,11 @@ if __name__=="__main__":
             for c in n.clades:
                 axs[row,1].plot([n.pos['x'], c.pos['x']], [n.pos['y'], c.pos['y']], c='k', lw=0.5, alpha=0.5)
 
-        axs[row,1].scatter([n.pos['x'] for n in phylo_tree.find_clades()], [n.pos['y'] for n in phylo_tree.find_clades()], 
+        axs[row,1].scatter([n.pos['x'] for n in phylo_tree.find_clades()], [n.pos['y'] for n in phylo_tree.find_clades()],
                 c=[n.t for n in phylo_tree.find_clades()], s=30)
         axs[row,1].axis('equal')
         axs[row,1].scatter([0], [0], c='r', s=100)
-    
+        axs[row,1].set_axis_off()
+
+    plt.tight_layout()
     plt.savefig('figures/illustration_tree.pdf')
-    

### 2024-06-21 src/sample_size_dependence.py

@@ -23,25 +23,31 @@ if __name__=="__main__":
 
     import matplotlib.pyplot as plt
 
-    for coal in ['yule', 'kingman']:
+    fs=12
+    fig, axs = plt.subplots(1,2,figsize=(10,4), sharey=True, sharex=True)
+    for ax, coal in zip(axs, ['yule', 'kingman']):
+        ax.text(10,100, "Kingman coalescent" if coal=='kingman' else "Yule tree", fontsize=fs)
         res = results[coal]
-        plt.figure()
         offset = 1.0/1.02
         for suffix, suffix_label in [('_total', 'weighted'), ('_branch', 'by branch')]:
             for p, quantity in [('Dxy', 'diffusion'), ('vxy', 'velocity')]:
                 q = f"{p}{suffix}"
                 label = f"{quantity}, {suffix_label}"
-                plt.errorbar([n*offset for n in res], [res[n][q] for n in res], [res[n][q+"_std"] for n in res], ls = '-', marker='o',
-                            label=label)
+                ax.errorbar([n*offset for n in res], [res[n][q] for n in res],
+                             [res[n][q+"_std"] for n in res],
+                             ls = '-', marker='o', label=label)
                 offset *= 1.02**2
-        plt.plot([10, 3000], [1,1], ls='--', color='k')
-        plt.plot([10, 3000], [10,10*300**0.5], ls='-.', color='k')
+        ax.plot([10, 3000], [1,1], ls='--', color='k')
+        ax.set_xlabel('number of samples', fontsize=fs)
+        if coal=='kingman':
+            ax.plot([10, 3000], [10,10*300**0.5], ls='-.', color='k')
+        if coal=='yule':
+            ax.set_ylabel('estimate', fontsize=fs)
+            ax.legend()
+        ax.set_ylim(0.5, 200)
+        ax.set_xlim(7, 3600)
+        ax.set_yscale('log')
+        ax.set_xscale('log')
+    plt.tight_layout()
 
-        plt.xlabel('Number of leaves')
-        plt.ylabel('estimate')
-        plt.legend()
-        plt.yscale('log')
-        plt.xscale('log')
-        plt.ylim(0.5, 200)
-        plt.xlim(7, 3600)
-        plt.savefig(f'figures/{coal}_dispersal.pdf')
+    plt.savefig(f'figures/dispersal_stats.pdf')

### 2024-06-22 manuscript/phylogeo.tex
edits

@@
 \newcommand{\rvec}{\mathbf{r}}
+\newcommand{\vfkpp}{v_f}
 
 \begin{document}
-\title{On the robustness and consistency of phylogeographic inference}
+\title{On the robustness and consistency of continuous phylogeographic inference}
 \author{Richard A.~Neher}
+\affiliation{Swiss Institute of Bioinformatics, and Biozentrum, University of Basel, Switzerland \looseness=-5}
+
 \date{\today}
@@
 While the underlying probabilistic models can be complex and their parameters are often computationally demanding to infer, these models typically ignore that replication and population growth are tightly coupled to spatial location: populations expand into fertile uninhabited areas and contract in regions with limited resources.
-Such coupling, in particular when habitats change over time, can result in biased and overconfident estimates of ancestral locations and dispersal parameters.
-These distortions depend in complicated ways on the past dynamics of habitats and underlying population dynamics and dispersal processes such that validity of phylogeographic inferences is hard to assess.
+Here, I first investigate the sampling consistency of popular summary statistics of dispersal and show that estimators of ``lineage velocities'' are ill-defined.
+I then use simulations to investigate how local density regulation or shifting habitats perturb phylogeographic inference and show that these can result in biased and overconfident estimates of ancestral locations and dispersal parameters.
+These, sometimes dramatic, distortions depend in complicated ways on the past dynamics of habitats and underlying population dynamics and dispersal processes.
+Confidence statements that depend on the tails of posterior distributions or likelihood surfaces are particularly sensitive to model mispecification and the validity of phylogeographic inferences is therefore hard to assess.
 \end{abstract}
@@
 My focus here is only on the ability and robustness to estimate dispersal parameters and ancestral locations and I will use simulations with perfect knowledge of the tree and the sampling locations.
-So no tree-sampling will be necessary.
-This simplification is justified since the tree tends to depend much more strongly on sequence information and spatial information and will allow stronger conclusions.
+So no tree reconstruction will be necessary and sampling is uniform across the population.
+This simplification is justified since the tree tends to depend much more strongly on sequence information and spatial information and will allow clearer conclusions.
+In practice, tree reconstruction and sampling biases can create additional uncertainty and biases.
 Furthermore, I will assume strictly diffusive dispersal without any long-range migrations.
@@
 \begin{figure}
-    \includegraphics[width=0.53\textwidth]{figures/stable_density}
+    \includegraphics[width=0.48\textwidth]{figures/stable_density}
     \caption{\label{fig:density_reg}  {\bf Effects of population density regulation.}
     When spatial motion is independent of population density and coalescence is sufficiently rapid, the realized population density fluctuates strongly across the habitat.
-Panel A shows the standard deviation of the population density in bins of linear dimension $L/5$ as a function of the diffusion coefficient relative to the expected value if individuals were independently distributed across the habitat.
+    {\bf Panel A} shows the standard deviation of the population density in bins of linear dimension $L/5$ as a function of the diffusion coefficient relative to the expected value if individuals were independently distributed across the habitat.
 The thick black line shows the case of free diffusion, the colored lines density regulation with different interaction radii.
 For $r\ll L$, density fluctuations are strongly suppressed.
-    At the same time, the time to the most recent common ancestor increases dramatically (panel B), indicating that the population fragments in subpopulations that don't interact.
-    Panel C shows the fold-error of the dispersal estimator $D_w$ for periodic boundary conditions (solid lines, like panels A\&B) and reflecting boundaries (dashed lines).
+    At the same time, the time to the most recent common ancestor increases dramatically ({\bf panel B}), indicating that the population fragments in subpopulations that don't interact.
+    {\bf Panel C} shows the fold-error of the dispersal estimator $D_w$ for periodic boundary conditions (triangles, solid lines, like panels A\&B) and reflecting boundaries (circles, dashed lines).
     }
@@
     \includegraphics[width=\textwidth]{figures/seasaw}
-    \caption{\label{fig:seasaw} {\bf Phylogeographic estimates in rapidly changing environments.} Panel A shows an illustration of the simulated set-up: The carrying capacity first peaks on the left, becomes flat, and then peaks on the right before shifting back.
+    \caption{\label{fig:seasaw} {\bf Phylogeographic estimates in rapidly changing environments.} {\bf Panel A} shows an illustration of the simulated set-up: The carrying capacity first peaks on the left, becomes flat, and then peaks on the right before shifting back.
 The habitat size is $L_x=3$, $L_y=1$.
-    Panel B shows a tree from a simulation under these conditions with $D=0.0008$ and $N=500$, while panel C right shows the inferred (triangles) and true positions (circles) of internal nodes, highlight the root with large markers.
-Blue and green colors correspond to older nodes, yellow to recent nodes.
-    Panels D show estimates of diffusion coefficients in directions $x$ (solid lines) and $y$ (dashed lines) relative to their true values as a function of the true value of $D$.
-Panel E shows the average $z_x^2$ and $z_y^2$ for the oldest 20\% of internal nodes.
+    {\bf Panel B} shows a tree from a simulation under these conditions with $D=0.0008$ and $N=500$, while {\bf panel C} on the right shows the inferred (triangles, green) and true positions (circles, red) of internal nodes, highlight the root with large markers.
+    The population is sampled at a time when the carrying capacity is flat after having moved the left, that is the population is about to spread to the right.
+    {\bf Panels D} show estimates of diffusion coefficients in directions $x$ (solid lines) and $y$ (dashed lines) relative to their true values as a function of the true value of $D$.
+{\bf Panel E} shows the average $z_x^2$ and $z_y^2$ for the oldest 20\% of internal nodes.
 Inference of the x-coordinate of ancestral nodes is drastically overconfident for some values of $D$.
@@
 Inference of ancestral locations in such situations can be misleading as illustrated in Fig.~\ref{fig:seasaw}B.
-The population was sampled at a time when it had just contracted to the left and the positions of older nodes are not estimated correctly.
+The population was sampled at a time when the carrying capacity is flat after having contracted to the left and the positions of older nodes are not estimated correctly.
 Their true locations are to the right of the current population, but the inferred locations are central in the shifted population.
@@
 The dispersal rates necessary to keep up with the shifting carrying capacity can be estimated from the Fisher-KPP equation (Eq.~\ref{eq:FKPP}) above.
-This equation admits solutions with a traveling front, i.e.~the population ``invades'' the empty territory with velocity $v_{FKPP} = 2\sqrt{D \alpha}$ \citep{fisher_wave_1937,KPP1937,hallatschek_life_2010}.
+This equation admits solutions with a traveling front, i.e.~the population ``invades'' the empty territory with velocity $\vfkpp = 2\sqrt{D \alpha}$ \citep{fisher_wave_1937,KPP1937,hallatschek_life_2010}.
 Note that the model itself does not have an explicit parameter that has dimensions of a velocity.
@@
     \caption{\label{fig:traveling_wave}{\bf Phylogeographic inference in a moving habitat.}
-    Panel A shows an illustration of the simulated set-up: a carrying capacity with a Gaussian profile with $\sigma=0.5$ that moves along the $x$-axis with velocity $v$ on a patch with $L_x=3$, $L_y=1$, and periodic boundary conditions.
-    Panels B\&C show estimates of diffusion coefficients and velocities, relative to their true values, as a function of the ratio of the FKPP velocity to the external velocity $v$. }
+    {\bf Panel A} shows an illustration of the simulated set-up: a carrying capacity with a Gaussian profile with $\sigma=0.5$ that moves along the $x$-axis with velocity $v$ on a patch with $L_x=3$, $L_y=1$, and periodic boundary conditions.
+    {\bf Panels B\&C} show estimates of diffusion coefficients and velocities, relative to their true values, as a function of the ratio of the FKPP velocity to the external velocity $v$. }
 \end{figure}
@@
 Fig \ref{fig:traveling_wave}B\&C show simulation results for this traveling habitat.
-In Panel B, estimates of $D_x$ using Eq.~\ref{eq:dispersal_parameters} are shown as a function of the ratio of the FKPP velocity to the external velocity $v$.
+In Panel B, estimates of $D_x$ using Eq.~\ref{eq:dispersal_parameters} are shown as a function of the ratio of the FKPP velocity $v_f$ to the external velocity $v$.
 As expected, estimates of $\hat{D}_x$ are often dramatically too high (by up to factors of 1000).
-They peak right around when the external velocity matches the FKPP velocity, corresponding to the situation where the population dynamics is maximally affected by the moving carrying capacity.
+They peak right around when the external velocity matches the FKPP velocity $\vfkpp$, corresponding to the situation where the population dynamics is maximally affected by the moving carrying capacity.
 If the external velocity was lower, undirected dispersal contributes more and more to the displacement of lineages, while the population would be unable to follow the moving carrying capacity if the external velocity was higher.
@@
 
-These behaviors are explained comparing the speed $v_{FKPP}$ at which the population can invade new territory with the external speed $v$, the ratio of which is used as the $x$-axis in Fig.~\ref{fig:traveling_wave} B\&C.
-For $v_{FKPP} \ll v$, the population cannot keep up with the moving carrying capacity and is saved from extinction only because selection is soft.
-In this regime, the estimated velocity of lineages is below $v$ and both the estimated $v$ and $D$ increase with $v_{FKPP}$.
-Once $v\sim v_{FKPP}$, the estimated lineage velocity coincides with $v_{FKPP}$, while the estimated $D$ peaks at values orders of magnitude above the true value.
-For values of $v_{FKPP}$ that are comparable the external velocity $v$, the estimated velocity is stable for some time and broadly agrees with $v$, before increasing again with $v_{FKKP}$.
+These behaviors are explained comparing the speed $\vfkpp$ at which the population can invade new territory with the external speed $v$, the ratio of which is used as the $x$-axis in Fig.~\ref{fig:traveling_wave} B\&C.
+For $\vfkpp \ll v$, the population cannot keep up with the moving carrying capacity and is saved from extinction only because selection is soft.
+In this regime, the estimated velocity of lineages is below $v$ and both the estimated $v$ and $D$ increase with $\vfkpp$.
+Once $v\sim \vfkpp$, the estimated lineage velocity coincides with $\vfkpp$, while the estimated $D$ peaks at values orders of magnitude above the true value.
+For values of $\vfkpp$ that are comparable the external velocity $v$, the estimated velocity is stable for some time and broadly agrees with $v$, before increasing again with $v_{FKKP}$.
 In this latter regime, diffusive motions dominates over directed motion and the estimates of $v$ suffer from the problems shown in Fig.~\ref{fig:D_and_v}.
@@
 \section*{Discussion}
-Phylogeography aims to reconstruct location and migration of ancestors from the spatial distribution of extant individuals and their genome sequences.
+Phylogeography aims to reconstruct locations and migrations of ancestors from the spatial distribution of sampled individuals and their genome sequences.
 The latter allow the reconstruction of the phylogenetic tree which encodes how closely different individuals are related to each other.
 Together with the spatial location of the samples -- the leaves of the tree -- one can reconstruct the likely locations of ancestors and estimate parameters of dispersal.
-Popular tools for such inference, however, make strong assumptions about the dispersal process, including assuming that dispersal of individuals is diffusive (i.e.~characterized by small displacements in random directions) and that spatial location is independent of the replication process.
+Popular tools for such inference, however, make strong assumptions about the dispersal process, including assuming that dispersal of individuals is diffusive (i.e.~characterized by small displacements in random directions) and that the replication process is independent of spatial location.
 In this note, I have explored the robustness of such inference to violations of the latter assumption while keeping microscopic dispersal strictly diffusive, that is without any long-range migration.
@@
 A more complex set of questions concerns how coupling between growth rate of the population and spatial density affects phylogeographic inferences and how robust these inferences are to shifting habitats.
-As has long been known, ignoring the coupling between population growth and spatial location can lead to unrealistic population densities \citep{felsenstein_pain_1975}.
-Density regulation leads to homogeneous population densities.
-However, the population dynamics strongly depends on whether dispersal is fast enough to explore the entire habitat during the coalescence time $T_c$ of the population in absence of spatial structure.
+As has long been known, ignoring the coupling between population growth and spatial location can lead to unrealistic population densities \citep{felsenstein_pain_1975} and that density regulation is necessary.
+However, once density regulation is in place, the population dynamics depends strongly on whether dispersal is fast enough to explore the entire habitat during the coalescence time $T_c$ of the population in absence of spatial structure.
 If $D\ll L^2/T_c$, the population fragments into subpopulations and the time to the most recent common ancestor increases dramatically.
 The resulting strong coupling between spatial location and coalescence process violates typical tree priors used in such inference.
-In the other case of $D\gg L^2/T_c$, dispersal is rapid enough that ancestral locations of deep nodes are very uncertain compared to the size of the habitat.
+In the other case of $D\gg L^2/T_c$, dispersal is rapid enough that uncertainties of ancestral locations of deep nodes are large compared to the size of the habitat and such inferences thus uninformative.
+We only considered spatially constant and static carrying capacities, but carrying capacities that vary strongly in space lead to unexpected effective lineage dynamics corresponding to sources at where population densities are high and sinks where they are low \citep{wilkins_coalescent_2002}.
 
-Interpretation of such inference becomes even more challenging when habitats shift in time.
+Interpretation of phylogeographic inference becomes even more challenging when habitats shift in time.
 Depending on parameters, dispersal rates $D$ can be overestimated or underestimated, sometimes by large factors, and estimates of ancestral locations can be overconfident or too uncertain.
-Simulations reveal that this behavior depends on the relative magnitude of the rate at which the habitat shifts and the FKPP velocity which depends on the diffusion constant $D$ and the growth rate in empty regions $\alpha$, as well as curtailed dispersal by the boundaries of the habitat.
+Simulations reveal that this behavior depends on the relative magnitude of the rate at which the habitat shifts and the FKPP velocity which depends on the diffusion constant $D$ and the growth rate in empty regions $\alpha$.
 The parameter range where both ancestral locations and dispersal parameters are inferred correctly is narrow.
 
-Phylogeographic methods are often studied using West Nile virus and Rabies virus.
-West-Nile virus was first detected on the east coast of the US in 1999 and reached the West Coast five years later in 2004 \citep{pybus_unifying_2012}, translating to a wave front velocity of about 1000 km/year.
+Phylogeographic methods are often studied using the spread of West Nile virus in North America and Rabies virus.
+West-Nile virus was first detected on the East Coast of the US in 1999 and reached the West Coast five years later in 2004 \citep{pybus_unifying_2012}, translating to a wave front velocity of about 1000 km/year.
 \citet{pybus_unifying_2012} estimated a diffusion coefficient of between 200 and 5000 km$^2$/day, where the latter refers to branches with particularly high diffusion constant.
-Using $v_{FKPP} = 2\sqrt{D\alpha}$ and equating it to 1000 km/year $\approx$ 3km/day, we obtain ranges of $\alpha=\frac{9 km^2}{4 D day^2} = 0.012/day \ldots 0.00045/day$ or between $0.003$ and $0.08$/week.
-A pathogen that produces seasonal outbreaks with many orders of magnitude in prevalence between peak and troughs typically requires a growth rate of 1/week, considerably more than what the estimates of $D$ and the FKPP equation imply.
-Alternatively, if we assume $\alpha = 0.1/day$, we would expect $D = \frac{10 km^2}{0.4 day} = 25 km^2/day$, implying a daily exploration radius of about 5 km.
-This picture is also consistent with the apparent ``slowing down'' of lineages after the initial expansion across North America \citep{dellicour_epidemiological_2020}: Once the habitat was fully explored, directed invasion with $v_{FKPP}$ ceases.
+These extremely high estimates of diffusive dispersal are implausible.
+Alternatively, one can interpret the spread from east to west as a range expansion.
+To this end, we need an estimate of the growth rate of the population at low density, which we can roughly obtain from the fact that the virus is seasonal: over the course of a few months, its prevalence increases by orders of magnitude, which typically requires a doubling time of about a week or a growth rate of $\alpha=0.1/$day.
+Using $\vfkpp = 2\sqrt{D\alpha}$ and equating it to 1000 km/year $\approx$ 3km/day, we obtain estimates of $D$ of $D=\frac{9 \mathrm{km}^2}{4 \alpha \mathrm{day}^2} \approx 22\mathrm{km}^2/\mathrm{day}$ -- around 10 to 200 times smaller than estimates from phylogeography.
+This lower estimate of $D$ implies a daily exploration radius of about 5 km.
+This picture is also consistent with the apparent ``slowing down'' of lineages after the initial expansion across North America \citep{dellicour_epidemiological_2020}: Once the habitat was fully explored, directed range expansion with $\vfkpp$ ceases.
 Furthermore, rare long range dispersal could have strongly influenced the spread of the virus in ways not captured by Brownian or relaxed random walks \citep{hallatschek_acceleration_2014}.
@@
 The most recent common ancestor of various populations is between 30 and 100 years in the past.
-This would translate into rather limited spread of 150 to 500km of individual clades from their common ancestor, suggesting that these populations are highly structured (isolated by distance) and that their long range dispersal is dominated by rare introductions of the virus into new populations.
+This would translate into rather limited spread of 150 to 500km of individual clades from their common ancestor, suggesting that these populations are highly structured (isolated by distance) and that their long range dispersal is dominated by rare introductions of the virus into new populations \citep{brunker_elucidating_2015}.
 
@@
 On short time scales, a diffusive model will mostly reduce to a parsimonious reconstruction of ancestral locations which is robust to details of the model.
-It is longer time scales, that is deeper nodes in the phylogeny, for which ancestral inferences can get problematic as the chance that the environment has changed and shifted, or that the ancestral locations are in poorly sampled regions, increases.
-Inferences, and in particular confidence statements, can be highly unreliable and misleading.
-In absence of a good understanding of the dispersal properties and how population dynamics varies and time and space, simple models that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models.
+Sampling at multiple time points further constrains ancestral locations sufficiently strong that there is limited uncertainty, at least within the period for which samples are available.
+It is longer time scales, that is deeper nodes in the phylogeny before the earliest samples, for which ancestral inferences can get problematic as the chance that the environment has changed and shifted, or that the ancestral locations are in poorly sampled regions, increases.
+And while phylogeographic inferences will still often yield plausible results by virtue of their tendency to produce parsimonious solutions that aggregate available spatial locations, they can also be highly unreliable and misleading.
+In particular confidence statements should be treated with extreme caution, because the tails of the inferred distributions are very sensitive to past environmental shifts and biased sampling.
+In absence of a good understanding of the dispersal properties and how population dynamics varied over time and space, simple models that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models.
+
+\section*{Materials and methods}
+All simulation were performed in Python and the associated code is available in the github repository \href{https://github.com/neherlab/phylogeography}{github.com/neherlab/phylogeography}.
 

### 2024-06-22 src/parse_and_plot.py
update scripts

@@ -111,7 +111,8 @@ def make_figure(func, params, Lx=3, Ly=1, time_points=5, fname=None, panel_label
     for i, ax in enumerate(axs.flatten()):
         grid = np.meshgrid(np.linspace(0,Lx,30*Lx), np.linspace(0,Ly,30))
         vals = f(grid[0], grid[1], i)
-        ax.matshow(vals/np.max(vals), vmax=1, vmin=0)
+        norm = np.sum(vals)/30/Lx/15
+        ax.matshow(vals/norm, vmax=1, vmin=0)
         ax.set_axis_off()
         if panel_label and i==0:
             add_panel_label(ax, panel_label, x=-0.1, y=1.1)

### 2024-06-22 src/seasaw_example.py

@@ -60,38 +60,52 @@ if __name__=="__main__":
         axs[0].set_axis_off()
         print(f"N={len(terminal_nodes)}")
 
-        # plot true and inferred positions of nodes
-        axs[1].scatter([n.pos['x'] for n in phylo_tree.get_nonterminals()],
-                       [n.pos['y'] for n in phylo_tree.get_nonterminals()],
-                        c=[n.t for n in phylo_tree.get_nonterminals()], s=30)
-
-        axs[1].scatter([n.inferred_pos['x']['mean'] for n in phylo_tree.get_nonterminals()],
-                       [n.inferred_pos['y']['mean'] for n in phylo_tree.get_nonterminals()],
-                        c=[n.t for n in phylo_tree.get_nonterminals()], s=20, marker='^')
-
-        # highlight true and inferred root positions
-        n = phylo_tree.root
-        axs[1].scatter([n.pos['x']], [n.pos['y']], c='g', s=100, label='True root')
-        axs[1].scatter([n.inferred_pos['x']['mean']], [n.inferred_pos['y']['mean']], c='r', s=100, marker='^', label='Inferred root')
-#                c=[n.t for n in phylo_tree.find_clades()], s=30, marker='d', ls=None)
-        # z = [(n.inferred_pos['x']['mean'] - n.pos['x'])/n.inferred_pos['x']['var']**0.5 for n in phylo_tree.get_nonterminals()]
-        # tps = [n.t for n in phylo_tree.get_nonterminals()]
-        # axs[2].scatter(tps, z, c='k', s=30)
-
-        # add arrows indicating the link between true and inferred positions
-        for n in phylo_tree.get_nonterminals():
-            axs[1].arrow(n.pos['x'], n.pos['y'], n.inferred_pos['x']['mean'] - n.pos['x'],
-                         n.inferred_pos['y']['mean'] - n.pos['y'], lw=0.5, alpha=0.5)
-
         # indicate density
         x = np.linspace(0, Lx,101)
         dens = target_density(x, 0, t)
-        axs[1].plot(x, dens/np.max(dens), label='Carrying capacity')
-        axs[1].legend()
+        axs[1].plot(x, dens/np.max(dens), label='Carrying capacity', lw=3)
         axs[1].set_xlim(0,Lx)
         axs[1].set_xlabel(r'habitat $x$ coordinate')
         axs[1].set_ylabel(r'habitat $y$ coordinate')
+
+
[…]

### 2024-06-22 manuscript/phylogeo.tex
edits

@@
 Instead, it is as high as 20 for the simulations shown and these over-confident and biased estimates depend in non-monotonic ways on the true diffusion constant and the period of environmental change.
-With decreasing $D$, $z_x^2$ increases largely because the expected uncertainty decreases, before peaking before the population can no longer follow the shifting carrying capacity.
+With decreasing $D$, $z_x^2$ increases largely because the expected uncertainty decreases and the strongest over-confidence is observed when the dispersal capacity of the population is just about sufficient to follow the shifting carrying capacity.
 A peak value of $z_x^2\approx 10-20$ indicates that the typical ancestral location of the deep nodes is 3-5 standard deviations away from the true location, an example of which is shown in Fig.~\ref{fig:seasaw}C.

### 2024-06-22 manuscript/phylogeo.tex
small edits

@@
 \usepackage{color}
+\usepackage{hyperref}
 %\usepackage{cite}
@@
     \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
-    While diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ (weighted) and $v_b$ (by branch) are not well-defined quantities.
-Their numerical value depends strongly on sample size and there is no ground truth to compare to.
-The expected scaling $\sim \sqrt{n}$ of $v$ estimates is indicated by the dashed line.}
+    While the diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ (weighted) and $v_b$ (by branch) are not well-defined quantities.
+For Kingman coalescent tree their numerical value depends strongly on sample size and there is no ground truth to compare to.
+The expected scaling $v_w \sim \sqrt{n}$ of $v$ estimates is indicated by the dashed line.}
 \end{figure*}
@@
 \begin{equation}
-    \ref{eq:density_estimator}
+    \label{eq:density_estimator}
     \hat{\phi}(\rvec, t) = \sum_{i\in N} \frac{e^{-|\rvec_i - \rvec|^2/2r^2}}{\sqrt{2\pi r^2}}
@@
 In addition to the true location of ancestral nodes, I also infer their locations from the locations of the extant samples.
-Since the model is purely diffusive, the marginal distributions of ancestral locations and their mean and variance can be calculated exactly using recursion relations.
+For this inference, I make the common assumption of continuous phylogeography that growth and location are uncoupled and that dispersal is diffusive.
+In this case, the marginal distributions of ancestral locations and their mean and variance can be calculated exactly using recursion relations.
 
@@
 
+\section*{Acknowledgements}
+I am grateful to Louis de Plessis and Oskar Hallatscheck for stimulating discussions and useful feedback on earlier versions of this manuscript.
+
 \bibliography{bib}

### 2024-06-22 notes/most_likely_diffusion.tex
small edits

@@
 
+\section{common examples}
+
+\subsection{WNV}
+West-Nile virus was first detected on the east coast of the US in 1999 and reached the West Coast 5 years later in 2004.
+This translates to a wave front velocity of about 1000 km/year.
+Previous analysis estimated a diffusion coefficient of between 200 and 10000 km^2/day (Pybus et al).
+Using $v_{FKPP} = \sqrt{2D\alpha}$ and equating it to 1000 km/year = 3km/day, we obtain ranges of $\alpha=\frac{10 km^2}{2D day^2} = 0.025/day \ldots 0.0005/day$ or between $0.0035$ and $0.14$/week.
+A pathogen that produces seasonal outbreaks with many orders of magnitude in prevalence between peak and troughs typically requires a growth rate of 1/week, considerably more than what the estimates of $D$ and the FKPP equation imply.
+Alternatively, if we assume $\alpha = 0.1/day$, we would expect $D = \frac{10 km^2}{0.2 day} = 50 km^2/day$, implying a daily exploration radius of about 10km.
+This picture is also consistent with the apparent ``slowing down" if lineages after the initial expansion across North America: Once the habitat was fully explored, directed invasion with $v_{FKPP}$ ceases.
+
+\subsection{rabies}
+Dispersal rates of rabies are typically estimated to be between 500 and 1500 km^2/year.
+The most recent common ancestor of various populations is between 50 and 100 years in the past.
+This would translate into rather limited spread of 200 to 500km of individual clades from their common ancestor, suggesting that these populations are highly fragmented and their long range dispersal is dominated by rare introductions of the virus into new populations.
+Within each population, phylogeographic reconstruction will often give plausible results, but it is hard to extrapolate to longer time and distance scales.
+
 \end{document}

### 2024-07-02 manuscript/phylogeo.tex
edits and figure tweaks

@@
 \begin{document}
-\title{On the robustness and consistency of continuous phylogeographic inference}
+%\title{On the robustness and consistency of continuous phylogeographic inference}
+%\title{Phylogeography: unsafe at any speed}
+\title{Lost in the woods: shifting habitats can lead phylogeography astray}
 \author{Richard A.~Neher}
@@
 These, sometimes dramatic, distortions depend in complicated ways on the past dynamics of habitats and underlying population dynamics and dispersal processes.
-Confidence statements that depend on the tails of posterior distributions or likelihood surfaces are particularly sensitive to model mispecification and the validity of phylogeographic inferences is therefore hard to assess.
+Consequently, the validity of phylogeographic inferences, in particular when involving poorly sampled locations or extrapolations far into the past, is hard to assess and confidence can be much lower than suggested by the inferred posterior distributions.
 \end{abstract}
@@
 It is also mathematically convenient as unknown ancestral locations can be integrated in closed form and the marginal distribution of each position can be calculated exactly for a fixed tree along with the spatial likelihood.
-The total likelihood of the configuration is then assumed to be the product of the spatial likelihood and the phylogenetic likelihood, that is the tree topology and branching rates of the tree are assumed to be independent of the spatial process.
+The total likelihood of the configuration is then assumed to be the product of the spatial likelihood and the phylogenetic likelihood.
+This product assumption is equivalent to assuming that the tree topology and the branching rates of the tree are independent of the spatial location.
 This independence assumption is for example manifest in coalescent models assuming ``exchangeability'' of individuals, or birth-death models assuming identical rates on contemporary branches of the tree.
@@
 Such coupling can be accommodated in inference methods based on structured models that divide the population into discrete demes that could correspond to spatial locations that exchange migrants.
-Such models are flexible enough to allow for different population dynamics in different demes \citep{vaughan_efficient_2014}.
+Such models are flexible enough to allow for different population dynamics in different demes \citep{vaughan_efficient_2014}, though in practice can struggle to represent the population in sufficient detail or suffer from identifiability problems \citep{layan_impact_2023}.
 But the extent to which violations of the independence assumption affect continuous phylogeographic inferences is not well understood.
@@
     The total tree length of Yule trees is proportional to the number of tips $n$ and the time to the MRCA is $\log(n)$.
-    Kingman trees have a total tree length that is proportional to $\log(n)$ and are characterized by rapid merging close to the present. }
+    Kingman trees have a total tree length that is proportional to $\log(n)$ and are characterized by rapid merging close to the present.
+    Temporal and spatial scales are determined by parameter choice, only tree shape and spatial patterns are meaningful. }
 \end{figure}
@@
 Here, I explore what dispersal quantities can be robustly estimated and how sensitive inferences are to the fact that coupling between population dynamics and spatial location is ignored.
-My focus here is only on the ability and robustness to estimate dispersal parameters and ancestral locations and I will use simulations with perfect knowledge of the tree and the sampling locations.
+My focus here is only on estimates of dispersal parameters and ancestral locations and I will use simulations with perfect knowledge of the tree and the sampling locations.
 So no tree reconstruction will be necessary and sampling is uniform across the population.
@@
 Before investigating the more complex issues of coupling between spatial locations and reproduction, I will first explore the properties of summary statistics in simple models where spatial location and population dynamics are independent.
-Popular summaries of the dispersal process are empirical estimates of the diffusion constant $D_w$ \citep{pybus_unifying_2012,trovao_bayesian_2015} and a so-called \emph{dispersal velocity} $v_w$ \citep{dellicour_using_2017}.
+Popular summaries of the dispersal process are empirical estimates of the diffusion constant $D_w$ \citep{pybus_unifying_2012,trovao_bayesian_2015} and a so-called \emph{dispersal velocity} $v_w$ \citep{lemey_phylogeography_2010,dellicour_using_2017}.
 These are calculated from the displacements along branches of the trees sampled from the posterior.
@@
 While the estimator $D_w$ directly estimates a parameter of the model and its behavior has been studied in simulations \citep{pybus_unifying_2012}, it is unclear what $v_w$ is measuring and how it behaves.
-I simulated diffusion along trees from a Kingman coalescent\citep{kingman_coalescent_1982} and Yule model \citep{yule_iimathematical_1925} tree ensembles and evaluated the different estimators using the known true ancestral locations.
-Kingman trees are the classical neutral model of a constant size population, while Yule trees capture features of exponentially growing populations (see Fig.~\ref{fig:illustration_tree}).
+I simulated diffusion along trees from Kingman coalescent \citep{kingman_coalescent_1982} and Yule model \citep{yule_iimathematical_1925} tree ensembles and evaluated the different estimators using the known true ancestral locations.
+Kingman trees correspond to the classical neutral model of a constant size population, while Yule trees capture features of exponentially growing populations (see Fig.~\ref{fig:illustration_tree}).
 Examples for each ensemble are shown in Fig.~\ref{fig:illustration_tree}.
@@
 If the probability of long jumps with distance $x<d<x+dx$ decays more slowly than $\sim dx/x^3$, the expectation value of the $D_w$ diverges.
-This may mean that even $D_w$ becomes problematic when very long-tailed prior distributions for $D$ are used in relaxed random walk models.
+In this case, the values of the above estimators for a dataset with long range dispersal are dominated by the longest displacements and a single number like $D_w$ is no longer a useful summary of the typical dispersal process.
+Estimators like $D_w$ might also become problematic under the assumptions of relaxed random walk models with long-tailed prior distributions for $D$.
 
@@
 where $ \Delta_{\rvec}$ is the two-dimensional Laplace operator, $\phi_0$ is the carrying capacity, and $\alpha$ is the population growth rate at zero density.
-Such logistic density regulation was proposed by \citet{bolker_using_1997} and further analyzed by \citet{etheridge_survival_2004}.
-The density has a stable fixed point at $\phi=\phi_0$ we thus expect the population density to settle at a constant density $\phi_0$ and density fluctuations will decay with rate $\alpha$.
+Such logistic density regulation was proposed by \citet{bolker_using_1997} and further analyzed by \citet{etheridge_survival_2004}, but goes back to much earlier work by \citet{fisher_wave_1937} and \citet{KPP1937}.
+The density has a stable fixed point at $\phi=\phi_0$ we thus expect the population density to settle at a constant density $\phi_0$ and density fluctuations to decay with rate $\alpha$.
 
@@
 
-Instead of the coalescent trees used above, I now used an individual based simulation forward in time where individuals are assigned a fitness $f=\alpha(1-\hat{\phi}/\phi_0) + (1-N/N_0)$ and produce a Poisson distributed number of offspring with mean $1+f$.
+Instead of the coalescent trees used above, I now used an individual based simulation forward in time.
+Individuals are assigned a fitness $f=\alpha(1-\hat{\phi}/\phi_0) + (1-N/N_0)$ and produce a Poisson distributed number of offspring with mean $1+f$.
 The first term determines the fitness due to the local population density and has the tendency to stabilize this density at $\phi_0$, the second term $(1-N/N_0)$ keeps the overall population close to the target population size $N_0$.
@@
 
-Despite the substantial reduction in density fluctuations, the estimates of dispersal rates using Eq.~\ref{eq:dispersal_parameters} are not dramatically affected by density regulation, see Fig.~\ref{fig:density_reg}C.
-If boundary conditions are periodic, that is the habitat has the topology of a torus, estimates $D_w$ of the diffusion constant are accurate for most values of $D$ but slightly inflated when the population is heavily fragmented.
+Despite the substantial reduction in density fluctuations and strong population structure, the estimates $D_w$ of the diffusion constant using Eq.~\ref{eq:dispersal_parameters} are not dramatically affected by density regulation, see Fig.~\ref{fig:density_reg}C.
+If boundary conditions are periodic, that is the habitat has the topology of a torus, estimates $D_w$ of the diffusion constant are accurate for most values of $D$ but slightly inflated and noisy  when the population is heavily fragmented (the increased noise is due to a smaller number of effective samples when tMRCA is large).
 If instead of periodic boundary conditions reflective boundary conditions are used, diffusion constants are underestimated once $\sqrt{2DT_c}$ becomes comparable to the habitat size.
@@
 A structured model would be more appropriate in this case.
-If $D\gg L^2/T_{mrca}$, the boundaries of the habitat restrict free diffusion, resulting in underestimated dispersal parameters, while estimates of deeper ancestral locations have uncertainties comparable to the habitat size.
+If $D\gg L^2/T_{mrca}$, the boundaries of the habitat restrict free diffusion, resulting in underestimated dispersal parameters, while estimates of deeper ancestral locations have uncertainties comparable to the habitat size and are thus uninformative.
 
 \begin{figure*}
-    \includegraphics[width=\textwidth]{figures/seasaw_illustration}
     \includegraphics[width=\textwidth]{figures/seasaw_example}
     \includegraphics[width=\textwidth]{figures/seasaw}
-    \caption{\label{fig:seasaw} {\bf Phylogeographic estimates in rapidly changing environments.} {\bf Panel A} shows an illustration of the simulated set-up: The carrying capacity first peaks on the left, becomes flat, and then peaks on the right before shifting back.
+    \caption{\label{fig:seasaw} {\bf Phylogeographic estimates in rapidly changing environments.} {\bf Panel A} shows the time course of the carrying capacity through half a period.
+The carrying capacity first peaks on the left, becomes flat, and then peaks on the right before shifting back in the second half of the period.
 The habitat size is $L_x=3$, $L_y=1$.
-    {\bf Panel B} shows a tree from a simulation under these conditions with $D=0.0008$ and $N=500$, while {\bf panel C} on the right shows the inferred (triangles, green) and true positions (circles, red) of internal nodes, highlight the root with large markers.
+    {\bf Panel B} shows a tree from a simulation under these conditions with $D=0.0008$ and $N=500$, while {\bf panel C} on the right shows the inferred (triangles, orange) and true positions (circles, blue) of internal nodes, highlighting the root with large markers and a red border.
     The population is sampled at a time when the carrying capacity is flat after having moved the left, that is the population is about to spread to the right.
@@
 Such variation happens on all time scales, ranging from geological timescales over millions of years, environmental changes over millennia to seasonal fluctuations.
-The circulation of seasonal influenza viruses, for example, varies by many orders of magnitude between summer and winter in temperate climates.
+The circulation of seasonal influenza viruses in temperate climates, for example, varies by many orders of magnitude between summer and winter.
+
 Fluctuations in carrying capacity mean that populations not only spread because individuals move, but also because populations grow in regions where the population density is below the carrying capacity.
 In such situations, the accuracy and interpretation of phylogeographic inferences are unclear.
-
 In Fig.~\ref{fig:seasaw} we consider a case where the region with highest carrying capacity shifts from the left to the right in a periodic manner.
-The carrying capacity is illustrated in Fig.~\ref{fig:seasaw}A for 5 time points covering half a period.
+The carrying capacity is illustrated in Fig.~\ref{fig:seasaw}A for 20 time points covering half a period.
 In between the two extreme left/right concentration of the carrying capacity, the carrying capacity is uniform.
@@
 If the dispersal rate is sufficiently high, however, the population follows the shifting carrying capacity and lineages in the tree undergo directed back-and-forth motion.
-Inference of ancestral locations in such situations can be misleading as illustrated in Fig.~\ref{fig:seasaw}B.
+Inference of ancestral locations in such situations can be misleading as illustrated in Fig.~\ref{fig:seasaw}C.
 The population was sampled at a time when the carrying capacity is flat after having contracted to the left and the positions of older nodes are not estimated correctly.
@@
 
-The severity of such biases depends on the period at which the carrying capacity oscillates and on the dispersal rate, see  Fig.~\ref{fig:seasaw}C.
+The severity of such biases depends on the period at which the carrying capacity oscillates and on the dispersal rate.
 Fig.~\ref{fig:seasaw}D shows the estimated diffusion constant (using Eq.~\ref{eq:dispersal_parameters}) in $x$ (solid lines) and $y$ (dashed) direction relative to the true value used in the simulations.
@@
 
-Fig.~\ref{fig:traveling_wave}C shows estimates of $v$ using Eq.~\ref{eq:dispersal_parameters} relative to the true velocity.
+I showed above that  estimates of $v$ using Eq.~\ref{eq:dispersal_parameters} are ill-defined in models with diffusive dispersal, but it is worth exploring how they behave in cases when one can define a velocity within the model.
+Fig.~\ref{fig:traveling_wave}C shows the estimator $v_w$ in $x$ direction relative to velocity of the habitat.
 This ratio is monotonically increasing with the diffusion coefficient and shows a range where it is close to 1, i.e.~where the estimated velocity of lineages agrees with the external velocity.
@@
 Once $v\sim \vfkpp$, the estimated lineage velocity coincides with $\vfkpp$, while the estimated $D$ peaks at values orders of magnitude above the true value.
-For values of $\vfkpp$ that are comparable the external velocity $v$, the estimated velocity is stable for some time and broadly agrees with $v$, before increasing again with $v_{FKKP}$.
-In this latter regime, diffusive motions dominates over directed motion and the estimates of $v$ suffer from the problems shown in Fig.~\ref{fig:D_and_v}.
+Once $\vfkpp$ increases beyond the narrow range of $v\approx\vfkpp$, the velocity estimate $v_w$ increases further.
+In this latter regime, diffusive motions dominates over directed motion and the estimates of $v$ suffer from the problems of sample size dependence and lack interpretation as shown in Fig.~\ref{fig:D_and_v}.
 
@@
 Popular tools for such inference, however, make strong assumptions about the dispersal process, including assuming that dispersal of individuals is diffusive (i.e.~characterized by small displacements in random directions) and that the replication process is independent of spatial location.
-In this note, I have explored the robustness of such inference to violations of the latter assumption while keeping microscopic dispersal strictly diffusive, that is without any long-range migration.
+Here, I have explored the robustness of such inference to violations of the latter assumption while keeping microscopic dispersal strictly diffusive, that is without any long-range migration.
 
@@
 However, as discussed above, this velocity is determined by the interplay of dispersal and population growth at the front.
-The velocity is better determined by modeling the position of the front, rather than by tracking individual lineages along the tree.
+Estimators of a ``lineage velocity'' as defined in Eq.~\ref{eq:dispersal_parameters} only agree with invasion speed in a narrow, a priori unknown, parameter regime.
+The speed of invasion into a new habitat is better determined by modeling the position of the front \citep{pybus_unifying_2012}, rather than by tracking individual lineages along the tree.
 
@@
 However, once density regulation is in place, the population dynamics depends strongly on whether dispersal is fast enough to explore the entire habitat during the coalescence time $T_c$ of the population in absence of spatial structure.
-If $D\ll L^2/T_c$, the population fragments into subpopulations and the time to the most recent common ancestor increases dramatically.
+If $D\ll L^2/T_c$, the population fragments into subpopulations and the time to the most recent common ancestor increases dramatically compared to a neutral panmictic population.
 The resulting strong coupling between spatial location and coalescence process violates typical tree priors used in such inference.
 In the other case of $D\gg L^2/T_c$, dispersal is rapid enough that uncertainties of ancestral locations of deep nodes are large compared to the size of the habitat and such inferences thus uninformative.
-We only considered spatially constant and static carrying capacities, but carrying capacities that vary strongly in space lead to unexpected effective lineage dynamics corresponding to sources at where population densities are high and sinks where they are low \citep{wilkins_coalescent_2002}.
+At the same time, this will lead to under-estimation of dispersal parameters since individuals.
+I did not consider additional effects that emerge from static carrying capacities that vary strongly in space, which cab lead to unexpected effective lineage dynamics corresponding to sources at where population densities are high and sinks where they are low \citep{wilkins_coalescent_2002}.
 
@@
 Simulations reveal that this behavior depends on the relative magnitude of the rate at which the habitat shifts and the FKPP velocity which depends on the diffusion constant $D$ and the growth rate in empty regions $\alpha$.
-The parameter range where both ancestral locations and dispersal parameters are inferred correctly is narrow.
+The parameter range where both ancestral locations and dispersal parameters are inferred correctly is narrow and a priori unknown.
 
@@
 Alternatively, one can interpret the spread from east to west as a range expansion.
-To this end, we need an estimate of the growth rate of the population at low density, which we can roughly obtain from the fact that the virus is seasonal: over the course of a few months, its prevalence increases by orders of magnitude, which typically requires a doubling time of about a week or a growth rate of $\alpha=0.1/$day.
-Using $\vfkpp = 2\sqrt{D\alpha}$ and equating it to 1000 km/year $\approx$ 3km/day, we obtain estimates of $D$ of $D=\frac{9 \mathrm{km}^2}{4 \alpha \mathrm{day}^2} \approx 22\mathrm{km}^2/\mathrm{day}$ -- around 10 to 200 times smaller than estimates from phylogeography.
+To this end, we need an estimate of the growth rate $\alpha$ of the population at low density, which we can roughly obtain from the fact that the virus is seasonal: over the course of a few months, its prevalence increases by orders of magnitude, which typically requires a doubling time of about a week or a growth rate of $\alpha=0.1/$day.
+Using $\vfkpp = 2\sqrt{D\alpha}$ and equating it to 1000 km/year ($\approx$ 3km/day), we obtain estimates of $D$ of $D=\frac{9 \mathrm{km}^2}{4 \alpha \mathrm{day}^2} \approx 22\mathrm{km}^2/\mathrm{day}$ -- around 10 to 200 times smaller than estimates from phylogeography.
 This lower estimate of $D$ implies a daily exploration radius of about 5 km.
@@
 
-Dispersal rates of rabies virus are typically estimated to be between 500 and 1500 km$^2$/year \citep{dellicour_using_2017}.
+Diffusion constants of rabies virus are typically estimated to be between 500 and 1500 km$^2$/year \citep{dellicour_using_2017}.
 The most recent common ancestor of various populations is between 30 and 100 years in the past.
@@
 On short time scales, a diffusive model will mostly reduce to a parsimonious reconstruction of ancestral locations which is robust to details of the model.
-Sampling at multiple time points further constrains ancestral locations sufficiently strong that there is limited uncertainty, at least within the period for which samples are available.
-It is longer time scales, that is deeper nodes in the phylogeny before the earliest samples, for which ancestral inferences can get problematic as the chance that the environment has changed and shifted, or that the ancestral locations are in poorly sampled regions, increases.
-And while phylogeographic inferences will still often yield plausible results by virtue of their tendency to produce parsimonious solutions that aggregate available spatial locations, they can also be highly unreliable and misleading.
-In particular confidence statements should be treated with extreme caution, because the tails of the inferred distributions are very sensitive to past environmental shifts and biased sampling.
-In absence of a good understanding of the dispersal properties and how population dynamics varied over time and space, simple models that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models.
+Sampling at multiple time points further constrains ancestral locations sufficiently strongly that there is limited uncertainty, at least within the period for which samples are available.
+But inferences for deeper nodes in the phylogeny before the earliest samples can be problematic as the chance that the environment has changed and shifted, or that the ancestral locations are in poorly sampled regions, increases.
+And while phylogeographic inferences will still often yield plausible results by virtue of their tendency to produce parsimonious solutions that aggregate available spatial information, they can also be highly unreliable and misleading.
+In particular confidence statements should be treated with extreme caution, because the tails of the inferred distributions are determined by properties of the model and the underlying assumption and typically do not reflect the wide spectrum of possible environmental histories.
+In absence of a good understanding of the dispersal properties and how population dynamics varied over time and space, simple models that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models that nevertheless lack relevant history.
 
@@
 \section*{Acknowledgements}
-I am grateful to Louis de Plessis and Oskar Hallatscheck for stimulating discussions and useful feedback on earlier versions of this manuscript.
+I am grateful to Louis de Plessis, Oskar Hallatscheck, Simon Dellicour, and Philippe Lemey for stimulating discussions and useful feedback on earlier versions of this manuscript.
 
@@
 \end{document}
-
-The above example assumed a habitat that is constant over times comparable to the time to the most recent common ancestor of the population.
-If instead the habitat shifts on shorter time scales, the population has to move and such population movement will skew estimates of dispersal.
-We simulated such changing habitats by assuming the carrying capacity of the habitat varies periodically in time and space
-\begin{equation}
-    \phi_0(\rvec, t) = N\left(1 + \sin(2\pi (r_x/L_x + t/T)\cos(2\pi (r_y/L_y + t/2T))) \right)
-\end{equation}
-This results in moving peaks and troughs in carrying capacity (and thus population density) as illustrated in
-
-\begin{figure}
-    \includegraphics[width=\textwidth]{figures/habitats.png}
-    \caption{\label{fig:moving_habitats}}
-\end{figure}

### 2024-07-02 src/illustration.py

@@ -7,6 +7,7 @@ from matplotlib.cm import viridis_r
 
 if __name__=="__main__":
     n_leaves = 100
+    np.random.seed(1234)
 
     fig, axs = plt.subplots(2,2, figsize=(6,8))
     for row, yule in [(0,True), (1,False)]:
@@ -14,8 +15,23 @@ if __name__=="__main__":
         tree = random_tree(n_leaves, yule=yule)
         add_positions(tree['tree'], 1)
         phylo_tree = dict_to_phylo_tree(tree['tree'], child_attr='clades')
-        for n in phylo_tree.find_clades():
+        phylo_tree.ladderize()
+        tip_count = 0
+        for n in phylo_tree.find_clades(order='postorder'):
+            if n.is_terminal():
+                n.xvalue=tip_count
+                tip_count+=1
+            else:
+                minx = min([c.xvalue for c in n])
+                maxx = max([c.xvalue for c in n])
+                n.xvalue = (minx+maxx)/2
+
             n.color = [int(x) for x in viridis_r(-n.t/scale, bytes=True)[:3]]
+        phylo_tree.root.div = phylo_tree.root.branch_length
+        for n in phylo_tree.get_nonterminals(order='preorder'):
+            for c in n:
+                c.div = n.div + c.branch_length
+
 
         Phylo.draw(phylo_tree, axes=axs[row,0], label_func=lambda x: '')
         axs[row,0].set_title(f"{'Yule' if yule else 'Kingman'} tree")
@@ -29,6 +45,11 @@ if __name__=="__main__":
         axs[row,1].axis('equal')
         axs[row,1].scatter([0], [0], c='r', s=100)
         axs[row,1].set_axis_off()
+
+
+        axs[row,0].scatter([n.div for n in phylo_tree.find_clades()], [n.xvalue for n in phylo_tree.find_clades()],
+                c=[n.t for n in phylo_tree.find_clades()], s=30)
+        axs[row,0].scatter([phylo_tree.root.div], [phylo_tree.root.xvalue], c='r', s=100)
 
     plt.tight_layout()
     plt.savefig('figures/illustration_tree.pdf')

### 2024-07-02 src/parse_and_plot.py

@@ -3,8 +3,7 @@ import matplotlib.pyplot as plt
 import pandas as pd
 
 plt.rcParams.update({
-    "text.usetex": True,
-    "font.family": "Helvetica"
+    "text.usetex": True
 })
 
 

### 2024-07-02 src/plot_habitat_diffusion.py

@@ -7,7 +7,6 @@ from habitat_shifts import cycling_patches
 
 plt.rcParams.update({
     "text.usetex": True,
-    "font.family": "Helvetica"
 })
 
 def make_figure(fname=None):

### 2024-07-02 src/plot_seasaw.py

@@ -22,6 +22,7 @@ if __name__=="__main__":
     args = parser.parse_args()
     data = pd.read_csv(args.data)
     Lx, Ly = 3, 1
+    fs=12
 
     make_figure(seasaw, {'period':8, 'amplitude':1.1}, fname="figures/seasaw_illustration.pdf", panel_label='A')
 
@@ -51,8 +52,8 @@ if __name__=="__main__":
                                 label=f'', ls='--', c=f"C{ti%10}")
                                 #label=f'D_y; N={N}, T={T}', ls='--', c=f"C{ti%10}")
 
-    ax.set_xlabel(r'$DN/L_y^2$')
-    ax.set_ylabel(r'$\hat{D} / D$')
+    ax.set_xlabel(r'diffusion constant $[L_y^2/N]$', fontsize=fs)
+    ax.set_ylabel(r'$\hat{D} / D$', fontsize=fs)
     #ax.set_yscale('log')
     ax.set_ylim(0,2)
     ax.set_xscale('log')
@@ -75,8 +76,8 @@ if __name__=="__main__":
                 label=f'', ls='--', c=f"C{ti%10}", marker=m)
                 #label=f'N={N}, T={T}', ls='--', c=f"C{ti%10}", marker=m)
 
-    ax.set_ylabel(r'$(x - \hat{x})^2/\sigma^2$')
-    ax.set_xlabel(r'$DN/L_y^2$')
+    ax.set_ylabel(r'$(x - \hat{x})^2/2\sigma^2$', fontsize=fs)
+    ax.set_xlabel(r'diffusion constant $[L_y^2/N]$', fontsize=fs)
     #ax.set_yscale('log')
     ax.set_ylim(0)
     ax.set_xscale('log')

### 2024-07-02 src/plot_waves.py

@@ -19,7 +19,7 @@ if __name__=="__main__":
 
     # make an illustration of the carrying capacity at different times
     make_figure(func=waves, params={'width':0.5, 'velocity':1}, fname = args.illustration, panel_label='A')
-
+    fs=12
 
     # read and organize data
     data = pd.read_csv(args.data)
@@ -56,7 +56,7 @@ if __name__=="__main__":
                     except:
                         pass
 
-            ax.set_ylabel(r'$\hat{D}_x / D$')
+            ax.set_ylabel(r'$\hat{D}_x / D$', fontsize=12)
             # ax.set_xlabel(r'$\frac{2\sqrt{\alpha D}}{v} = \frac{v_{FKPP}}{v}$')
             ax.set_yscale('log')
             ax.set_xscale('log')
@@ -82,10 +82,10 @@ if __name__=="__main__":
                     except:
                         pass
 
-            ax.set_xlabel(r'$\frac{2\sqrt{\alpha D}}{v} = \frac{v_{FKPP}}{v}$')
-            ax.set_ylabel(r'$\hat{v}_x / v$')
-            ax.set_yscale('log')
-            ax.set_xscale('log')
+        ax.set_xlabel(r'$\frac{2\sqrt{\alpha D}}{v} = \frac{v_{FKPP}}{v}$', fontsize=14)
+        ax.set_ylabel(r'$\hat{v}_x / v$', fontsize=12)
+        ax.set_yscale('log')
+        ax.set_xscale('log')
             #ax.legend()
         #plt.colorbar()
         #plt.legend()
@@ -93,7 +93,7 @@ if __name__=="__main__":
 
     add_panel_label(ax, 'C')
     plt.tight_layout()
-
+    plt.savefig('figures/waves.pdf')
     ## ADDITIONAL FIGURES
 
     for v in velocity:

### 2024-07-02 src/seasaw_example.py

@@ -5,7 +5,7 @@ from density_regulation import make_node, evolve, subsample_tree, dict_to_phylo_
 from estimate_diffusion_from_tree import estimate_diffusion, estimate_ancestral_positions
 import matplotlib.pyplot as plt
 from Bio import Phylo
-from matplotlib.cm import viridis_r
+from matplotlib.cm import viridis_r,plasma
 
 
 if __name__=="__main__":
@@ -54,46 +54,55 @@ if __name__=="__main__":
                                         child_attr='clades')
         phylo_tree.ladderize()
 
-        fig, axs = plt.subplots(1,2, figsize=(12,3))
-        Phylo.draw(phylo_tree, axes=axs[0], label_func=lambda x: '')
-        add_panel_label(axs[0], 'B', x=-0.05, y=1.1)
-        axs[0].set_axis_off()
+        fig, axs = plt.subplots(1,3, figsize=(12,3))
+        Phylo.draw(phylo_tree, axes=axs[1], label_func=lambda x: '')
+        add_panel_label(axs[1], 'B', x=-0.05, y=1.1)
+        axs[1].set_axis_off()
         print(f"N={len(terminal_nodes)}")
 
-        # indicate density
+        add_panel_label(axs[0], 'A', x=-0.05, y=1.1)
         x = np.linspace(0, Lx,101)
-        dens = target_density(x, 0, t)
-        axs[1].plot(x, dens/np.max(dens), label='Carrying capacity', lw=3)
-        axs[1].set_xlim(0,Lx)
-        axs[1].set_xlabel(r'habitat $x$ coordinate')
-        axs[1].set_ylabel(r'habitat $y$ coordinate')
+        for t_illustrate in np.linspace(0,period/2,21):
+            dens = target_density(x, 0, t_illustrate)
+            axs[0].plot(x, dens/np.sum(dens)*30, label='Carrying capacity', lw=2, c=plasma(t_illustrate/period*2))
+
+        axs[0].set_xlim(0,Lx)
+        axs[0].set_xlabel(r'habitat $x$ coordinate')
+        axs[0].set_ylabel(r'carrying capacity')
+
+        # indicate density
+        # dens = target_density(x, 0, t)
+        # axs[2].plot(x, dens/np.max(dens), label='Carrying capacity', lw=3)
+        axs[2].set_xlim(0,Lx)
+        axs[2].set_xlabel(r'habitat $x$ coordinate')
+        axs[2].set_ylabel(r'habitat $y$ coordinate')
 
 
[…]

### 2024-07-03 manuscript/phylogeo.tex
final edits

@@
 It is also mathematically convenient as unknown ancestral locations can be integrated in closed form and the marginal distribution of each position can be calculated exactly for a fixed tree along with the spatial likelihood.
-The total likelihood of the configuration is then assumed to be the product of the spatial likelihood and the phylogenetic likelihood.
-This product assumption is equivalent to assuming that the tree topology and the branching rates of the tree are independent of the spatial location.
+But this model assumes dispersal of one lineage is independent of other lineages, and that dispersal properties are homogeneous across the habitat, and that the habitat does not change over time.
+Furthermore, it assumes that the tree topology and the branching rates of the tree are independent of the spatial location.
 This independence assumption is for example manifest in coalescent models assuming ``exchangeability'' of individuals, or birth-death models assuming identical rates on contemporary branches of the tree.
@@
 It has been known for a long time that ignoring such coupling leads to unrealistically heterogeneous population densities \citep{felsenstein_pain_1975}, as is evident in the example of the Kingman tree in Fig.~\ref{fig:illustration_tree}.
-If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D<L^2/T_c$, where $T_c$ is the coalescent timescale of the population.
+If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D\ll L^2/T_c$, where $T_c$ is the coalescent timescale of the population ($T_c=N$ in the case of a neutral model with a population of size $N$).
 Fig.~\ref{fig:density_reg}A (black line) quantifies this heterogeneity as the standard deviation of population density in two-dimensional bins of size $L/5$ relative to the expectation in the case of a spatially well-mixed population when each individual picks its location at random.
-For $D<L^2/2T_c$, the time to the most recent common ancestor $T_{mrca}$ is too small for the population to spread evenly in the habitat.
+For $D< L^2/2T_c$, diffusion is too slow to explore the entire habitat during the time since the MRCA, leading to clustering of individuals.
 This heterogeneity increases substantially for even lower $D$, when the population is essentially a cloud of width $\sqrt{4DT_c}\ll L$.
+The comparison of $D$ to the ratio $L^2/T_c$ will continue to be important below, and I therefore use diffusion constants in units of $L^2/T_c$.
 
@@
 For $r\ll L$, density fluctuations are strongly suppressed.
-    At the same time, the time to the most recent common ancestor increases dramatically ({\bf panel B}), indicating that the population fragments in subpopulations that don't interact.
+    At the same time, the time to the most recent common ancestor increases dramatically to values several fold higher than the neutral coalescence timescale $T_c=N=1000$ ({\bf panel B}), indicating that the population fragments in subpopulations that don't interact.
     {\bf Panel C} shows the fold-error of the dispersal estimator $D_w$ for periodic boundary conditions (triangles, solid lines, like panels A\&B) and reflecting boundaries (circles, dashed lines).
@@
 
-These simulations reveal that both at high and at low diffusion constants, assumptions of phylogeographic inference can be problematic: if diffusion constants are much smaller than $L^2/T_{mrca}$, the population is fragmented into effective subpopulations and this fragmentation violates the assumptions of phylodynamic models.
+These simulations reveal that both at high and at low diffusion constants, assumptions of phylogeographic inference can be problematic: if diffusion constants are much smaller than $L^2/T_{c}$, the population is fragmented into effective subpopulations and this fragmentation violates the assumptions of phylodynamic models.
 A structured model would be more appropriate in this case.
-If $D\gg L^2/T_{mrca}$, the boundaries of the habitat restrict free diffusion, resulting in underestimated dispersal parameters, while estimates of deeper ancestral locations have uncertainties comparable to the habitat size and are thus uninformative.
+If $D\gg L^2/T_{c}$, the boundaries of the habitat restrict free diffusion, resulting in underestimated dispersal parameters, while estimates of deeper ancestral locations have uncertainties comparable to the habitat size and are thus uninformative.
 
@@
 The habitat size is $L_x=3$, $L_y=1$.
-    {\bf Panel B} shows a tree from a simulation under these conditions with $D=0.0008$ and $N=500$, while {\bf panel C} on the right shows the inferred (triangles, orange) and true positions (circles, blue) of internal nodes, highlighting the root with large markers and a red border.
+    {\bf Panel B} shows a tree from a simulation under these conditions with $T=500$, $D=0.0008$ and $N=500$, while {\bf panel C} on the right shows the inferred (triangles, orange) and true positions (circles, blue) of internal nodes, highlighting the root with large markers and a red border.
     The population is sampled at a time when the carrying capacity is flat after having moved the left, that is the population is about to spread to the right.
@@
 In such situations, the accuracy and interpretation of phylogeographic inferences are unclear.
-In Fig.~\ref{fig:seasaw} we consider a case where the region with highest carrying capacity shifts from the left to the right in a periodic manner.
+In Fig.~\ref{fig:seasaw} we consider a case where the region with highest carrying capacity shifts from the left to the right in a periodic manner with period $T$.
 The carrying capacity is illustrated in Fig.~\ref{fig:seasaw}A for 20 time points covering half a period.
@@
 
-The severity of such biases depends on the period at which the carrying capacity oscillates and on the dispersal rate.
+The severity of such biases depends on the period $T$ at which the carrying capacity oscillates and on the dispersal rate and tends to be strongest when the period is of the same magnitude as the coalescent time scale.
 Fig.~\ref{fig:seasaw}D shows the estimated diffusion constant (using Eq.~\ref{eq:dispersal_parameters}) in $x$ (solid lines) and $y$ (dashed) direction relative to the true value used in the simulations.
@@
 In the other case of $D\gg L^2/T_c$, dispersal is rapid enough that uncertainties of ancestral locations of deep nodes are large compared to the size of the habitat and such inferences thus uninformative.
-At the same time, this will lead to under-estimation of dispersal parameters since individuals.
-I did not consider additional effects that emerge from static carrying capacities that vary strongly in space, which cab lead to unexpected effective lineage dynamics corresponding to sources at where population densities are high and sinks where they are low \citep{wilkins_coalescent_2002}.
+At the same time, this will lead to under-estimation of dispersal parameters since ancestral locations are constrained by the boundaries of the habitat.
+I did not consider additional effects that emerge from static carrying capacities that vary strongly in space, which cab lead to unexpected effective lineage dynamics corresponding to sources where population densities are high and sinks where they are low \citep{wilkins_coalescent_2002}.
 
@@
 West-Nile virus was first detected on the East Coast of the US in 1999 and reached the West Coast five years later in 2004 \citep{pybus_unifying_2012}, translating to a wave front velocity of about 1000 km/year.
-\citet{pybus_unifying_2012} estimated a diffusion coefficient of between 200 and 5000 km$^2$/day, where the latter refers to branches with particularly high diffusion constant.
+\citet{pybus_unifying_2012} estimated a diffusion coefficient between 200 and 5000 km$^2$/day, where the latter refers to branches with particularly high diffusion constant.
 These extremely high estimates of diffusive dispersal are implausible.
@@
 To this end, we need an estimate of the growth rate $\alpha$ of the population at low density, which we can roughly obtain from the fact that the virus is seasonal: over the course of a few months, its prevalence increases by orders of magnitude, which typically requires a doubling time of about a week or a growth rate of $\alpha=0.1/$day.
-Using $\vfkpp = 2\sqrt{D\alpha}$ and equating it to 1000 km/year ($\approx$ 3km/day), we obtain estimates of $D$ of $D=\frac{9 \mathrm{km}^2}{4 \alpha \mathrm{day}^2} \approx 22\mathrm{km}^2/\mathrm{day}$ -- around 10 to 200 times smaller than estimates from phylogeography.
+Using $\vfkpp = 2\sqrt{D\alpha}$ and equating it to 1000 km/year ($\approx$ 3km/day), we obtain estimates of $D$ of $D\approx \frac{9 \mathrm{km}^2}{4 \alpha \mathrm{day}^2} \approx 22\mathrm{km}^2/\mathrm{day}$ -- around 10 to 200 times smaller than estimates from phylogeography.
 This lower estimate of $D$ implies a daily exploration radius of about 5 km.
@@
 And while phylogeographic inferences will still often yield plausible results by virtue of their tendency to produce parsimonious solutions that aggregate available spatial information, they can also be highly unreliable and misleading.
-In particular confidence statements should be treated with extreme caution, because the tails of the inferred distributions are determined by properties of the model and the underlying assumption and typically do not reflect the wide spectrum of possible environmental histories.
+In particular confidence statements should be treated with extreme caution, because the tails of the inferred distributions are determined by properties of the model and the underlying assumptions typically do not reflect the wide spectrum of possible environmental histories.
 In absence of a good understanding of the dispersal properties and how population dynamics varied over time and space, simple models that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models that nevertheless lack relevant history.
@@
 \section*{Acknowledgements}
-I am grateful to Louis de Plessis, Oskar Hallatscheck, Simon Dellicour, and Philippe Lemey for stimulating discussions and useful feedback on earlier versions of this manuscript.
+I am grateful to Louis de Plessis, Oskar Hallatscheck, Emma Hodcroft, Simon Dellicour, and Philippe Lemey for stimulating discussions and useful feedback on earlier versions of this manuscript.
 

### 2024-07-03 src/plot_seasaw.py

@@ -52,7 +52,7 @@ if __name__=="__main__":
                                 label=f'', ls='--', c=f"C{ti%10}")
                                 #label=f'D_y; N={N}, T={T}', ls='--', c=f"C{ti%10}")
 
-    ax.set_xlabel(r'diffusion constant $[L_y^2/N]$', fontsize=fs)
+    ax.set_xlabel(r'diffusion constant $[L_y^2/T_c]$', fontsize=fs)
     ax.set_ylabel(r'$\hat{D} / D$', fontsize=fs)
     #ax.set_yscale('log')
     ax.set_ylim(0,2)
@@ -77,7 +77,7 @@ if __name__=="__main__":
                 #label=f'N={N}, T={T}', ls='--', c=f"C{ti%10}", marker=m)
 
     ax.set_ylabel(r'$(x - \hat{x})^2/2\sigma^2$', fontsize=fs)
-    ax.set_xlabel(r'diffusion constant $[L_y^2/N]$', fontsize=fs)
+    ax.set_xlabel(r'diffusion constant $[L_y^2/T_c]$', fontsize=fs)
     #ax.set_yscale('log')
     ax.set_ylim(0)
     ax.set_xscale('log')

### 2024-07-03 src/stable_density_figure.py

@@ -34,18 +34,19 @@ if __name__=="__main__":
     interaction_radius = data.interaction_radius.unique()
     density_reg = data.density_reg.unique()
     N_vals = data.N.unique()
+    interaction_radius = [0.1, 0.2, 0.3, 0.4, 0.5]
 
     # figure
     fig, axs = plt.subplots(3, 1, figsize=(5, 7.5), sharex=True)
 
     # Plot the heterogeneity
     ls = ['-', '-.', "--", ":"] #[:len(density_reg_to_plot)]
-    N=500
+    N=1000
     dr = density_reg[1]
     ax=axs[0]
-    for i, ir in enumerate(interaction_radius[1:]):
+    for i, ir in enumerate(interaction_radius):
         D_array = np.array(resPBC["density_variation"][ir, dr, N, :].index)
-        label = f'r={ir}' if N==N_vals[1] else ''
+        label = f'r={ir}'
         ax.plot(D_array*N/Lx/Ly, resPBC["density_variation"][ir, dr, N, :]/np.sqrt(nbins/N),
                 label=label, ls='-', c=f"C{i%10}")
         # ax.plot(D_array*N/Lx/Ly, resPBC["density_variation"][ir, dr, N, :]/np.sqrt(nbins/N),
@@ -64,7 +65,7 @@ if __name__=="__main__":
 
     ## Plot the diffusion estiamtes
     ax=axs[2]
-    for i, ir in enumerate(interaction_radius[1:]):
+    for i, ir in enumerate(interaction_radius):
         D_array = np.array(resPBC["diffusion_mean"][ir, dr, N, :].index)
         label = f'r={ir}, a={dr}' if N==N_vals[0] else ''
         ax.errorbar(D_array*N/Lx/Ly, res["diffusion_mean"][ir, dr, N, :]/D_array,
@@ -76,7 +77,7 @@ if __name__=="__main__":
 
     ax.plot(N*D_array/Lx/Ly, np.ones_like(D_array), c='k')
 #    ax.set_xlabel('true $ND/L^2$')
-    ax.set_xlabel(r'diffusion constant $[L^2/N]$')
+    ax.set_xlabel(r'diffusion constant $[L^2/T_c]$')
     ax.set_ylabel(r'fold-error in $D$ estimate')
     # plt.yscale('log')
     ax.set_xscale('log')
@@ -103,7 +104,7 @@ if __name__=="__main__":
 
[…]

### 2025-03-09 manuscript/phylogeo.tex
change from last year

@@
 It is also mathematically convenient as unknown ancestral locations can be integrated in closed form and the marginal distribution of each position can be calculated exactly for a fixed tree along with the spatial likelihood.
-But this model assumes dispersal of one lineage is independent of other lineages, and that dispersal properties are homogeneous across the habitat, and that the habitat does not change over time.
+But this model assumes dispersal of one lineage is independent of other lineages, that dispersal properties are homogeneous across the habitat, and that the habitat does not change over time.
 Furthermore, it assumes that the tree topology and the branching rates of the tree are independent of the spatial location.
@@
     \label{eq:density_estimator}
-    \hat{\phi}(\rvec, t) = \sum_{i\in N} \frac{e^{-|\rvec_i - \rvec|^2/2r^2}}{\sqrt{2\pi r^2}}
+    \hat{\phi}(\rvec, t) = \sum_{i\in N} \frac{e^{-|\rvec_i - \rvec|^2/2r^2}}{\sqrt{4\pi r^2}}
 \end{equation}

### 2025-03-24 manuscript/ResponseToReviews.tex
work on response

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{a4paper}                   		% ... or a4paper or a5paper or ...
\usepackage{color}
\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex

\usepackage{hyperref}

\newcommand{\nt}[1]{\texttt{#1}}
\newcommand{\mut}[2]{\nt{#1}$\to$\nt{#2}}
\usepackage{amssymb}
\newcommand{\comment}[1]{{\bf Comment:} #1}
\newcommand{\response}[1]{{\color{black}{\bf Response:} #1}}

\title{Response to reviews of ``Lost in the woods: shifting habitats can lead phylogeography astray'' for \textit{Virus Evolution}}
\author{
Richard A. Neher
}

\begin{document}
%\raggedright
\maketitle

\emph{Below, the original comments {\color{blue} are in blue}, and our responses are in black.}

\color{blue}

\response{
Thank you for the opportunity to revise our manuscript and for the thoughtful comments from the reviewers.
I particularly appreciate the detailed constructive feedback from Simon Dellicour.
I hope I have addressed all points in a satisfactory manner and that this manuscript will be a valuable contribution to the discussions of the value and pitfalls of phylogeographic inference.

Below are point-by-point responses to each reviewer.
}

\section*{Reviewer 1 comments}

{\bf Comments for the Authors.}
n this study, the author rigorously critiques continuous phylogeographic models and shows how their posterior inferences are unreliable due to shifts/changes in the habitats or environmental conditions, particularly in poorly sampled areas. I enjoyed reading this well-written paper and analyses, and I might agree with many points raised by the author. However, I have also issues with his illustration summarized in the following points:

\response{Thank you very much for this assessment and for highlighting where the manuscript falls short. I address these points individually below.}

{\bf Major concerns:}

1. He is basing his conclusions on simulated data, not actual examples of real sequences for specific species. In fact, I think the author needs to illustrate his points using sequence data from different species.

\response{The purpose of this manuscript is to show that the assumptions of continuous phylogeographic methods can be problematic in several ways. To show that, I need scenarios where the ground truth is known and this is not the case for observational data.
Furthermore, we are not proposing an alternative method to analyze data.
I discuss the expansion of West Nile virus in the US as an example where the assumptions of continuous phylogeographic models are likely violated and show how a different set of assumptions leads to dramatically different conclusions.}

2. Also, it is very important to note that most of the published results from phylogeographic models do exhibit acceptable epidemiological plausibility to biology and the distribution of species in space and time.

\response{I discuss this towards the end of the manuscript. Phylogeographic methods tend to produce a parsimonious reconstruction of past migrations which is often plausible (and will often also cover the actual history). However, these inferences depend on certain assumptions about the past and extrapolations into the past are dangerous. }

3. Also, the journal is specialized in virus evolution, so it is important to shift the scope of the paper to the implications of such quantitative findings to the biology and genomics of viruses or similar rapidly evolving pathogens.

\response{The methods that discussed in this paper are widely used in the analysis of viral sequences. I discuss examples of Rabius virus and West Nile virus. }

\section*{Reviewer 1 comments}
- general comment: given the popularity of phylogeographic approaches, especially in the field of molecular epidemiology to infer the dispersal history of viral lineages, the limitations addressed and explored in this study are of importance for the community of users of those approaches. In addition to a series of more minor comments and suggestions listed below, I would have a comment about a potential ambiguity on the possibility to apply an alternative approach based on a structured coalescent model that would allow to circumvent the coupling issue illustrated in the present study focusing on the continuous phylogeographic approach (see the next two comments). The focus of this study is indeed on the continuous diffusion model, and I think that this is important to underline that the author did not demonstrate that in practice the use of such a alternative approach would allow avoiding this issue (even if I'm not saying that he should, see my related comment just below). While a criticism of these limitations inherent to the current implementation of the continuous diffusion model implemented in BEAST are totally relevant, the other phylogeographic approaches/models are also associated with a number of assumptions and limitations (for instance, any outcome of phylogeographic inference will be impacted by sampling bias).

\response{I agree -- see response to the next points.}

- p1, related to the above comment, "though in practice can struggle to represent the population in sufficient detail or suffer from identifiability problems (Layan et al., 2023)": indeed, while structured coalescent models appear in theory like an interesting alternative to accommodate the coupling between the replication process and local environmental conditions, I however cannot think of a practical example. Coalescent models are based on demes (or "population"/"subpopulation", depending on the terminology used) within which panmixia is hypothesised, and the broader the deme is on the map, the more this assumption will be difficult to meet in practice. Furthermore, to allow such a relevant coupling, the study area should be divided into an important number of distinct (adjacent) demes of small sizes, close to a fine-scale grid setting. As the use of structured coalescent models becomes computationally demanding when increasing the number of considered demes, my impression here is that we are far from being able to apply this alternative. Of course, I don't pretend having an exhaustive view on past studies applying those models, but it seemed to be that most of the time a relatively restricted number of demes are considered with today genomic datasets (made of, e.g., hundreds of viral genomes). Overall, my comment here is that in practice, and at least at the moment, structured coalescent models do not offer an alternative to the continuous diffusion model to circumvent the coupling issue. As the present study aims to address this issue, I think that inherent limitations associated with structured coalescent models, here presented as an alternative, should be more explicitly developed in the Introduction (or Discussion), even if the focus of the study is on the continuous diffusion model. Otherwise, and of course again in my opinion, it could sound like a recommendation of not using that particular phylogeographic approach because of the coupling issue while other approach could handle it, which is as far as I know not the case in practice.

\response{I fully agree with the reviewer that structured coalescent models are typically not a practical alternative. They are difficult to parameterize, suffer from identifiability problems, and any simple model structure will likely suffer from similar issues as pointed out here for the continuous diffusion model.
I merely meant to point out than in principle they could accommodate the coupling between population dynamics and spatial structure.

In the manuscript, I have now emphasized that structured models can be considered an alternative in principle, but that in practice don't capture the real world complexities.
}

- p3, also related to the above comment, "These simulations reveal that both at high and at low diffusion constants, assumptions of phylogeographic inference can be problematic: if diffusion constants are much smaller than L2/Tc, the population is fragmented into effective subpopulations and this fragmentation violates the assumptions of phylodynamic models. A structured model would be more appropriate in this case": in theory yes, but in practice is it even something that could be considered at the moment? In other words, could someone use the exact same simulations and reproduce the assessments presented in Figure 3 but with a structured coalescent model? Without any additional development, the above statement suggests that it could be the case... In my view, it should either be demonstrated in the present study, or such a statement clearly be nuanced due to practical limitations.

\response{I have rephrased this to make clear that a structured model can in principle accommodate the coalescence dynamics and provide a better tree prior, but since the demes are unknown, this is not a practical alternative.}

- Abstract, "Continuous phylogeographic inference is a popular method to reconstruct the spatial distribution of ancestral populations and estimate parameters of the dispersal process": I wouldn't talk about "spatial distribution of ancestral populations" as I think that it could be read as the spatial distribution of the population of the target organism at some points in the past; while phylogeographic inference will only aim at inferring the past position of the ancestors of the sampled individuals. I'm of course not saying that's what the author meant, just a minor comment on the terminology used. For instance, the definition/description given in the first paragraph of the Introduction or in the first sentence of the Discussion section seems more accurate to me.

\response{Thank you for pointing out this imprecise phrasing. I have reworded this sentence to ``\emph{... reconstruct the spatial location of ancestors of extant populations...}''}

- Abstract: in it, the author never mentioned that his investigations solely focused on the continuous phylogeographic inference. Imo, without this precision (that is explicitly stated in the first sentence of the second paragraph of the Introduction section), it could sound that most popular phylogeographic models/approaches were assess in regard to those limitations.

\response{I agree that it is important to make this point clear. The first word of the abstract already read `Continuous`, but I have added additional qualifiers to clarify that the focus is on continuous phylogeographic inference.}

- p1, Introduction section: some developments associated with or related to the continuous diffusion model are not mentioned or addressed in that section. For instance: Fitzjohn et al. (2009, doi: 10.1093/sysbio/syp067), Guindon et al. (2016, doi: 10.1016/j.tpb.2016.05.002), Gill et al. (2017, doi: 10.1093/sysbio/syw093).

\response{thank you for pointing out these references. However, I could not see the relevance of 10.1093/sysbio/syp067 to the present study. Fitzjohn et al study estimation of speciation rates in incompletely resolved phylogenies. I have added the reference Guidon et al when discussing challenges of defining consistent spatial models for inference. Gill et al seem mostly relevant in the context of directional random walks models. I have added a reference to this work when discussing range expansions.}

- p1, "But this model assumes dispersal of one lineage is independent of other lineages, that dispersal properties are homogeneous across the habitat, and that the habitat does not change over time": not completely true for the RRW model allowing branch dispersal velocity to vary across branches (which could potentially lead to its heterogeneity across the habitat).

\response{This statement refers to the simplest model in the equation just above -- but point taken. My understanding is that RRW models assume a distribution of dispersal rates that can vary along the tree, but that this distribution is the same everywhere. This can indeed result in different inferred dispersal properties in different parts of the habitat. I added a clarifying statement. }

- p2, "Popular summaries of the dispersal process are empirical estimates of the diffusion constant Dw (Pybus et al., 2012; Trovao et al., 2015) and a so called dispersal velocity vw (Dellicour et al., 2017; Lemey et al., 2010)": the Dellicour et al. (2017) reference is irrelevant here as this paper is not introducing this metric (nor using it for the comparison of dispersal capacities). In Dellicour et al. (2017), the authors used two diffusion coefficients, and not a lineage dispersal metric, to measure and compare the dispersal capacities of RABV lineages in various settings. In addition to Lemey et al. (2010), the author could e.g. cite Raghwani et al. (2011, PLoS Path., PMID: 21655108), which is, in my knowledge, one of the first (or the first?) study using a lineage dispersal velocity metric in practice.

\response{Thank you for pointing out the inaccuracies in the referencing and suggesting additional references. I have removed Dellicour 2017 and added Raghwani et al. 2011.}

- p2, "While the estimator Dw directly estimates a parameter of the model and its behavior has been studied in simulations (Pybus et al., 2012)": Pybus et al. didn't use any simulation in their study. Except if the author aimed to refer to a different study, in my knowledge, an analysis of the diffusion coefficient(s) behaviour based on simulations was also lacking (until recent related works mentioned above and the present study).

\response{Thank you for pointing out this error. I had confused myself. Pybus et al discuss how this estimator uses fundamental properties of diffusion -- with a reference to Einstein!. I reworded this sentence.}

- p3, "While the estimator Dw as defined in Eq. 2 is well behaved when the model is exactly diffusive, long range dispersal can affect Dw dramatically. If the probability of long jumps with distance $x < d < x+dx$ decays more slowly than $\sim dx/x^3$, the expectation value of the Dw diverges": as it could be of interest for future works involving the estimation of $D_w$, I was wondering if the author could further develop the later statement to explain why this is actually expected. It is maybe obvious, but at least not for me, so, if relevant, a more detailed development could be inserted (?). Also, and my apologies if I missed something here, but how do you define a "long jump"?

\response{
The estimator $D_w$ is an empirical average over the squared displacement along different branches, scaled by the inverse elapsed time.
However, RRW models sample jumps from a Cauchy distribution which does not have a mean nor a variance. The central limit theorem does not apply for sums of many such variables and estimators based on such sums will be dominated by the largest jumps. Estimates of $D_w$ will therefore be very noisy and often dominated by largest jumps possible in the systems.
I think the broad distribution are fine as fairly uninformative priors for inference, the estimate $D_w$ of a forward process with long jumps is not well defined.
}

- Figure 3: I had a hard time understanding it. I would for instance suggest to be explicit with what you mean by "rel. heterogeneity" on the y-axis of panel A (which is "the standard deviation of population density in two dimensional bins of size L/5"), or at least to re-state it in the figure's legend. Also, I would re-define, within the figure's legend, to what "r" refers to (and its unit?). Finally, in panel C, this is difficult to distinguish the curves and their associated symbols (symbols that could also be re-defined in a colour legend included within the box of the graph).
\response{TODO}

- p3, "For $D < L2/2Tc$, diffusion is too slow to explore the entire habitat during the time since the MRCA, leading to clustering of individuals": maybe mark it with a vertical mark (/dashed line) in Figure 3?

\response{TODO}

- p3, "Phylogeographic inference typically assumes that the spatial diffusion process is independent of the branching pattern of the tree" (and in other parts of the text): maybe better to explicitly refer to continuous phylogeographic inference here?

\response{Done.}

- p3, regarding the definition of Tc ("the coalescent timescale of the population"): do we agree that you meant the coalescent timescale of the population as we consider a panmictic one (which makes it different from the TMRCA that can be larger if, e.g., the population is rather spatially structured)?

\response{Yes, I have added `panmictic'.}

- p5, "For this inference, I make the common assumption of continuous phylogeography that growth and location are uncoupled and that dispersal is diffusive": the RRW is far more common that the strict BM in phylogeographic analyses (personally, I cannot think of a single application of the strict Brownian motion model).

\response{I would argue as stated, these assumptions are true for both the BM and RRW model. The latter allows the diffusion coefficient to vary which allows to accommodate a lot of scenarios in inference, but it does make the tree prior in anyway depend on location. }.

- Figure 4: a colour scale is missing in panel A, which I didn't find intuitive. Also, in the legend: "moved [to?] the left"?

\response{Thank you for pointing out the missing word. I didn't think that adding a color scale would help much, but I have added sentence to the caption detailing how yellow and purple correspond to the most extreme habitats.}

- p5, "This is expected as there is no information in the current sample to suggest otherwise, but it highlights the potential of misleading inference, in particular when extrapolating far into the past": (i) do you get the same trend while considering a RRW model? (ii) Is the process only sample at the end? (What happens if we have dated samples?) (iii) Without a complex time varying environment, a RRW model with drift (Gill et al. 2017) might potentially achieve the same kind of result where the root does not lie in between the samples.

\response{This statements refers explicitly to 'extrapolation' into the past and the inference in this case was made form data gathered at one time point. If there were samples from the time of the $T_{mrca}$ to the present, the task would be interpolation rather than extrapolation and I would expect the biases to disappear. Integrating over a distribution of diffusion constants will likely broaden the confidence intervals and thus make the estimates more consistent with reality, but I don't expect the bias to go away.}

- similar questions related to the results presented in Figure 5: (i) What would be the outcome if a RRW model was used? (ii) What about the "RRW with drift" (Gill et al. 2017), which appears particularly suited here? (iii) How is the process sampled in time? (All at the end or dynamically?)

\response{Note that the dispersal model used to generate these data is purely diffusive with no drift. The only part that is time varying is the habitat. The point of this analysis is to illustrate the importance of coupling between location and growth. It would probably be possible to capture this in a model with drift, but one would again estimate an effective parameter that does not describe the dispersal process. Instead, these data should be analysed using a wavefront model. As before, this analysis is focussed on snapshots from a single time point. I have added a statement to the legend to make this clear.}

- p6, "This equation admits solutions with a traveling front, i.e. ...": not clear to me to which equation this is referring to (?).

\response{I have made the equation reference explicit.}

- p7, "The common practice to estimate dispersal “velocities” using displacements of lineages along the tree (Dellicour et al., 2021) is problematic": the cited paper (Dellicour et al., 2021) is a "protocol" dedicated to the use of the RRW model, not a paper on dispersal metrics (they are neither described or used in it). I therefore don't think that it is an appropriate reference for the above statement. Furthermore, and personally, while I do acknowledge having reported and compared lineage dispersal velocity metrics in previous works (frequently along estimates and comparison of diffusion coefficient), I'm not sure it is correct to refer to this as a "common" practice. (In addition to the non-appropriate reference, I would suggest dropping the "common").

\response{As a protocol paper that gives an overview of methods and has a paragraph listing these statistics (albeit in a section on related methods), I felt this was an ok reference. Dellicour et al, MBE, 2021 refers to range of other papers by Dellicour in this context. (Dellicour 2017 seems to be mislabeling diffusion coefficients as accelerations). But I am happy to drop that reference and reword. }

- p7, "Firstly, the numerical value of such estimates depends strongly on the sample size and tree ensemble. Secondly, the underlying model is diffusive and has no parameters that have dimensions of a velocity and even if the quantity could be estimated robustly, there is no interpretation of the quantity within the model". These two points have also been treated extensively in our recent study (Dellicour et al. 2024, PLoS Biol., in press; preprint available here: https://www.biorxiv.org/content/10.1101/2024.04.10.588821v2). While I think it might be relevant to cite this work here, I would of course be in some sort of conflict of interest if 'insisting' on this. Moreover, I am aware that both works were conducted at the same time and while mutually sharing ideas and feedback.

\response{I am aware of these works, and I am happy to include a reference. But for context, I think it is important to note that I pointed out these problems to the authors in the summer of 2023, including figures that illustrate the problem and an interpretation why that is. They invited me to be a co-author on a study into which their analysis of the problem was integrated, but I ultimately didn't feel comfortable with some of the other content of the work and declined to be a co-author. I gave them the go-ahead to publish this, and as I see now I am acknowledged in this work for giving 'useful feedback'. I decided to write up my thoughts on this problem myself, resulting in this work that has been on biorxiv since July last year.

I have now included a reference and a note.}

- p8. "Furthermore, rare long range dispersal could have strongly influenced the spread of the virus in ways not captured by Brownian or relaxed random walks": a RRW model can, in principle, capture long-range dispersal events.

\response{It can accommodate such jumps, but the inferred parameters still won't be meaningful. }

- p8, "This picture is also consistent with the apparent “slowing down” of lineages after the initial expansion across North America (Dellicour et al., 2020): Once the habitat was fully explored, directed range expansion with vf ceases": agree, and purely FYI this is somehow something already mentioned in that study ("which could reflect a consequence of increased bird immunity through time slowing down spatial dispersal").

\response{Thanks for highlighting.}

\section*{Reviewer 3 comments}

Phylogeography aims to embed phylogenetic trees in geographic space to infer past locations and movements. This manuscript critically examines the assumptions underlying the most influential phylogeographic inference methods and assesses their robustness. Neher demonstrates that a widely used summary statistic, “lineage speed,” derived from diffusion-based models, is fundamentally flawed as it depends on sample size. More critically, under realistic demographic scenarios—such as expanding populations or those not strictly governed by local density regulation—lineages exhibit strong backward-in-time biases, despite unbiased forward-time movements. This discrepancy can result in biased and overconfident phylogeographic inferences, which often assume unbiased lineage movement backward in time. Based on these findings, Neher concludes that the validity of phylogeographic inference methods is difficult to assess and should not be trusted without additional information, such as detailed time-series data.

This paper offers an important and long-overdue critique of phylogeographic methods, which have gained significant popularity due to their visual appeal—a popularity that, in light of this critique, appears questionable. The manuscript is exceptionally well-written and pedagogical, making the limitations of diffusion-based methods strikingly clear. The discussion is both illuminating and practical, with helpful examples.

\response{I thank the reviewer for this positive assessment.}

1.  ``This model assumes that the dispersal of one lineage is independent of other lineages, that dispersal properties are homogeneous across the habitat, and that the habitat does not change over time. Furthermore, it assumes that the tree topology and branching rates are independent of spatial location.'' I agree with these points, but it would be helpful to explain why these assumptions are implied by the seemingly innocuous Eq. 1. Users of diffusion-based phylogeographic inference methods may not be aware that these assumptions are inherent.

\response{Thank you for this suggestion. I have expanded this section.}

2. The section on density regulation seems to assume global or habitat-wide density regulation. However, it is plausible that many species are density-regulated over a characteristic length scale, such as their dispersal distance during their lifetime (or from birth to reproductive age). In that case, wouldn’t the parameter  L  (habitat size) be replaced with the scale over which density is regulated? If so, this should be mentioned. A static, homogeneous, and locally density-regulated population should then approximately follow Eq. 1.

\response{TODO}

3. ``which cab lead to unexpected effective lineage dynamics corresponding to sources where population densities are high and sinks where they are low'': Replace ``cab'' with ``can''. Additionally, in this context, beyond Wilkins and Wakeley, it may be worthwhile to cite the Appendix B of Hallatschek, O., & Nelson, D. R. (2008). Gene surfing in expanding populations. Theoretical population biology, 73(1), 158-170.

\response{Done.}

4. It might be helpful to provide guidance on whether users can make quantitative estimates or perform tests to assess the reliability of their phylogeographic inferences. For example, how dense does a time series need to be to sufficiently constrain the spatiotemporal embedding? Alternatively, this manuscript could call for studies that address this issue. Following this work, the onus is on the phylogeographic community to develop best practices for these methods.

\response{TODO}

\end{document}

### 2025-03-24 manuscript/phylogeo.tex
work on response

@@
 \begin{abstract}
-Continuous phylogeographic inference is a popular method to reconstruct the spatial distribution of ancestral populations and estimate parameters of the dispersal process.
+Continuous phylogeographic inference is a popular method to reconstruct the spatial location of ancestors of extant populations and estimate parameters of the dispersal process.
 While the underlying probabilistic models can be complex and their parameters are often computationally demanding to infer, these models typically ignore that replication and population growth are tightly coupled to spatial location: populations expand into fertile uninhabited areas and contract in regions with limited resources.
 Here, I first investigate the sampling consistency of popular summary statistics of dispersal and show that estimators of ``lineage velocities'' are ill-defined.
-I then use simulations to investigate how local density regulation or shifting habitats perturb phylogeographic inference and show that these can result in biased and overconfident estimates of ancestral locations and dispersal parameters.
+I then use simulations to investigate how local density regulation or shifting habitats perturb phylogeographic inference in continuous space and show that these can result in biased and overconfident estimates of ancestral locations and dispersal parameters.
 These, sometimes dramatic, distortions depend in complicated ways on the past dynamics of habitats and underlying population dynamics and dispersal processes.
-Consequently, the validity of phylogeographic inferences, in particular when involving poorly sampled locations or extrapolations far into the past, is hard to assess and confidence can be much lower than suggested by the inferred posterior distributions.
+Consequently, the validity of continuous phylogeographic inferences, in particular when involving poorly sampled locations or extrapolations far into the past, is hard to assess and confidence can be much lower than suggested by the inferred posterior distributions.
 \end{abstract}
@@
 As organisms replicate, they also disperse in space, resulting in mixing of the population and exploration of new habitats.
-The reconstruction of ancestral locations and past migrations from sampling locations of extant individuals, typically in combination with genome sequence information to infer the phylogeny, is known as phylogeography.
+The reconstruction of ancestral locations, past migrations, and dispersal parameters from sampling locations of extant individuals, typically in combination with genome sequence information to infer the phylogeny, is known as phylogeography.
 Phylogeographic methods are implemented in popular evolutionary analysis software such as BEAST \citep{lemey_bayesian_2009,lemey_phylogeography_2010}.
+However, defining consistent models for demographic inference in continuous space is challenging \citep{felsenstein_pain_1975,guindon_demographic_2016}.
 
@@
     P(\rvec_c| \rvec_p, t) = \frac{1}{4\pi D t}e^{\frac{|r_c - r_p|^2}{4Dt}}
+    \label{eq:diffusion}
 \end{equation}
@@
 Furthermore, it assumes that the tree topology and the branching rates of the tree are independent of the spatial location.
-This independence assumption is for example manifest in coalescent models assuming ``exchangeability'' of individuals, or birth-death models assuming identical rates on contemporary branches of the tree.
+Other model component, for example those modeling the tree topology, coalescence properties, and molecular evolution are independent of Eq.~\ref{eq:diffusion}, that is their contributions to the likelihood are multiplicative factors.
+A coalescent model, for example, would assume that each pair of lineage merges with equal rates, or a birth-death models would assume identical rates on contemporary branches of the tree.
+Variation of dispersal across the tree can be accounted for in relaxed random walk models, but coupling between growth and location can only be captured in indirect ways.
 
@@
 
-Such coupling can be accommodated in inference methods based on structured models that divide the population into discrete demes that could correspond to spatial locations that exchange migrants.
+Such coupling can in principle be accommodated in inference methods based on structured models that divide the population into discrete demes that could correspond to spatial locations that exchange migrants.
 Such models are flexible enough to allow for different population dynamics in different demes \citep{vaughan_efficient_2014}, though in practice can struggle to represent the population in sufficient detail or suffer from identifiability problems \citep{layan_impact_2023}.
@@
 So no tree reconstruction will be necessary and sampling is uniform across the population.
-This simplification is justified since the tree tends to depend much more strongly on sequence information and spatial information and will allow clearer conclusions.
-In practice, tree reconstruction and sampling biases can create additional uncertainty and biases.
+This simplification is justified since the tree topology tends to be informed much more strongly by sequence information.
+Using the true trees will result in clearer demonstrations of the limits of phylogeographic inference without the additional complication of tree uncertainty.
 Furthermore, I will assume strictly diffusive dispersal without any long-range migrations.
+In practice, long-range migrations, errors in tree reconstruction, and sampling biases will add to the problems investigated here.
 
@@
 Before investigating the more complex issues of coupling between spatial locations and reproduction, I will first explore the properties of summary statistics in simple models where spatial location and population dynamics are independent.
-Popular summaries of the dispersal process are empirical estimates of the diffusion constant $D_w$ \citep{pybus_unifying_2012,trovao_bayesian_2015} and a so-called \emph{dispersal velocity} $v_w$ \citep{lemey_phylogeography_2010,dellicour_using_2017}.
+Popular summaries of the dispersal process are empirical estimates of the diffusion constant $D_w$ \citep{pybus_unifying_2012,trovao_bayesian_2015} and a so-called \emph{dispersal velocity} $v_w$ \citep{lemey_phylogeography_2010,raghwani_endemic_2011}.
 These are calculated from the displacements along branches of the trees sampled from the posterior.
@@
 The estimators in Eq.~\ref{eq:dispersal_parameters} are commonly used to summarize dispersal of lineages.
-While the estimator $D_w$ directly estimates a parameter of the model and its behavior has been studied in simulations \citep{pybus_unifying_2012}, it is unclear what $v_w$ is measuring and how it behaves.
+While the estimator $D_w$ directly estimates a parameter of the model using fundamental properties of diffusion \citep{pybus_unifying_2012}, it is unclear what $v_w$ is measuring and how it behaves.
 I simulated diffusion along trees from Kingman coalescent \citep{kingman_coalescent_1982} and Yule model \citep{yule_iimathematical_1925} tree ensembles and evaluated the different estimators using the known true ancestral locations.
-Kingman trees correspond to the classical neutral model of a constant size population, while Yule trees capture features of exponentially growing populations (see Fig.~\ref{fig:illustration_tree}).
+Kingman trees correspond to the classical neutral model of a constant size population, while Yule trees capture features of exponentially growing populations.
 Examples for each ensemble are shown in Fig.~\ref{fig:illustration_tree}.
@@
 There is no ground truth to compare the estimates to and $v_w$ and $v_b$ are just descriptive summaries of spatial spread that can not be compared between datasets.
+Similar observations have recently been reported by \citet{dellicour_how_2024}\footnote{I had alerted \citet{dellicour_how_2024} in May 2023 to the problems with the velocity estimators and illustrated the behavior using figures qualitatively similar to Fig.~\ref{fig:D_and_v}}.
 
@@
 \section*{Effects of density regulation on population structure and dispersal}
-Phylogeographic inference typically assumes that the spatial diffusion process is independent of the branching pattern of the tree and ignores coupling between the two.
+Continuous phylogeographic inference typically assumes that the spatial diffusion process is independent of the branching pattern of the tree and ignores coupling between the two.
 Such a coupling, however, is expected since organisms compete for resources and growth rates naturally depend on local population density.
 It has been known for a long time that ignoring such coupling leads to unrealistically heterogeneous population densities \citep{felsenstein_pain_1975}, as is evident in the example of the Kingman tree in Fig.~\ref{fig:illustration_tree}.
-If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D\ll L^2/T_c$, where $T_c$ is the coalescent timescale of the population ($T_c=N$ in the case of a neutral model with a population of size $N$).
+If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D\ll L^2/T_c$, where $T_c$ is the coalescent timescale of a panmictic population ($T_c=N$ in the case of a neutral model with a population of size $N$).
 Fig.~\ref{fig:density_reg}A (black line) quantifies this heterogeneity as the standard deviation of population density in two-dimensional bins of size $L/5$ relative to the expectation in the case of a spatially well-mixed population when each individual picks its location at random.
@@
 These simulations reveal that both at high and at low diffusion constants, assumptions of phylogeographic inference can be problematic: if diffusion constants are much smaller than $L^2/T_{c}$, the population is fragmented into effective subpopulations and this fragmentation violates the assumptions of phylodynamic models.
-A structured model would be more appropriate in this case.
+This could potentially be accounted for in structured coalescent model, but demes are not discrete and not known a priori such that these models are unlikely a practical alternative.
 If $D\gg L^2/T_{c}$, the boundaries of the habitat restrict free diffusion, resulting in underestimated dispersal parameters, while estimates of deeper ancestral locations have uncertainties comparable to the habitat size and are thus uninformative.
@@
     \caption{\label{fig:seasaw} {\bf Phylogeographic estimates in rapidly changing environments.} {\bf Panel A} shows the time course of the carrying capacity through half a period.
+The color indicates time along the periodic habitat oscillations with yellow and purple representing most turning points of the oscillations.
 The carrying capacity first peaks on the left, becomes flat, and then peaks on the right before shifting back in the second half of the period.
@@
     {\bf Panel B} shows a tree from a simulation under these conditions with $T=500$, $D=0.0008$ and $N=500$, while {\bf panel C} on the right shows the inferred (triangles, orange) and true positions (circles, blue) of internal nodes, highlighting the root with large markers and a red border.
-    The population is sampled at a time when the carrying capacity is flat after having moved the left, that is the population is about to spread to the right.
-    {\bf Panels D} show estimates of diffusion coefficients in directions $x$ (solid lines) and $y$ (dashed lines) relative to their true values as a function of the true value of $D$.
+    The population is sampled at a time when the carrying capacity is flat after having moved to the left, that is the population is about to spread to the right.
+    {\bf Panel D} show estimates of diffusion coefficients in directions $x$ (solid lines) and $y$ (dashed lines) relative to their true values as a function of the true value of $D$.
 {\bf Panel E} shows the average $z_x^2$ and $z_y^2$ for the oldest 20\% of internal nodes.
 Inference of the x-coordinate of ancestral nodes is drastically overconfident for some values of $D$.
+The summary inferences in panels D \& E are made from trees of samples from a single time point.
 }
@@
 Inference of ancestral locations in such situations can be misleading as illustrated in Fig.~\ref{fig:seasaw}C.
-The population was sampled at a time when the carrying capacity is flat after having contracted to the left and the positions of older nodes are not estimated correctly.
+The population was sampled at a single time point when the carrying capacity is flat after having contracted to the left and the positions of older nodes are not estimated correctly.
 Their true locations are to the right of the current population, but the inferred locations are central in the shifted population.
@@
 The dispersal rates necessary to keep up with the shifting carrying capacity can be estimated from the Fisher-KPP equation (Eq.~\ref{eq:FKPP}) above.
-This equation admits solutions with a traveling front, i.e.~the population ``invades'' the empty territory with velocity $\vfkpp = 2\sqrt{D \alpha}$ \citep{fisher_wave_1937,KPP1937,hallatschek_life_2010}.
+Eq.~\ref{eq:FKPP} admits solutions with a traveling front, i.e.~the population ``invades'' the empty territory with velocity $\vfkpp = 2\sqrt{D \alpha}$ \citep{fisher_wave_1937,KPP1937,hallatschek_life_2010}.
 Note that the model itself does not have an explicit parameter that has dimensions of a velocity.
@@
 
-The common practice to estimate dispersal ``velocities'' using displacements of lineages along the tree \citep{dellicour_relax_2021} is problematic.
+Estimating dispersal ``velocities'' using displacements of lineages along the tree is problematic.
 Firstly, the numerical value of such estimates depends strongly on the sample size and tree ensemble.
@@
 In cases where a population invades a novel habitat, e.g.~the invasion of North America by the West-Nile virus \citep{pybus_unifying_2012}, the moving front of the population defines a bona fide velocity.
-However, as discussed above, this velocity is determined by the interplay of dispersal and population growth at the front.
+However, as discussed above, this velocity is determined by the interplay of dispersal and population growth at the front, not a directional motion of individuals.
+It might, however, be possible to capture this effective directional motions by random walks models that allow for directed motion \citep{gill_relaxed_2017}.
 Estimators of a ``lineage velocity'' as defined in Eq.~\ref{eq:dispersal_parameters} only agree with invasion speed in a narrow, a priori unknown, parameter regime.
@@
 At the same time, this will lead to under-estimation of dispersal parameters since ancestral locations are constrained by the boundaries of the habitat.
-I did not consider additional effects that emerge from static carrying capacities that vary strongly in space, which cab lead to unexpected effective lineage dynamics corresponding to sources where population densities are high and sinks where they are low \citep{wilkins_coalescent_2002}.
+I did not consider additional effects that emerge from static carrying capacities that vary strongly in space, which can lead to unexpected effective lineage dynamics corresponding to sources where population densities are high and sinks where they are low \citep{wilkins_coalescent_2002, hallatschek_gene_2008}.
 
@@
 In particular confidence statements should be treated with extreme caution, because the tails of the inferred distributions are determined by properties of the model and the underlying assumptions typically do not reflect the wide spectrum of possible environmental histories.
-In absence of a good understanding of the dispersal properties and how population dynamics varied over time and space, simple models that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models that nevertheless lack relevant history.
+In absence of a good understanding of the dispersal properties and how population dynamics varied over time and space, simple models that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models that nevertheless fail to faithfully infer history or essential parameters.
 

### 2025-03-30 manuscript/ResponseToReviews.tex
work on response

@@
 {\bf Comments for the Authors.}
-n this study, the author rigorously critiques continuous phylogeographic models and shows how their posterior inferences are unreliable due to shifts/changes in the habitats or environmental conditions, particularly in poorly sampled areas.
+In this study, the author rigorously critiques continuous phylogeographic models and shows how their posterior inferences are unreliable due to shifts/changes in the habitats or environmental conditions, particularly in poorly sampled areas.
 I enjoyed reading this well-written paper and analyses, and I might agree with many points raised by the author.
@@
 To show that, I need scenarios where the ground truth is known and this is not the case for observational data.
-Furthermore, we are not proposing an alternative method to analyze data.
-I discuss the expansion of West Nile virus in the US as an example where the assumptions of continuous phylogeographic models are likely violated and show how a different set of assumptions leads to dramatically different conclusions.}
+Furthermore, we are not proposing an alternative method to analyze data -- so it is unclear to me what exactly the reviewer is asking me to do.
+I discuss the expansion of West Nile virus in the US as an example where the assumptions of continuous phylogeographic models are likely violated and show how a different set of assumptions leads to dramatically different conclusions.
+}
 
@@
 However, these inferences depend on certain assumptions about the past and extrapolations into the past are dangerous.
+
+I have extended this section in the discussion to make clear when plausible inferences are expected, and when inferences should interpreted with care.
 }
@@
 
-\response{The methods that discussed in this paper are widely used in the analysis of viral sequences. I discuss examples of Rabius virus and West Nile virus. }
+\response{The methods that discussed in this paper are widely used in the analysis of viral sequences and often were developed in the context of viral evolution.
+While the implications of some of the issues might apply more broadly, I feel that phylogeographic inference is sufficiently central to the analysis of viral genomic data and the readership of Virus Evolution, that the current focus is justified.
+I discuss examples of Rabius virus and West Nile virus. }
 
-\section*{Reviewer 1 comments}
+\section*{Reviewer 2 comments}
 - general comment: given the popularity of phylogeographic approaches, especially in the field of molecular epidemiology to infer the dispersal history of viral lineages, the limitations addressed and explored in this study are of importance for the community of users of those approaches.
@@
 
-\response{thank you for pointing out these references.
-However, I could not see the relevance of 10.1093/sysbio/syp067 to the present study.
+\response{Thank you for pointing out these references.
+However, I could not see the relevance of Fitzjohn et al 10.1093/sysbio/syp067 to the present study.
 Fitzjohn et al study estimation of speciation rates in incompletely resolved phylogenies.
@@
 The estimator $D_w$ is an empirical average over the squared displacement along different branches, scaled by the inverse elapsed time.
-However, RRW models sample jumps from a Cauchy distribution which does not have a mean nor a variance.
+However, the Cauchy distribution is often used to model long range dispersal, but it does not have a finite mean nor a variance.
 The central limit theorem does not apply for sums of many such variables and estimators based on such sums will be dominated by the largest jumps.
 Estimates of $D_w$ will therefore be very noisy and often dominated by largest jumps possible in the systems.
-I think the broad distribution are fine as fairly uninformative priors for inference, the estimate $D_w$ of a forward process with long jumps is not well defined.
+In the concrete case of a Cauchy distribution, the largest jump will typically be twice the distance of the second largest jump, 3 times as large as the third largest jump, etc.
+The sum in the estimator $D_w$ is them dominated by the largest jump (and hence hugely variable).
+I have added a footnote elaborating this issue.
+
+I think broad distribution are fine as fairly uninformative priors for inference, but the estimate $D_w$ of a forward process with long jumps is not well defined.
+
 }
@@
 Finally, in panel C, this is difficult to distinguish the curves and their associated symbols (symbols that could also be re-defined in a colour legend included within the box of the graph).
-\response{TODO}
 
@@
 
-\response{TODO}
+\response{Thank you for these suggestions.
+I have renamed the y-axis to "density fluctuations" and explained how they are defined in more detail in the text.
+Their definition is also given in the caption.
+I have also specified in the caption that $r$ is the interaction radius of individuals (and thus is a length) and added a vertical line indicating $D=L^2/2T_c$.
+}
 
@@
 
-\response{This statements refers explicitly to 'extrapolation' into the past and the inference in this case was made form data gathered at one time point.
+\response{This statement refers explicitly to 'extrapolation' into the past and the inference in this case was made form data gathered at one time point.
 If there were samples from the time of the $T_{mrca}$ to the present, the task would be interpolation rather than extrapolation and I would expect the biases to disappear.
@@
 
-I have now included a reference and a note.}
+I have now included a reference along with a footnote.}
 
@@
 
-\response{TODO}
+\response{In the individual based simulation, fitness of individuals is determined by a local density estimator which is given by a sum of Gaussians with characteristic radius $r$ for each individual in the population (see Eq(4)).
+However, even when local density fluctuates much less, the population still deviates from the assumption of the typical phylodynamic/phylogeographic model since the rate of coalescence between any two lineages strongly depends on their spatial separation.}
 
@@
 ``which cab lead to unexpected effective lineage dynamics corresponding to sources where population densities are high and sinks where they are low'': Replace ``cab'' with ``can''.
-Additionally, in this context, beyond Wilkins and Wakeley, it may be worthwhile to cite the Appendix B of Hallatschek, O., & Nelson, D.
+Additionally, in this context, beyond Wilkins and Wakeley, it may be worthwhile to cite the Appendix B of Hallatschek, O., \& Nelson, D.
 R.
@@
 
-\response{TODO}
+\response{Thank you for this suggestion.
+My gut feeling in this context is "interpolation is mostly fine, extrapolation problematic", while parameter estimates are typically hard to interpret.
+I have elaborated on this a bit in the final paragraph of the discussion.}
 

### 2025-03-30 manuscript/phylogeo.tex
work on response

@@
 There is no ground truth to compare the estimates to and $v_w$ and $v_b$ are just descriptive summaries of spatial spread that can not be compared between datasets.
-Similar observations have recently been reported by \citet{dellicour_how_2024}\footnote{I had alerted \citet{dellicour_how_2024} in May 2023 to the problems with the velocity estimators and illustrated the behavior using figures qualitatively similar to Fig.~\ref{fig:D_and_v}}.
+Similar observations have recently been reported by \citet{dellicour_how_2024}\footnote{I had alerted \citet{dellicour_how_2024} in May 2023 to the problems with the velocity estimators and illustrated the sample size dependence using figures qualitatively similar to Fig.~\ref{fig:D_and_v}}.
 
@@
 If the probability of long jumps with distance $x<d<x+dx$ decays more slowly than $\sim dx/x^3$, the expectation value of the $D_w$ diverges.
-In this case, the values of the above estimators for a dataset with long range dispersal are dominated by the longest displacements and a single number like $D_w$ is no longer a useful summary of the typical dispersal process.
-Estimators like $D_w$ might also become problematic under the assumptions of relaxed random walk models with long-tailed prior distributions for $D$.
+When jumps are drawn from a Cauchy distribution, as for example used in \citet{dellicour_how_2024}, the estimator $D_w$ will be dominated by the longest jumps (or the size of the system)\footnote{The largest jump in a sample of $n$ jumps is typically on the order of $2\gamma n/\pi$, where $\gamma$ is the scale of the Cauchy distribution.
+The second largest jump is typically half, the third largest a third, etc.~of the largest value.
+The estimator $D_w$ sums their squares, which in many cases will be dominated by a few very largest values.
+}.
+In this case, $D_w$ is no longer a useful summary of the typical dispersal process as it is determined by the most extreme events that are will fluctuate widely between different realizations of the process or even different samples from the same realization.
 
@@
 Such a coupling, however, is expected since organisms compete for resources and growth rates naturally depend on local population density.
-It has been known for a long time that ignoring such coupling leads to unrealistically heterogeneous population densities \citep{felsenstein_pain_1975}, as is evident in the example of the Kingman tree in Fig.~\ref{fig:illustration_tree}.
+It has been known for a long time that ignoring such coupling leads to unrealistically heterogeneous population densities \citep{felsenstein_pain_1975}, as is evident in the example of the Kingman tree in Fig.~\ref{fig:illustration_tree}:
+The clades defined by long branches in the tree correspond to tight clusters of individuals in space.
+A more realistic population would spread out more evenly across the habitat.
+
 If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D\ll L^2/T_c$, where $T_c$ is the coalescent timescale of a panmictic population ($T_c=N$ in the case of a neutral model with a population of size $N$).
-Fig.~\ref{fig:density_reg}A (black line) quantifies this heterogeneity as the standard deviation of population density in two-dimensional bins of size $L/5$ relative to the expectation in the case of a spatially well-mixed population when each individual picks its location at random.
-For $D< L^2/2T_c$, diffusion is too slow to explore the entire habitat during the time since the MRCA, leading to clustering of individuals.
+I quantified this heterogeneity using by binning individuals into two-dimensional bins with linear dimension $L/5$ (25 bins in total) and calculated the standard deviation of the number of individuals in these bins, normalized by the expectation variation in a well-mixed population when each individual picks its location at random.
+The black line in Fig.~\ref{fig:density_reg}A shows this heterogeneity measured as this standard deviation for a purely diffusive model as a function of the diffusion constant.
+For $D < L^2/2T_c$, diffusion is too slow to explore the entire habitat during the time since the MRCA, leading to clustering of individuals and high heterogeneity.
 This heterogeneity increases substantially for even lower $D$, when the population is essentially a cloud of width $\sqrt{4DT_c}\ll L$.
@@
     {\bf Panel A} shows the standard deviation of the population density in bins of linear dimension $L/5$ as a function of the diffusion coefficient relative to the expected value if individuals were independently distributed across the habitat.
-The thick black line shows the case of free diffusion, the colored lines density regulation with different interaction radii.
+The thick black line shows the case of free diffusion, the colored lines density regulation with different interaction radii $r$.
 For $r\ll L$, density fluctuations are strongly suppressed.
@@
     {\bf Panel C} shows the fold-error of the dispersal estimator $D_w$ for periodic boundary conditions (triangles, solid lines, like panels A\&B) and reflecting boundaries (circles, dashed lines).
-    }
+    The dashed vertical line indicates $D=L^2/2T_c$.}
 \end{figure}
@@
 \citet{pybus_unifying_2012} estimated a diffusion coefficient between 200 and 5000 km$^2$/day, where the latter refers to branches with particularly high diffusion constant.
+\citet{dellicour_how_2024} estimated a diffusion constant of 310$km^2$/day for the expansion phase of the virus in North America and 2-3-fold lower rate after the initial expansion.
 These extremely high estimates of diffusive dispersal are implausible.
@@
 This lower estimate of $D$ implies a daily exploration radius of about 5 km.
-This picture is also consistent with the apparent ``slowing down'' of lineages after the initial expansion across North America \citep{dellicour_epidemiological_2020}: Once the habitat was fully explored, directed range expansion with $\vfkpp$ ceases.
+This picture is also consistent with the apparent ``slowing down'' of lineages after the initial expansion across North America \citep{dellicour_epidemiological_2020, dellicour_how_2024}: Once the habitat was fully explored, directed range expansion with $\vfkpp$ ceases.
 Furthermore, rare long range dispersal could have strongly influenced the spread of the virus in ways not captured by Brownian or relaxed random walks \citep{hallatschek_acceleration_2014}.
+West Nile virus might for example by carried by migratory birds, but their directed migration should not be described by a diffusion constant.
 
@@
 Sampling at multiple time points further constrains ancestral locations sufficiently strongly that there is limited uncertainty, at least within the period for which samples are available.
-But inferences for deeper nodes in the phylogeny before the earliest samples can be problematic as the chance that the environment has changed and shifted, or that the ancestral locations are in poorly sampled regions, increases.
+In these cases, phylogeographic reconstruction effectively interpolates been sampled locations along the relationship structure encoded in the tree and the results tend to be both plausible and well constrained by the data.
+But inferences for deeper nodes in the phylogeny long before the earliest samples can be problematic as the chance that the environment has changed and shifted, or that the true ancestral locations are in poorly sampled regions, increases.
 And while phylogeographic inferences will still often yield plausible results by virtue of their tendency to produce parsimonious solutions that aggregate available spatial information, they can also be highly unreliable and misleading.
 In particular confidence statements should be treated with extreme caution, because the tails of the inferred distributions are determined by properties of the model and the underlying assumptions typically do not reflect the wide spectrum of possible environmental histories.
+This concern also applies to scenarios where the spatial component of the likelihood strongly influences the posterior of phylogenetic or phylodynamic parameters.
+In such cases, chances are that model misspecification such as missing interaction between spatial location and population dynamics will lead to biased inferences.
+Integrating over broad distributions of diffusion constants will make the model more permissive to absorb such misspecifications, but does not address the underlying problem.
+
 In absence of a good understanding of the dispersal properties and how population dynamics varied over time and space, simple models that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models that nevertheless fail to faithfully infer history or essential parameters.

### 2025-03-30 src/stable_density_figure.py

@@ -44,6 +44,7 @@ if __name__=="__main__":
     N=1000
     dr = density_reg[1]
     ax=axs[0]
+    ax.axvline(0.5, c='k', ls='--')
     for i, ir in enumerate(interaction_radius):
         D_array = np.array(resPBC["density_variation"][ir, dr, N, :].index)
         label = f'r={ir}'
@@ -56,7 +57,7 @@ if __name__=="__main__":
     ax.plot(D_array*N/Lx/Ly, np.ones_like(D_array), c='k', ls='--')
     ax.set_xscale('log')
 #    ax.set_xlabel('diffusion constant $[L^2/N]$')
-    ax.set_ylabel('rel. heterogeneity')
+    ax.set_ylabel('density fluctuations')
     ax.legend(fontsize=8)
     ax.set_xlim(0)
     add_panel_label(ax, 'A')
@@ -65,6 +66,7 @@ if __name__=="__main__":
 
     ## Plot the diffusion estiamtes
     ax=axs[2]
+    ax.axvline(0.5, c='k', ls='--')
     for i, ir in enumerate(interaction_radius):
         D_array = np.array(resPBC["diffusion_mean"][ir, dr, N, :].index)
         label = f'r={ir}, a={dr}' if N==N_vals[0] else ''
@@ -104,6 +106,7 @@ if __name__=="__main__":
 
     # Figure showing the TMRCA of the population
     ax=axs[1]
+    ax.axvline(0.5, c='k', ls='--')
     for i, ir in enumerate(interaction_radius):
         D_array = np.array(res["tmrca_mean"][ir, dr, N, :].index)
         label = f'r={ir}' if N==N_vals[1] else ''

### 2025-04-20 src/free_diffusion.py
add gitignore

@@ -27,16 +27,33 @@ def random_tree(n, yule=False):
 
     return {"tree": tree[0], "terminals": terminals}
 
-def add_positions(tree, D):
+def add_positions(tree, D, gauss=True, exponent=3, generation_time=0.01):
     '''
     recursively add positions to the nodes by adding diffusive increments
     '''
+    from scipy.stats import genpareto
+    pareto_mean = genpareto.mean(1/(exponent-1))
     def add_positions_rec(node, D):
         if node["type"] == "internal":
             for child in node["clades"]:
                 dt = child["time"] - node["time"]
-                dx = np.random.randn()*np.sqrt(2*D*dt)
-                dy = np.random.randn()*np.sqrt(2*D*dt)
+                if gauss:
+                    dx = np.random.randn()*np.sqrt(2*D*dt)
+                    dy = np.random.randn()*np.sqrt(2*D*dt)
+                else:
+                    # (1+cx)^(-1 - 1/c) --> e = 1 + 1/c --> c = 1/(e-1)
+                    # mean = int_0^inty x (1+x)^-e dx =  int_0^inty (1+x)/(1+x)^e - 1/(1+x)^e dx
+                    # = 1/(e-2) - 1/(e-1) = ((e-1) - (e-2))/(e-1)/(e-2) = 1/(e-1)/(e-2)
+                    ngen = np.random.poisson(dt/generation_time)
+                    dx=0
+                    dy=0
+                    for i in range(ngen):
+                        # generate a pareto distributed step
+                        angle = np.random.rand()*2*np.pi
+                        d = genpareto.rvs(1/(exponent-1))*np.sqrt(2*D*generation_time)/pareto_mean
+                        dx += np.cos(angle)*d
+                        dy += np.sin(angle)*d
+
                 child['dt'] = dt
                 child['dx'] = dx
                 child['dy'] = dy

### 2025-04-20 src/sample_size_dependence.py

@@ -1,7 +1,6 @@
 import numpy as np
-from free_diffusion import *
-from estimate_diffusion_from_tree import *
-from itertools import product
+from free_diffusion import random_tree, add_positions
+from estimate_diffusion_from_tree import estimate_diffusion
 
 if __name__=="__main__":
     replicates = 100
@@ -13,7 +12,7 @@ if __name__=="__main__":
             res = []
             for i in range(replicates):
                 tree = random_tree(n, yule=coal=='yule')['tree']
-                add_positions(tree, 1)
+                add_positions(tree, 1, gauss=False, exponent=2.5, generation_time=0.1)
                 D = estimate_diffusion(tree)
                 res.append(D)
             print(n, np.mean([x['t_total'] for x in res]))
@@ -50,4 +49,4 @@ if __name__=="__main__":
         ax.set_xscale('log')
     plt.tight_layout()
 
-    plt.savefig(f'figures/dispersal_stats.pdf')
+    plt.savefig('figures/dispersal_stats.pdf')

### 2025-04-20 manuscript/phylogeo.tex
edit round

@@
 \begin{abstract}
-Continuous phylogeographic inference is a popular method to reconstruct the spatial location of ancestors of extant populations and estimate parameters of the dispersal process.
-While the underlying probabilistic models can be complex and their parameters are often computationally demanding to infer, these models typically ignore that replication and population growth are tightly coupled to spatial location: populations expand into fertile uninhabited areas and contract in regions with limited resources.
+Continuous phylogeographic inference is a popular method to estimate parameters of the dispersal process and to reconstruct the spatial location of ancestors of extant populations from samples viral genome sequences.
+However, these models typically ignore that replication and population growth are tightly coupled to spatial location: viral populations expand in areas with abundant susceptible hosts and contract in regions with limited resources.
 Here, I first investigate the sampling consistency of popular summary statistics of dispersal and show that estimators of ``lineage velocities'' are ill-defined.
@@
 These, sometimes dramatic, distortions depend in complicated ways on the past dynamics of habitats and underlying population dynamics and dispersal processes.
-Consequently, the validity of continuous phylogeographic inferences, in particular when involving poorly sampled locations or extrapolations far into the past, is hard to assess and confidence can be much lower than suggested by the inferred posterior distributions.
+Consequently, the validity of continuous phylogeographic inferences is hard to assess and confidence can be much lower than suggested by the inferred posterior distributions, in particular when involving poorly sampled locations or extrapolations far into the past.
 \end{abstract}
@@
 The reconstruction of ancestral locations, past migrations, and dispersal parameters from sampling locations of extant individuals, typically in combination with genome sequence information to infer the phylogeny, is known as phylogeography.
-Phylogeographic methods are implemented in popular evolutionary analysis software such as BEAST \citep{lemey_bayesian_2009,lemey_phylogeography_2010}.
-However, defining consistent models for demographic inference in continuous space is challenging \citep{felsenstein_pain_1975,guindon_demographic_2016}.
+Phylogeographic methods are implemented in popular evolutionary analysis software such as BEAST \citep{lemey_bayesian_2009,lemey_phylogeography_2010} and are widely used to study the spatial dynamics of viral populations \citep{dellicour_epidemiological_2020}.
+However, it has long been known that defining consistent models for demographic inference in continuous space is challenging \citep{felsenstein_pain_1975,guindon_demographic_2016}.
 
@@
 Furthermore, it assumes that the tree topology and the branching rates of the tree are independent of the spatial location.
-Other model component, for example those modeling the tree topology, coalescence properties, and molecular evolution are independent of Eq.~\ref{eq:diffusion}, that is their contributions to the likelihood are multiplicative factors.
+In these models, other components of the model -- for example those modeling the tree topology, coalescence properties, or molecular evolution -- are independent of spatial likelihood and represented as multiplicative factors in posterior.
 A coalescent model, for example, would assume that each pair of lineage merges with equal rates, or a birth-death models would assume identical rates on contemporary branches of the tree.
-Variation of dispersal across the tree can be accounted for in relaxed random walk models, but coupling between growth and location can only be captured in indirect ways.
+Variation of dispersal properties across the tree can be accommodated in relaxed random walk models that allow $D$ in Eq.~\ref{eq:diffusion} to vary, but even in that case coupling between growth, location, and density is only be captured in indirect ways.
 
@@
 Changes of the environment could mean that a habitat of a population is shifting, resulting in population growth and rapid branching at the expanding edge of the habitat and/or contraction in other areas.
-Similarly, invasive species or pathogens spread into areas that support high population densities but were previously unoccupied.
-Examples of such shifting habitats are common and range from geological time scales, over decades (climate change), to weeks (seasonal fluctuations).
+Similarly, pathogens spread into areas that support high population densities but were previously unoccupied.
+Examples of such shifting habitats are common and range from geological time scales, over decades (climate change), to months (seasonal fluctuations).
 Such changes are not restricted to the physical environment, but ecological shifts can similarly affect the habitat.
 In all such cases, the replication process determining the phylogeny of the sample and the spatial location are strongly coupled.
+The extent to which violations of the independence assumption affect continuous phylogeographic inferences is not well understood.
 
@@
 Such models are flexible enough to allow for different population dynamics in different demes \citep{vaughan_efficient_2014}, though in practice can struggle to represent the population in sufficient detail or suffer from identifiability problems \citep{layan_impact_2023}.
-But the extent to which violations of the independence assumption affect continuous phylogeographic inferences is not well understood.
 
@@
 
+\begin{figure*}[tb]
+    \includegraphics*[width=0.99\textwidth]{figures/dispersal_stats.pdf}
+    % \includegraphics*[width=0.48\textwidth]{figures/yule_dispersal.pdf}
+    \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
+    While the diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ (weighted) and $v_b$ (by branch) are not well-defined quantities.
+For Kingman coalescent tree their numerical value depends strongly on sample size and there is no ground truth to compare to.
+The expected scaling $v_w \sim \sqrt{n}$ of $v$ estimates is indicated by the dashed line.}
+\end{figure*}
+
 \section*{Consistency of summary statistics of dispersal parameters.}
@@
 We refer to these here as `by branch' estimates.
-The latter tend to be dominated by short branches and thus noisier \citep{trovao_bayesian_2015}.
+The latter tend to be dominated by short branches and are thus noisier \citep{trovao_bayesian_2015}.
 
@@
 I simulated diffusion along trees from Kingman coalescent \citep{kingman_coalescent_1982} and Yule model \citep{yule_iimathematical_1925} tree ensembles and evaluated the different estimators using the known true ancestral locations.
-Kingman trees correspond to the classical neutral model of a constant size population, while Yule trees capture features of exponentially growing populations.
+Kingman trees correspond to the classical neutral model of a constant size population and are characterized by many short branches and a few deep splits.
+Consequently, the spatial locations of samples from a Kingman tree tend to be clustered.
+Yule trees capture features of exponentially growing populations and tend to have a more even distribution of splits through time and long terminal branches.
 Examples for each ensemble are shown in Fig.~\ref{fig:illustration_tree}.
 
-\begin{figure*}[tb]
-    \includegraphics*[width=0.99\textwidth]{figures/dispersal_stats.pdf}
-    % \includegraphics*[width=0.48\textwidth]{figures/yule_dispersal.pdf}
-    \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
-    While the diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ (weighted) and $v_b$ (by branch) are not well-defined quantities.
-For Kingman coalescent tree their numerical value depends strongly on sample size and there is no ground truth to compare to.
-The expected scaling $v_w \sim \sqrt{n}$ of $v$ estimates is indicated by the dashed line.}
-\end{figure*}
-
 Fig.~\ref{fig:D_and_v} shows empirical estimates of diffusion constants and velocity as defined in Eq.~\ref{eq:dispersal_parameters} from simulated data using freely diffusing locations along Yule and Kingman trees.
-The diffusion constant $D$ can be estimated robustly from these data and the estimates are compatible with the diffusion constant used to simulate the data $D=1$.
+The diffusion constant $D$ can be estimated robustly from these data and the estimates are compatible with the diffusion constant $D=1$ used to simulate the data.
 Values of $v_w$ and $v_b$ estimated from Kingman trees, however, depend strongly on the sample size and are incompatible with each other.
 That the velocity estimates are ill-defined is not surprising: By changing the sample size, the average length of branches in Kingman coalescent trees changes.
-Diffusion along each branch results in a displacement of $d_i \sim \sqrt{4D\Delta t_i}$ such that a quantity that depends on $d_i / \Delta t_i \sim 1/\Delta t_i^{1/2}$ will change with the time $\Delta t_i$ that elapsed along the branch.
+Diffusion along a branch $i$ over time $\Delta t_i$ results in a displacement of $d_i \sim \sqrt{4D\Delta t_i}$ such that a quantity that depends on $d_i / \Delta t_i \sim 1/\Delta t_i^{1/2}$ will change with typical length of branches $\Delta t_i$.
 With more samples in the tree, the average branch length decreases, and the velocity estimate increases.
@@
 There is no ground truth to compare the estimates to and $v_w$ and $v_b$ are just descriptive summaries of spatial spread that can not be compared between datasets.
-Similar observations have recently been reported by \citet{dellicour_how_2024}\footnote{I had alerted \citet{dellicour_how_2024} in May 2023 to the problems with the velocity estimators and illustrated the sample size dependence using figures qualitatively similar to Fig.~\ref{fig:D_and_v}}.
+Similar observations have recently been reported by \citet{dellicour_how_2024}\footnote{I had alerted some of the authors of \citep{dellicour_how_2024} in May 2023 to the problems with the velocity estimators and illustrated the sample size dependence using figures qualitatively similar to Fig.~\ref{fig:D_and_v}}.
 
-While the estimator $D_w$ as defined in Eq.~\ref{eq:dispersal_parameters} is well-behaved when the model is exactly diffusive, long range dispersal can affect $D_w$ dramatically.
+While the estimator $D_w$ as defined in Eq.~\ref{eq:dispersal_parameters} is well-behaved when the model is exactly diffusive, the estimator $D_w$ is not necessarily well-behaved when dispersal occasionally proceeds via long-range migrations.
 If the probability of long jumps with distance $x<d<x+dx$ decays more slowly than $\sim dx/x^3$, the expectation value of the $D_w$ diverges.
-When jumps are drawn from a Cauchy distribution, as for example used in \citet{dellicour_how_2024}, the estimator $D_w$ will be dominated by the longest jumps (or the size of the system)\footnote{The largest jump in a sample of $n$ jumps is typically on the order of $2\gamma n/\pi$, where $\gamma$ is the scale of the Cauchy distribution.
+A popular model for long-range dispersal -- as for example used in \citet{dellicour_how_2024} -- are jumps drawn from a Cauchy distribution.
+The Cauchy distributions decays asymptotically as $1/x^2$ and does not have a finite mean nor variance.
+In this case the estimator $D_w$ will be dominated by the longest jumps (or the size of the system) and fluctuate wildly\footnote{The largest jump in a sample of $n$ jumps from a Cauchy distribution is typically on the order of $2\gamma n/\pi$, where $\gamma$ is the scale of the Cauchy distribution.
 The second largest jump is typically half, the third largest a third, etc.~of the largest value.
-The estimator $D_w$ sums their squares, which in many cases will be dominated by a few very largest values.
+The estimator $D_w$ sums their squares, which in many cases will be dominated by a few of the very largest values.
 }.
-In this case, $D_w$ is no longer a useful summary of the typical dispersal process as it is determined by the most extreme events that are will fluctuate widely between different realizations of the process or even different samples from the same realization.
+When expectation of $D_w$ diverges, the empirical value of $D_w$ is no longer a useful summary of the typical dispersal process as it is determined by the most extreme events that fluctuate widely between different realizations of the process or even different samples from the same realization.
+
+\begin{figure}[!h]
+    \includegraphics[width=0.48\textwidth]{figures/stable_density}
+    \caption{\label{fig:density_reg}  {\bf Effects of population density regulation.}
+    When spatial motion is independent of population density and coalescence is sufficiently rapid, the realized population density fluctuates strongly across the habitat.
+    {\bf Panel A} shows the standard deviation of the population density in bins of linear dimension $L/5$ as a function of the diffusion coefficient relative to the expected value if individuals were independently distributed across the habitat.
+The thick black line shows the case of free diffusion, the colored lines show results from simulations with density regulation with different interaction radii $r$.
+    For $r\ll L$, density fluctuations are strongly suppressed.
+    At the same time, the time to the most recent common ancestor increases dramatically to values several fold higher than the neutral coalescence timescale $T_c=N=1000$ ({\bf panel B}), indicating that the population fragments in subpopulations that don't interact.
+    {\bf Panel C} shows the fold-error of the dispersal estimator $D_w$ for periodic boundary conditions (triangles, solid lines, like panels A\&B) and reflecting boundaries (circles, dashed lines).
+    The dashed vertical line indicates $D=L^2/2T_c$.}
+\end{figure}
 
@@
 Continuous phylogeographic inference typically assumes that the spatial diffusion process is independent of the branching pattern of the tree and ignores coupling between the two.
-Such a coupling, however, is expected since organisms compete for resources and growth rates naturally depend on local population density.
+Such a coupling, however, is expected since organisms compete for resources (e.g. susceptible hosts) and growth rates naturally depend on local population density.
 It has been known for a long time that ignoring such coupling leads to unrealistically heterogeneous population densities \citep{felsenstein_pain_1975}, as is evident in the example of the Kingman tree in Fig.~\ref{fig:illustration_tree}:
@@
 If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D\ll L^2/T_c$, where $T_c$ is the coalescent timescale of a panmictic population ($T_c=N$ in the case of a neutral model with a population of size $N$).
-I quantified this heterogeneity using by binning individuals into two-dimensional bins with linear dimension $L/5$ (25 bins in total) and calculated the standard deviation of the number of individuals in these bins, normalized by the expectation variation in a well-mixed population when each individual picks its location at random.
-The black line in Fig.~\ref{fig:density_reg}A shows this heterogeneity measured as this standard deviation for a purely diffusive model as a function of the diffusion constant.
+I quantified this heterogeneity using by binning individuals into two-dimensional bins with linear dimension $L/5$ (25 bins in total) and calculated the standard deviation of the number of individuals in these bins, normalized by the expected variation in a well-mixed population when each individual picks its location at random.
+The black line in Fig.~\ref{fig:density_reg}A shows this heterogeneity measured by this standard deviation for a purely diffusive model as a function of the diffusion constant.
 For $D < L^2/2T_c$, diffusion is too slow to explore the entire habitat during the time since the MRCA, leading to clustering of individuals and high heterogeneity.
@@
 The comparison of $D$ to the ratio $L^2/T_c$ will continue to be important below, and I therefore use diffusion constants in units of $L^2/T_c$.
-
-\begin{figure}
-    \includegraphics[width=0.48\textwidth]{figures/stable_density}
-    \caption{\label{fig:density_reg}  {\bf Effects of population density regulation.}
-    When spatial motion is independent of population density and coalescence is sufficiently rapid, the realized population density fluctuates strongly across the habitat.
-    {\bf Panel A} shows the standard deviation of the population density in bins of linear dimension $L/5$ as a function of the diffusion coefficient relative to the expected value if individuals were independently distributed across the habitat.
-The thick black line shows the case of free diffusion, the colored lines density regulation with different interaction radii $r$.
-For $r\ll L$, density fluctuations are strongly suppressed.
-    At the same time, the time to the most recent common ancestor increases dramatically to values several fold higher than the neutral coalescence timescale $T_c=N=1000$ ({\bf panel B}), indicating that the population fragments in subpopulations that don't interact.
-    {\bf Panel C} shows the fold-error of the dispersal estimator $D_w$ for periodic boundary conditions (triangles, solid lines, like panels A\&B) and reflecting boundaries (circles, dashed lines).
-    The dashed vertical line indicates $D=L^2/2T_c$.}
-\end{figure}
 
@@
 
-\begin{figure*}
+\begin{figure*}[tb]
     \includegraphics[width=\textwidth]{figures/seasaw_example}
@@
     {\bf Panel E} shows the average $z_x^2$ and $z_y^2$ for the oldest 20\% of internal nodes.
-Inference of the x-coordinate of ancestral nodes is drastically overconfident for some values of $D$.
-The summary inferences in panels D \& E are made from trees of samples from a single time point.
-}
+Inference of the x-coordinate of ancestral nodes is drastically overconfident for some values of $D$ (data for $z_y^2$ are hard to see in the figure as $z_y^2$ is of order 1 as expected).
+    The summary inferences in panels D \& E are made from many trees each of which represents samples from a single time point.  }
 \end{figure*}
@@
 As discussed above, even a static carrying capacity in an equilibrium population can strongly affect population structure, though this structure has limited effects on estimates of ancestral locations and diffusion coefficients beyond boundary effects.
-In reality, carrying capacities vary through time.
+In reality, carrying capacities will often vary through time.
 Such variation happens on all time scales, ranging from geological timescales over millions of years, environmental changes over millennia to seasonal fluctuations.
-The circulation of seasonal influenza viruses in temperate climates, for example, varies by many orders of magnitude between summer and winter.
+The circulation of seasonal influenza viruses in temperate climates, for example, varies by several orders of magnitude between summer and winter.
 
@@
 The carrying capacity is illustrated in Fig.~\ref{fig:seasaw}A for 20 time points covering half a period.
-In between the two extreme left/right concentration of the carrying capacity, the carrying capacity is uniform.
-To avoid extinction at low dispersal rates, selection is soft, meaning that the overall population size is kept approximately constant even if the entire population is stuck in the unfavorable region.
+In between the two extreme left/right concentration of the carrying capacity (indicated by purple and yellow colors, respectively), the carrying capacity is uniform at a time between these two extremes.
+To avoid extinction at low dispersal rates, I implemented soft selection, meaning that the overall population size is kept approximately constant even if the entire population is stuck in the unfavorable region.
 If the dispersal rate is sufficiently high, however, the population follows the shifting carrying capacity and lineages in the tree undergo directed back-and-forth motion.
 Inference of ancestral locations in such situations can be misleading as illustrated in Fig.~\ref{fig:seasaw}C.
-The population was sampled at a single time point when the carrying capacity is flat after having contracted to the left and the positions of older nodes are not estimated correctly.
+The population was sampled at a single time point when the carrying capacity is flat after having contracted to the left.
+The positions of older nodes and in particular of the root (symbols with bold red outline) are not estimated correctly.
 Their true locations are to the right of the current population, but the inferred locations are central in the shifted population.
@@
 
-The severity of such biases depends on the period $T$ at which the carrying capacity oscillates and on the dispersal rate and tends to be strongest when the period is of the same magnitude as the coalescent time scale.
+The severity of such biases depends on the period $T$ at which the carrying capacity oscillates and on the dispersal rate and tends to be strongest when the period is of the same magnitude as the coalescent timescale.
 Fig.~\ref{fig:seasaw}D shows the estimated diffusion constant (using Eq.~\ref{eq:dispersal_parameters}) in $x$ (solid lines) and $y$ (dashed) direction relative to the true value used in the simulations.
@@
 In this range of dispersal rates, the population can no longer follow the shifting carrying capacity and the $T_{mrca}$ increases sharply (not shown).
-This decoupling happens at lower values of $D$ for longer periods of environmental change.
+This decoupling happens at lower values of $D$ for longer periods $T$ of environmental change.
 With the parameters used in the simulation, the estimates of $D_x$ deviate by a factor of 2 from its true value, which is not dramatic compared to the variation of dispersal rates in nature, but clearly shows that spatial inference can be substantially affected if the habitat changes on time scales comparable to the $T_{mrca}$.
@@
 Conversely, at high dispersal rates, the average $z_x^2$ and $z_y^2$ are both smaller than one, i.e.~the inference is not confident enough.
-In this range, the estimated uncertainty is larger than the true uncertainty since the diffusion is constrained by the boundaries of the habitat.
+In this range, the estimated uncertainty is larger than the true uncertainty since the diffusion is constrained by the boundaries of the habitat while the estimated uncertainty extends beyond the habitat.
 
 The above results show that behavior of a population in shifting habitats and the interpretation of phylogeographic inferences can depend strongly on the rate of environmental change and how it compares to the dispersal rate.
-If the population can not follow shifting habitat, the situation is similar to the static case often leads to fragmented populations.
-Once the population starts following, inferred ancestral locations of deep nodes become unreliable, while at high dispersal rates expected uncertainties are larger than the habitat size and there is little information on deep ancestral locations.
+If the population can not follow shifting habitat, the situation is similar to the static case and often leads to fragmented populations.
+Once the population starts following the shifting habitat, inferred ancestral locations of deep nodes become unreliable, while at high dispersal rates expected uncertainties are larger than the habitat size and there is little information on deep ancestral locations.
 
 To study the interplay of moving habitats and phylogeographic inference more systematically, I will now consider a situation where the population is constrained to a stripe with a Gaussian profile that moves at a constant velocity $v$ along the $x$-axis (Fig.~\ref{fig:traveling_wave}).
+This set-up is sometimes referred to as ``moving oasis'' in the population genetics literature \citep{desai_quasispecies_2005}.
 The dispersal rates necessary to keep up with the shifting carrying capacity can be estimated from the Fisher-KPP equation (Eq.~\ref{eq:FKPP}) above.
@@
 
-\begin{figure}
+\begin{figure}[tb]
     \includegraphics[width=0.3\textwidth]{figures/traveling_wave}
@@
 These behaviors are explained comparing the speed $\vfkpp$ at which the population can invade new territory with the external speed $v$, the ratio of which is used as the $x$-axis in Fig.~\ref{fig:traveling_wave} B\&C.
-For $\vfkpp \ll v$, the population cannot keep up with the moving carrying capacity and is saved from extinction only because selection is soft.
+For $\vfkpp \ll v$, the population cannot keep up with the moving carrying capacity and is saved from extinction only because selection is soft \citep{desai_quasispecies_2005}.
 In this regime, the estimated velocity of lineages is below $v$ and both the estimated $v$ and $D$ increase with $\vfkpp$.
@@
 Popular tools for such inference, however, make strong assumptions about the dispersal process, including assuming that dispersal of individuals is diffusive (i.e.~characterized by small displacements in random directions) and that the replication process is independent of spatial location.
-Here, I have explored the robustness of such inference to violations of the latter assumption while keeping microscopic dispersal strictly diffusive, that is without any long-range migration.
+Here, I have explored the robustness of such inference to violations of the latter assumption while keeping microscopic dispersal strictly diffusive, i.e.~I have avoided additional complications arising from long-range migration.
 
-Estimating dispersal ``velocities'' using displacements of lineages along the tree is problematic.
+My first observation is that estimating dispersal ``velocities'' using displacements of lineages along the tree is problematic.
 Firstly, the numerical value of such estimates depends strongly on the sample size and tree ensemble.
@@
 As has long been known, ignoring the coupling between population growth and spatial location can lead to unrealistic population densities \citep{felsenstein_pain_1975} and that density regulation is necessary.
-However, once density regulation is in place, the population dynamics depends strongly on whether dispersal is fast enough to explore the entire habitat during the coalescence time $T_c$ of the population in absence of spatial structure.
+However, once density regulation is in place, the population diversity and history depends strongly on whether dispersal is fast enough to explore the entire habitat during the coalescence time $T_c$ of the population in absence of spatial structure.
 If $D\ll L^2/T_c$, the population fragments into subpopulations and the time to the most recent common ancestor increases dramatically compared to a neutral panmictic population.
 The resulting strong coupling between spatial location and coalescence process violates typical tree priors used in such inference.
-In the other case of $D\gg L^2/T_c$, dispersal is rapid enough that uncertainties of ancestral locations of deep nodes are large compared to the size of the habitat and such inferences thus uninformative.
-At the same time, this will lead to under-estimation of dispersal parameters since ancestral locations are constrained by the boundaries of the habitat.
+In the other case of $D\gg L^2/T_c$, dispersal is rapid enough that uncertainties of ancestral locations of deep nodes are large compared to the size of the habitat and such inferences are thus uninformative.
+At the same time, dispersal parameters are likely underestimated for $D\gg L^2/T_c$ since ancestral locations are constrained by the boundaries of the habitat.
 I did not consider additional effects that emerge from static carrying capacities that vary strongly in space, which can lead to unexpected effective lineage dynamics corresponding to sources where population densities are high and sinks where they are low \citep{wilkins_coalescent_2002, hallatschek_gene_2008}.
@@
 \citet{dellicour_how_2024} estimated a diffusion constant of 310$km^2$/day for the expansion phase of the virus in North America and 2-3-fold lower rate after the initial expansion.
-These extremely high estimates of diffusive dispersal are implausible.
+These extremely high estimates of diffusive dispersal are implausible and likely indicate that directed elements of the spread of the virus are interpreted as undirected dispersal.
 Alternatively, one can interpret the spread from east to west as a range expansion.
@@
 And while phylogeographic inferences will still often yield plausible results by virtue of their tendency to produce parsimonious solutions that aggregate available spatial information, they can also be highly unreliable and misleading.
-In particular confidence statements should be treated with extreme caution, because the tails of the inferred distributions are determined by properties of the model and the underlying assumptions typically do not reflect the wide spectrum of possible environmental histories.
+When population densities are not at equilibrium, there is no guarantee (and indeed we should not expect) that the posterior distributions suggested by such models accurately capture the uncertainty of ancestral events prior to the time interval covered by the data.
+While typical or most likely scenarios might often be predicted correctly, confidence statements should be treated with extreme caution.
+The tails of the inferred distributions are determined by properties of the model and the underlying assumptions typically do not reflect the wide spectrum of possible environmental histories.
+Model misspecification of the kind discussed here can strongly affect the probability mass in the tails of distributions.
 This concern also applies to scenarios where the spatial component of the likelihood strongly influences the posterior of phylogenetic or phylodynamic parameters.
 In such cases, chances are that model misspecification such as missing interaction between spatial location and population dynamics will lead to biased inferences.
-Integrating over broad distributions of diffusion constants will make the model more permissive to absorb such misspecifications, but does not address the underlying problem.
 
-In absence of a good understanding of the dispersal properties and how population dynamics varied over time and space, simple models that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models that nevertheless fail to faithfully infer history or essential parameters.
+A common approach in the field is to make models more complex and sample a larger parameter space, for example allow for a broad distribution of diffusion constants in relaxed random walk models.
+Such expansion of parameter space and integration over additional phenomenological parameters like $D$ will result in better fits and make the model more permissive to absorb  misspecifications, but does not address the underlying problem.
+The computational resources necessary to infer such models increase rapidly with complexity, and it becomes hard to understand how the data inform and constrain the inferences.
+In absence of a good understanding of the dispersal properties past how population dynamics in time and space, simple models that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models that nevertheless can fail to faithfully infer history or essential parameters.
 

### 2025-04-20 manuscript/phylogeo.tex
edits

@@
 
-\begin{figure*}[tb]
-    \includegraphics*[width=0.99\textwidth]{figures/dispersal_stats.pdf}
-    % \includegraphics*[width=0.48\textwidth]{figures/yule_dispersal.pdf}
-    \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
-    While the diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ (weighted) and $v_b$ (by branch) are not well-defined quantities.
-For Kingman coalescent tree their numerical value depends strongly on sample size and there is no ground truth to compare to.
-The expected scaling $v_w \sim \sqrt{n}$ of $v$ estimates is indicated by the dashed line.}
-\end{figure*}
-
 \section*{Consistency of summary statistics of dispersal parameters.}
@@
 Examples for each ensemble are shown in Fig.~\ref{fig:illustration_tree}.
+
+\begin{figure*}[tb]
+    \includegraphics*[width=0.99\textwidth]{figures/dispersal_stats.pdf}
+    \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
+    While the diffusion constant $D$ (equal to 1 in this example) can be robustly estimated from simulation data of the underlying model, the dispersal rate $v_w$ (weighted) and $v_b$ (by branch) are not well-defined quantities.
+For Kingman coalescent tree their numerical value depends strongly on sample size and there is no ground truth to compare to.
+The expected scaling $v_w \sim \sqrt{n}$ of $v$ estimates is indicated by the dashed line.}
+\end{figure*}
 
@@
 
-\begin{figure}[!h]
-    \includegraphics[width=0.48\textwidth]{figures/stable_density}
-    \caption{\label{fig:density_reg}  {\bf Effects of population density regulation.}
-    When spatial motion is independent of population density and coalescence is sufficiently rapid, the realized population density fluctuates strongly across the habitat.
-    {\bf Panel A} shows the standard deviation of the population density in bins of linear dimension $L/5$ as a function of the diffusion coefficient relative to the expected value if individuals were independently distributed across the habitat.
-The thick black line shows the case of free diffusion, the colored lines show results from simulations with density regulation with different interaction radii $r$.
-    For $r\ll L$, density fluctuations are strongly suppressed.
-    At the same time, the time to the most recent common ancestor increases dramatically to values several fold higher than the neutral coalescence timescale $T_c=N=1000$ ({\bf panel B}), indicating that the population fragments in subpopulations that don't interact.
-    {\bf Panel C} shows the fold-error of the dispersal estimator $D_w$ for periodic boundary conditions (triangles, solid lines, like panels A\&B) and reflecting boundaries (circles, dashed lines).
-    The dashed vertical line indicates $D=L^2/2T_c$.}
-\end{figure}
-
 \section*{Effects of density regulation on population structure and dispersal}
@@
 The comparison of $D$ to the ratio $L^2/T_c$ will continue to be important below, and I therefore use diffusion constants in units of $L^2/T_c$.
+
+\begin{figure}[!h]
+    \includegraphics[width=0.48\textwidth]{figures/stable_density}
+    \caption{\label{fig:density_reg}  {\bf Effects of population density regulation.}
+    When spatial motion is independent of population density and coalescence is sufficiently rapid, the realized population density fluctuates strongly across the habitat.
+    {\bf Panel A} shows the standard deviation of the population density in bins of linear dimension $L/5$ as a function of the diffusion coefficient relative to the expected value if individuals were independently distributed across the habitat.
+The thick black line shows the case of free diffusion, the colored lines show results from simulations with density regulation with different interaction radii $r$.
+    For $r\ll L$, density fluctuations are strongly suppressed.
+    At the same time, the time to the most recent common ancestor increases dramatically to values several fold higher than the neutral coalescence timescale $T_c=N=1000$ ({\bf panel B}), indicating that the population fragments in subpopulations that don't interact.
+    {\bf Panel C} shows the fold-error of the dispersal estimator $D_w$ for periodic boundary conditions (triangles, solid lines, like panels A\&B) and reflecting boundaries (circles, dashed lines).
+    The dashed vertical line indicates $D=L^2/2T_c$.}
+\end{figure}
 
@@
 
-A common approach in the field is to make models more complex and sample a larger parameter space, for example allow for a broad distribution of diffusion constants in relaxed random walk models.
+A common approach in the field is to make models more complex and sample a larger parameter space, for example by allowing for a broad distribution of diffusion constants in relaxed random walk models.
 Such expansion of parameter space and integration over additional phenomenological parameters like $D$ will result in better fits and make the model more permissive to absorb misspecifications, but does not address the underlying problem.
-The computational resources necessary to infer such models increase rapidly with complexity, and it becomes hard to understand how the data inform and constrain the inferences.
-In absence of a good understanding of the dispersal properties past how population dynamics in time and space, simple models that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models that nevertheless can fail to faithfully infer history or essential parameters.
+The computational resources necessary to infer such models increase rapidly with complexity, it becomes harder to understand how the data inform the inferences, and interpretation of inferred parameter values becomes ambiguous as they might have absorbed effects not explicitly modeled.
+In absence of a good understanding of the dispersal properties past how population dynamics in time and space that defines model structure, simple approaches that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models that can nevertheless fail to faithfully infer history or essential parameters.
 

### 2025-04-20 manuscript/ResponseToReviews.tex
add to response

@@
 
-\response{Thank you very much for this assessment and for highlighting where the manuscript falls short. I address these points individually below.}
+\response{Thank you very much for this assessment and for highlighting where the manuscript could be improved. I address these points individually below.}
 
@@
 
-\response{I discuss this towards the end of the manuscript.
+\response{It is true that phylogeographic reconstructions often produce plausible results.
+I discuss this towards the end of the manuscript.
 Phylogeographic methods tend to produce a parsimonious reconstruction of past migrations which is often plausible (and will often also cover the actual history).
 However, these inferences depend on certain assumptions about the past and extrapolations into the past are dangerous.
+In particular, these reconstructions can be overconfident, don't explore scenarios that can arise when habitats shift, and summary statistics used in phylogeography don't correspond to biological parameters or are not faithfully estimated.
 
-I have extended this section in the discussion to make clear when plausible inferences are expected, and when inferences should interpreted with care.
+I have extended this section in the discussion to make clear when plausible inferences are expected, and when inferences should be interpreted with care.
 }
@@
 While the implications of some of the issues might apply more broadly, I feel that phylogeographic inference is sufficiently central to the analysis of viral genomic data and the readership of Virus Evolution, that the current focus is justified.
-I discuss examples of Rabius virus and West Nile virus. }
+I discuss examples of Rabies virus and West Nile virus. }
 
@@
 
-\response{This statement refers to the simplest model in the equation just above -- but point taken.
+\response{This statement referred to the simplest model in the equation just above -- but point taken.
 My understanding is that RRW models assume a distribution of dispersal rates that can vary along the tree, but that this distribution is the same everywhere.
-This can indeed result in different inferred dispersal properties in different parts of the habitat.
+This can indeed result in different inferred dispersal properties in different parts of the habitat, though does not generate an explicit coupling between density and tree statistics.
 I added a clarifying statement.
@@
 The estimator $D_w$ is an empirical average over the squared displacement along different branches, scaled by the inverse elapsed time.
-However, the Cauchy distribution is often used to model long range dispersal, but it does not have a finite mean nor a variance.
-The central limit theorem does not apply for sums of many such variables and estimators based on such sums will be dominated by the largest jumps.
-Estimates of $D_w$ will therefore be very noisy and often dominated by largest jumps possible in the systems.
+For this estimator to be well behaved, the second moment of the displacements along branches needs to be well defined.
+
+However, if these displacements are drawn from a Cauchy distribution  (a popular model for long range dispersal), this second moment does not exist.
+In fact the Cauchy distribution has neither a finite mean nor a variance.
+For any distribution where the probability of large jumps over distance $d$ decays more slowly than $d^{-3}$, the central limit theorem does not apply for sums of many such variables and estimators based on such sums will be dominated by the largest terms.
+Estimates of $D_w$ in such cases will therefore be very noisy and often dominated by largest jumps possible in the systems.
+
 In the concrete case of a Cauchy distribution, the largest jump will typically be twice the distance of the second largest jump, 3 times as large as the third largest jump, etc.
@@
 \response{I would argue as stated, these assumptions are true for both the BM and RRW model.
-The latter allows the diffusion coefficient to vary which allows to accommodate a lot of scenarios in inference, but it does make the tree prior in anyway depend on location.
+The latter allows the diffusion coefficient to vary which allows to accommodate more flexible scenarios during inference, but it does make the tree prior in anyway depend on location or density.
 }.
@@
 
-\response{This statement refers explicitly to 'extrapolation' into the past and the inference in this case was made form data gathered at one time point.
+\response{This statement refers explicitly to 'extrapolation' into the past and the inference in this case was made from data gathered at one time point.
 If there were samples from the time of the $T_{mrca}$ to the present, the task would be interpolation rather than extrapolation and I would expect the biases to disappear.
@@
 
-\response{Note that the dispersal model used to generate these data is purely diffusive with no drift.
+\response{Note that the dispersal model used to generate these data is purely diffusive with no drift or any variation in the diffusion process.
 The only part that is time varying is the habitat.
@@
 
-\response{It can accommodate such jumps, but the inferred parameters still won't be meaningful. }
+\response{It can accommodate such jumps will lead to much better fits (in the sense of a higher posterior probability), but the inferred parameters still won't be meaningful and extrapolation back time can not be trusted.
+}
 
@@
 
-\response{Thanks for highlighting.}
+\response{Thanks for highlighting.
+But the wording ``[...] increased bird immunity through time slowing down spatial dispersal'' highlights the core point of this manuscript and its critique of phylogeographic inference: immunity does not affect the dispersal process -- birds and mosquitos fly around just the same way.
+But immunity affects how fast the viral population can expand.
+If this differs spatially, dispersal and growth lead to a wave-like expansion.
+While the resulting data can be fit with a (maybe RRW) diffusion model, it is fundamentally the wrong model for the expansion phase.
+As a result, parameter estimates don't have a biological interpretation and confidence intervals, even of interpretable quantities like locations, can not be trusted.}
 

### 2025-04-20 manuscript/ResponseToReviews.tex
edit to response

@@
 While the implications of some of the issues might apply more broadly, I feel that phylogeographic inference is sufficiently central to the analysis of viral genomic data and the readership of Virus Evolution, that the current focus is justified.
-I discuss examples of Rabies virus and West Nile virus. }
+I discuss examples of Rabies virus and West Nile virus. I now made it clear in the abstract and introduction that many applications are in the domain of viral evolution.}
 
@@
 \response{I am aware of these works, and I am happy to include a reference.
-But for context, I think it is important to note that I pointed out these problems to the authors in the summer of 2023, including figures that illustrate the problem and an interpretation why that is.
+But for context, I think it is important to note that I pointed out these problems to Simon Dellicour and Philippe Lemey in the summer of 2023, including figures that illustrate the problem and an interpretation why that is.
 They invited me to be a co-author on a study into which their analysis of the problem was integrated, but I ultimately didn't feel comfortable with some of the other content of the work and declined to be a co-author.
 I gave them the go-ahead to publish this, and as I see now I am acknowledged in this work for giving 'useful feedback'.
-I decided to write up my thoughts on this problem myself, resulting in this work that has been on biorxiv since July last year.
+I decided to write up my thoughts on this problem myself, resulting in this work that I shared with Simon and Phillipe June 2024 and posted on biorxiv in July last year.
 
-I have now included a reference along with a footnote.}
+I have now included a reference to Dellicour et al 2024 along with a footnote describing the relation of their work to mine.}
 

### 2025-05-17 manuscript/phylogeo.tex
minor edits

@@
 Continuous phylogeographic inference is a popular method to estimate parameters of the dispersal process and to reconstruct the spatial location of ancestors of extant populations from samples viral genome sequences.
-However, these models typically ignore that replication and population growth are tightly coupled to spatial location: viral populations expand in areas with abundant susceptible hosts and contract in regions with limited resources.
+However, these models typically ignore that replication and population growth are tightly coupled to spatial location: populations expand in areas with abundant susceptible hosts and contract in regions with limited resources.
 Here, I first investigate the sampling consistency of popular summary statistics of dispersal and show that estimators of ``lineage velocities'' are ill-defined.
@@
 In phylogeographic inference, one can either assume that individuals migrate between discrete locations or disperse in continuous space.
-Here, we consider migration in continuous space, where the underlying model is typically assumed to be diffusive, that is the probability of sampling a descendant at position $\rvec_c$ after time $t$ when the parent was at position $\rvec_p$ is given by
+Here, we consider migration in continuous space, where the underlying model is typically assumed to be diffusive, i.e.~the probability of sampling a descendant at position $\rvec_c$ after time $t$ when the parent was at position $\rvec_p$ is given by
 \begin{equation}
@@
 Furthermore, it assumes that the tree topology and the branching rates of the tree are independent of the spatial location.
-In these models, other components of the model -- for example those modeling the tree topology, coalescence properties, or molecular evolution -- are independent of spatial likelihood and represented as multiplicative factors in posterior.
+Specifically, different components of the model -- for example those describing the tree topology, coalescence properties, or molecular evolution -- are independent of spatial likelihood and represented as multiplicative factors in posterior.
 A coalescent model, for example, would assume that each pair of lineage merges with equal rates, or a birth-death models would assume identical rates on contemporary branches of the tree.
@@
 However, population dynamics are often tightly coupled to spatial location.
-Changes of the environment could mean that a habitat of a population is shifting, resulting in population growth and rapid branching at the expanding edge of the habitat and/or contraction in other areas.
-Similarly, pathogens spread into areas that support high population densities but were previously unoccupied.
+Changes of the environment, for example, could mean that a habitat of a population is shifting, resulting in population growth and rapid branching at the expanding edge of the habitat and/or contraction in other areas.
+Similarly, pathogens spread in susceptible host populations that had not been previously exposed or have lost immunity.
 Examples of such shifting habitats are common and range from geological time scales, over decades (climate change), to months (seasonal fluctuations).
-Such changes are not restricted to the physical environment, but ecological shifts can similarly affect the habitat.
+Such changes are not restricted to the physical environment or population immunity, but ecological shifts can similarly affect the habitat.
 In all such cases, the replication process determining the phylogeny of the sample and the spatial location are strongly coupled.
@@
 
-In a typical phylogeographic inference problem, the input data are genome sequences as well as their sampling times and sampling locations.
+In a typical phylogeographic inference problem, the input data are genome sequences along with their sampling times and sampling locations.
 Software like BEAST will then sample the posterior distribution of the phylogenetic tree topology, the sampling times and locations of ancestral nodes in the tree, and model parameters like the diffusion constant $D$ and the evolutionary rate $\mu$.
@@
 Using the true trees will result in clearer demonstrations of the limits of phylogeographic inference without the additional complication of tree uncertainty.
-Furthermore, I will assume strictly diffusive dispersal without any long-range migrations.
-In practice, long-range migrations, errors in tree reconstruction, and sampling biases will add to the problems investigated here.
+Furthermore, I will assume strictly diffusive dispersal without any long-range migrations -- the simulation and inference therefore use the same model.
+In practice, model violations through long-range migrations, errors in tree reconstruction, and sampling biases will add to the problems investigated here.
 
@@
 There is no ground truth to compare the estimates to and $v_w$ and $v_b$ are just descriptive summaries of spatial spread that can not be compared between datasets.
-Similar observations have recently been reported by \citet{dellicour_how_2024}\footnote{I had alerted some of the authors of \citep{dellicour_how_2024} in May 2023 to the problems with the velocity estimators and illustrated the sample size dependence using figures qualitatively similar to Fig.~\ref{fig:D_and_v}}.
+Similar observations have recently been reported by \citet{dellicour_how_2024}\footnote{I had alerted some of the authors of \citep{dellicour_how_2024} in May 2023 to the problems with the velocity estimators.
+We independently explored the behavior of velocity estimators.}.
 
 While the estimator $D_w$ as defined in Eq.~\ref{eq:dispersal_parameters} is well-behaved when the model is exactly diffusive, the estimator $D_w$ is not necessarily well-behaved when dispersal occasionally proceeds via long-range migrations.
-If the probability of long jumps with distance $x<d<x+dx$ decays more slowly than $\sim dx/x^3$, the expectation value of the $D_w$ diverges.
+If the probability of long jumps with distance $d$ in the interval $x<d<x+dx$ decays more slowly than $\sim dx/x^3$, the expectation value of the $D_w$ diverges.
 A popular model for long-range dispersal -- as for example used in \citet{dellicour_how_2024} -- are jumps drawn from a Cauchy distribution.
@@
 In this case the estimator $D_w$ will be dominated by the longest jumps (or the size of the system) and fluctuate wildly\footnote{The largest jump in a sample of $n$ jumps from a Cauchy distribution is typically on the order of $2\gamma n/\pi$, where $\gamma$ is the scale of the Cauchy distribution.
-The second largest jump is typically half, the third largest a third, etc.~of the largest value.
+The second-largest jump is typically half, the third largest a third, etc.~of the largest value.
 The estimator $D_w$ sums their squares, which in many cases will be dominated by a few of the very largest values.
@@
 If one simulates a diffusive spatial process in a square of length $L$ with periodic boundary conditions, the population density shows strong fluctuations whenever the diffusion constant $D\ll L^2/T_c$, where $T_c$ is the coalescent timescale of a panmictic population ($T_c=N$ in the case of a neutral model with a population of size $N$).
-I quantified this heterogeneity using by binning individuals into two-dimensional bins with linear dimension $L/5$ (25 bins in total) and calculated the standard deviation of the number of individuals in these bins, normalized by the expected variation in a well-mixed population when each individual picks its location at random.
+I quantified this heterogeneity by binning individuals into two-dimensional bins with linear dimension $L/5$ (25 bins in total) and calculated the standard deviation of the number of individuals in these bins, normalized by the expected variation in a well-mixed population when each individual picks its location at random.
 The black line in Fig.~\ref{fig:density_reg}A shows this heterogeneity measured by this standard deviation for a purely diffusive model as a function of the diffusion constant.
@@
     When spatial motion is independent of population density and coalescence is sufficiently rapid, the realized population density fluctuates strongly across the habitat.
-    {\bf Panel A} shows the standard deviation of the population density in bins of linear dimension $L/5$ as a function of the diffusion coefficient relative to the expected value if individuals were independently distributed across the habitat.
+    {\bf Panel A} shows the standard deviation of the population density in bins of linear dimension $L/5$ as a function of the diffusion coefficient relative to the expected value if individuals were independently distributed across the habitat (dashed horizontal line).
 The thick black line shows the case of free diffusion, the colored lines show results from simulations with density regulation with different interaction radii $r$.
     For $r\ll L$, density fluctuations are strongly suppressed.
-    At the same time, the time to the most recent common ancestor increases dramatically to values several fold higher than the neutral coalescence timescale $T_c=N=1000$ ({\bf panel B}), indicating that the population fragments in subpopulations that don't interact.
+    When density regulation homogenizes density, the time to the most recent common ancestor increases dramatically to values several fold higher than the neutral coalescence timescale $T_c=N=1000$ ({\bf panel B}), indicating that the population fragments in subpopulations that don't interact.
     {\bf Panel C} shows the fold-error of the dispersal estimator $D_w$ for periodic boundary conditions (triangles, solid lines, like panels A\&B) and reflecting boundaries (circles, dashed lines).
@@
 
-Instead of the coalescent trees used above, I now used an individual based simulation forward in time.
+Instead of the coalescent trees used above, I used an individual based simulation forward in time.
 Individuals are assigned a fitness $f=\alpha(1-\hat{\phi}/\phi_0) + (1-N/N_0)$ and produce a Poisson distributed number of offspring with mean $1+f$.
@@
 
-These simulations reveal that both at high and at low diffusion constants, assumptions of phylogeographic inference can be problematic: if diffusion constants are much smaller than $L^2/T_{c}$, the population is fragmented into effective subpopulations and this fragmentation violates the assumptions of phylodynamic models.
-This could potentially be accounted for in structured coalescent model, but demes are not discrete and not known a priori such that these models are unlikely a practical alternative.
+These simulations reveal that both at high and at low diffusion constants, assumptions of phylogeographic inference can be problematic: if diffusion constants are much smaller than $L^2/T_{c}$, the population is fragmented into effective subpopulations.
+This fragmentation violates the assumptions of phylodynamic models and in particular will affect estimates of population sizes.
+Such fragmentation could potentially be accounted for in structured coalescent model, but demes are not discrete and not known a priori such that these models are unlikely a practical alternative.
 If $D\gg L^2/T_{c}$, the boundaries of the habitat restrict free diffusion, resulting in underestimated dispersal parameters, while estimates of deeper ancestral locations have uncertainties comparable to the habitat size and are thus uninformative.
@@
 Inference of the x-coordinate of ancestral nodes is drastically overconfident for some values of $D$ (data for $z_y^2$ are hard to see in the figure as $z_y^2$ is of order 1 as expected).
-    The summary inferences in panels D \& E are made from many trees each of which represents samples from a single time point.  }
+    The summary inferences in panels D \& E are made from many trees each of which represents samples from a single time point.}
 \end{figure*}
@@
 
-The severity of such biases depends on the period $T$ at which the carrying capacity oscillates and on the dispersal rate and tends to be strongest when the period is of the same magnitude as the coalescent timescale.
+The severity of such biases depends on the period $T$ at which the carrying capacity oscillates, the population size, and on the dispersal rate.
+The bias tends to be strongest when the period $T$ is of the same magnitude as the coalescent timescale.
 Fig.~\ref{fig:seasaw}D shows the estimated diffusion constant (using Eq.~\ref{eq:dispersal_parameters}) in $x$ (solid lines) and $y$ (dashed) direction relative to the true value used in the simulations.
@@
 The above results show that behavior of a population in shifting habitats and the interpretation of phylogeographic inferences can depend strongly on the rate of environmental change and how it compares to the dispersal rate.
-If the population can not follow shifting habitat, the situation is similar to the static case and often leads to fragmented populations.
-Once the population starts following the shifting habitat, inferred ancestral locations of deep nodes become unreliable, while at high dispersal rates expected uncertainties are larger than the habitat size and there is little information on deep ancestral locations.
+At very low $D$ when the population can not follow shifting habitat, the situation is similar to the static case and often leads to fragmented populations.
+With increasing $D$, the population starts following the shifting habitat, but inferred ancestral locations of deep nodes become unreliable.
+At high dispersal rates the expected uncertainties are larger than the habitat size and there is little information on deep ancestral locations.
 
-To study the interplay of moving habitats and phylogeographic inference more systematically, I will now consider a situation where the population is constrained to a stripe with a Gaussian profile that moves at a constant velocity $v$ along the $x$-axis (Fig.~\ref{fig:traveling_wave}).
+To study the interplay of moving habitats and phylogeographic inference more systematically, I will now consider a situation where the population is constrained to a stripe with a Gaussian profile that moves at a constant velocity $v$ along the $x$-axis (Fig.~\ref{fig:traveling_wave}A).
 This set-up is sometimes referred to as ``moving oasis'' in the population genetics literature \citep{desai_quasispecies_2005}.
@@
     \includegraphics[width=0.48\textwidth]{figures/waves}
-    % \includegraphics[width=0.48\textwidth]{figures/waves_v}
     \caption{\label{fig:traveling_wave}{\bf Phylogeographic inference in a moving habitat.}
@@
 They peak right around when the external velocity matches the FKPP velocity $\vfkpp$, corresponding to the situation where the population dynamics is maximally affected by the moving carrying capacity.
-If the external velocity was lower, undirected dispersal contributes more and more to the displacement of lineages, while the population would be unable to follow the moving carrying capacity if the external velocity was higher.
+At higher diffusion constants (to the right of the peak in Fig \ref{fig:traveling_wave}B), the effects of the moving habitat are less pronounced compared to undirected diffusion and the estimates $\hat{D}$ are close to the true value $D$.
+At lower diffusion (to the left of the peak) the population is unable to follow the moving carrying capacity.
 
-I showed above that estimates of $v$ using Eq.~\ref{eq:dispersal_parameters} are ill-defined in models with diffusive dispersal, but it is worth exploring how they behave in cases when one can define a velocity within the model.
+I showed above that estimates of $v$ using Eq.~\ref{eq:dispersal_parameters} are ill-defined in models with diffusive dispersal, but it is worth exploring how they behave in cases when the habitat is moving with an external velocity.
 Fig.~\ref{fig:traveling_wave}C shows the estimator $v_w$ in $x$ direction relative to velocity of the habitat.
 This ratio is monotonically increasing with the diffusion coefficient and shows a range where it is close to 1, i.e.~where the estimated velocity of lineages agrees with the external velocity.
-This range coincides with the location of the peak of $D_w$.
 
@@
 However, as discussed above, this velocity is determined by the interplay of dispersal and population growth at the front, not a directional motion of individuals.
-It might, however, be possible to capture this effective directional motions by random walks models that allow for directed motion \citep{gill_relaxed_2017}.
+It might be possible to capture this effective directional motions by random walks models that allow for directed motion \citep{gill_relaxed_2017}.
 Estimators of a ``lineage velocity'' as defined in Eq.~\ref{eq:dispersal_parameters} only agree with invasion speed in a narrow, a priori unknown, parameter regime.
-The speed of invasion into a new habitat is better determined by modeling the position of the front \citep{pybus_unifying_2012}, rather than by tracking individual lineages along the tree.
+In summary, the speed of invasion into a new habitat is better determined by modeling the position of the front \citep{pybus_unifying_2012}, rather than by tracking individual lineages along the tree.
 
@@
 As has long been known, ignoring the coupling between population growth and spatial location can lead to unrealistic population densities \citep{felsenstein_pain_1975} and that density regulation is necessary.
-However, once density regulation is in place, the population diversity and history depends strongly on whether dispersal is fast enough to explore the entire habitat during the coalescence time $T_c$ of the population in absence of spatial structure.
+However, once density regulation is in place, the population diversity and history depends strongly on whether dispersal is fast enough to explore the entire habitat during the time $T_c$ it would take the population to coalesce in absence of spatial structure.
 If $D\ll L^2/T_c$, the population fragments into subpopulations and the time to the most recent common ancestor increases dramatically compared to a neutral panmictic population.
-The resulting strong coupling between spatial location and coalescence process violates typical tree priors used in such inference.
+The resulting strong coupling between spatial location and coalescence process violates typical tree priors used in such inference and distorts estimates of ancestral population sizes.
 In the other case of $D\gg L^2/T_c$, dispersal is rapid enough that uncertainties of ancestral locations of deep nodes are large compared to the size of the habitat and such inferences are thus uninformative.
@@
 \citet{pybus_unifying_2012} estimated a diffusion coefficient between 200 and 5000 km$^2$/day, where the latter refers to branches with particularly high diffusion constant.
-\citet{dellicour_how_2024} estimated a diffusion constant of 310$km^2$/day for the expansion phase of the virus in North America and 2-3-fold lower rate after the initial expansion.
+\citet{dellicour_how_2024} estimated a comparable diffusion constant of 310$km^2$/day ($D_w$ rather than $D_b$) for the expansion phase of the virus in North America and 2-3-fold lower rate after the initial expansion.
 These extremely high estimates of diffusive dispersal are implausible and likely indicate that directed elements of the spread of the virus are interpreted as undirected dispersal.
@@
 To this end, we need an estimate of the growth rate $\alpha$ of the population at low density, which we can roughly obtain from the fact that the virus is seasonal: over the course of a few months, its prevalence increases by orders of magnitude, which typically requires a doubling time of about a week or a growth rate of $\alpha=0.1/$day.
-Using $\vfkpp = 2\sqrt{D\alpha}$ and equating it to 1000 km/year ($\approx$ 3km/day), we obtain estimates of $D$ of $D\approx \frac{9 \mathrm{km}^2}{4 \alpha \mathrm{day}^2} \approx 22\mathrm{km}^2/\mathrm{day}$ -- around 10 to 200 times smaller than estimates from phylogeography.
+Using $\vfkpp = 2\sqrt{D\alpha}$ and equating it to 1000 km/year ($\approx$ 3km/day), we obtain the estimate $D\approx \frac{9 \mathrm{km}^2}{4 \alpha \mathrm{day}^2} \approx 22\mathrm{km}^2/\mathrm{day}$ -- around 10 to 200 times smaller than estimates from phylogeography.
 This lower estimate of $D$ implies a daily exploration radius of about 5 km.
@@
 Furthermore, rare long range dispersal could have strongly influenced the spread of the virus in ways not captured by Brownian or relaxed random walks \citep{hallatschek_acceleration_2014}.
-West Nile virus might for example by carried by migratory birds, but their directed migration should not be described by a diffusion constant.
+West Nile virus might for example by carried by migratory birds and their directed migration should not be described by a diffusion constant.
 
@@
 Sampling at multiple time points further constrains ancestral locations sufficiently strongly that there is limited uncertainty, at least within the period for which samples are available.
-In these cases, phylogeographic reconstruction effectively interpolates been sampled locations along the relationship structure encoded in the tree and the results tend to be both plausible and well constrained by the data.
+In these cases, phylogeographic reconstruction effectively interpolates between sampled locations along the relationship structure encoded in the tree and the results tend to be both plausible and well constrained by the data.
 But inferences for deeper nodes in the phylogeny long before the earliest samples can be problematic as the chance that the environment has changed and shifted, or that the true ancestral locations are in poorly sampled regions, increases.
@@
 This concern also applies to scenarios where the spatial component of the likelihood strongly influences the posterior of phylogenetic or phylodynamic parameters.
-In such cases, chances are that model misspecification such as missing interaction between spatial location and population dynamics will lead to biased inferences.
+In such cases, model misspecification such as missing interaction between spatial location and population dynamics can lead to biased inferences.
 
@@
 Such expansion of parameter space and integration over additional phenomenological parameters like $D$ will result in better fits and make the model more permissive to absorb misspecifications, but does not address the underlying problem.
-The computational resources necessary to infer such models increase rapidly with complexity, it becomes harder to understand how the data inform the inferences, and interpretation of inferred parameter values becomes ambiguous as they might have absorbed effects not explicitly modeled.
-In absence of a good understanding of the dispersal properties past how population dynamics in time and space that defines model structure, simple approaches that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models that can nevertheless fail to faithfully infer history or essential parameters.
+The computational resources necessary to infer such models increase rapidly with complexity, it becomes harder to understand how the data inform the inferences, and interpretation of inferred parameter values becomes ambiguous as they might have absorbed effects that were not explicitly modeled.
+In absence of a good understanding of the dispersal properties and the population history in time and space, simple approaches that allow transparent interpretation of how results are informed by features of the data are preferable to complex and computationally demanding models that can nevertheless fail to faithfully infer history or essential parameters.
 
@@
 \section*{Acknowledgements}
-I am grateful to Louis de Plessis, Oskar Hallatscheck, Emma Hodcroft, Simon Dellicour, and Philippe Lemey for stimulating discussions and useful feedback on earlier versions of this manuscript.
+I am grateful to Louis de Plessis, Oskar Hallatscheck, Emma Hodcroft, Philippe Lemey, and particularly Simon Dellicour for stimulating discussions and useful feedback on earlier versions of this manuscript.
 

### 2025-05-17 manuscript/phylogeo.tex
inverse affiliation
inline bib

@@
 \begin{document}
-%\title{On the robustness and consistency of continuous phylogeographic inference}
-%\title{Phylogeography: unsafe at any speed}
 \title{Lost in the woods: shifting habitats can lead phylogeography astray}
 \author{Richard A.~Neher}
-\affiliation{Swiss Institute of Bioinformatics, and Biozentrum, University of Basel, Switzerland \looseness=-5}
+\affiliation{Biozentrum, University of Basel and Swiss Institute of Bioinformatics, Switzerland \looseness=-5}
+\email{richard.neher@unibas.ch}
 
@@
 \begin{figure}[tb]
-    \includegraphics*[width=\columnwidth]{figures/illustration_tree.pdf}
+    \includegraphics*[width=\columnwidth]{figures/Fig1}
     \caption{\label{fig:illustration_tree}{\bf Illustration Yule and Kingman trees with $N=100$ tips and the spatial location of the tips.}
@@
 \begin{figure*}[tb]
-    \includegraphics*[width=0.99\textwidth]{figures/dispersal_stats.pdf}
+    \includegraphics*[width=0.99\textwidth]{figures/Fig2}
     \caption{\label{fig:D_and_v}{\bf Empirical summary statistics of $D$ and ``dispersal velocity'' for different sample sizes.}
@@
 \begin{figure}[!h]
-    \includegraphics[width=0.48\textwidth]{figures/stable_density}
+    \includegraphics[width=0.48\textwidth]{figures/Fig3}
     \caption{\label{fig:density_reg}  {\bf Effects of population density regulation.}
@@
 \begin{figure*}[tb]
-    \includegraphics[width=\textwidth]{figures/seasaw_example}
-    \includegraphics[width=\textwidth]{figures/seasaw}
+    \includegraphics[width=\textwidth]{figures/Fig4}
     \caption{\label{fig:seasaw} {\bf Phylogeographic estimates in rapidly changing environments.} {\bf Panel A} shows the time course of the carrying capacity through half a period.
@@
 \begin{figure}[tb]
-    \includegraphics[width=0.3\textwidth]{figures/traveling_wave}
-    \includegraphics[width=0.48\textwidth]{figures/waves}
+    \includegraphics[width=\columnwidth]{figures/Fig5}
     \caption{\label{fig:traveling_wave}{\bf Phylogeographic inference in a moving habitat.}
@@
 
-\bibliography{bib}
+\begin{thebibliography}{27}
+    \expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi
+    \expandafter\ifx\csname bibnamefont\endcsname\relax
+      \def\bibnamefont#1{#1}\fi
+    \expandafter\ifx\csname bibfnamefont\endcsname\relax
+      \def\bibfnamefont#1{#1}\fi
+    \expandafter\ifx\csname citenamefont\endcsname\relax
+      \def\citenamefont#1{#1}\fi
+    \expandafter\ifx\csname url\endcsname\relax
+      \def\url#1{\texttt{#1}}\fi
+    \expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
+    \providecommand{\bibinfo}[2]{#2}
+    \providecommand{\eprint}[2][]{\url{#2}}
+
+    \bibitem[{\citenamefont{Bolker and Pacala}(1997)}]{bolker_using_1997}
+    \bibinfo{author}{\bibnamefont{Bolker}, \bibfnamefont{B.}}, and
+      \bibinfo{author}{\bibfnamefont{S.~W.} \bibnamefont{Pacala}},
+      \bibinfo{year}{1997}, \bibinfo{journal}{Theoretical Population Biology}
+      \textbf{\bibinfo{volume}{52}}(\bibinfo{number}{3}), \bibinfo{pages}{179},
+      ISSN \bibinfo{issn}{0040-5809}.
+
+    \bibitem[{\citenamefont{Brunker} \emph{et~al.}(2015)\citenamefont{Brunker,
+      Marston, Horton, Cleaveland, Fooks, Kazwala, Ngeleja, Lembo, Sambo, Mtema,
+      Sikana, Wilkie} \emph{et~al.}}]{brunker_elucidating_2015}
+    \bibinfo{author}{\bibnamefont{Brunker}, \bibfnamefont{K.}},
+      \bibinfo{author}{\bibfnamefont{D.~A.} \bibnamefont{Marston}},
+      \bibinfo{author}{\bibfnamefont{D.~L.} \bibnamefont{Horton}},
+      \bibinfo{author}{\bibfnamefont{S.}~\bibnamefont{Cleaveland}},
+      \bibinfo{author}{\bibfnamefont{A.~R.} \bibnamefont{Fooks}},
+      \bibinfo{author}{\bibfnamefont{R.}~\bibnamefont{Kazwala}},
+      \bibinfo{author}{\bibfnamefont{C.}~\bibnamefont{Ngeleja}},
+      \bibinfo{author}{\bibfnamefont{T.}~\bibnamefont{Lembo}},
+      \bibinfo{author}{\bibfnamefont{M.}~\bibnamefont{Sambo}},
+      \bibinfo{author}{\bibfnamefont{Z.~J.} \bibnamefont{Mtema}},
+      \bibinfo{author}{\bibfnamefont{L.}~\bibnamefont{Sikana}},
+      \bibinfo{author}{\bibfnamefont{G.}~\bibnamefont{Wilkie}}, \emph{et~al.},
+      \bibinfo{year}{2015}, \bibinfo{journal}{Virus Evolution}
+      \textbf{\bibinfo{volume}{1}}(\bibinfo{number}{1}), \bibinfo{pages}{vev011},
+      ISSN \bibinfo{issn}{2057-1577}.
+
+    \bibitem[{\citenamefont{Dellicour} \emph{et~al.}(2024)\citenamefont{Dellicour,
+      Bastide, Rocu, Fargette, Hardy, Suchard, Guindon, and
+      Lemey}}]{dellicour_how_2024}
+    \bibinfo{author}{\bibnamefont{Dellicour}, \bibfnamefont{S.}},
+      \bibinfo{author}{\bibfnamefont{P.}~\bibnamefont{Bastide}},
+      \bibinfo{author}{\bibfnamefont{P.}~\bibnamefont{Rocu}},
+      \bibinfo{author}{\bibfnamefont{D.}~\bibnamefont{Fargette}},
+      \bibinfo{author}{\bibfnamefont{O.~J.} \bibnamefont{Hardy}},
+      \bibinfo{author}{\bibfnamefont{M.~A.} \bibnamefont{Suchard}},
+      \bibinfo{author}{\bibfnamefont{S.}~\bibnamefont{Guindon}}, and
+      \bibinfo{author}{\bibfnamefont{P.}~\bibnamefont{Lemey}},
+      \bibinfo{year}{2024}, \bibinfo{journal}{PLOS Biology}
+      \textbf{\bibinfo{volume}{22}}(\bibinfo{number}{12}),
+      \bibinfo{pages}{e3002914}, ISSN \bibinfo{issn}{1545-7885},
+      \bibinfo{note}{publisher: Public Library of Science}.
+
+    \bibitem[{\citenamefont{Dellicour} \emph{et~al.}(2021)\citenamefont{Dellicour,
+      Gill, Faria, Rambaut, Pybus, Suchard, and Lemey}}]{dellicour_relax_2021}
+    \bibinfo{author}{\bibnamefont{Dellicour}, \bibfnamefont{S.}},
+      \bibinfo{author}{\bibfnamefont{M.~S.} \bibnamefont{Gill}},
+      \bibinfo{author}{\bibfnamefont{N.~R.} \bibnamefont{Faria}},
+      \bibinfo{author}{\bibfnamefont{A.}~\bibnamefont{Rambaut}},
+      \bibinfo{author}{\bibfnamefont{O.~G.} \bibnamefont{Pybus}},
+      \bibinfo{author}{\bibfnamefont{M.~A.} \bibnamefont{Suchard}}, and
+      \bibinfo{author}{\bibfnamefont{P.}~\bibnamefont{Lemey}},
+      \bibinfo{year}{2021}, \bibinfo{journal}{Molecular Biology and Evolution}
+      \textbf{\bibinfo{volume}{38}}(\bibinfo{number}{8}), \bibinfo{pages}{3486},
+      ISSN \bibinfo{issn}{1537-1719}.
+
+    \bibitem[{\citenamefont{Dellicour} \emph{et~al.}(2020)\citenamefont{Dellicour,
+      Lequime, Vrancken, Gill, Bastide, Gangavarapu, Matteson, Tan, du~Plessis,
+      Fisher, Nelson, Gilbert} \emph{et~al.}}]{dellicour_epidemiological_2020}
+    \bibinfo{author}{\bibnamefont{Dellicour}, \bibfnamefont{S.}},
+      \bibinfo{author}{\bibfnamefont{S.}~\bibnamefont{Lequime}},
+      \bibinfo{author}{\bibfnamefont{B.}~\bibnamefont{Vrancken}},
+      \bibinfo{author}{\bibfnamefont{M.~S.} \bibnamefont{Gill}},
+      \bibinfo{author}{\bibfnamefont{P.}~\bibnamefont{Bastide}},
+      \bibinfo{author}{\bibfnamefont{K.}~\bibnamefont{Gangavarapu}},
+      \bibinfo{author}{\bibfnamefont{N.~L.} \bibnamefont{Matteson}},
+      \bibinfo{author}{\bibfnamefont{Y.}~\bibnamefont{Tan}},
+      \bibinfo{author}{\bibfnamefont{L.}~\bibnamefont{du~Plessis}},
+      \bibinfo{author}{\bibfnamefont{A.~A.} \bibnamefont{Fisher}},
+      \bibinfo{author}{\bibfnamefont{M.~I.} \bibnamefont{Nelson}},
+      \bibinfo{author}{\bibfnamefont{M.}~\bibnamefont{Gilbert}}, \emph{et~al.},
+      \bibinfo{year}{2020}, \bibinfo{journal}{Nature Communications}
+      \textbf{\bibinfo{volume}{11}}(\bibinfo{number}{1}), \bibinfo{pages}{5620},
+      ISSN \bibinfo{issn}{2041-1723}, \bibinfo{note}{publisher: Nature Publishing
+      Group}.
+
+    \bibitem[{\citenamefont{Dellicour} \emph{et~al.}(2017)\citenamefont{Dellicour,
+      Rose, Faria, Vieira, Bourhy, Gilbert, Lemey, and
+      Pybus}}]{dellicour_using_2017}
+    \bibinfo{author}{\bibnamefont{Dellicour}, \bibfnamefont{S.}},
+      \bibinfo{author}{\bibfnamefont{R.}~\bibnamefont{Rose}},
+      \bibinfo{author}{\bibfnamefont{N.~R.} \bibnamefont{Faria}},
+      \bibinfo{author}{\bibfnamefont{L.~F.~P.} \bibnamefont{Vieira}},
+      \bibinfo{author}{\bibfnamefont{H.}~\bibnamefont{Bourhy}},
+      \bibinfo{author}{\bibfnamefont{M.}~\bibnamefont{Gilbert}},
+      \bibinfo{author}{\bibfnamefont{P.}~\bibnamefont{Lemey}}, and
+      \bibinfo{author}{\bibfnamefont{O.~G.} \bibnamefont{Pybus}},
+      \bibinfo{year}{2017}, \bibinfo{journal}{Molecular Biology and Evolution}
+      \textbf{\bibinfo{volume}{34}}(\bibinfo{number}{10}), \bibinfo{pages}{2563},
+      ISSN \bibinfo{issn}{0737-4038}.
+
+    \bibitem[{\citenamefont{Desai and Nelson}(2005)}]{desai_quasispecies_2005}
+    \bibinfo{author}{\bibnamefont{Desai}, \bibfnamefont{M.~M.}}, and
+      \bibinfo{author}{\bibfnamefont{D.~R.} \bibnamefont{Nelson}},
+      \bibinfo{year}{2005}, \bibinfo{journal}{Theoretical Population Biology}
+      \textbf{\bibinfo{volume}{67}}(\bibinfo{number}{1}), \bibinfo{pages}{33}, ISSN
+      \bibinfo{issn}{0040-5809}.
+
+    \bibitem[{\citenamefont{Etheridge}(2004)}]{etheridge_survival_2004}
+    \bibinfo{author}{\bibnamefont{Etheridge}, \bibfnamefont{A.~M.}},
+      \bibinfo{year}{2004}, \bibinfo{journal}{The Annals of Applied Probability}
+      \textbf{\bibinfo{volume}{14}}(\bibinfo{number}{1}), \bibinfo{pages}{188},
+      ISSN \bibinfo{issn}{1050-5164}, \bibinfo{note}{publisher: Institute of
+      Mathematical Statistics}.
+
+    \bibitem[{\citenamefont{Felsenstein}(1975)}]{felsenstein_pain_1975}
+    \bibinfo{author}{\bibnamefont{Felsenstein}, \bibfnamefont{J.}},
+      \bibinfo{year}{1975}, \bibinfo{journal}{The American Naturalist}
+      \textbf{\bibinfo{volume}{109}}(\bibinfo{number}{967}), \bibinfo{pages}{359},
+      ISSN \bibinfo{issn}{0003-0147}, \bibinfo{note}{publisher: The University of
+      Chicago Press}.
+
+    \bibitem[{\citenamefont{Fisher}(1937)}]{fisher_wave_1937}
+    \bibinfo{author}{\bibnamefont{Fisher}, \bibfnamefont{R.~A.}},
+      \bibinfo{year}{1937}, \bibinfo{journal}{Annals of Eugenics}
+      \textbf{\bibinfo{volume}{7}}(\bibinfo{number}{4}), \bibinfo{pages}{355}, ISSN
+      \bibinfo{issn}{2050-1439}.
+
+    \bibitem[{\citenamefont{Gill} \emph{et~al.}(2017)\citenamefont{Gill, Tung~Ho,
+      Baele, Lemey, and Suchard}}]{gill_relaxed_2017}
+    \bibinfo{author}{\bibnamefont{Gill}, \bibfnamefont{M.~S.}},
+      \bibinfo{author}{\bibfnamefont{L.~S.} \bibnamefont{Tung~Ho}},
+      \bibinfo{author}{\bibfnamefont{G.}~\bibnamefont{Baele}},
+      \bibinfo{author}{\bibfnamefont{P.}~\bibnamefont{Lemey}}, and
+      \bibinfo{author}{\bibfnamefont{M.~A.} \bibnamefont{Suchard}},
+      \bibinfo{year}{2017}, \bibinfo{journal}{Systematic Biology}
+      \textbf{\bibinfo{volume}{66}}(\bibinfo{number}{3}), \bibinfo{pages}{299},
+      ISSN \bibinfo{issn}{1063-5157}.
+
+    \bibitem[{\citenamefont{Guindon} \emph{et~al.}(2016)\citenamefont{Guindon, Guo,
+      and Welch}}]{guindon_demographic_2016}
+    \bibinfo{author}{\bibnamefont{Guindon}, \bibfnamefont{S.}},
+      \bibinfo{author}{\bibfnamefont{H.}~\bibnamefont{Guo}}, and
+      \bibinfo{author}{\bibfnamefont{D.}~\bibnamefont{Welch}},
+      \bibinfo{year}{2016}, \bibinfo{journal}{Theoretical Population Biology}
+      \textbf{\bibinfo{volume}{111}}, \bibinfo{pages}{43}, ISSN
+      \bibinfo{issn}{0040-5809}.
+
+    \bibitem[{\citenamefont{Hallatschek and
+      Fisher}(2014)}]{hallatschek_acceleration_2014}
+    \bibinfo{author}{\bibnamefont{Hallatschek}, \bibfnamefont{O.}}, and
+      \bibinfo{author}{\bibfnamefont{D.~S.} \bibnamefont{Fisher}},
+      \bibinfo{year}{2014}, \bibinfo{journal}{Proceedings of the National Academy
+      of Sciences} \textbf{\bibinfo{volume}{111}}(\bibinfo{number}{46}),
+      \bibinfo{pages}{E4911}, \bibinfo{note}{publisher: Proceedings of the National
+      Academy of Sciences}.
+
+    \bibitem[{\citenamefont{Hallatschek and Nelson}(2008)}]{hallatschek_gene_2008}
+    \bibinfo{author}{\bibnamefont{Hallatschek}, \bibfnamefont{O.}}, and
+      \bibinfo{author}{\bibfnamefont{D.~R.} \bibnamefont{Nelson}},
+      \bibinfo{year}{2008}, \bibinfo{journal}{Theoretical Population Biology}
+      \textbf{\bibinfo{volume}{73}}(\bibinfo{number}{1}), \bibinfo{pages}{158},
+      ISSN \bibinfo{issn}{0040-5809}.
+
+    \bibitem[{\citenamefont{Hallatschek and Nelson}(2010)}]{hallatschek_life_2010}
+    \bibinfo{author}{\bibnamefont{Hallatschek}, \bibfnamefont{O.}}, and
+      \bibinfo{author}{\bibfnamefont{D.~R.} \bibnamefont{Nelson}},
+      \bibinfo{year}{2010}, \bibinfo{journal}{Evolution}
+      \textbf{\bibinfo{volume}{64}}(\bibinfo{number}{1}), \bibinfo{pages}{193},
+      ISSN \bibinfo{issn}{0014-3820}.
+
+    \bibitem[{\citenamefont{Kalkauskas}
+      \emph{et~al.}(2021)\citenamefont{Kalkauskas, Perron, Sun, Goldman, Baele,
+      Guindon, and Maio}}]{kalkauskas_sampling_2021}
+    \bibinfo{author}{\bibnamefont{Kalkauskas}, \bibfnamefont{A.}},
+      \bibinfo{author}{\bibfnamefont{U.}~\bibnamefont{Perron}},
+      \bibinfo{author}{\bibfnamefont{Y.}~\bibnamefont{Sun}},
+      \bibinfo{author}{\bibfnamefont{N.}~\bibnamefont{Goldman}},
+      \bibinfo{author}{\bibfnamefont{G.}~\bibnamefont{Baele}},
+      \bibinfo{author}{\bibfnamefont{S.}~\bibnamefont{Guindon}}, and
+      \bibinfo{author}{\bibfnamefont{N.~D.} \bibnamefont{Maio}},
+      \bibinfo{year}{2021}, \bibinfo{journal}{PLOS Computational Biology}
+      \textbf{\bibinfo{volume}{17}}(\bibinfo{number}{1}),
+      \bibinfo{pages}{e1008561}, ISSN \bibinfo{issn}{1553-7358},
+      \bibinfo{note}{publisher: Public Library of Science}.
+
+    \bibitem[{\citenamefont{Kingman}(1982)}]{kingman_coalescent_1982}
+    \bibinfo{author}{\bibnamefont{Kingman}, \bibfnamefont{J.~F.~C.}},
+      \bibinfo{year}{1982}, \bibinfo{journal}{Stochastic Processes and their
+      Applications} \textbf{\bibinfo{volume}{13}}(\bibinfo{number}{3}),
+      \bibinfo{pages}{235}, ISSN \bibinfo{issn}{0304-4149}.
+
+    \bibitem[{\citenamefont{Kolmogorov}
+      \emph{et~al.}(1937)\citenamefont{Kolmogorov, Petrovskii, and
+      Piscunov}}]{KPP1937}
+    \bibinfo{author}{\bibnamefont{Kolmogorov}, \bibfnamefont{A.}},
+      \bibinfo{author}{\bibfnamefont{I.}~\bibnamefont{Petrovskii}}, and
+      \bibinfo{author}{\bibfnamefont{N.}~\bibnamefont{Piscunov}},
+      \bibinfo{year}{1937}, \bibinfo{journal}{Byul. Moskovskogo Gos. Univ.}
+      \textbf{\bibinfo{volume}{1}}(\bibinfo{number}{6}), \bibinfo{pages}{1}.
+
+    \bibitem[{\citenamefont{Layan} \emph{et~al.}(2023)\citenamefont{Layan, Müller,
+      Dellicour, De~Maio, Bourhy, Cauchemez, and Baele}}]{layan_impact_2023}
+    \bibinfo{author}{\bibnamefont{Layan}, \bibfnamefont{M.}},
+      \bibinfo{author}{\bibfnamefont{N.~F.} \bibnamefont{Müller}},
+      \bibinfo{author}{\bibfnamefont{S.}~\bibnamefont{Dellicour}},
+      \bibinfo{author}{\bibfnamefont{N.}~\bibnamefont{De~Maio}},
+      \bibinfo{author}{\bibfnamefont{H.}~\bibnamefont{Bourhy}},
+      \bibinfo{author}{\bibfnamefont{S.}~\bibnamefont{Cauchemez}}, and
+      \bibinfo{author}{\bibfnamefont{G.}~\bibnamefont{Baele}},
+      \bibinfo{year}{2023}, \bibinfo{journal}{Virus Evolution}
+      \textbf{\bibinfo{volume}{9}}(\bibinfo{number}{1}), \bibinfo{pages}{vead010},
+      ISSN \bibinfo{issn}{2057-1577}.
+
+    \bibitem[{\citenamefont{Lemey} \emph{et~al.}(2009)\citenamefont{Lemey, Rambaut,
+      Drummond, and Suchard}}]{lemey_bayesian_2009}
+    \bibinfo{author}{\bibnamefont{Lemey}, \bibfnamefont{P.}},
+      \bibinfo{author}{\bibfnamefont{A.}~\bibnamefont{Rambaut}},
+      \bibinfo{author}{\bibfnamefont{A.~J.} \bibnamefont{Drummond}}, and
+      \bibinfo{author}{\bibfnamefont{M.~A.} \bibnamefont{Suchard}},
+      \bibinfo{year}{2009}, \bibinfo{journal}{PLOS Computational Biology}
+      \textbf{\bibinfo{volume}{5}}(\bibinfo{number}{9}), \bibinfo{pages}{e1000520},
+      ISSN \bibinfo{issn}{1553-7358}, \bibinfo{note}{publisher: Public Library of
+      Science}.
+
+    \bibitem[{\citenamefont{Lemey} \emph{et~al.}(2010)\citenamefont{Lemey, Rambaut,
+      Welch, and Suchard}}]{lemey_phylogeography_2010}
+    \bibinfo{author}{\bibnamefont{Lemey}, \bibfnamefont{P.}},
+      \bibinfo{author}{\bibfnamefont{A.}~\bibnamefont{Rambaut}},
+      \bibinfo{author}{\bibfnamefont{J.~J.} \bibnamefont{Welch}}, and
+      \bibinfo{author}{\bibfnamefont{M.~A.} \bibnamefont{Suchard}},
+      \bibinfo{year}{2010}, \bibinfo{journal}{Molecular Biology and Evolution}
+      \textbf{\bibinfo{volume}{27}}(\bibinfo{number}{8}), \bibinfo{pages}{1877},
+      ISSN \bibinfo{issn}{0737-4038}.
+
+    \bibitem[{\citenamefont{Pybus} \emph{et~al.}(2012)\citenamefont{Pybus, Suchard,
+      Lemey, Bernardin, Rambaut, Crawford, Gray, Arinaminpathy, Stramer, Busch, and
+      Delwart}}]{pybus_unifying_2012}
+    \bibinfo{author}{\bibnamefont{Pybus}, \bibfnamefont{O.~G.}},
+      \bibinfo{author}{\bibfnamefont{M.~A.} \bibnamefont{Suchard}},
+      \bibinfo{author}{\bibfnamefont{P.}~\bibnamefont{Lemey}},
+      \bibinfo{author}{\bibfnamefont{F.~J.} \bibnamefont{Bernardin}},
+      \bibinfo{author}{\bibfnamefont{A.}~\bibnamefont{Rambaut}},
+      \bibinfo{author}{\bibfnamefont{F.~W.} \bibnamefont{Crawford}},
+      \bibinfo{author}{\bibfnamefont{R.~R.} \bibnamefont{Gray}},
+      \bibinfo{author}{\bibfnamefont{N.}~\bibnamefont{Arinaminpathy}},
+      \bibinfo{author}{\bibfnamefont{S.~L.} \bibnamefont{Stramer}},
+      \bibinfo{author}{\bibfnamefont{M.~P.} \bibnamefont{Busch}}, and
+      \bibinfo{author}{\bibfnamefont{E.~L.} \bibnamefont{Delwart}},
+      \bibinfo{year}{2012}, \bibinfo{journal}{Proceedings of the National Academy
+      of Sciences} \textbf{\bibinfo{volume}{109}}(\bibinfo{number}{37}),
+      \bibinfo{pages}{15066}, \bibinfo{note}{publisher: Proceedings of the National
+      Academy of Sciences}.
+
+    \bibitem[{\citenamefont{Raghwani} \emph{et~al.}(2011)\citenamefont{Raghwani,
+      Rambaut, Holmes, Hang, Hien, Farrar, Wills, Lennon, Birren, Henn, and
+      Simmons}}]{raghwani_endemic_2011}
+    \bibinfo{author}{\bibnamefont{Raghwani}, \bibfnamefont{J.}},
+      \bibinfo{author}{\bibfnamefont{A.}~\bibnamefont{Rambaut}},
+      \bibinfo{author}{\bibfnamefont{E.~C.} \bibnamefont{Holmes}},
+      \bibinfo{author}{\bibfnamefont{V.~T.} \bibnamefont{Hang}},
+      \bibinfo{author}{\bibfnamefont{T.~T.} \bibnamefont{Hien}},
+      \bibinfo{author}{\bibfnamefont{J.}~\bibnamefont{Farrar}},
+      \bibinfo{author}{\bibfnamefont{B.}~\bibnamefont{Wills}},
+      \bibinfo{author}{\bibfnamefont{N.~J.} \bibnamefont{Lennon}},
+      \bibinfo{author}{\bibfnamefont{B.~W.} \bibnamefont{Birren}},
+      \bibinfo{author}{\bibfnamefont{M.~R.} \bibnamefont{Henn}}, and
+      \bibinfo{author}{\bibfnamefont{C.~P.} \bibnamefont{Simmons}},
+      \bibinfo{year}{2011}, \bibinfo{journal}{PLOS Pathogens}
+      \textbf{\bibinfo{volume}{7}}(\bibinfo{number}{6}), \bibinfo{pages}{e1002064},
+      ISSN \bibinfo{issn}{1553-7374}, \bibinfo{note}{publisher: Public Library of
+      Science}.
+
+    \bibitem[{\citenamefont{Trovão} \emph{et~al.}(2015)\citenamefont{Trovão,
+      Suchard, Baele, Gilbert, and Lemey}}]{trovao_bayesian_2015}
+    \bibinfo{author}{\bibnamefont{Trovão}, \bibfnamefont{N.~S.}},
+      \bibinfo{author}{\bibfnamefont{M.~A.} \bibnamefont{Suchard}},
+      \bibinfo{author}{\bibfnamefont{G.}~\bibnamefont{Baele}},
+      \bibinfo{author}{\bibfnamefont{M.}~\bibnamefont{Gilbert}}, and
+      \bibinfo{author}{\bibfnamefont{P.}~\bibnamefont{Lemey}},
+      \bibinfo{year}{2015}, \bibinfo{journal}{Molecular Biology and Evolution}
+      \textbf{\bibinfo{volume}{32}}(\bibinfo{number}{12}), \bibinfo{pages}{3264},
+      ISSN \bibinfo{issn}{0737-4038}.
+
+    \bibitem[{\citenamefont{Vaughan} \emph{et~al.}(2014)\citenamefont{Vaughan,
+      Kühnert, Popinga, Welch, and Drummond}}]{vaughan_efficient_2014}
+    \bibinfo{author}{\bibnamefont{Vaughan}, \bibfnamefont{T.~G.}},
+      \bibinfo{author}{\bibfnamefont{D.}~\bibnamefont{Kühnert}},
+      \bibinfo{author}{\bibfnamefont{A.}~\bibnamefont{Popinga}},
+      \bibinfo{author}{\bibfnamefont{D.}~\bibnamefont{Welch}}, and
+      \bibinfo{author}{\bibfnamefont{A.~J.} \bibnamefont{Drummond}},
+      \bibinfo{year}{2014}, \bibinfo{journal}{Bioinformatics}
+      \textbf{\bibinfo{volume}{30}}(\bibinfo{number}{16}), \bibinfo{pages}{2272},
+      ISSN \bibinfo{issn}{1367-4803}.
+
+    \bibitem[{\citenamefont{Wilkins and Wakeley}(2002)}]{wilkins_coalescent_2002}
+    \bibinfo{author}{\bibnamefont{Wilkins}, \bibfnamefont{J.~F.}}, and
+      \bibinfo{author}{\bibfnamefont{J.}~\bibnamefont{Wakeley}},
+      \bibinfo{year}{2002}, \bibinfo{journal}{Genetics}
+      \textbf{\bibinfo{volume}{161}}(\bibinfo{number}{2}), \bibinfo{pages}{873},
+      ISSN \bibinfo{issn}{0016-6731}.
+
+    \bibitem[{\citenamefont{Yule}(1925)}]{yule_iimathematical_1925}
+    \bibinfo{author}{\bibnamefont{Yule}, \bibfnamefont{G.~U.}},
+      \bibinfo{year}{1925}, \bibinfo{journal}{Philosophical Transactions of the
+      Royal Society of London. Series B, Containing Papers of a Biological
+      Character} \textbf{\bibinfo{volume}{213}}(\bibinfo{number}{402-410}),
+      \bibinfo{pages}{21}, \bibinfo{note}{publisher: Royal Society}.
+
+    \end{thebibliography}
 

### 2025-05-17 src/sample_size_dependence.py
revert long jumps mode

@@ -12,7 +12,8 @@ if __name__=="__main__":
             res = []
             for i in range(replicates):
                 tree = random_tree(n, yule=coal=='yule')['tree']
-                add_positions(tree, 1, gauss=False, exponent=2.5, generation_time=0.1)
+                add_positions(tree, 1, gauss=True)
+                #add_positions(tree, 1, gauss=False, exponent=2.5, generation_time=0.1)
                 D = estimate_diffusion(tree)
                 res.append(D)
             print(n, np.mean([x['t_total'] for x in res]))

### 2025-05-18 manuscript/phylogeo.tex
edit pass

@@
     For $r\ll L$, density fluctuations are strongly suppressed.
-    When density regulation homogenizes density, the time to the most recent common ancestor increases dramatically to values several fold higher than the neutral coalescence timescale $T_c=N=1000$ ({\bf panel B}), indicating that the population fragments in subpopulations that don't interact.
+    When density regulation homogenizes the population distribution, the time to the most recent common ancestor increases dramatically to values several fold higher than the neutral coalescence timescale $T_c=N=1000$ ({\bf panel B}), indicating that the population fragments in subpopulations that don't interact.
     {\bf Panel C} shows the fold-error of the dispersal estimator $D_w$ for periodic boundary conditions (triangles, solid lines, like panels A\&B) and reflecting boundaries (circles, dashed lines).
@@
 
-The severity of such biases depends on the period $T$ at which the carrying capacity oscillates, the population size, and on the dispersal rate.
+The severity of such biases depends on the period $T$ at which the carrying capacity oscillates, on the population size, and on the dispersal rate.
 The bias tends to be strongest when the period $T$ is of the same magnitude as the coalescent timescale.
@@
 The above results show that behavior of a population in shifting habitats and the interpretation of phylogeographic inferences can depend strongly on the rate of environmental change and how it compares to the dispersal rate.
-At very low $D$ when the population can not follow shifting habitat, the situation is similar to the static case and often leads to fragmented populations.
+At very low $D$ the population can not follow shifting habitat and the situation is similar to the static case discussed above, which often leads to fragmented populations.
 With increasing $D$, the population starts following the shifting habitat, but inferred ancestral locations of deep nodes become unreliable.
@@
 They peak right around when the external velocity matches the FKPP velocity $\vfkpp$, corresponding to the situation where the population dynamics is maximally affected by the moving carrying capacity.
-At higher diffusion constants (to the right of the peak in Fig \ref{fig:traveling_wave}B), the effects of the moving habitat are less pronounced compared to undirected diffusion and the estimates $\hat{D}$ are close to the true value $D$.
-At lower diffusion (to the left of the peak) the population is unable to follow the moving carrying capacity.
+At higher diffusion constants (to the right of the peak in Fig \ref{fig:traveling_wave}B), the effects of the moving habitat are less pronounced compared to the more vigorous undirected diffusion and the estimates $\hat{D}$ approach the true value as $D$ increases.
+At lower diffusion constants (to the left of the peak) the population is unable to follow the moving carrying capacity.
 
@@
 \citet{pybus_unifying_2012} estimated a diffusion coefficient between 200 and 5000 km$^2$/day, where the latter refers to branches with particularly high diffusion constant.
-\citet{dellicour_how_2024} estimated a comparable diffusion constant of 310$km^2$/day ($D_w$ rather than $D_b$) for the expansion phase of the virus in North America and 2-3-fold lower rate after the initial expansion.
+\citet{dellicour_how_2024} estimated a comparable diffusion constant of 310$km^2$/day (via $D_w$ rather than $D_b$) for the expansion phase of the virus in North America and 2-3-fold lower rate after the initial expansion.
 These extremely high estimates of diffusive dispersal are implausible and likely indicate that directed elements of the spread of the virus are interpreted as undirected dispersal.
@@
 Model misspecification of the kind discussed here can strongly affect the probability mass in the tails of distributions.
-This concern also applies to scenarios where the spatial component of the likelihood strongly influences the posterior of phylogenetic or phylodynamic parameters.
+This concern particularly applies to scenarios population density, expansion, or contraction varies strongly in space and time.
 In such cases, model misspecification such as missing interaction between spatial location and population dynamics can lead to biased inferences.
 
-A common approach in the field is to make models more complex and sample a larger parameter space, for example by allowing for a broad distribution of diffusion constants in relaxed random walk models.
+A common approach in the field is to make models more complex and increase model flexibility by adding more parameters, for example by allowing for a broad distribution of diffusion constants in relaxed random walk models.
 Such expansion of parameter space and integration over additional phenomenological parameters like $D$ will result in better fits and make the model more permissive to absorb misspecifications, but does not address the underlying problem.
@@
 \section*{Materials and methods}
-All simulation were performed in Python and the associated code is available in the github repository \href{https://github.com/neherlab/phylogeography}{github.com/neherlab/phylogeography}.
+All simulations were performed in Python and the associated code is available in the github repository \href{https://github.com/neherlab/phylogeography}{github.com/neherlab/phylogeography}.
 

### 2025-05-19 manuscript/phylogeo.tex
added affiliation and data availability

@@
 \author{Richard A.~Neher}
-\affiliation{Biozentrum, University of Basel and Swiss Institute of Bioinformatics, Switzerland \looseness=-5}
+\affiliation{Biozentrum, University of Basel and Swiss Institute of Bioinformatics, Spitalstrasse 41, 4053 Basel, Switzerland \looseness=-5}
 \email{richard.neher@unibas.ch}
@@
 All simulations were performed in Python and the associated code is available in the github repository \href{https://github.com/neherlab/phylogeography}{github.com/neherlab/phylogeography}.
+\subsection*{Data availability}
+The manuscript reports exclusively simulation results.
+Aggregate output of the simulations are available in the github repository \href{https://github.com/neherlab/phylogeography}{github.com/neherlab/phylogeography} archived at \href{https://doi.org/10.5281/zenodo.15462925}{10.5281/zenodo.15462925}.
 
